### 2012-03-02 draft1_VRsmoothing.tex
Starting merge.

\documentclass[journal]{IEEEtran}

\usepackage{cite}
\usepackage[pdftex]{graphicx}
\usepackage[cmex10]{amsmath}
\usepackage{amsfonts}
\usepackage{algorithmic}
\usepackage{array}
\usepackage{mdwmath}
\usepackage{mdwtab}
\usepackage{eqparbox}
\usepackage[caption=false,font=footnotesize]{subfig}
\usepackage{fixltx2e}
\usepackage{url}
\usepackage{color}

\newenvironment{meta}[0]{\color{red} \em}{}

% correct bad hyphenation here
\hyphenation{}

\begin{document}

\title{Smoothing Algorithms for Variable Rate Models}

\author{Pete~Bunch,~\IEEEmembership{}
        Simon~Godsill,~\IEEEmembership{Member,~IEEE,}% <-this % stops a space
\thanks{P. Bunch and S. Godsill are with the Department
of Engineering, Cambridge University, UK. email: \{pb404,sjg30\}@cam.ac.uk}% <-this % stops a space
\thanks{Manuscript received January 01, 1901; revised January 02, 1901.}}

% The paper headers
\markboth{IEEE Transaction in Signal Processing,~Vol.~1, No.~1, January~1901}%
{Bunch \& Godsill: Smoothing Algorithms for Variable Rate Models}

% make the title area
\maketitle

\begin{abstract}
The abstract goes here.
\end{abstract}

\begin{IEEEkeywords}

\end{IEEEkeywords}

\section{Introduction}

\IEEEPARstart{T}{he} objective of sequential Bayesian inference is to estimate an imperfectly observed quantity as it varies over time. This is accomplished through the use of probabilistic models for the state evolution and measurement processes. Often, the latent state is a continuously varying quantity, whereas the observations are made at a discrete set of times. In these circumstances, it is simplest to discretise the state onto the same set of times as the observations. When the system is also Markovian, this leads to the standard ``fixed rate'' hidden Markov model (HMM). The standard HMM is poorly suited to systems where the state evolution contains discontinuities; for example, the price of a financial asset which may display large jumps at random times between periods of diffusion-like behaviour, or the kinematic state of a manoeuvring vehicle which may have sudden changes in the acceleration when turns begin or end. Such problems can be handled more naturally using a ``variable rate'' model, in which the state dynamics are conditioned upon a set of random changepoints which characterise transitions in behaviour.

In a variable rate model, the set of changepoints and associated parameters are modelled as a marked point process (MPP), the mathematical properties of which are thoroughly set out in \cite{Jacobsen2006}. Conditional upon this MPP the state evolves according to some benign dynamics. In \cite{Godsill2007,Whiteley2011}, the conditional state evolution is treated as deterministic, while in \cite{Godsill2007a,Christensen2012} a conditionally linear-Gaussian state model is considered.

The posterior distribution for the changepoint MPP is inherently nonlinear, and cannot be calculated analytically. Instead, inference must be conducted using numerical approximations. The particle filter (introduced by \cite{Gordon1993}) and smoother (see \cite{Doucet2000a,Godsill2004}) are schemes which approximate a posterior distribution using a set of samples, or ``particles'', drawn sequentially from it.  A thorough introduction to particle filtering and smoothing methods can be found in \cite{Cappe2007,Doucet2009}. In \cite{Godsill2007a,Godsill2007,Whiteley2011}, the particle filter was adapted for use with variable rate models, resulting in the variable rate particle filter (VRPF).

The VRPF allows the changepoint sequence -- and hence the current state -- to be estimated sequentially as observations are received. However, estimates can often be improved later once further observations have been made. In this paper, we address the problem of smoothing in variable models, i.e. the estimation of the changepoint sequence and latent state given all the observations. This is achieved with an efficient backward sweep through the observations, in a similar manner to the method for standard HMMs described in \cite{Godsill2004}. Two new schemes are introduced: one for conditionally linear-Gaussian models which exploits the method of Rao-Blackwellisation; and a second for use with conditionally-deterministic models which uses an augmented target distribution in the style of an SMC sampler \cite{DelMoral2006}.

{\meta Add a bit about structure of paper.}

\section{Variable Rate Models} \label{sec:vr_models}

We consider a general model from time $0$ to $T$, between which observations, $\{y_1 \dots y_N\}$, are made at times $\{t_1 \dots t_N = T\}$. During this period, an unknown number of changepoints, $K$, occur at times $\{\tau_0 = 0, \tau_1 \dots \tau_K \}$, each with associated changepoint parameters, $\{ u_0, u_1 \dots u_K \}$. The pairs $\{\tau_k, u_k\}$ are the elements of an marked point process (MPP). We will refer to the times of the MPP as the ``change-time sequence'' and marks as ``parameter sequence'', and both together as the ``changepoint sequence''. The latent state is a continuous-time process denoted $x(t)$. Discrete sets containing multiple values over time will be written as, e.g. $y_{1:n} = \{y_1 \dots y_n\}$.

The objective for inference will be to estimate the changepoint sequence. This will be denoted as $\theta = \{\tau_{0:K}, u_{0:K}\}$. At a particular time $t_n$, the sequence will be divided into past $\theta_n = \{\tau_{j}, u_{j} \forall j : 0 \leq \tau_j < t_n \}$, and future $\theta_n^+ = \{\tau_{j}, u_{j} \forall j : t_n \leq \tau_j < T \}$. It will also be useful to define a variable for the changepoints which occur in the interval $[t_{n-1},t_n)$, $\theta_{n \setminus n-1} = \{\tau_{j}, u_{j} \forall j : t_{n-1} \leq \tau_j < t_n \}$.

 which has values $\{ x_0, x_1 \dots x_K \}$ at the changepoints. 
 Conditional upon a realisation of this process, $x(t)$ may be calculated deterministically.
   The cardinality of these sets (i.e. the number of changepoints occurring in the appropriate range) is denoted $K_n$, $K_n^+$ and $K_{n \setminus n-1}$ respectively.
%Note that $\theta_n$ not only conveys information about where changepoints occur, but also where they do not. i.e. it is implicit that $\tau_{K_n+1} > t_n$.

The changepoint sequence is assumed to be a Markov process.

\begin{IEEEeqnarray}{rCl}
 \{\tau_k, u_k\} & \sim & p(u_k|\tau_k, \tau_{k-1}, u_{k-1}) p(\tau_k|\tau_{k-1}, u_{k-1}) \label{eq:cp_model}
\end{IEEEeqnarray}

In the manner of \cite{Whiteley2011}, a survivor function is defined as the probability that no new changepoint occurs before a given time.

\begin{IEEEeqnarray}{rCl}
 S(\tau_k, u_k, t) &=& P(\tau_{k+1}>t|\tau_k, u_k) \nonumber \\
              &=& 1 - \int_{\tau_k}^{t} p(\xi|\tau_{k}, u_k) d\xi
\end{IEEEeqnarray}

It is now possible to write down a prior for the changepoint sequence, where we use the convention that $\tau_0 = 0$. The existence of such a density for a MPP is addressed in \cite{Jacobsen2006}.

\begin{IEEEeqnarray}{rCl}
p(\theta_n) & = & S(\tau_{K_n},t_n) p(u_0) \prod_{k=1}^{K_n} p(\tau_k, u_k| \tau_{k-1}, u_{k-1}) \label{eq:cp_sequence_prior}  .
\end{IEEEeqnarray}

Next we consider the state dynamics. A counting variable is introduced to keep track of the most recent changepoint to have occurred.

\begin{IEEEeqnarray}{rCl}
 \nu_t = \max(k : \tau_k<t)
\end{IEEEeqnarray}

Conditional upon the sequence of past changepoints, $\theta_n$, the state dynamics are given by a deterministic differential equation.

\begin{IEEEeqnarray}{rCl}
 dx(t) & = & \mathrm{f}(x(t), \tau_{\nu_t}, u_{\nu_t})
\end{IEEEeqnarray}

Assuming an analytic solution exists, a state transition function may be calculated.

\begin{IEEEeqnarray}{rCll}
 x(t) & = & f(x_{\nu_t}, v_{\nu_t}, \tau_{\nu_t}, t) &, \tau_{\nu_t} < t <= \tau_{\nu_t+1}    \label{eq:disc_time_state_diff_eq}
\end{IEEEeqnarray}

By choosing $t = \tau_{\nu_t+1}$, this equation specifies the state at the next changepoint time. Similarly, by choosing $t=t_n$, the state at the observation times is specified. These points will be denoted $\hat{x}_n$. Note than these points can be calculated deterministically given the initial state and past changepoint sequence.

\begin{IEEEeqnarray}{rCll}
 \hat{x}_n & = & f(x_0, \theta_n, t)   \label{eq:obs_time_state}
\end{IEEEeqnarray}

It is assumed for notational simplicity in the following sections that the initial state, $x_0$, is known a priori. This condition is easily relaxed.

In a target tracking example, $\tau_{1:K}$ might specify the times at which manoeuvres start, and $u_{1:K}$ the accelerations or other motion parameters associated with each manoeuvre. $x(t)$ would be the kinematic state of the target.

We consider a general model from time $0$ to $T$, between which observations, $\{y_1 \dots y_N\}$, are made at times $\{t_1 \dots t_N = T\}$. The linear state at these times is written as $\{x_1 \dots x_N\}$. During this period, an unknown number of changepoints, $K$, occur at times $\{ \tau_1 \dots \tau_K \}$, each with associated changepoint parameters, $\{ u_1 \dots u_K \}$. These are the marks of the MPP. Discrete sets containing multiple values over time will be written as, e.g. $y_{1:n} = \{y_1 \dots y_n\}$.

The objective for inference will be to estimate the sequence of changepoints. This will be denoted as $\theta = \{\tau_{1:K}, u_{1:K}\}$. In addition, at a particular time $t_n$, the sequence will be divided into past $\theta_n = \{\tau_{j}, u_{j} \forall j : 0 \leq \tau_j < t_n \}$, and future $\theta_n^+ = \{\tau_{j}, u_{j} \forall j : t_n \leq \tau_j < T \}$. It will also be useful to define a variable for the changepoints which occur in the interval $[t_{n-1},t_n)$, $\theta_{n \setminus n-1} = \{\tau_{j}, u_{j} \forall j : t_{n-1} \leq \tau_j < t_n \}$. The cardinality of these sets (i.e. the number of changepoints occurring in the appropriate range) is denoted $K_n$, $K_n^+$ and $K_{n \setminus n-1}$ respectively.

Note that $\theta_n$ not only conveys information about where changepoints occur, but also where they do not. i.e. it is implicit that $\tau_{K_n+1} > t_n$.%; furthermore that $\theta_n$ conveys less information than the two sets $\theta_{n-1}$ and $\theta_{n \setminus n-1}$ separately. The latter also impose the condition that $\tau_{K_{n-1}} < t_{n-1} < \tau_{K_{n-1}+1}$.

Assuming a Markovian changepoint sequence, a conditionally linear-Gaussian variable rate model can now be expressed by the following system equations:
%
\begin{IEEEeqnarray}{rCl}
 \{\tau_k, u_k\} & \sim & p(u_k|\tau_k, \tau_{k-1}, u_{k-1}) p(\tau_k|\tau_{k-1}) \label{eq:cp_model} \\
 x_n & = & A_n(\theta_{n})x_{n-1} + w_n \\
 y_n & = & C_n(\theta_{n})x_n + v_n  .
\end{IEEEeqnarray}

The random variables $w_n$ and $v_n$ have a zero-mean Gaussian distribution with covariance matrices $Q_n$ and $R_n$ respectively. The changepoint density will be constructed such that $P(\tau_k < \tau_{k-1}) = 0$. The factorisation of (\ref{eq:cp_model}) may be relaxed, but is assumed throughout as it simplifies the subsequent derivations.

It will be useful to define the survivor function in the same manner as \cite{Whiteley2011}.
%
\begin{IEEEeqnarray}{rCl}
 S(\tau_k, t) &=& P(\tau_{k+1}>t|\tau_k) \nonumber \\
              &=& 1 - \int_{\tau_k}^{t} p(\xi|\tau_{k}) d\xi
\end{IEEEeqnarray}

Using the convention that $\tau_0 = 0$ and $u_0$ is known a priori, the prior distribution over the changepoint sequence, $\theta_n$, may now be expressed as,
%
\begin{IEEEeqnarray}{rCl}
p(\theta_n) & = & S(\tau_{K_n},t_n) \prod_{k=1}^{K_n} p(\tau_k, u_k| \tau_{k-1}, u_{k-1}) \label{eq:cp_sequence_prior}  .
\end{IEEEeqnarray}

See \cite{Jacobsen2006} for a rigorous derivation of MPP probability measures.

The tasks of filtering and smoothing may now be considered as estimation of the posterior distributions of $\theta$ and $x_{1:N}$.

\end{document}

Target tracking algorithms are commonly based upon fixed rate models (see, e.g. \cite{Li2003} for a thorough survey), in which the target kinematics (position, velocity, etc.) are estimated at a set of fixed times at which observations (e.g. radar measurements) are made. In \cite{Godsill2007a,Godsill2007}, variable rate models were introduced for tracking, in which the state trajectory is divided up by a set of changepoints between which the motion follows a deterministic path governed by motion parameters (accelerations, etc.) which are fixed for that division. The inference problem is hence reduced to estimation of the changepoint times and associated motion parameters. As noted in \cite{Whiteley2011}, these comprise a marked point process (MPP), the mathematical properties of which are thoroughly set out in \cite{Jacobsen2006}.

We introduce piecewise-deterministic variable rate models in section~\ref{sec:pd_vr_models}, and review the VRPF in section~\ref{sec:pd_vrpf}. The new variable rate particle smoother (VRPS) algorithm is described in section~\ref{sec:pd_vrps}. In section~\ref{sec:tracking_models}, we adapt the intrinsic-coordinate models of \cite{Godsill2007a,Godsill2007} for use with the new smoothing algorithm. Simulations are presented in section~\ref{sec:simuations}.

The posterior distribution of the changepoint sequence is analytically intractable, but may be estimated using a particle filter. When the conditional state dynamics are linear-Gaussian, the efficiency of this filter may be greatly improved by using the method of Rao-Blackwellisation (see, e.g. \cite{Casella1996,Doucet2000}). Rather than targeting the entire posterior distribution over changepoints and states, the particle filter is used to estimate only the nonlinear changepoint sequence, after which Kalman filtering and smoothing may be used to infer the linear states. This is the Rao-Blackwellised variable rate particle filter (RBVRPF) of \cite{Godsill2007a,Christensen2012}.

In this paper, a new algorithm is described for use with conditionally linear-Gaussian variable rate models for estimation of the smoothing distribution, i.e. the distribution over the sequence of changepoints and linear states given all the observations. This uses a similar derivation to that of the fixed rate Rao-Blackwellised particle smoother of \cite{Sarkka2012}. The new algorithm is called the Rao-Blackwellised variable rate particle smoother (RBVRPS).

We review the structure of conditionally linear-Gaussian variable rate models in section~\ref{sec:rbvr_models} and revise the RBVRPF in section~\ref{sec:rbvrpf}. The new smoothing algorithm is presented in section~\ref{sec:rbvrps} with supporting simulations in section~\ref{sec:simulations}.

### 2012-03-05 draft1_VRsmoothing.tex
merged filtering and rb smoothing

@@
 \usepackage{mdwtab}
-\usepackage{eqparbox}
+%\usepackage{eqparbox}
 \usepackage[caption=false,font=footnotesize]{subfig}
@@
 
-{\meta Add a bit about structure of paper.}
+We introduce the general structure of variable rate models in section~\ref{sec:vr_models} and discuss two particular types, those with conditionally linear-Gaussian and conditionally deterministic dynamics.
+The VRPF is reviewed in section~\ref{sec:vrpf}, and new variable rate smoothing algorithms are described in section~\ref{sec:vrps}.
+In section~\ref{sec:simuations}, particular examples of variable rate models are presented and their performance demonstrated in a series of simulations.
 
@@
 
- which has values $\{ x_0, x_1 \dots x_K \}$ at the changepoints. 
- Conditional upon a realisation of this process, $x(t)$ may be calculated deterministically.
-   The cardinality of these sets (i.e. the number of changepoints occurring in the appropriate range) is denoted $K_n$, $K_n^+$ and $K_{n \setminus n-1}$ respectively.
-%Note that $\theta_n$ not only conveys information about where changepoints occur, but also where they do not. i.e. it is implicit that $\tau_{K_n+1} > t_n$.
+For notational simplicity, the following counting variables are introduced to keep track of the most recent changepoint to have occurred,
+%
+\begin{IEEEeqnarray}{rCl}
+ K(t)  & = & \max(k : \tau_k<t) \\
+ K_n   & = & K(t_n)     .
+\end{IEEEeqnarray}
 
 The changepoint sequence is assumed to be a Markov process.
-
+%
 \begin{IEEEeqnarray}{rCl}
@@
 
-In the manner of \cite{Whiteley2011}, a survivor function is defined as the probability that no new changepoint occurs before a given time.
+The changepoint density will be constructed such that $P(\tau_k < \tau_{k-1}) = 0$.
 
+In the manner of \cite{Whiteley2011}, a survivor function is defined as the probability that no new changepoint occurs before a given time,
+%
 \begin{IEEEeqnarray}{rCl}
  S(\tau_k, u_k, t) &=& P(\tau_{k+1}>t|\tau_k, u_k) \nonumber \\
-              &=& 1 - \int_{\tau_k}^{t} p(\xi|\tau_{k}, u_k) d\xi
+              &=& 1 - \int_{\tau_k}^{t} p(\xi|\tau_{k}, u_k) d\xi     .
 \end{IEEEeqnarray}
@@
 \begin{IEEEeqnarray}{rCl}
-p(\theta_n) & = & S(\tau_{K_n},t_n) p(u_0) \prod_{k=1}^{K_n} p(\tau_k, u_k| \tau_{k-1}, u_{k-1}) \label{eq:cp_sequence_prior}  .
+p(\theta_n) & = & S(\tau_{K_n},t_n) p(u_0) \prod_{k=1}^{K_n} p(\tau_k, u_k| \tau_{k-1}, u_{k-1}) \label{eq:cp_sequence_prior}
 \end{IEEEeqnarray}
 
-Next we consider the state dynamics. A counting variable is introduced to keep track of the most recent changepoint to have occurred.
+\subsection{Conditionally Linear-Gaussian Models}
 
+The first class of variable rate models to be considered is those whose state dynamics are linear-Gaussian conditional on the changepoint sequence.
+Such a model may be discretised onto the set of observation times in exactly the same manner as a standard the standard HMM.
+%
 \begin{IEEEeqnarray}{rCl}
- \nu_t = \max(k : \tau_k<t)
+ x_n & = & A_n(\theta_{n})x_{n-1} + w_n \\
+ y_n & = & C_n(\theta_{n})x_n + v_n       ,
 \end{IEEEeqnarray}
 
-Conditional upon the sequence of past changepoints, $\theta_n$, the state dynamics are given by a deterministic differential equation.
-
+where,
+%
 \begin{IEEEeqnarray}{rCl}
- dx(t) & = & \mathrm{f}(x(t), \tau_{\nu_t}, u_{\nu_t})
+ w_n & \sim & \mathcal{N}(w_n|0, Q_n(\theta_{n})) \\
+ v_n & \sim & \mathcal{N}(v_n|0, R_n(\theta_{n}))       .
 \end{IEEEeqnarray}
 
-Assuming an analytic solution exists, a state transition function may be calculated.
-
-\begin{IEEEeqnarray}{rCll}
- x(t) & = & f(x_{\nu_t}, v_{\nu_t}, \tau_{\nu_t}, t) &, \tau_{\nu_t} < t <= \tau_{\nu_t+1}    \label{eq:disc_time_state_diff_eq}
+In addition, the prior state distribution should be Gaussian, with known mean and variance.
+%
+\begin{IEEEeqnarray}{rCl}
+ x_0 & \sim & \mathcal{N}(x_0|m_0, P_0)       .
 \end{IEEEeqnarray}
 
-By choosing $t = \tau_{\nu_t+1}$, this equation specifies the state at the next changepoint time.
-Similarly, by choosing $t=t_n$, the state at the observation times is specified.
-These points will be denoted $\hat{x}_n$.
-Note than these points can be calculated deterministically given the initial state and past changepoint sequence.
+If the changepoint sequence is known, or has been estimated, then the state values, $x_n$ may be inferred using optimal Kalman filtering and smoothing recursions.
+As well as the basic Kalman filter \cite{Kalman1960}, the Rauch-Tung-Striebel (RTS) smoother \cite{Rauch1965} and two-filter smoother \cite{Fraser1969} will prove useful for this step.
 
-\begin{IEEEeqnarray}{rCll}
- \hat{x}_n & = & f(x_0, \theta_n, t)   \label{eq:obs_time_state}
+Conditionally linear-Gaussian variable rate models were introduced in \cite{Godsill2007a} for a financial inference algorithm.
+Changepoints correspond to jumps in the value or trend of a security, at which points the process covariance, $Q_n(\theta_n)$, is inflated.
+
+\subsection{Conditionally Deterministic Models}
+
+The second class of variable rate models for consideration is those in which the state is completely specified by the changepoint sequence, with no additional random components.
+Such a process is commonly referred to as ``piecewise-deterministic'', as the latent state follows a deterministic path between changepoints.
+In this case, it is not necessary to discretise the state -- it may be kept as a continuous variable.
+
+In general, the state dynamics will be governed by a differential equation which may depend on the entire changepoint sequence. Here we assume that only the most recent changepoint is significant.
+%
+\begin{IEEEeqnarray}{rCl}
+ dx(t) & = & \mathrm{f}(x(t), \tau_{K(t)}, u_{K(t)})     .
 \end{IEEEeqnarray}
 
-It is assumed for notational simplicity in the following sections that the initial state, $x_0$, is known a priori. This condition is easily relaxed.
-
-In a target tracking example, $\tau_{1:K}$ might specify the times at which manoeuvres start, and $u_{1:K}$ the accelerations or other motion parameters associated with each manoeuvre.
-$x(t)$ would be the kinematic state of the target.
-
-We consider a general model from time $0$ to $T$, between which observations, $\{y_1 \dots y_N\}$, are made at times $\{t_1 \dots t_N = T\}$.
-The linear state at these times is written as $\{x_1 \dots x_N\}$.
-During this period, an unknown number of changepoints, $K$, occur at times $\{ \tau_1 \dots \tau_K \}$, each with associated changepoint parameters, $\{ u_1 \dots u_K \}$.
-These are the marks of the MPP.
-Discrete sets containing multiple values over time will be written as, e.g.
-$y_{1:n} = \{y_1 \dots y_n\}$.
-
-The objective for inference will be to estimate the sequence of changepoints.
-This will be denoted as $\theta = \{\tau_{1:K}, u_{1:K}\}$.
-In addition, at a particular time $t_n$, the sequence will be divided into past $\theta_n = \{\tau_{j}, u_{j} \forall j : 0 \leq \tau_j < t_n \}$, and future $\theta_n^+ = \{\tau_{j}, u_{j} \forall j : t_n \leq \tau_j < T \}$.
-It will also be useful to define a variable for the changepoints which occur in the interval $[t_{n-1},t_n)$, $\theta_{n \setminus n-1} = \{\tau_{j}, u_{j} \forall j : t_{n-1} \leq \tau_j < t_n \}$.
-The cardinality of these sets (i.e.
-the number of changepoints occurring in the appropriate range) is denoted $K_n$, $K_n^+$ and $K_{n \setminus n-1}$ respectively.
-
-Note that $\theta_n$ not only conveys information about where changepoints occur, but also where they do not.
-i.e.
-it is implicit that $\tau_{K_n+1} > t_n$.%; furthermore that $\theta_n$ conveys less information than the two sets $\theta_{n-1}$ and $\theta_{n \setminus n-1}$ separately.
-The latter also impose the condition that $\tau_{K_{n-1}} < t_{n-1} < \tau_{K_{n-1}+1}$.
-
-Assuming a Markovian changepoint sequence, a conditionally linear-Gaussian variable rate model can now be expressed by the following system equations:
+By introducing a new sequence, $\{ x_0, x_1 \dots x_K \}$, which denotes the value of the state at each changepoint (i.e.
+$x(\tau_k)$), and assuming that an analytic solution exists, a state transition function may be found,
 %
-\begin{IEEEeqnarray}{rCl}
- \{\tau_k, u_k\} & \sim & p(u_k|\tau_k, \tau_{k-1}, u_{k-1}) p(\tau_k|\tau_{k-1}) \label{eq:cp_model} \\
- x_n & = & A_n(\theta_{n})x_{n-1} + w_n \\
- y_n & = & C_n(\theta_{n})x_n + v_n  .
+\begin{IEEEeqnarray}{rCll}
+ x(t) & = & f(x_{K_n}, v_{K_n}, \tau_{K_n}, t) &, \tau_{K_n} < t <= \tau_{K_{n}+1}    \label{eq:disc_time_state_diff_eq}     .
 \end{IEEEeqnarray}
 
-The random variables $w_n$ and $v_n$ have a zero-mean Gaussian distribution with covariance matrices $Q_n$ and $R_n$ respectively.
-The changepoint density will be constructed such that $P(\tau_k < \tau_{k-1}) = 0$.
-The factorisation of (\ref{eq:cp_model}) may be relaxed, but is assumed throughout as it simplifies the subsequent derivations.
+By choosing $t = \tau_{K_{n}+1}$, this equation specifies the state at the next changepoint time.
+Similarly, by choosing $t=t_n$, the state at the observation times may be calculated.
+These points will be denoted $\hat{x}_n$.
+To complete the model, a probabilistic measurement model must be chosen for the observation process, $p(y_n|\hat{x}_n)$.
 
-It will be useful to define the survivor function in the same manner as \cite{Whiteley2011}.
-%
-\begin{IEEEeqnarray}{rCl}
- S(\tau_k, t) &=& P(\tau_{k+1}>t|\tau_k) \nonumber \\
-              &=& 1 - \int_{\tau_k}^{t} p(\xi|\tau_{k}) d\xi
-\end{IEEEeqnarray}
-
-Using the convention that $\tau_0 = 0$ and $u_0$ is known a priori, the prior distribution over the changepoint sequence, $\theta_n$, may now be expressed as,
-%
-\begin{IEEEeqnarray}{rCl}
-p(\theta_n) & = & S(\tau_{K_n},t_n) \prod_{k=1}^{K_n} p(\tau_k, u_k| \tau_{k-1}, u_{k-1}) \label{eq:cp_sequence_prior}  .
-\end{IEEEeqnarray}
-
-See \cite{Jacobsen2006} for a rigorous derivation of MPP probability measures.
-
-The tasks of filtering and smoothing may now be considered as estimation of the posterior distributions of $\theta$ and $x_{1:N}$.
-
-\end{document}
+For convenience, we assume that $x_0$ is known in the following sections.
+This means that $x(t)$ may be calculated deterministically for all $t$ given $\theta$.
+However, this condition is easily relaxed by including $x_0$ as a random variable in the posterior distribution.
 
@@
 In \cite{Godsill2007a,Godsill2007}, variable rate models were introduced for tracking, in which the state trajectory is divided up by a set of changepoints between which the motion follows a deterministic path governed by motion parameters (accelerations, etc.) which are fixed for that division.
-The inference problem is hence reduced to estimation of the changepoint times and associated motion parameters.
-As noted in \cite{Whiteley2011}, these comprise a marked point process (MPP), the mathematical properties of which are thoroughly set out in \cite{Jacobsen2006}.
 
-We introduce piecewise-deterministic variable rate models in section~\ref{sec:pd_vr_models}, and review the VRPF in section~\ref{sec:pd_vrpf}.
-The new variable rate particle smoother (VRPS) algorithm is described in section~\ref{sec:pd_vrps}.
-In section~\ref{sec:tracking_models}, we adapt the intrinsic-coordinate models of \cite{Godsill2007a,Godsill2007} for use with the new smoothing algorithm.
-Simulations are presented in section~\ref{sec:simuations}.
+\section{The Variable Rate Particle Filter} \label{sec:vrpf}
 
-The posterior distribution of the changepoint sequence is analytically intractable, but may be estimated using a particle filter.
-When the conditional state dynamics are linear-Gaussian, the efficiency of this filter may be greatly improved by using the method of Rao-Blackwellisation (see, e.g.
-\cite{Casella1996,Doucet2000}).
-Rather than targeting the entire posterior distribution over changepoints and states, the particle filter is used to estimate only the nonlinear changepoint sequence, after which Kalman filtering and smoothing may be used to infer the linear states.
-This is the Rao-Blackwellised variable rate particle filter (RBVRPF) of \cite{Godsill2007a,Christensen2012}.
+The variable rate particle filter (VRPF) is described in \cite{Godsill2007,Godsill2007a}.
+The objective of the algorithm is to sequentially estimate the posterior distribution of the changepoint sequence, $p(\theta_{n}| y_{1:n})$, at each time $t_n$.
+This distribution may be expanded using Bayes' rule.
+%
+\begin{IEEEeqnarray}{rCl}
+\IEEEeqnarraymulticol{3}{l}{p(\theta_{n}|y_{1:n})} \nonumber \\
+ & \propto & p(y_n|\theta_{n}, y_{1:n-1}) p(\theta_{n \setminus n-1}|\theta_{n-1}) p(\theta_{n-1}|y_{1:n-1}) \label{eq:vrpf_target}
+\end{IEEEeqnarray}
 
-In this paper, a new algorithm is described for use with conditionally linear-Gaussian variable rate models for estimation of the smoothing distribution, i.e.
-the distribution over the sequence of changepoints and linear states given all the observations.
-This uses a similar derivation to that of the fixed rate Rao-Blackwellised particle smoother of \cite{Sarkka2012}.
-The new algorithm is called the Rao-Blackwellised variable rate particle smoother (RBVRPS).
+The transition term, $p(\theta_{n \setminus n-1} | \theta_{n-1})$, can be constructed in a similar manner to (\ref{eq:cp_sequence_prior}) \cite{Jacobsen2006}, as a product of density terms for each new changepoint and a survivor function term.
+In the particular (but common) case that no new changepoints occur within the interval, the density consists of a probability mass on the empty set, with weight $P(\tau_{K_n+1} > t_n | \tau_{K_n+1} > t_{n-1}, \tau_{K_n} )$.
+%
+\begin{IEEEeqnarray}{rCl}
+\IEEEeqnarraymulticol{3}{l}{p(\theta_{n \setminus n-1} | \theta_{n-1})} \nonumber \\
+    & = & \begin{cases} S(\tau_{K_n}, t_n) \prod\limits_{j:t_{n-1} \leq \tau_j < t_n} p(\tau_j, u_j| \tau_{j-1}, u_{j-1}, \tau_j>t_{n-1}) & K_n > K_{n-1} \\
+                        S(\tau_{K_n}, t_n) / S(\tau_{K_n}, t_{n-1}) & K_n = K_{n-1} \end{cases} \IEEEeqnarraynumspace \label{eq:cp_sequence_trandens}
+\end{IEEEeqnarray}
 
-We review the structure of conditionally linear-Gaussian variable rate models in section~\ref{sec:rbvr_models} and revise the RBVRPF in section~\ref{sec:rbvrpf}.
-The new smoothing algorithm is presented in section~\ref{sec:rbvrps} with supporting simulations in section~\ref{sec:simulations}.
+For all but the first changepoint in the interval, the density is given by the prior model of (\ref{eq:cp_model}).
+For the first changepoint, indexed by $k=K_{n-1}+1$, we must account for the fact that a changepoint cannot occur before $t_{n-1}$,
+%
+\begin{IEEEeqnarray}{rCl}
+\IEEEeqnarraymulticol{3}{l}{p(\tau_{k}, u_{k}| \tau_{k-1}, u_{k-1}, \tau_{k}>t_{n-1})} \nonumber \\
+  & = & \frac{1}{S(\tau_{k-1}, t_{n-1})} \begin{cases} p(\tau_{k}, u_{k}| \tau_{k-1}, u_{k-1}) & \tau_{k} > t_{n-1} \\ 0 & \tau_{k} < t_{n-1} \end{cases}  \IEEEeqnarraynumspace \label{eq:cp_cond_model}   .
+\end{IEEEeqnarray}
+
+Practically, because changepoints will be relatively rare events, it is not likely that more than one new changepoint will occur between $t_{n-1}$ and $t_n$.
+
+The target distribution of (\ref{eq:vrpf_target}) cannot be calculated analytically, but may be approximated numerically.
+A particle filter is an algorithm for approximating a probability distribution using a set of weighted samples (or ``particles'') drawn from that distribution.
+In this case, each particle will be a set of changepoint times and parameters.
+%
+\begin{equation}
+ \hat{p}(\theta_{n}|y_{1:n}) = \sum_j w_n^{(j)} \delta_{\theta_{n}^{(j)}}(\theta_{n})
+\end{equation}
+
+where $\delta_x(X)$ is a dirac probability mass at $X=x$.
+
+The particle filter works recursively.
+At the $n$th step, a particle, $\theta_{n-1}^{(i)}$, is first resampled from those approximating the filtering distribution at the $(n-1)$th step, using an appropriately chosen set of proposal weights.
+%
+\begin{equation}
+ q(\theta_{n-1}) = \sum_j v_{n-1}^{(j)} \delta_{\theta_{n-1}^{(j)}}(\theta_{n-1})
+\end{equation}
+
+The choice of weights determines the type of resampling used.
+The simplest choice, $v_{n-1}^{(j)} = 1/N_F$ (where $N_F$ is the number of filter particles) may be achieved by simply omitting this step all together and using the particles of $\hat{p}(\theta_{n-1}|y_{1:n-1})$.
+This, however, leads to degeneracy of the particle weights over time.
+Conventional resampling is achieved by using $v_{n-1}^{(j)} = w_{n-1}^{(j)}$.
+Any other choice results in an auxiliary particle filter \cite{Pitt1999}.
+For further discussion of resampling, see \cite{Cappe2007,Doucet2009,Douc2005}.
+
+Next, an extension to the changepoint sequence, $\theta_{n \setminus n-1}^{(i)}$, is proposed from an importance distribution, $q(\theta_{n \setminus n-1}|\theta_{n-1}^{(i)}, y_n)$, and concatenated to $\theta_{n-1}$ to create an estimate of $\theta_n$.
+Finally, the particle is weighted according to the ratio of the target and proposal densities.
+%
+\begin{IEEEeqnarray}{rCl}
+w_n^{(i)} & = & \frac{ p(\theta_{n}^{(i)}|y_{1:n}) }{ q(\theta_{n}) } \nonumber \\
+    & \propto & \frac{ p(y_n|\theta_n, y_{1:n-1}) p(\theta_{n \setminus n-1}|\theta_{n-1}) p(\theta_{n-1}|y_{1:n-1}) }{ q(\theta_{n-1}) q(\theta_{n \setminus n-1}|\theta_{n-1}^{(i)}, y_n) } \nonumber \\
+    & =       & \frac{w_{n-1}^{(i)}}{v_{n-1}^{(i)}} \times \frac{ p(y_n|\theta_n, y_{1:n-1}) p(\theta_{n \setminus n-1}|\theta_{n-1}) }{ q(\theta_{n \setminus n-1}|\theta_{n-1}^{(i)}, y_n) } \label{eq:vrpf_weights}
+\end{IEEEeqnarray}
+
+The normalisation may be enforced by scaling the weights so that they sum to $1$.
+
+For the most basic ``bootstrap'' \cite{Gordon1993} form of the VRPF, $\theta_{n \setminus n-1}$ may be proposed from the prior transition density (\ref{eq:cp_sequence_trandens}).
+This can be achieved by sampling new changepoints sequentially from the transition model (\ref{eq:cp_model}) (apart from the first which is sampled from (\ref{eq:cp_cond_model})) until one falls after the current time, $t_n$.
+This final future changepoint is discarded.
+(This process can be thought of as sampling the entire future changepoint sequence from $t_{n-1}$ onwards, and then marginalising those which fall after $t_n$.) The bootstrap proposal leads to the usual simplification of the weight formula.
+%
+\begin{IEEEeqnarray}{rCl}
+w_n^{(i)} & = & \frac{w_{n-1}^{(i)}}{v_{n-1}^{(i)}} \times p(y_n|\hat{x}_n) \label{eq:bootstrap_vrpf_weights}
+\end{IEEEeqnarray}
+
+The choice of proposal weights, $\{v_{n-1}^{(i)}\}$, requires particular attention in the design of VRPFs.
+In some models a changepoint may not have an immediate effect on the observations, especially if a jump occurs in some quantity which is only observed via its integral, e.g.
+if there is a jump in the acceleration of a moving object, yet only the position is measured, the change will not be apparent until several more observations have been made.
+In the meantime, particles which contain a changepoint at the correct time may all have been removed by the resampling process.
+To avoid this loss of good particles, proposal weights should be chosen which preserve a significant number of low-weight particles.
+One scheme which has been found to work well is described in \cite{Godsill2007}, in which proposal weights are given by:
+%
+\begin{IEEEeqnarray}{rCl}
+v_{n-1}^{(i)} & \propto & \max ( 1, N_F w_{n-1}^{(i)} )
+\end{IEEEeqnarray}
+
+It only remains to consider the likelihood term required for evaluation of the importance weights, $p(y_n|\theta_n, y_{1:n-1})$.
+The form of this term depends on the particular model under consideration.
+In the following sections, the likelihood expressions for the conditionally linear-Gaussian and deterministic cases are considered.
+
+\subsection{Conditionally Linear-Gaussian Likelihoods} \label{sec:rb-vrpf}
+
+For a conditionally linear-Gaussian model, the required likelihood term, $p(y_n|\theta_n, y_{1:n-1})$ is the predictive distribution estimated by the Kalman filter.
+Conveniently, the Kalman filter also provides us with an estimate of the current state given the changepoint sequence and all the preceding observations.
+It follows from the the Gaussian dynamics and prior that these distributions are all Gaussian as well, \cite{Anderson1979},
+%
+\begin{IEEEeqnarray}{rCl}
+ p(x_n|\theta_{n}, y_{1:n-1}) & = & \mathcal{N}(x_n|m_n^-, P_n^-) \\
+ p(x_n|\theta_{n}, y_{1:n}) & = & \mathcal{N}(x_n|m_n, P_n) \\
+ p(y_n|\theta_{n}, y_{1:n-1}) & = & \mathcal{N}(y_n|\mu_n, S_n)     ,
+\end{IEEEeqnarray}
+
+with means and variances given by the following standard recursions (dependence on $\theta_{n}$ suppressed for clarity),
+%
+\begin{IEEEeqnarray}{rCl}
+ m_n^- & = & A_n m_{n-1} \label{eq:kf_predict_start} \\
+ P_n^- & = & A_n P_{n-1} A_n^T + Q_n \\
+ \mu_n & = & C_n m_n^- \\
+ S_n   & = & C_n P_n^- C_n^T + R_n \label{eq:kf_predict_stop} \\
+ K_n   & = & P_n^- C_n^T S_n^{-1} \label{eq:kf_update_start}\\
+ m_n   & = & m_n^- + K_n (y_n - \mu_n) \\
+ P_n   & = & P_n^- - K_n S_n K_n^T \label{eq:kf_update_stop}    .
+\end{IEEEeqnarray}
+
+This completes the requirements for the particle filter, resulting in the final algorithm shown in Fig.~\ref{alg:RBVRPF}.
+The concept of estimating the posterior for just the changepoint sequence with a particle filter rather than the joint posterior over changepoints and linear states is an example of Rao-Blackwellisation (see, e.g.
+\cite{Casella1996,Doucet2000}), so this algorithm is known as the Rao-Blackwellised variable rate particle filter (RB-VRPF) \cite{Godsill2007a,Christensen2012}.
+
+\begin{figure}
+\fbox{\parbox{\columnwidth}{
+\begin{algorithmic}[1]
+\STATE For each $i$, initialise particle changepoint sequence with $\{\theta_{0}^{(i)}\} \gets \{\tau_0^{(i)}, u_0^{(i)}$, where $\tau_0^{(i)}=0$ and $\{u_0^{(i)}\} \sim p(u_0)$.
+\STATE For each $i$, initialise particle sufficient statistics, $m_0^{(i)}$ and $P_0^{(i)}$ with prior values.
+\FOR{$n=1 \dots N$}
+  \FOR{$i=1 \dots N_F$}
+  	\STATE Sample a changepoint history $\theta_{n-1}^{(i)} \sim \sum_j v_{n-1}^{(j)} \delta_{\theta_{n-1}^{(j)}}(\theta_{n-1})$.
+    \STATE Propose a sequence extension $\theta_{n \setminus n-1}^{(i)} \sim q(\theta_{n \setminus n-1} | \theta_{n-1}^{(i)})$.
+    \STATE Add extension to changepoint sequence $\theta_n^{(i)} \gets \theta_{n-1}^{(i)} \cup \theta_{n \setminus n-1}^{(i)}$.
+    \STATE Predict state mean and covariance $\mu_n^{(i)}$ and $S_n^{(i)}$ using (\ref{eq:kf_predict_start})~to~(\ref{eq:kf_predict_stop}).
+    \STATE Calculate weight $w_n^{(i)}$ using (\ref{eq:vrpf_weights}).
+    \STATE Update state mean and covariance $m_n^{(i)}$ and $P_n^{(i)}$ using (\ref{eq:kf_update_start})~to~(\ref{eq:kf_update_stop}).
+  \ENDFOR
+  \STATE Scale weights such that $\sum_i w_n^{(i)}=1$.
+\ENDFOR
+\end{algorithmic}
+}}
+\label{alg:RBVRPF}
+\caption{Rao-Blackwellised variable rate particle filter}
+\end{figure}
+
+\subsection{Conditionally Deterministic Likelihoods} \label{sec:pd-vrpf}
+
+When a conditionally deterministic model is used, the state at observation time $t_n$ is specified by the changepoint sequence $\theta_n$ (plus the initial state, $x_0$), using (\ref{eq:disc_time_state_diff_eq}).
+Thus, the required likelihood term is simply given by,
+%
+\begin{IEEEeqnarray}{rCl}
+ p(y_n|\theta_{n}, y_{1:n-1}) & = & p(y_n|\hat{x}_n)     .
+\end{IEEEeqnarray}
+
+This leads to a particle filter for variable rate models with conditionally deterministic dynamics, summarised in Fig.~\ref{alg:VRPF}.
+The resulting trajectories $x(t)^{(i)}$ for each particle are realisations from a piecewise-deterministic process, so this algorithm is named the piecewise-deterministic variable rate particle filter (PD-VRPF).
+
+\begin{figure}
+\fbox{\parbox{\columnwidth}{
+\begin{algorithmic}[1]
+\STATE For each $i$, initialise particle changepoint sequence with $\{\theta_{0}^{(i)}\} \gets \{\tau_0^{(i)}, u_0^{(i)}$, where $\tau_0^{(i)}=0$ and $\{u_0^{(i)}\} \sim p(u_0)$.
+\FOR{$n=1 \dots N$}
+  \FOR{$i=1 \dots N_F$}
+  	\STATE Sample a history $\theta_{n-1}^{(i)} \sim \sum_j v_{n-1}^{(j)} \delta_{\theta_{n-1}^{(j)}}(\theta_{n-1})$.
+    \STATE Propose a sequence extension $\theta_{n \setminus n-1}^{(i)} \sim q(\theta_{n \setminus n-1}|\theta_{n-1}^{(i)}, y_{n})$.
+    \STATE Add extension to changepoint sequence $\theta_n^{(i)} \gets \theta_{n-1}^{(i)} \cup \theta_{n \setminus n-1}^{(i)}$.
+    \STATE Calculate state $\hat{x}_n$ using (\ref{eq:disc_time_state_diff_eq}).
+    \STATE Calculate particle weight using (\ref{eq:vrpf_weights}).
+  \ENDFOR
+  \STATE Scale weights such that $\sum_i w_n^{(i)}=1$.
+\ENDFOR
+\end{algorithmic}
+}}
+\label{alg:VRPF}
+\caption{Piecewise deterministic variable rate particle filter}
+\end{figure}
+
+\subsection{Improving the Variable Rate Particle Filter}
+
+The bootstrap versions of the VRPFs described in sections~\ref{sec:rb-vrpf} and~\ref{sec:pd-vrpf} may perform poorly if changepoints are not obvious until significantly after they occur.
+For example, in a tracking example, if a jump occurs in the acceleration, but only the position is observed, then this change may not be obvious until a number of observations have arrived.
+In this case, the estimation may be improved by the introduction of resample-move (RM) steps \cite{Gilks2001}.
+In an RM scheme, optional Metropolis-Hastings (MH) moves are conducted to alter the particle states after the importance sampling has taken place.
+For variable rate models, any one of the previous changepoints, $\tau_k$, or associated parameters, $u_k$, could be adjusted.
+Because more observations are available than when the changepoint was first proposed, it may be possible to construct more informed proposals and so move the changepoints towards regions with higher posterior probability.
+It is even possible to retrospectively add or remove changepoints, using reversible jump MH moves \cite{Green1995}.
+Variable rate particle filters using RM with piecewise deterministic models are described in \cite{Whiteley2011,Gilholm2008}.
+
+Rather than conducting the IS and MH steps separately, it is possible to combine them using the framework of SMC samplers \cite{DelMoral2006}.
+This was suggested in \cite{Whiteley2011}, again for piecewise deterministic dynamics, but the extension to conditionally linear-Gaussian models is straightforward.
+
+Filtering schemes which alter past changepoints -- whether using an SMC sampler or RM -- are computationally expensive, because many likelihood calculations must be conducted, for each observation from the time of the change onwards.
+In some cases, it may be simpler to just use a bootstrap filter with more particles.
+
+\section{Variable Rate Particle Smoothing} \label{sec:vrps}
+
+A filter conducts inference sequentially as new observations are introduced.
+The purpose of a smoother is to produce a second estimate once all the observations have been made, using future values to improve upon the filter performance.
+Estimating changepoints online is a challenging task because the presence of a change may not be obvious until after it has happened.
+Thus, it is anticipated that a smoothing algorithm will provide significantly improved performance at changepoint estimation.
+
+The target distribution for a variable rate smoothing algorithm is the posterior distribution over the entire changepoint sequence, $p(\theta|y_{1:N})$.
+A particle approximation to this distribution is generated by the final step of the VRPF.
+However, in the same manner as the fixed rate filter-smoother of \cite{Kitagawa1996}, this approximation is likely to lack path-space diversity -- because of the necessary resampling step in the filtering algorithm, the particles all share the same set of changepoints before a particular time, with variation only appearing for changepoints closer to $T$.
+For a good characterisation of the smoothing distribution, it is necessary to rejuvenate the set of particles.
+This is achieved with a backward pass through the observations in a similar manner to the forward-backward method described in \cite{Godsill2004}.
+
+The target distribution may be factorised as follows,
+%
+\begin{IEEEeqnarray}{rCl}
+ p(\theta|y_{1:N}) = p(\theta_{n}^{+}|y_{1:N}) p(\theta_{n}|\theta_{n}^{+}, y_{1:N})     .
+\end{IEEEeqnarray}
+
+This suggests a sequential strategy for smoothing.
+A particle drawn from $p(\theta_{n}^{+}|y_{1:N})$ may be extended backwards by sampling from the backwards conditional distribution, $p(\theta_{n}|\theta_{n}^{+}, y_{1:N})$ and concatenating the two partial sequences.
+The new particle will be a sample from the target distribution, but the resulting approximation will still suffer from low path-space diversity before time $t_{n-1}$.
+Therefore, $\theta_{n-1}$ is marginalised by simply discarding the changepoints which occur before this time.
+This yields in a particle drawn from $p(\theta_{n-1}^{+}|y_{1:N})$.
+The procedure then continues recursively.
+After a complete backwards pass through the observations, a single particle from the target distribution will have been generated.
+The process is then repeated until sufficient samples are obtained.
+
+At step $n$ for a given particle, the future changepoint sequence may be considered to be fixed, and is denoted $\tilde{\theta}_n^+$.
+It remains to devise a method for drawing samples from the backwards conditional distribution, $p(\theta_{n}|\theta_{n}^{+}, y_{1:N})$.
+Such methods are discussed in the following sections for the two classes of models.
+
+\subsection{Conditionally Linear-Gaussian Smoothing} \label{sec:rb-vrps}
+
+When the state dynamics are linear-Gaussian conditional on the changepoint sequence, it is possible split the dependence on the observations into past and future by introducing the current state as an additional variable.
+This trick was used in \cite{Sarkka2012} in the derivation of the Rao-Blackwellised particle filter for fixed rate conditionally linear-Gaussian models.
+%
+\begin{IEEEeqnarray}{rCl}
+ \IEEEeqnarraymulticol{3}{l}{p(\theta_{n}|\tilde{\theta}_{n}^+, y_{1:N})} \nonumber \\
+\qquad & \propto & p(\theta_{n}, \tilde{\theta}_{n}^+| y_{1:N}) \nonumber  \\
+       & =       & \int p(x_n, \theta_{n}, \tilde{\theta}_{n}^+| y_{1:N}) dx_n \nonumber  \\
+       & \propto & \int p(y_{n+1:N}|x_n, \theta_{n}, \tilde{\theta}_{n}^+, y_{1:n}) p(x_n, \theta_{n}, \tilde{\theta}_{n}^+| y_{1:n}) dx_n \nonumber \\
+       & = & \int p(y_{n+1:N}|x_n, \tilde{\theta}_{n}^+) p(x_n|\theta_{n}, y_{1:n}) dx_n \nonumber \\
+       &   & \times p(\tilde{\theta}_{n}^+|\theta_{n}) p(\theta_{n}|y_{1:n})
+\end{IEEEeqnarray}
+
+Finally, the particle approximation is substituted for the filtering distribution.
+%
+\begin{IEEEeqnarray}{rCl}
+\hat{p}(\theta_{n}|\tilde{\theta}_{n}^+, y_{1:N}) & = & \sum_i \tilde{w}_{n}^{(i)} \delta_{\theta_{n}^{(i)}}(\theta_{n})  \label{eq:rb-vrps_back_cond}
+\end{IEEEeqnarray}
+
+where the backwards conditional weights are given by
+%
+\begin{IEEEeqnarray}{rCl}
+ \tilde{w}_n & \propto & \int p(y_{n+1:N}|x_n, \tilde{\theta}_{n}^+) \nonumber \\
+             &         & \times p(x_n|\theta_{n}^{(i)}, y_{1:n}) dx_n p(\tilde{\theta}_{n}^+|\theta_{n}^{(i)}) \label{eq:rb-vrps_back_cond_weight}
+\end{IEEEeqnarray}
+
+As before, normalisation is enforced by scaling the weights so that they sum to $1$.
+
+The changepoint transition density may be expressed as,
+%
+\begin{IEEEeqnarray}{rCl}
+ p(\tilde{\theta}_{n}^+|\theta_{n}) &=      & p(\tilde{\tau}_{K_n+1:K}, \tilde{u}_{K_n+1:K}|\tau_{K_n}, u_{K_n}, \tau_{K_n+1}>t_n) \nonumber \\
+                                    &\propto& p(\tilde{u}_{K_n+1}|\tilde{\tau}_{K_n+1}, \tau_{K_n}, u_{K_n}) p(\tilde{\tau}_{K_n+1}|\tau_{K_n}, u_{K_n})     .
+\end{IEEEeqnarray}
+
+The integral in~(\ref{eq:RBVRPS_back_cond_weight}) contains two terms involving the current state, $x_n$.
+The first is the familiar state posterior generated by the Kalman filter using (\ref{eq:kf_predict_start}-\ref{eq:kf_update_stop}).
+%
+\begin{IEEEeqnarray}{rCl}
+p(x_n|\theta_{n}^{(i)}, y_{1:n}) & = & \mathcal{N}(x_n|m_n^{(i)}, P_n^{(i)})
+\end{IEEEeqnarray}
+
+The mean and covariance of this distribution will have been calculated during the filtering stage, and can be stored for use now in the smoother.
+
+The second state term in (\ref{eq:rb-vrps_back_cond_weight}) is the likelihood, $p(y_{n+1:N}|x_n, \tilde{\theta}_{n}^+)$.
+This may be considered to be an improper density over $x_n$, and may be calculated analytically using a backwards Kalman filter, in a similar manner to that used in the two-filter smoother \cite{Fraser1969,Anderson1979,Sarkka2012}.
+Such a backwards Kalman filter uses the following recursions.
+Details are provided in appendix~\ref{app:backward_filter}, and in the aforesaid references.
+%
+\begin{IEEEeqnarray}{rCl}
+ p(y_{n+1:N}|x_n, \tilde{\theta}_{n}^+) & \propto & \mathcal{N}(x_n|\tilde{m}_n^-, \tilde{P}_n^-) \\
+ p(y_{n:N}|x_n, \tilde{\theta}_{n}^+) & \propto & \mathcal{N}(x_n|\tilde{m}_n, \tilde{P}_n)
+\end{IEEEeqnarray}
+
+\begin{IEEEeqnarray}{rCl}
+ \tilde{m}_n^- & = & A_{n+1}^{-1} \tilde{m}_{n+1} \label{eq:backward_kf_predict_start} \\
+ \tilde{P}_n^- & = & A_{n+1}^{-1} (\tilde{P}_{n+1} + Q_{n+1}) A_{n+1}^{-T} \label{eq:backward_kf_predict_stop} \\
+ \tilde{\mu}_n & = & C_n \tilde{m}_n^- \label{eq:backward_kf_update_start} \\
+ \tilde{S}_n   & = & C_n \tilde{P}_n^- C_n^T + R_n \\
+ \tilde{K}_n   & = & \tilde{P}_n^- C_n^T \tilde{S}_n^{-1} \\
+ \tilde{m}_n   & = & \tilde{m}_n^- + \tilde{K}_n (y_n - \tilde{\mu}_n) \\
+ \tilde{P}_n   & = & \tilde{P}_n^- - \tilde{K}_n \tilde{S}_n \tilde{K}_n^T \label{eq:backward_kf_update_stop}
+\end{IEEEeqnarray}
+
+Exact methods for initialising this recursion are discussed in appendix~\ref{app:backward_filter}.
+However, it is often sufficient to use an approximation, for example $\tilde{m}_N^{(i)} = m_N^{(i)}$ and $\tilde{P}_N^{(i)} = P_N^{(i)}$.
+
+Substituting into (\ref{eq:rb-vrps_back_cond_weight}), the backwards conditional weights are given by,
+%
+\begin{equation}
+ \tilde{w}_n \propto p(\theta_{n}^+|\theta_{n}^{(i)}) \mathcal{N}(\tilde{m}_n^-|m_n, \tilde{P}_n^- + P_n)     .
+\label{eq:rb-vrps_back_cond_weight2}
+\end{equation}
+
+Using these weights, a sample of $\theta_{n}$ may be drawn from the particle distribution of (\ref{eq:rb-vrps_back_cond}), completing the smoothing algorithm.
+Once sampling has progressed backwards from $n=N \dots 1$, a particle from the smoothing distribution will have been generated.
+This procedure may then be repeated until sufficient particles have been obtained.
+The procedure is summarised in algorithm~\ref{alg:RBVRPS}.
+
+\begin{figure}
+\fbox{\parbox{\columnwidth}{
+\begin{algorithmic}[1]
+  \STATE Run Rao-Blackwellised particle filter to approximate $p(\theta_{n}|y_{1:n})$ with weighted particles $\{\theta_{n}^{(i)}, w_{n}^{(i)}\}$ and $\{p(x_n|\theta_{n}^{(i)},y_{1:n})\}$ as normal distributions with moments $\{m_{n}^{(i)}\}$ and $\{P_{n}^{(i)}\}$.
+Store all results.
+  \FOR{$i=1 \dots N_S$}
+  	\STATE Initialise particle using $\tilde{\theta}^{(i)} \sim \sum_j w_N^{(j)} \delta_{\theta^{(j)}}(\theta)$.
+    \STATE Initialise sufficient statistics $\tilde{m}_N^{(i)}$ and $\tilde{P}_N^{(i)}$ (see text).
+    \FOR{$n=N \dots 1$}
+      \STATE Predict state mean and covariance $\tilde{m}_n^{-(i)}$ and $\tilde{P}_n^{-(i)}$ using \ref{eq:backward_kf_predict_start} to \ref{eq:backward_kf_predict_stop}.
+      \FOR{$j=1 \dots N_P$}
+	      \STATE Calculate weight $\tilde{w}_n^{(j)}$ using (\ref{eq:rb-vrps_back_cond_weight2}).
+      \ENDFOR
+      \STATE Sample $\tilde{\theta}_{n}^{(i)} \sim \sum_j \tilde{w}_n^{(j)} \delta_{\theta_{n}^{(j)}}(\theta_{n})$.
+      \STATE Update state mean and covariance $\tilde{m}_n^{(i)}$ and $\tilde{P}_n^{(i)}$ using \ref{eq:backward_kf_update_start} to \ref{eq:backward_kf_update_stop}.
+      \STATE Discard $\tilde{\theta}_{n-1}^{(i)}$.
+    \ENDFOR
+  \ENDFOR
+\end{algorithmic}
+}}
+\caption{Rao-Blackwellised Variable Rate Particle Smoother}
+\label{alg:RBVRPS}
+\end{figure}
+
+\subsection{Conditionally Deterministic Smoothing} \label{sec:pd-vrps}
+
+In a piecewise-deterministic system, the trick of introducing a state variable to split the set of observations into past and future does not work.
+Furthermore, the simplifications used by the fixed rate particle smoother which exploit the Markovian nature of the state sequence \cite{Godsill2004} are of no help, because past changepoints are not independent of future observations (i.e.
+$p(\theta_{n}|\theta_{n}^{+}, y_{1:N}) \ne p(\theta_{n}|\theta_{n}^{+}, y_{1:n}))$.
+A new approach is required.
+
+\bibliographystyle{IEEEtran}
+\bibliography{D:/pb404/Dropbox/PhD/OTbib}
+\end{document}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+ For ordinary fixed rate models, a particle approximation of the joint smoothing distribution may be constructed by resampling values from the filter approximations backwards through time \cite{Godsill2004}.
+
+Formulating an effective smoother for piecewise deterministic variable rate models proves to be a challenging task, because changing an estimate of the innovations or the state in the past will change the state, and thus the observation likelihood, in the future.
+This obstacle prevents the use of direct backward sampling smoothers, such as that of \cite{Godsill2004}, for which an intractable integral would be required in the weight calculations.
+In this section, a smoother which uses MCMC to sample the changepoint history is developed, using similar principles to \cite{Bunch2012}.
+
+In the derivation of the VRPF, the filter targetted the joint distribution over changepoint times, $\tau_{[0,t_n]}$, and innovations, $v_{[0,t_n]}$.
+This was desirable because once these sequences have been estimated, the state can be calculated deterministically for all times up to $t_n$, an essential property for measurement likelihood evaluations.
+For a smoother, such a scheme is problematic because changing one of the innovations at an early time will alter the state at all subsequent times.
+Instead, the sequence of innovations, $v_{[0,T]}$, is replaced by the sequence of states occuring at the changepoint times, $x_{[0,T]} = \{ x(\tau_k) \forall k \}$, which may be calculated deterministically from the changepoint time and innovation sequences.
+Targetting a distribution over this state sequence, if a change is made to an estimate of $x_k$, it is no longer necessary for the rest of the states to change, only to alter the preceeding and following innovations, $v_{k-1}$ and $v_{k}$.
+For such a scheme to be workable, it must be possible to calculate an innovation from the states and times of the bounding changepoints.
+In other words, (\ref{eq:disc_time_state_diff_eq}) must be ``invertible''.
+
+\begin{IEEEeqnarray}{rCl}
+ v_k & = & f^{-1}(x_{k-1}, x_k, \tau_{k-1}, \tau_{k})  \label{eq:state_invert}
+\end{IEEEeqnarray}
+
+Analagously with $\theta$, a new variable is introduced to represent changepoint time-state pairs, $\zeta_k = \{\tau_k, x_k\}$.
+Thus if $\zeta_{0:K}$ is known (with $\tau_0 = 0$), it is possible to calculate the innovation sequence using (\ref{eq:state_invert}) and thence to calculate the state at any time.
+Thus, for piecewise-deterministic variable rate models, the posterior distribution targetted by a smoother is $p(\zeta_{[0,T]}|y_{1:N})$.
+An approximation to this distibution may be derived from the final filtering results, by tranforming from the innovation sequence, $v_{[0,T]}$, to the changepoint state sequence, $x_{[0,T]}$.
+This approximation will tend to suffer from particle degeneracy due to the resampling procedure, i.e.
+the number of unique particles decreases for earlier times, giving a poor representation of the distribution \cite{Kitagawa1996}.
+An improved smoothing algorithm is needed to rejeuvenate this particle diversity.
+
+For an ordinary fixed rate model, particle smoothing may be conducted by sequentially resampling states from the filtering approximations backwards in time.
+With the nonlinear variable rate models under consideration here, this process is not tractable.
+Instead, particles are sampled from the joint smoothing distribution using MCMC.
+The posterior distribution is first expanded using Bayes theorem around observation time $t_n$.
+
+\begin{IEEEeqnarray}{rCl}
+ \IEEEeqnarraymulticol{3}{l}{p(\zeta_{[0,T]}|y_{1:N})} \nonumber \\
+\qquad & =       & \frac{ p(y_{1:N}|\zeta_{[0,T]}) p(\zeta_{[0,T]}) }{ p(y_{1:N}) } \\
+       & \propto & p(\zeta_{[t_n,T]}|\zeta_{[0,t_n]}) p(\zeta_{[0,t_n]}) \nonumber \\
+       &         &  \times p(y_{r_k^+ +1:N}|\hat{x}_{r_k^+ +1:N}) p(y_{r_k^-:r_k^+}|\hat{x}_{r_k^-:r_k^+}) p(y_{1:r_k^- -1}|\hat{x}_{1:r_k^- -1})  \IEEEeqnarraynumspace
+\end{IEEEeqnarray}
+
+The likelihood term has been split up into three parts, divided at $r_k^-$ and $r_k^+$, the earliest and latest observations to occur between the successive changepoints $\tau_k$ and $\tau_{k+1}$, which lie before and after $t_n$ respectively.
+
+A Markov chain is initialised by sampling a particle from the approximation generated by the final filtering step.
+Because this is an approximation of the target distribution, no burn-in is requiired.
+Proceeding backwards in time, MH steps are conducted by proposing a replacement for the changepoint sequence history from the appropriate filtering approximation.
+At the $n$th step, the future changepoints are considered to be fixed, $\tilde{\zeta}_{[t_n,T]}$.
+A replacement for the past changepoints is proposed from the filtering particles.
+
+\begin{equation}
+ \zeta_{[0,t_n]}^* \sim q(\zeta_{[0,t_n]}|y_{1:n}) = \sum_j \gamma_n^{(j)} \delta_{(\zeta_{[0,t_n]}^{(j)})}(\zeta_{[0,t_n]})
+\end{equation}
+
+This proposed move is accepted with the following acceptance probability, where $\zeta_{[0,t_n]}^{(m)}$ is the previous sequence of past changepoints in the Markov chain.
+
+\begin{IEEEeqnarray}{rCl}
+ \alpha & = & \frac{ p(\zeta_{[0,t_n]}^*, \tilde{\zeta}_{[t_n,T]}|y_{1:N}) q(\zeta_{[0,t_n]}^{(m)}|y_{1:n}) }{ p(\zeta_{[0,t_n]}^{(m)}, \tilde{\zeta}_{[t_n,T]}|y_{1:N}) q(\zeta_{[0,t_n]}^*|y_{1:n}) } \nonumber \\
+        & = & \frac{ \gamma_n^{(m)} }{ \gamma_n^* } \times \frac{ p(\tilde{\zeta}_{[t_n,T]}|\zeta_{[0,t_n]}^*) p(\zeta_{[0,t_n]}^*) }{ p(\tilde{\zeta}_{[t_n,T]}|\zeta_{[0,t_n]}^{(m)}) p(\zeta_{[0,t_n]}^{(m)}) } \nonumber \\
+        &   &  \times \frac{ p(y_{r_k^-:r_k^+}|\hat{x}_{r_k^-:r_k^+}^*) p(y_{1:r_k^- -1}|\hat{x}_{1:r_k^- -1}^*) }{ p(y_{r_k^-:r_k^+}|\hat{x}_{r_k^-:r_k^+}^{(m)}) p(y_{1:r_k^- -1}|\hat{x}_{1:r_k^- -1}^{(m)}) } \nonumber \\
+        & = & \frac{ \gamma_n^{(m)} }{ \gamma_n^* } \times \frac{ p(\zeta_{[0,t_n]}^*) p(y_{1:r_k^- -1}|\hat{x}_{1:r_k^- -1}^*) }{ p(\zeta_{[0,t_n]}^{(m)}) p(y_{1:r_k^- -1}|\hat{x}_{1:r_k^- -1}^{(m)}) } \nonumber \\
+        &   &  \times \frac{ p(\tilde{\zeta}_{k+1}|\zeta_{k}^*) p(y_{r_k^-:r_k^+}|\hat{x}_{r_k^-:r_k^+}^*) }{ p(\tilde{\zeta}_{k+1}|\zeta_{k}^{(m)}) p(y_{r_k^-:r_k^+}|\hat{x}_{r_k^-:r_k^+}^{(m)}) }     \label{eq:MCMC-VRPS_ap}
+\end{IEEEeqnarray}
+
+The first fraction is simply a ratio of proposal weights for the current and proposed changepoint histories.
+The second ratio comprises probabilities which depend solely on the past changepoints, and which are thus the same as values calculated during the filtering stage.
+These may be stored when first calculated, meaning they can simply be fetched from memory when calculating the acceptance probability.
+
+The third ratio consists of the transition probability between two known changepoints, $\zeta_k$ and $\zeta_{k+1}$, and the likelihood of observations occurring between them.
+These may be found by calculating the innovation, $v_k$, to get from $\zeta_k$ to $\zeta_{k+1}$.
+The continuous state trajectory may then be calculated allowing the likelihood terms to be evaluated.
+The transition density term is given by:
+
+\begin{IEEEeqnarray}{rCl}
+ p(\zeta_{k+1}|\zeta_{k}) & = & p(\tau_{k+1}|\tau_{k}) p(v_{k}|v_{k-1})
+\end{IEEEeqnarray}
+
+The variable rate particle smoother (VRPS) is summarised below.%in algorithm~\ref{alg:MCMC-VRPS}.
+
+%\begin{algorithm}
+ \begin{algorithmic}
+  \STATE Run a particle filter to approximate $p(\theta_{[0,t_n]}, x_0|y_{1:n})$ for each $n$.
+  \STATE Convert each particle innovation sequence to a changepoint state sequence, to produce approximations of $p(\zeta_{[0,t_n]}|y_{1:n})$.
+  \FOR{$i=1 \dots S$}
+    \STATE Initialise sampler with $\tilde{\zeta}_{[0,T]}^{(i)} \sim \sum_{j} w_N^{(j)} \delta_{\zeta_{[0,T]}^{(j)}}(\zeta_{[0,T]})$
+    \FOR{$n=N \dots 1$}
+      \STATE $\zeta_{[0,T]}^{(i)(0)} \gets \tilde{\zeta}_{[0,T]}^{(i)}$.
+      \FOR{$m=1 \dots M$}
+        \STATE Propose a new past $\zeta_{[0,t_n]}^{(i)*} \sim \sum_j \gamma_n^{(j)} \delta_{(\zeta_{[0,t_n]}^{(j)})}(\zeta_{[0,t_n]})$.
+	      \STATE Calculate $\alpha$ using equation~\ref{eq:MCMC-VRPS_ap}
+	      \STATE With probability $\alpha$, $\zeta_{[0,t_n]}^{(i)(m)} \gets \zeta_{[0,t_n]}^{(i)*}$. Otherwise $\zeta_{[0,t_n]}^{(i)(m)} \gets \zeta_{[0,t_n]}^{(i)(m-1)}$.
+      \ENDFOR
+      \STATE $\tilde{\zeta}_{[0,t_n]}^{(i)} \gets \zeta_{[0,t]}^{(i)(M)}$.
+    \ENDFOR
+  \ENDFOR
+ \end{algorithmic}
+%\label{alg:MCMC-VRPS}
+%\caption{MCMC variable rate particle smoother}
+%\end{algorithm}
+
+The number of MH steps, $M$, at each observation time, $n$, allows a trade-off of performance against time.
+Larger values of $M$ will ensure more unique samples in the smoothing approximation, but will also take longer to execute.
+In the simulations discussed in this report, $M=1$ was used throughout.
+
+The recursion is initialised by inversion of the observation model.
+By stacking up the last $L+1$ observations into a single vector, it is possible to write down a joint Gaussian distribution conditional on the $(N-L)$th state.
+This may be inverted to produce an improper Gaussian density for $x_{N-L}$,
+%
+\begin{IEEEeqnarray}{rCl}
+ \IEEEeqnarraymulticol{3}{l}{p(y_{N-L:N}|x_{N-L}) = \mathcal{N}( \mathbf{y}_{N-L:N} | H_{N-L} x_{N-L}, \Gamma_{N-L} ) } \\
+ & \propto & \mathcal{N}(x_{N-L} | (H_{N-L}^T \Gamma_{N-L} H_{N-L})^{-1} H_{N-L}^T \Gamma_{N-L}, (H_{N-L}^T \Gamma_{N-L} H_{N-L})^{-1} )      ,
+\end{IEEEeqnarray}
+
+where $H_{N-L}$ and $\Gamma_{N-L}$ may be constructed from the system models and,
+%
+\begin{IEEEeqnarray}{rCl}
+\mathbf{y}_{N-L:N} & = & \begin{bmatrix}y_N \\ y_{N-1} \\ \vdots \\ y_{N-L} \end{bmatrix}     .
+\end{IEEEeqnarray}
+
+This should effected for the smallest value of $L$ such that the matrix inversions possible.
+Further details are provided in appendix~\ref{app:backward_filter}.
+Alternatively, the recursion could be initialised with an approximation.

### 2012-03-06 draft1_VRsmoothing.tex
Written pd-vrps section

@@
 Therefore, $\theta_{n-1}$ is marginalised by simply discarding the changepoints which occur before this time.
-This yields in a particle drawn from $p(\theta_{n-1}^{+}|y_{1:N})$.
+This results in a particle drawn from $p(\theta_{n-1}^{+}|y_{1:N})$.
 The procedure then continues recursively.
@@
 
-At step $n$ for a given particle, the future changepoint sequence may be considered to be fixed, and is denoted $\tilde{\theta}_n^+$.
+At step $n$ for a given particle, the future changepoint sequence may be considered to be fixed.
 It remains to devise a method for drawing samples from the backwards conditional distribution, $p(\theta_{n}|\theta_{n}^{+}, y_{1:N})$.
@@
 \begin{IEEEeqnarray}{rCl}
- \IEEEeqnarraymulticol{3}{l}{p(\theta_{n}|\tilde{\theta}_{n}^+, y_{1:N})} \nonumber \\
-\qquad & \propto & p(\theta_{n}, \tilde{\theta}_{n}^+| y_{1:N}) \nonumber  \\
-       & =       & \int p(x_n, \theta_{n}, \tilde{\theta}_{n}^+| y_{1:N}) dx_n \nonumber  \\
-       & \propto & \int p(y_{n+1:N}|x_n, \theta_{n}, \tilde{\theta}_{n}^+, y_{1:n}) p(x_n, \theta_{n}, \tilde{\theta}_{n}^+| y_{1:n}) dx_n \nonumber \\
-       & = & \int p(y_{n+1:N}|x_n, \tilde{\theta}_{n}^+) p(x_n|\theta_{n}, y_{1:n}) dx_n \nonumber \\
-       &   & \times p(\tilde{\theta}_{n}^+|\theta_{n}) p(\theta_{n}|y_{1:n})
+ \IEEEeqnarraymulticol{3}{l}{p(\theta_{n}|\theta_{n}^+, y_{1:N})} \nonumber \\
+\qquad & \propto & p(\theta_{n}, \theta_{n}^+| y_{1:N}) \nonumber  \\
+       & =       & \int p(x_n, \theta_{n}, \theta_{n}^+| y_{1:N}) dx_n \nonumber  \\
+       & \propto & \int p(y_{n+1:N}|x_n, \theta_{n}, \theta_{n}^+, y_{1:n}) p(x_n, \theta_{n}, \theta_{n}^+| y_{1:n}) dx_n \nonumber \\
+       & = & \int p(y_{n+1:N}|x_n, \theta_{n}^+) p(x_n|\theta_{n}, y_{1:n}) dx_n \nonumber \\
+       &   & \times p(\theta_{n}^+|\theta_{n}) p(\theta_{n}|y_{1:n})
 \end{IEEEeqnarray}
@@
 \begin{IEEEeqnarray}{rCl}
-\hat{p}(\theta_{n}|\tilde{\theta}_{n}^+, y_{1:N}) & = & \sum_i \tilde{w}_{n}^{(i)} \delta_{\theta_{n}^{(i)}}(\theta_{n})  \label{eq:rb-vrps_back_cond}
+\hat{p}(\theta_{n}|\theta_{n}^+, y_{1:N}) & = & \sum_i \tilde{w}_{n}^{(i)} \delta_{\theta_{n}^{(i)}}(\theta_{n})  \label{eq:rb-vrps_back_cond}
 \end{IEEEeqnarray}
@@
 \begin{IEEEeqnarray}{rCl}
- \tilde{w}_n & \propto & \int p(y_{n+1:N}|x_n, \tilde{\theta}_{n}^+) \nonumber \\
-             &         & \times p(x_n|\theta_{n}^{(i)}, y_{1:n}) dx_n p(\tilde{\theta}_{n}^+|\theta_{n}^{(i)}) \label{eq:rb-vrps_back_cond_weight}
+ \tilde{w}_n & \propto & \int p(y_{n+1:N}|x_n, \theta_{n}^+) \nonumber \\
+             &         & \times p(x_n|\theta_{n}^{(i)}, y_{1:n}) dx_n p(\theta_{n}^+|\theta_{n}^{(i)}) \label{eq:rb-vrps_back_cond_weight}
 \end{IEEEeqnarray}
@@
 \begin{IEEEeqnarray}{rCl}
- p(\tilde{\theta}_{n}^+|\theta_{n}) &=      & p(\tilde{\tau}_{K_n+1:K}, \tilde{u}_{K_n+1:K}|\tau_{K_n}, u_{K_n}, \tau_{K_n+1}>t_n) \nonumber \\
-                                    &\propto& p(\tilde{u}_{K_n+1}|\tilde{\tau}_{K_n+1}, \tau_{K_n}, u_{K_n}) p(\tilde{\tau}_{K_n+1}|\tau_{K_n}, u_{K_n})     .
+ p(\theta_{n}^+|\theta_{n}) &=      & p(\tau_{K_n+1:K}, u_{K_n+1:K}|\tau_{K_n}, u_{K_n}, \tau_{K_n+1}>t_n) \nonumber \\
+                                    &\propto& p(u_{K_n+1}|\tau_{K_n+1}, \tau_{K_n}, u_{K_n}) p(\tau_{K_n+1}|\tau_{K_n}, u_{K_n})     .
 \end{IEEEeqnarray}
 
-The integral in~(\ref{eq:RBVRPS_back_cond_weight}) contains two terms involving the current state, $x_n$.
-The first is the familiar state posterior generated by the Kalman filter using (\ref{eq:kf_predict_start}-\ref{eq:kf_update_stop}).
+The integral in~(\ref{eq:rb-vrps_back_cond_weight}) contains two terms involving the current state, $x_n$.
+The first is the familiar state posterior generated by the Kalman filter using (\ref{eq:kf_predict_start}--\ref{eq:kf_update_stop}).
 %
@@
 \begin{IEEEeqnarray}{rCl}
- p(y_{n+1:N}|x_n, \tilde{\theta}_{n}^+) & \propto & \mathcal{N}(x_n|\tilde{m}_n^-, \tilde{P}_n^-) \\
- p(y_{n:N}|x_n, \tilde{\theta}_{n}^+) & \propto & \mathcal{N}(x_n|\tilde{m}_n, \tilde{P}_n)
+ p(y_{n+1:N}|x_n, \theta_{n}^+) & \propto & \mathcal{N}(x_n|\tilde{m}_n^-, \tilde{P}_n^-) \\
+ p(y_{n:N}|x_n, \theta_{n}^+) & \propto & \mathcal{N}(x_n|\tilde{m}_n, \tilde{P}_n)
 \end{IEEEeqnarray}
@@
 This procedure may then be repeated until sufficient particles have been obtained.
-The procedure is summarised in algorithm~\ref{alg:RBVRPS}.
+The algorithm is summarised in Fig.~\ref{alg:RBVRPS}.
 
@@
 \begin{algorithmic}[1]
-  \STATE Run Rao-Blackwellised particle filter to approximate $p(\theta_{n}|y_{1:n})$ with weighted particles $\{\theta_{n}^{(i)}, w_{n}^{(i)}\}$ and $\{p(x_n|\theta_{n}^{(i)},y_{1:n})\}$ as normal distributions with moments $\{m_{n}^{(i)}\}$ and $\{P_{n}^{(i)}\}$.
+  \STATE Run Rao-Blackwellised variable rate particle filter to approximate $p(\theta_{n}|y_{1:n})$ with weighted particles $\{\theta_{n}^{(i)}, w_{n}^{(i)}\}$ and $\{p(x_n|\theta_{n}^{(i)},y_{1:n})\}$ as normal distributions with moments $\{m_{n}^{(i)}\}$ and $\{P_{n}^{(i)}\}$.
 Store all results.
   \FOR{$i=1 \dots N_S$}
-  	\STATE Initialise particle using $\tilde{\theta}^{(i)} \sim \sum_j w_N^{(j)} \delta_{\theta^{(j)}}(\theta)$.
+  	\STATE Initialise particle using $\theta^{(i)} \sim \sum_j w_N^{(j)} \delta_{\theta^{(j)}}(\theta)$.
     \STATE Initialise sufficient statistics $\tilde{m}_N^{(i)}$ and $\tilde{P}_N^{(i)}$ (see text).
@@
       \ENDFOR
-      \STATE Sample $\tilde{\theta}_{n}^{(i)} \sim \sum_j \tilde{w}_n^{(j)} \delta_{\theta_{n}^{(j)}}(\theta_{n})$.
+      \STATE Sample $\theta_{n}^{(i)} \sim \sum_j \tilde{w}_n^{(j)} \delta_{\theta_{n}^{(j)}}(\theta_{n})$.
       \STATE Update state mean and covariance $\tilde{m}_n^{(i)}$ and $\tilde{P}_n^{(i)}$ using \ref{eq:backward_kf_update_start} to \ref{eq:backward_kf_update_stop}.
-      \STATE Discard $\tilde{\theta}_{n-1}^{(i)}$.
+      \STATE Discard $\theta_{n-1}^{(i)}$.
     \ENDFOR
@@
 
+The algorithmic complexity of this variable rate particle smoother for conditionally linear-Gaussian models is $O(N_F \times N_S \times N)$.
+It is possible to reduce this by using an MCMC sampling scheme in the style of \cite{Bunch2012}, in which case it is no longer necessary to calculate the backwards sampling weights for every filter particle.
+
 \subsection{Conditionally Deterministic Smoothing} \label{sec:pd-vrps}
@@
 
+Consider the augmented MPP consisting of $\tilde{\theta} = \{\tau_{0:K}, u_{0:K}, x_{0:K}\}$ ($\tilde{\theta}_n$ and $\tilde{\theta}_n^+$ defined in the same manner as before).
+The mark corresponding to $\tau_k$ is now the parameter-state pair $(u_k, x_k)$, i.e.
+the state at the time of the change plus the parameters for the next division.
+The smoother is formulated by factorising the augmented sequence posterior distribution as before,
+%
+\begin{IEEEeqnarray}{rCl}
+ p(\tilde{\theta}|y_{1:N}) = p(\tilde{\theta_{n}}^{+}|y_{1:N}) p(\tilde{\theta}_{n}|\tilde{\theta}_{n}^{+}, y_{1:N})     .
+\end{IEEEeqnarray}
+
+The backwards conditional term may be expanded as,
+%
+\begin{IEEEeqnarray}{rCl}
+p(\tilde{\theta}_{n}|\tilde{\theta}_{n}^{+}, y_{1:N}) & \propto & p(y_{1:N}|\tilde{\theta}_{n}, \tilde{\theta}_{n}^{+}) p(\tilde{\theta}_{n}^{+}|\tilde{\theta}_{n}) p(\tilde{\theta}_{n})    .
+\end{IEEEeqnarray}
+
+For each particle in the filtering approximations, the changepoint state values $x_{0:K_n}$ may be calculated deterministically from $\theta_n$ and stored during the filtering procedure.
+The obvious strategy then is to recursively sample $\tilde{\theta}_{n}$ from the backwards conditional term, using the re-weighted particles of the $n$th filtering approximation.
+However, the conditionally deterministic structure prevents this from working.
+$x(t)$ is fully determined by $\tilde{\theta}_{n}$ up to $\tau_{K_n+1}$, but this will not usually match up with $x_{K_n+1}$ as specified in $\tilde{\theta}_{n}^+$.
+The result is that $p(\tilde{\theta}_{n}^{+}|\tilde{\theta}_{n})=0$ for (almost) all particles.
+In other words, we are trying to join two deterministic state trajectories together, but they do not meet up in the middle.
+
+The solution to this problem is provided by an idea from the Sequential Monte Carlo (SMC) samplers of \cite{DelMoral2006}.
+Rather than just using a value of $\tilde{\theta}_{n}$ from the filtering particles, a replacement, $u_{K_n}'$, is proposed for the existing parameter corresponding to the final changepoint, $u_{K_n}$, with a value chosen such that the past and future paths meet up.
+The modified sequence is denoted $\tilde{\theta}_{n}'$.
+This requires a new condition to be imposed on the state dynamics.
+For any pair of adjacent changepoints, $(\tau_k,x_k)$, and $(\tau_{k+1},x_{k+1})$ it must be possible to calculate a parameter value which results in the transition from the former to the latter.
+As a rule of thumb, this requires the number of dimensions of $u_k$ to be greater or equal to the number of dimensions of $x_k$.
+
+The target distribution is augmented by introduction of an artificial density to cover the new changepoint sequence and the discarded parameter. It is clear that this new target distribution,
+
+\begin{IEEEeqnarray}{rCl}
+ p(\tilde{\theta}_{n}'|\tilde{\theta}_{n}^{+}, y_{1:N}) L(u_{K_n}|\tilde{\theta}_{n}', \tilde{\theta}_{n}^{+}, y_{1:N}) \label{eq:pd-vrps_target}     ,
+\end{IEEEeqnarray}
+
+admits the desired posterior term as a marginal.
+
+The augmented target distribution may be sampled using Markov chain Monte Carlo (MCMC) \cite{Gilks1996}, in a similar manner to the fixed rate smoother of \cite{Bunch2012}.
+Metropolis-Hastings (MH) steps are conducted to draw samples, $\{\tilde{\theta}_{n}, u_{K_n}'\}$, from the target distribution by sampling a proposal and accepting the new values with a given probability.
+The proposal is conducted by first resampling a particle from the $n$th filtering approximation and then sampling the replacement parameter from an appropriate proposal density.
+Thus a sample is drawn from,
+
+\begin{IEEEeqnarray}{rCl}
+\IEEEeqnarraymulticol{3}{l}{p(\tilde{\theta}_{n}|y_{1:n}) q(u_{K_n}|\tilde{\theta}_{n}, \tilde{\theta}_{n}^{+}, y_{1:N})} \nonumber \\
+ \qquad & \propto & p(y_{1:n}|\tilde{\theta}_{n}) p(\tilde{\theta}_{n}) q(u_{K_n}'|\tilde{\theta}_{n}, \tilde{\theta}_{n}^{+}, y_{1:N}) \label{eq:pd-vrps_proposal}     .
+\end{IEEEeqnarray}
+
+Hence the ratio, of target (\ref{eq:pd-vrps_target}) and proposal (\ref{eq:pd-vrps_proposal}) densities is,
+
+\begin{IEEEeqnarray}{rCl}
+\beta_n &\propto& \frac{ p(y_{r_n^-:r_n^+}|\tilde{\theta}_{n}') }{ p(y_{r_n^-:n}|\tilde{\theta}_{n}) } \nonumber \\
+  & & \times \frac{ p(\tilde{\theta}_{n}^{+}|\tilde{\theta}_{n}') p(u_{K_n}'|\tilde{\theta}_n \setminus u_{K_n}) }{ p(u_{K_n}|\tilde{\theta}_n \setminus u_{K_n}) } \nonumber \\
+  & & \times \frac{ L(u_{K_n}|\tilde{\theta}_{n}, \tilde{\theta}_{n}^{+}, y_{1:N}) }{ q(u_{K_n}'|\tilde{\theta}_{n}, \tilde{\theta}_{n}^{+}, y_{1:N}) } \label{eq:pd-vrps_tpr}    ,
+\end{IEEEeqnarray}
+
+where $r_n^- = \min( m : t_m > \tau_{K_n} )$ and $r_n^+ = \max( m : t_m < \tau_{K_n+1} )$.
+
+The transition density term will include a delta function with unit probability mass at the point for which the past and future state trajectories match up.
+
+\begin{IEEEeqnarray}{rCl}
+\IEEEeqnarraymulticol{3}{l}{p(\tilde{\theta}_{n}^{+}|\tilde{\theta}_{n}') \propto p(u_{K_n+1}|\tau_{K_n+1}, \tau_{K_n}, u_{K_n})} \nonumber \\ \times p(\tau_{K_n+1}|\tau_{K_n}, u_{K_n}) \delta_{f(x_{K_n}, v_{K_n}, \tau_{K_n}, \tau_{K_n+1})}(x_{K_n+1})
+\end{IEEEeqnarray}
+
+If there is only one value of $u_{K_n}'$ for which the two trajectories match up, then the proposal density $q(u_{K_n}'|\tilde{\theta}_{n}, \tilde{\theta}_{n}^{+}, y_{1:N})$ is also a delta function, with unit probability mass at the one valid value.
+The remaining terms are likelihoods and parameter transition density terms, defined by the system equations.
+The simplest choice for $L(.|.)$ is simply as a uniform distribution so that it cancels out in the acceptance probability.
+
+If $\beta_n^{(m-1)}$ is the target-proposal ratio for the current state in the chain, and $\beta_n^*$ that for the proposed state, then the acceptance probability is given by,
+
+\begin{IEEEeqnarray}{rCl}
+\alpha_n^{(m)} = \min \left( 1, \frac{\beta_n^*}{\beta_n^{(m-1)}} \right) \label{eq:pd-vrps_ap}     .
+\end{IEEEeqnarray}
+
+The algorithm is summarised in Fig.~\ref{alg:VRPS}.
+
+\begin{figure}
+\fbox{\parbox{\columnwidth}{
+\begin{algorithmic}[1]
+  \STATE Run variable rate particle filter to approximate $p(\tilde{\theta}_{n}|y_{1:n})$ with weighted particles $\{\tilde{\theta}_{n}^{(i)}, w_{n}^{(i)}\}$. Store all results.
+  \FOR{$i=1 \dots N_S$}
+  	\STATE Initialise particle from $N$th filter approximation $\tilde{\theta}^{(i)} \sim \sum_j w_N^{(j)} \delta_{\tilde{\theta}^{(j)}}(\tilde{\theta})$.
+    \FOR{$n=N \dots 1$}
+      \STATE Get the current changepoint history $\tilde{\theta}_n^{(i)(0)} \gets \tilde{\theta}_n^{(i)}$.
+      \FOR{$m=1 \dots M$}
+        \STATE Propose a new changepoint history from $n$th filter approximation $\tilde{\theta}_n^{(i)*} \sim \sum_j w_n^{(j)} \delta_{\tilde{\theta}_{n}^{(j)}}(\tilde{\theta}_{n})$.
+        \STATE Propose a replacement parameter $u_{K_n}'^* \sim q(u_{K_n}|\tilde{\theta}_{n}, \tilde{\theta}_{n}^{+}, y_{1:N})$.
+        \STATE Construct a new changepoint history, $\tilde{\theta}_n'^{(i)*}$, by replacing $u_{K_n}^*$ with $u_{K_n}'^*$.
+        \STATE Calculate acceptance probability $\alpha_n^{(m)}$ using (\ref{eq:pd-vrps_tpr}) and (\ref{eq:pd-vrps_ap}).
+        \STATE With probability $\alpha_n^{(m)}$, $\tilde{\theta}_n^{(i)(m)} \gets \tilde{\theta}_n^{(i)*}$. Otherwise, $\tilde{\theta}_n^{(i)(m)} \gets \tilde{\theta}_n^{(i)(m-1)}$.
+      \ENDFOR
+      \STATE Keep final sample $\tilde{\theta}_n^{(i)} \gets \tilde{\theta}_n^{(i)(M)}$.
+    \ENDFOR
+  \ENDFOR
+\end{algorithmic}
+}}
+\caption{MCMC piecewise-deterministic variable rate particle smoother}
+\label{alg:VRPS}
+\end{figure}
+
+The Markov chains require no burn-in, because the original value is itself a sample from the target distribution.
+The number of MH steps, $M$, at each observation time, $n$, allows a trade-off of performance against time.
+Larger values of $M$ will ensure more unique samples in the smoothing approximation, but will also take longer to execute.
+In the simulations discussed in this report, $M=1$ was used throughout.
+
 \bibliographystyle{IEEEtran}
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
- For ordinary fixed rate models, a particle approximation of the joint smoothing distribution may be constructed by resampling values from the filter approximations backwards through time \cite{Godsill2004}.
-
-Formulating an effective smoother for piecewise deterministic variable rate models proves to be a challenging task, because changing an estimate of the innovations or the state in the past will change the state, and thus the observation likelihood, in the future.
-This obstacle prevents the use of direct backward sampling smoothers, such as that of \cite{Godsill2004}, for which an intractable integral would be required in the weight calculations.
-In this section, a smoother which uses MCMC to sample the changepoint history is developed, using similar principles to \cite{Bunch2012}.
-
-In the derivation of the VRPF, the filter targetted the joint distribution over changepoint times, $\tau_{[0,t_n]}$, and innovations, $v_{[0,t_n]}$.
-This was desirable because once these sequences have been estimated, the state can be calculated deterministically for all times up to $t_n$, an essential property for measurement likelihood evaluations.
-For a smoother, such a scheme is problematic because changing one of the innovations at an early time will alter the state at all subsequent times.
-Instead, the sequence of innovations, $v_{[0,T]}$, is replaced by the sequence of states occuring at the changepoint times, $x_{[0,T]} = \{ x(\tau_k) \forall k \}$, which may be calculated deterministically from the changepoint time and innovation sequences.
-Targetting a distribution over this state sequence, if a change is made to an estimate of $x_k$, it is no longer necessary for the rest of the states to change, only to alter the preceeding and following innovations, $v_{k-1}$ and $v_{k}$.
-For such a scheme to be workable, it must be possible to calculate an innovation from the states and times of the bounding changepoints.
-In other words, (\ref{eq:disc_time_state_diff_eq}) must be ``invertible''.
-
-\begin{IEEEeqnarray}{rCl}
- v_k & = & f^{-1}(x_{k-1}, x_k, \tau_{k-1}, \tau_{k})  \label{eq:state_invert}
-\end{IEEEeqnarray}
-
-Analagously with $\theta$, a new variable is introduced to represent changepoint time-state pairs, $\zeta_k = \{\tau_k, x_k\}$.
-Thus if $\zeta_{0:K}$ is known (with $\tau_0 = 0$), it is possible to calculate the innovation sequence using (\ref{eq:state_invert}) and thence to calculate the state at any time.
-Thus, for piecewise-deterministic variable rate models, the posterior distribution targetted by a smoother is $p(\zeta_{[0,T]}|y_{1:N})$.
-An approximation to this distibution may be derived from the final filtering results, by tranforming from the innovation sequence, $v_{[0,T]}$, to the changepoint state sequence, $x_{[0,T]}$.
-This approximation will tend to suffer from particle degeneracy due to the resampling procedure, i.e.
-the number of unique particles decreases for earlier times, giving a poor representation of the distribution \cite{Kitagawa1996}.
-An improved smoothing algorithm is needed to rejeuvenate this particle diversity.
-
-For an ordinary fixed rate model, particle smoothing may be conducted by sequentially resampling states from the filtering approximations backwards in time.
-With the nonlinear variable rate models under consideration here, this process is not tractable.
-Instead, particles are sampled from the joint smoothing distribution using MCMC.
-The posterior distribution is first expanded using Bayes theorem around observation time $t_n$.
-
-\begin{IEEEeqnarray}{rCl}
- \IEEEeqnarraymulticol{3}{l}{p(\zeta_{[0,T]}|y_{1:N})} \nonumber \\
-\qquad & =       & \frac{ p(y_{1:N}|\zeta_{[0,T]}) p(\zeta_{[0,T]}) }{ p(y_{1:N}) } \\
-       & \propto & p(\zeta_{[t_n,T]}|\zeta_{[0,t_n]}) p(\zeta_{[0,t_n]}) \nonumber \\
-       &         &  \times p(y_{r_k^+ +1:N}|\hat{x}_{r_k^+ +1:N}) p(y_{r_k^-:r_k^+}|\hat{x}_{r_k^-:r_k^+}) p(y_{1:r_k^- -1}|\hat{x}_{1:r_k^- -1})  \IEEEeqnarraynumspace
-\end{IEEEeqnarray}
-
-The likelihood term has been split up into three parts, divided at $r_k^-$ and $r_k^+$, the earliest and latest observations to occur between the successive changepoints $\tau_k$ and $\tau_{k+1}$, which lie before and after $t_n$ respectively.
 

### 2012-03-16 draft1_VRsmoothing.tex
Added some results for RB version.

@@
 \usepackage{color}
+
+\graphicspath{{./images/}{./images/finance/}{./images/tracking/}}
 
@@
 }}
+\caption{Rao-Blackwellised variable rate particle filter}
 \label{alg:RBVRPF}
-\caption{Rao-Blackwellised variable rate particle filter}
 \end{figure}
@@
 }}
+\caption{Piecewise deterministic variable rate particle filter}
 \label{alg:VRPF}
-\caption{Piecewise deterministic variable rate particle filter}
 \end{figure}
@@
 The solution to this problem is provided by an idea from the Sequential Monte Carlo (SMC) samplers of \cite{DelMoral2006}.
-Rather than just using a value of $\tilde{\theta}_{n}$ from the filtering particles, a replacement, $u_{K_n}'$, is proposed for the existing parameter corresponding to the final changepoint, $u_{K_n}$, with a value chosen such that the past and future paths meet up.
+Rather than just using a value of $\tilde{\theta}_{n}$ from the filtering particles, a replacement, $u_{K_n}'$, is proposed for the existing parameter corresponding to the final changepoint with a value chosen such that the past and future paths meet up.
 The modified sequence is denoted $\tilde{\theta}_{n}'$.
-This requires a new condition to be imposed on the state dynamics.
-For any pair of adjacent changepoints, $(\tau_k,x_k)$, and $(\tau_{k+1},x_{k+1})$ it must be possible to calculate a parameter value which results in the transition from the former to the latter.
-As a rule of thumb, this requires the number of dimensions of $u_k$ to be greater or equal to the number of dimensions of $x_k$.
-
-The target distribution is augmented by introduction of an artificial density to cover the new changepoint sequence and the discarded parameter. It is clear that this new target distribution,
-
-\begin{IEEEeqnarray}{rCl}
- p(\tilde{\theta}_{n}'|\tilde{\theta}_{n}^{+}, y_{1:N}) L(u_{K_n}|\tilde{\theta}_{n}', \tilde{\theta}_{n}^{+}, y_{1:N}) \label{eq:pd-vrps_target}     ,
-\end{IEEEeqnarray}
-
-admits the desired posterior term as a marginal.
-
-The augmented target distribution may be sampled using Markov chain Monte Carlo (MCMC) \cite{Gilks1996}, in a similar manner to the fixed rate smoother of \cite{Bunch2012}.
-Metropolis-Hastings (MH) steps are conducted to draw samples, $\{\tilde{\theta}_{n}, u_{K_n}'\}$, from the target distribution by sampling a proposal and accepting the new values with a given probability.
-The proposal is conducted by first resampling a particle from the $n$th filtering approximation and then sampling the replacement parameter from an appropriate proposal density.
-Thus a sample is drawn from,
+Such a proposal may be written as,
 
@@
 
-Hence the ratio, of target (\ref{eq:pd-vrps_target}) and proposal (\ref{eq:pd-vrps_proposal}) densities is,
+For this proposal to work, a new condition to be imposed on the state dynamics.
+For any pair of adjacent changepoints, $(\tau_k,x_k)$, and $(\tau_{k+1},x_{k+1})$ it must be possible to calculate a parameter value which results in the transition from the former to the latter.
+As a rule of thumb, this requires the number of dimensions of $u_k$ to be greater or equal to the number of dimensions of $x_k$.
+
+The target distribution is augmented by introduction of an artificial density to cover the new changepoint sequence and the discarded parameter.
+
+\begin{IEEEeqnarray}{rCl}
+ p(\tilde{\theta}_{n}'|\tilde{\theta}_{n}^{+}, y_{1:N}) L(u_{K_n}|\tilde{\theta}_{n}', \tilde{\theta}_{n}^{+}, y_{1:N}) \label{eq:pd-vrps_target}     ,
+\end{IEEEeqnarray}
+
+It is clear that this new target distribution admits the desired posterior term as a marginal.
+
+The augmented target distribution may be sampled using Markov chain Monte Carlo (MCMC) \cite{Gilks1996}, in a similar manner to the fixed rate smoother of \cite{Bunch2012}.
+Metropolis-Hastings (MH) steps are conducted to draw samples, $\{\tilde{\theta}_{n}, u_{K_n}'\}$, from the target distribution by sampling the proposal and accepting the new values with a given probability.
+
+The ratio of target (\ref{eq:pd-vrps_target}) and proposal (\ref{eq:pd-vrps_proposal}) densities is given by,
 
@@
 In the simulations discussed in this report, $M=1$ was used throughout.
+
+\section{Simulations}
+
+\subsection{Conditionally Linear-Gaussian Model} \label{sec:finance}
+
+The Rao-Blackwellised VRPS algorithm was tested on the financial time series model of \cite{Godsill2007a,Christensen2012}, in which prices of an asset are treated as noisy observations of a latent state, which evolves according to a drift-diffusion with occasional jumps.
+
+The latent state is a vector with two elements, the underlying value of the asset, and the trend followed by this value.
+%
+\begin{equation}
+ \mathbf{x}(t) = [ x(t), \dot{x}(t)]^T
+\end{equation}
+
+This evolves continuously according to a jump-diffusion model:
+%
+\begin{IEEEeqnarray}{c}
+ d\mathbf{x}(t) = \begin{bmatrix}0 & 1 \\ 0 & -\lambda \end{bmatrix} \mathbf{x}(t) dt + \begin{bmatrix}0 \\ \sigma \end{bmatrix} d\mathbf{\beta}(t) + d\mathbf{J}(t)
+\end{IEEEeqnarray}
+
+where $\lambda$ introduces a mean regression effect on the trend and $\mathbf{\beta}(t)$ is standard Brownian motion (with unit diffusion constant).
+The jump term, $d\mathbf{J}(t)$, is zero everywhere except where jumps occur.
+%
+\begin{IEEEeqnarray}{rCl}
+ d\mathbf{J}(t) & = & \begin{cases} \mathbf{J}_k & t \in \{\tau_k\} \\ 0 & \text{elsewhere} \end{cases} \\
+ \mathbf{J}_k  & \sim & \mathcal{N}(\mathbf{J}_k| \mathbf{0}, Q_{J,u_k})
+\end{IEEEeqnarray}
+
+Jumps occur at a random set of times, $\{\tau_k\}$, and are one of two types: value jumps, indicated by $u_k = 1$, and trend jumps, indicated by $u_k=2$.
+
+The jump covariance matrices are,
+%
+\begin{IEEEeqnarray}{c}
+Q_{J,u_k} = \begin{cases} \begin{bmatrix}\sigma_{J1}^2 & 0 \\ 0 & 0 \end{bmatrix} & u_k = 1 \\
+                          \begin{bmatrix}0 & 0 \\ 0 & \sigma_{J2}^2 \end{bmatrix} & u_k = 2  \end{cases}   .
+\end{IEEEeqnarray}
+
+This model may be discretised at the observation times using standard methods (see e.g.
+\cite{Grewal2002}).
+Assuming observations of value only and Gaussian observation noise with standard deviation $\sigma_y^2$, the resulting discrete time dynamics are described by the following equations:% (see appendix \ref{app:model_discretisation}):
+%
+\begin{IEEEeqnarray}{rCl}
+ \mathbf{x}_n &=& A \mathbf{x}_{n-1} + \mathbf{w}_n \\
+ y_n          &=& C \mathbf{x}_{n} + v_n
+\end{IEEEeqnarray}
+
+where the $\mathbf{w}_b$ and $v_n$ are Gaussian random variables with covariance matrixes $Q_n$ and $R$ respectively. The time between observations times is denotes $\Delta t = t_{n} - t_{n-1}$.
+%
+\begin{IEEEeqnarray}{rCl}
+ A     &=& \begin{bmatrix}1 & \frac{1}{\lambda}(1-e^{(-\lambda \Delta t)} \\ 0 & e^{(-\lambda \Delta t)}\end{bmatrix} \\
+ C     &=& \begin{bmatrix}1 & 0\end{bmatrix}
+\end{IEEEeqnarray}
+\begin{IEEEeqnarray}{rCl}
+ Q_n   &=& \begin{cases}Q_{D} + Q_{J,u_k} & \exists k : \tau_k \in [t_{n-1},t_n]\\
+                        Q_{D}             & \text{otherwise} \end{cases} \\
+ Q_{D} &=& \frac{\sigma^2}{2 \lambda}\begin{bmatrix} q_1 & q_2 \\ q_2 & q_3\end{bmatrix} \\
+ q_1   &=& \frac{1}{\lambda^2}(2 \lambda \Delta t - (3 - e^{(-\lambda \Delta t)})(1 - e^{(-\lambda \Delta t)})~) \\
+ q_2   &=& \frac{1}{\lambda} (1-e^{(-\lambda \Delta t)})^2 \\
+ q_3   &=& 1-e^{(-2 \lambda \Delta t)} \\
+ R     &=& [\sigma_y^2]
+\end{IEEEeqnarray}
+
+The times between changepoints were assumed to be exponentially distributed, with equal probability of value and trend jumps.
+%
+\begin{IEEEeqnarray}{rCl}
+ \IEEEeqnarraymulticol{3}{l}{p(\tau_k, u_k|\tau_{k-1}, u_{k-1}) = P(u_k) p(\tau_k|\tau_{k-1})} \\
+ p(\tau_k|\tau_{k-1})               &=& \begin{cases}\frac{1}{\alpha} \exp ( \alpha (\tau_k-\tau_{k-1}) ) & \tau_k>\tau_{k-1} \\
+                                                     0 & \tau_k < \tau_{k-1} \end{cases} \\
+ P(u_k)                             &=& \begin{cases}0.5 & u_k = 1 \\ 0.5 & u_k = 2\end{cases}
+\end{IEEEeqnarray}
+
+The algorithms were first tested on artificial data simulated from this model.
+The following parameters were used: $\Delta t = 0.0017$, $N = 1000$, $\alpha = 20$, $\lambda = 5$, $\sigma = 0.05$, $\sigma_{J1} = 0.005$, $\sigma_{J2} = 0.05$, $\sigma_{y} = 0.001$.
+
+The filter used $N_P = 100$ particles, and the smoother resampled $N_S = 100$ sequences. Bootstrap proposals were used for the filter.
+
+An example realisation simulated from the model is shown in figure~\ref{fig:example_data}.
+
+\begin{figure}[!t]
+\centering
+\includegraphics[width=0.9\columnwidth]{example_data.pdf}
+\caption{An example simulated data set, showing value (top) and trend (bottom). Value observations are overlayed. Jump times are shown as dotted vertical lines.}
+\label{fig:example_data}
+\end{figure}
+
+Both the smoother and the final processing step of the filter produce a particle approximation to the changepoint sequence posterior, $p(\theta|y_{1:N})$.
+It is actually no simple task to compare the quality of these approximations.
+Each particle consists of a list of changepoint times and types, and there is no obvious way to take an average over the set to produce a single estimate.
+Furthermore, once a best estimate has been calculated, it is not obvious how the error should be assessed, as it is not trivial to define a distance metric between lists of varying lengths.
+
+For a first, qualitative comparison of the changepoint sequence distributions, it is possible to generate kernel density estimates for the changepoint times.
+These are constructed by simply adding a (unit amplitude) Gaussian kernel for each changepoint present in each particle and scaling by the number of particles.
+The resulting function provides an approximate measure of the likelihood of finding a jump at a given time.
+These are shown in figure~\ref{fig:example_kdest} for the example data set.
+
+\begin{figure}[!t]
+\centering
+\subfloat[]{\includegraphics[width=0.95\columnwidth]{example_filter_kdest.pdf}} \\
+\subfloat[]{\includegraphics[width=0.95\columnwidth]{example_smoother_kdest.pdf}}
+\caption{Filter (a) and smoother (b) kernel density estimates for value (top) and trend (bottom) jump times. Correct times overlayed as dotted vertical lines.}
+\label{fig:example_kdest}
+\end{figure}
+
+It is immediately apparent that the approximation produced by the filter lacks diversity.
+The jump time kernel density estimate comprises peaks with magnitude $1$, with regions of $0$ between.
+This is a result of every particle containing an almost identical list of changepoints, as a result of resampling.
+In contrast, the smoother is clearly providing a less degenerate particle representation, and generally contains larger kernel density peaks at the times of larger jumps.
+Furthermore, the kernel density for the trend jump times contains broader peaks, capturing the fact that because the trend is not observed directly, the jumps are harder to localise precisely.
+
+A further comparison of the algorithms is possible via the state estimates they generate.
+Here we can compare three options: the filtering results, using the VRPF and a Kalman filter for the state estimates; the filter-smoother results, using the final VRPF approximation for the changepoint sequence and a Rauch-Tung-Striebel (RTS) smoother for the state estimates; and the full smoothing results, using the VRPS followed by an RTS smoother.
+For the example data set, the trend estimates are shown in figure~\ref{fig:example_state} (The estimates of value are less informative, as this quantity is observed).
+Again, the improved particle diversity of the smoother is apparent.
+
+\begin{figure}[!t]
+\centering
+\subfloat[]{\includegraphics[width=0.95\columnwidth]{example_filter_state.pdf}} \\
+\subfloat[]{\includegraphics[width=0.95\columnwidth]{example_filtersmoother_state.pdf}} \\
+\subfloat[]{\includegraphics[width=0.95\columnwidth]{example_smoother_state.pdf}}
+\caption{Trend estimates using RBVRPF + Kalman filter (a) RBVRPF + RTS smoother (b) and RBVRPS + RTS smoother.
+Solid lines show means of each particle.
+Dashed lines show mean $\pm$2 standard deviations.}
+\label{fig:example_state}
+\end{figure}
+
+For a quantitative comparison, the algorithms were tested on $10$ realisations from the model, each of $1000$ time steps, and the following statistics were calculated for each:
+\begin{itemize}
+	\item The number of unique changepoint sequences. This is a measure of particle diversity of the approximation.
+	\item The number of unique changepoint times. Another measure of particle diversity.
+	%\item The optimal sub-pattern assignment (OSPA) distance between the maximum a posteriori (MAP) changepoint sequence and the true sequence.
+The OSPA is a concept from multiple target tracking introduced in \cite{Schuhmacher2008} to measure the distance between sets of points of varying cardinality.
+It is thus well suited to measuring the error between changepoint sequences of varying length.
+For each approximation, the sequence from the MAP particle was used as a point estimate.
+	%\item The root-mean-square error (RMSE) of the MAP state estimate.
+The MAP state estimate is taken to be the Gaussian mean of the conditional state distribution, from the Kalman smoother or RTS, selected from the particle with the highest posterior probability.
+	\item The root-mean-square error (RMSE) of the mean state estimate. The mean state estimate is the average of the Gaussian means from all the particles.
+\end{itemize}
+
+%Note that the MAP particle may be found by calculating the posterior probability of each. The terms required for this calculation will already have been evaluated during the filtering procedure.
+
+%\begin{IEEEeqnarray}{rCl}
+% p(\theta|y_{1:N}) & \propto & p(y_{1:N}|\theta) p(\theta) \nonumber \\
+%                         &         & S(\tau_K, T) \prod_{k=1}^K p(u_k) p(\tau_k|\tau_{k-1}) \prod_{n=1}^{N} p(y_n|\theta, y_{1:n-1})
+%\end{IEEEeqnarray}
+
+The results are shown in tables~\ref{tab:cp_performance} and ~\ref{tab:state_performance}.% The OSPA parameters are: exponent $p=1$, threshold $c=0.01$.
+
+\begin{table}%
+\caption{Changepoint sequence estimation performance.}
+\label{tab:cp_performance}
+\centering
+\renewcommand{\arraystretch}{1.5}
+\begin{tabular}{|c|c|c|}
+\hline
+ & RBVRPF & RBVRPS \\
+\hline \hline
+mean unique no. sequences   & 7.8      & 100       \\
+\hline
+mean unique no. jump times  & 47.8     & 1076.7    \\
+%MAP sequence OSPA    & $5.36 \times 10^{-3}$ & $4.98 \times 10^{-3}$  \\
+\hline
+\end{tabular}
+\end{table}
+
+\begin{table}%
+\caption{State estimation performance.}
+\label{tab:state_performance}
+\centering
+\renewcommand{\arraystretch}{1.5}
+\begin{tabular}{|c|c|c|c|}
+\hline
+ &\raisebox{0cm}[0.4cm][0.3cm]{\parbox[c]{1.5cm}{\centering RBVRPF and KF}} & \raisebox{0cm}[0.4cm][0.3cm]{\parbox[c]{1.5cm}{\centering RBVRPF and RTS}} & \raisebox{0cm}[0.4cm][0.3cm]{\parbox[c]{1.5cm}{\centering RBVRPS and RTS}} \\
+% & \begin{minipage}[c]{1.5cm}\centering RBVRPF + KF\end{minipage} & \begin{minipage}[c]{1.5cm}\centering RBVRPF + RTS\end{minipage} & \begin{minipage}[c]{1.5cm}\centering RBVRPS + RTS\end{minipage} \\
+\hline \hline
+\raisebox{0cm}[0.4cm][0.3cm]{\parbox[c]{2cm}{\centering mean value estimate RMSE}}   & $5.31 \times 10^{-4}$ & $4.48 \times 10^{-4}$ & $4.16 \times 10^{-4}$ \\
+\hline
+\raisebox{0cm}[0.4cm][0.3cm]{\parbox[c]{2cm}{\centering mean trend estimate RMSE}}   & $2.56 \times 10^{-2}$ & $1.79 \times 10^{-2}$ & $1.49 \times 10^{-2}$ \\
+%MAP value estimate RMSE    & $5.98 \times 10^{-4}$ & $4.48 \times 10^{-4}$ & $4.49 \times 10^{-4}$ \\
+%MAP trend estimate RMSE    & $3.13 \times 10^{-2}$ & $1.79 \times 10^{-2}$ & $1.74 \times 10^{-2}$ \\
+\hline
+\end{tabular}
+\end{table}
+
+The new RBVRPS algorithm outperforms the filter in terms of both state estimation accuracy and particle diversity.
+
+%\begin{meta}
+%Another interesting question is how the posterior distribution over the number of changepoints is expected to behave.
+The prior is poisson distributed with mean $\alpha T$.
+The means of both the filter and smoother particle posterior are consistently greater than the true number (by about 50\%).
+Is this expected and why?
+%\end{meta}
+
+Finally, we demonstrate the algorithms on a real financial data set, 1000 data points representing 1.7s of USD-GBP foreign exchange rate data from April 2008, using the same parameters as for the previous simulations.
+The data shown in Fig.~\ref{fig:fx_results}, along with jump time kernel density estimates from the filter and smoother.
+In this example, there is no ground truth against which to judge the results.
+However, the smoother does a pleasing job of estimating jump times in the same areas that one would if analysing the data ``by eye''.
+We submit this simply to illustrate the practical applications of the model and algorithm.
+
+\begin{figure}[!t]
+\centering
+\subfloat[]{\includegraphics[width=0.95\columnwidth]{fx_data.pdf}} \\
+\subfloat[]{\includegraphics[width=0.95\columnwidth]{fx_filter_kdest.pdf}} \\
+\subfloat[]{\includegraphics[width=0.95\columnwidth]{fx_smoother_kdest.pdf}}
+\caption{Foreign exchange data (a), along with filter (a) and smoother (b) kernel density estimates for the state and trend jump times.}
+\label{fig:fx_results}
+\end{figure}
 

### 2012-04-06 draft2_VRsmoothing.tex
Still polishing

--- 2012-03-16 draft1_VRsmoothing.tex
@@
 
-\author{Pete~Bunch,~\IEEEmembership{}
+\author{Pete~Bunch,~\IEEEmembership{Member,~IEEE,}
         Simon~Godsill,~\IEEEmembership{Member,~IEEE,}% <-this % stops a space
@@
 The abstract goes here.
+Blah, blah, blah, blah, blah, blah, blah, blah, blah, blah.
+Blah, blah, blah, blah, blah, blah, blah, blah, blah, blah.
+Blah, blah, blah, blah, blah, blah, blah, blah, blah, blah.
+Blah, blah, blah, blah, blah, blah, blah, blah, blah, blah.
+Blah, blah, blah, blah, blah, blah, blah, blah, blah, blah.
+Blah, blah, blah, blah, blah, blah, blah, blah, blah, blah.
+Blah, blah, blah, blah, blah, blah, blah, blah, blah, blah.
+Blah, blah, blah, blah, blah, blah, blah, blah, blah, blah.
+Blah, blah, blah, blah, blah, blah, blah, blah, blah, blah.
+Blah, blah, blah, blah, blah, blah, blah, blah, blah, blah.
 \end{abstract}
@@
 The standard HMM is poorly suited to systems where the state evolution contains discontinuities; for example, the price of a financial asset which may display large jumps at random times between periods of diffusion-like behaviour, or the kinematic state of a manoeuvring vehicle which may have sudden changes in the acceleration when turns begin or end.
-Such problems can be handled more naturally using a ``variable rate'' model, in which the state dynamics are conditioned upon a set of random changepoints which characterise transitions in behaviour.
+Such problems can be handled more naturally using a ``variable rate'' model, in which the state dynamics are conditioned upon a set of changepoints which characterise transitions in behaviour.
 
-In a variable rate model, the set of changepoints and associated parameters are modelled as a marked point process (MPP), the mathematical properties of which are thoroughly set out in \cite{Jacobsen2006}.
+In a variable rate model, the set of changepoints and associated parameters are modelled as a marked point process (MPP), the mathematical properties of which are set out thoroughly in \cite{Jacobsen2006}.
 Conditional upon this MPP the state evolves according to some benign dynamics.
@@
 The posterior distribution for the changepoint MPP is inherently nonlinear, and cannot be calculated analytically.
-Instead, inference must be conducted using numerical approximations.
+Instead, inference may be conducted using numerical approximations.
 The particle filter (introduced by \cite{Gordon1993}) and smoother (see \cite{Doucet2000a,Godsill2004}) are schemes which approximate a posterior distribution using a set of samples, or ``particles'', drawn sequentially from it.
@@
 
-We introduce the general structure of variable rate models in section~\ref{sec:vr_models} and discuss two particular types, those with conditionally linear-Gaussian and conditionally deterministic dynamics.
+We introduce the general structure of variable rate models in section~\ref{sec:vr_models}.
 The VRPF is reviewed in section~\ref{sec:vrpf}, and new variable rate smoothing algorithms are described in section~\ref{sec:vrps}.
-In section~\ref{sec:simuations}, particular examples of variable rate models are presented and their performance demonstrated in a series of simulations.
+In section~\ref{sec:simulations}, particular examples of variable rate models are presented and algorithm performance is demonstrated in a series of simulations.
 
@@
 The pairs $\{\tau_k, u_k\}$ are the elements of an marked point process (MPP).
-We will refer to the times of the MPP as the ``change-time sequence'' and marks as ``parameter sequence'', and both together as the ``changepoint sequence''.
 The latent state is a continuous-time process denoted $x(t)$.
@@
 \begin{IEEEeqnarray}{rCl}
- \{\tau_k, u_k\} & \sim & p(u_k|\tau_k, \tau_{k-1}, u_{k-1}) p(\tau_k|\tau_{k-1}, u_{k-1}) \label{eq:cp_model}
+ \{\tau_k, u_k\} & \sim & p(\tau_k, u_k|\tau_{k-1}, u_{k-1}) \label{eq:cp_model}
 \end{IEEEeqnarray}
 
-The changepoint density will be constructed such that $P(\tau_k < \tau_{k-1}) = 0$.
+This density will be constructed such that $P(\tau_k < \tau_{k-1}) = 0$.
 
@@
 \begin{IEEEeqnarray}{rCll}
- x(t) & = & f(x_{K_n}, v_{K_n}, \tau_{K_n}, t) &, \tau_{K_n} < t <= \tau_{K_{n}+1}    \label{eq:disc_time_state_diff_eq}     .
+ x(t) & = & f(x_{K_n}, u_{K_n}, \tau_{K_n}, t) &, \tau_{K_n} < t \leq \tau_{K_{n}+1}    \label{eq:disc_time_state_diff_eq}     .
 \end{IEEEeqnarray}
@@
 By choosing $t = \tau_{K_{n}+1}$, this equation specifies the state at the next changepoint time.
-Similarly, by choosing $t=t_n$, the state at the observation times may be calculated.
-These points will be denoted $\hat{x}_n$.
-To complete the model, a probabilistic measurement model must be chosen for the observation process, $p(y_n|\hat{x}_n)$.
+Similarly, by choosing $t=t_n$, the state at the observation times may be calculated --- these points will be denoted $\hat{x}_n$.
+To complete the model, a probabilistic measurement model must be devised for the observation process, $p(y_n|\hat{x}_n)$.
 
@@
 This means that $x(t)$ may be calculated deterministically for all $t$ given $\theta$.
-However, this condition is easily relaxed by including $x_0$ as a random variable in the posterior distribution.
+This condition is easily relaxed by including $x_0$ as a random variable in the posterior distribution.
 
@@
 radar measurements) are made.
-In \cite{Godsill2007a,Godsill2007}, variable rate models were introduced for tracking, in which the state trajectory is divided up by a set of changepoints between which the motion follows a deterministic path governed by motion parameters (accelerations, etc.) which are fixed for that division.
+In \cite{Godsill2007a,Godsill2007,Whiteley2011}, variable rate models were introduced for tracking, in which the state trajectory is divided up by a set of changepoints between which the motion follows a deterministic path governed by motion parameters (accelerations, etc.) which are fixed for that division.
 
@@
 
-The variable rate particle filter (VRPF) is described in \cite{Godsill2007,Godsill2007a}.
+The variable rate particle filter (VRPF) is described in \cite{Godsill2007,Godsill2007a,Whiteley2011}.
 The objective of the algorithm is to sequentially estimate the posterior distribution of the changepoint sequence, $p(\theta_{n}| y_{1:n})$, at each time $t_n$.
-This distribution may be expanded using Bayes' rule.
+This distribution may be expanded using Bayes' rule,
 %
 \begin{IEEEeqnarray}{rCl}
-\IEEEeqnarraymulticol{3}{l}{p(\theta_{n}|y_{1:n})} \nonumber \\
- & \propto & p(y_n|\theta_{n}, y_{1:n-1}) p(\theta_{n \setminus n-1}|\theta_{n-1}) p(\theta_{n-1}|y_{1:n-1}) \label{eq:vrpf_target}
+\IEEEeqnarraymulticol{3}{l}{ p(\theta_{n}|y_{1:n}) \propto p(y_n|\theta_{n}, y_{1:n-1}) } \nonumber \\
+ \qquad & & \times p(\theta_{n \setminus n-1}|\theta_{n-1}) p(\theta_{n-1}|y_{1:n-1}) \label{eq:vrpf_target}     .
 \end{IEEEeqnarray}
 
-The transition term, $p(\theta_{n \setminus n-1} | \theta_{n-1})$, can be constructed in a similar manner to (\ref{eq:cp_sequence_prior}) \cite{Jacobsen2006}, as a product of density terms for each new changepoint and a survivor function term.
-In the particular (but common) case that no new changepoints occur within the interval, the density consists of a probability mass on the empty set, with weight $P(\tau_{K_n+1} > t_n | \tau_{K_n+1} > t_{n-1}, \tau_{K_n} )$.
+The transition term, $p(\theta_{n \setminus n-1} | \theta_{n-1})$, has a similar form to the changepoint prior of (\ref{eq:cp_sequence_prior}) \cite{Jacobsen2006}, comprising a product of density terms (one for each new changepoint) and a survivor function term.% In the particular (but common) case that no new changepoints occur within the interval, the density consists of a probability mass on the empty set, with weight $P(\tau_{K_n+1} > t_n | \tau_{K_n+1} > t_{n-1}, \tau_{K_n} )$.
 %
@@
 \IEEEeqnarraymulticol{3}{l}{p(\theta_{n \setminus n-1} | \theta_{n-1})} \nonumber \\
+    & = & p(\tau_{K_{n-1}+1:K_n}, u_{K_{n-1}+1:K_n}, \tau_{K_n+1}>t_n|\tau_{K_{n-1}+1}>t_{n-1}, \tau_{K_{n-1}}, u_{K_{n-1}}) \nonumber \\
     & = & \begin{cases} S(\tau_{K_n}, t_n) \prod\limits_{j:t_{n-1} \leq \tau_j < t_n} p(\tau_j, u_j| \tau_{j-1}, u_{j-1}, \tau_j>t_{n-1}) & K_n > K_{n-1} \\
@@
 For all but the first changepoint in the interval, the density is given by the prior model of (\ref{eq:cp_model}).
-For the first changepoint, indexed by $k=K_{n-1}+1$, we must account for the fact that a changepoint cannot occur before $t_{n-1}$,
+For the first changepoint, indexed by $k=K_{n-1}+1$, we must account for the fact that it cannot occur before $t_{n-1}$,
 %
@@
 
-Next, an extension to the changepoint sequence, $\theta_{n \setminus n-1}^{(i)}$, is proposed from an importance distribution, $q(\theta_{n \setminus n-1}|\theta_{n-1}^{(i)}, y_n)$, and concatenated to $\theta_{n-1}$ to create an estimate of $\theta_n$.
+Next, an extension to the changepoint sequence, $\theta_{n \setminus n-1}^{(i)}$, is proposed from an importance distribution, $q(\theta_{n \setminus n-1}|\theta_{n-1}^{(i)}, y_n)$, and concatenated with $\theta_{n-1}$ to create an estimate of $\theta_n$.
 Finally, the particle is weighted according to the ratio of the target and proposal densities.
@@
 
-The choice of proposal weights, $\{v_{n-1}^{(i)}\}$, requires particular attention in the design of VRPFs.
+%The choice of proposal weights, $\{v_{n-1}^{(i)}\}$, requires particular attention in the design of VRPFs.
 In some models a changepoint may not have an immediate effect on the observations, especially if a jump occurs in some quantity which is only observed via its integral, e.g.
@@
 One scheme which has been found to work well is described in \cite{Godsill2007}, in which proposal weights are given by:
+
+In \cite{Godsill2007}, the following choice of proposal weights was found to work well, as it preserves low probability particles which might turn out to be good estimates at later time steps,
 %
 \begin{IEEEeqnarray}{rCl}
-v_{n-1}^{(i)} & \propto & \max ( 1, N_F w_{n-1}^{(i)} )
+v_{n-1}^{(i)} & \propto & \max ( 1, N_F w_{n-1}^{(i)} )     .
 \end{IEEEeqnarray}
@@
 Conveniently, the Kalman filter also provides us with an estimate of the current state given the changepoint sequence and all the preceding observations.
-It follows from the the Gaussian dynamics and prior that these distributions are all Gaussian as well, \cite{Anderson1979},
+It follows from the the Gaussian dynamics and prior that these distributions are all Gaussian as well, \cite{Grewal2002},
 %
@@
 
-This completes the requirements for the particle filter, resulting in the final algorithm shown in Fig.~\ref{alg:RBVRPF}.
-The concept of estimating the posterior for just the changepoint sequence with a particle filter rather than the joint posterior over changepoints and linear states is an example of Rao-Blackwellisation (see, e.g.
+This completes the requirements for the particle filter, resulting in the final algorithm shown in Fig.~\ref{alg:RBVRPF}.% The concept of estimating the posterior for just the changepoint sequence with a particle filter rather than the joint posterior over changepoints and linear states is an example of Rao-Blackwellisation (see, e.g.
 \cite{Casella1996,Doucet2000}), so this algorithm is known as the Rao-Blackwellised variable rate particle filter (RB-VRPF) \cite{Godsill2007a,Christensen2012}.
@@
 \begin{algorithmic}[1]
-\STATE For each $i$, initialise particle changepoint sequence with $\{\theta_{0}^{(i)}\} \gets \{\tau_0^{(i)}, u_0^{(i)}$, where $\tau_0^{(i)}=0$ and $\{u_0^{(i)}\} \sim p(u_0)$.
+\linespread{1.7} \selectfont
+\STATE For each $i$, initialise particle changepoint sequence with $\{\theta_{0}^{(i)}\} \gets \{\tau_0^{(i)}, u_0^{(i)}\}$, where $\tau_0^{(i)}=0$ and $u_0^{(i)} \sim p(u_0)$.
 \STATE For each $i$, initialise particle sufficient statistics, $m_0^{(i)}$ and $P_0^{(i)}$ with prior values.
@@
   \FOR{$i=1 \dots N_F$}
-  	\STATE Sample a changepoint history $\theta_{n-1}^{(i)} \sim \sum_j v_{n-1}^{(j)} \delta_{\theta_{n-1}^{(j)}}(\theta_{n-1})$.
-    \STATE Propose a sequence extension $\theta_{n \setminus n-1}^{(i)} \sim q(\theta_{n \setminus n-1} | \theta_{n-1}^{(i)})$.
-    \STATE Add extension to changepoint sequence $\theta_n^{(i)} \gets \theta_{n-1}^{(i)} \cup \theta_{n \setminus n-1}^{(i)}$.
-    \STATE Predict state mean and covariance $\mu_n^{(i)}$ and $S_n^{(i)}$ using (\ref{eq:kf_predict_start})~to~(\ref{eq:kf_predict_stop}).
+  	\STATE Sample a history $\theta_{n-1}^{(i)} \sim \sum_j v_{n-1}^{(j)} \delta_{\theta_{n-1}^{(j)}}(\theta_{n-1})$.
+    \STATE Propose an extension $\theta_{n \setminus n-1}^{(i)} \sim q(\theta_{n \setminus n-1} | \theta_{n-1}^{(i)})$.
+    \STATE Add extension to sequence $\theta_n^{(i)} \gets \theta_{n-1}^{(i)} \cup \theta_{n \setminus n-1}^{(i)}$.
+    \STATE Predict observation mean and covariance $\mu_n^{(i)}$ and $S_n^{(i)}$ using (\ref{eq:kf_predict_start})~to~(\ref{eq:kf_predict_stop}).
     \STATE Calculate weight $w_n^{(i)}$ using (\ref{eq:vrpf_weights}).
@@
 
-This leads to a particle filter for variable rate models with conditionally deterministic dynamics, summarised in Fig.~\ref{alg:VRPF}.
-The resulting trajectories $x(t)^{(i)}$ for each particle are realisations from a piecewise-deterministic process, so this algorithm is named the piecewise-deterministic variable rate particle filter (PD-VRPF).
+This leads to a particle filter for variable rate models with conditionally deterministic dynamics, summarised in Fig.~\ref{alg:VRPF}.% The resulting trajectories $x(t)^{(i)}$ for each particle are realisations from a piecewise-deterministic process, so this algorithm is named the piecewise-deterministic variable rate particle filter (PD-VRPF).
 
@@
 \begin{algorithmic}[1]
-\STATE For each $i$, initialise particle changepoint sequence with $\{\theta_{0}^{(i)}\} \gets \{\tau_0^{(i)}, u_0^{(i)}$, where $\tau_0^{(i)}=0$ and $\{u_0^{(i)}\} \sim p(u_0)$.
+\linespread{1.7} \selectfont
+\STATE For each $i$, initialise particle changepoint sequence with $\{\theta_{0}^{(i)}\} \gets \{\tau_0^{(i)}, u_0^{(i)}\}$, where $\tau_0^{(i)}=0$ and $u_0^{(i)} \sim p(u_0)$.
 \FOR{$n=1 \dots N$}
@@
   	\STATE Sample a history $\theta_{n-1}^{(i)} \sim \sum_j v_{n-1}^{(j)} \delta_{\theta_{n-1}^{(j)}}(\theta_{n-1})$.
-    \STATE Propose a sequence extension $\theta_{n \setminus n-1}^{(i)} \sim q(\theta_{n \setminus n-1}|\theta_{n-1}^{(i)}, y_{n})$.
-    \STATE Add extension to changepoint sequence $\theta_n^{(i)} \gets \theta_{n-1}^{(i)} \cup \theta_{n \setminus n-1}^{(i)}$.
+    \STATE Propose an extension $\theta_{n \setminus n-1}^{(i)} \sim q(\theta_{n \setminus n-1}|\theta_{n-1}^{(i)}, y_{n})$.
+    \STATE Add extension to sequence $\theta_n^{(i)} \gets \theta_{n-1}^{(i)} \cup \theta_{n \setminus n-1}^{(i)}$.
     \STATE Calculate state $\hat{x}_n$ using (\ref{eq:disc_time_state_diff_eq}).
-    \STATE Calculate particle weight using (\ref{eq:vrpf_weights}).
+    \STATE Calculate weight $w_n^{(i)}$ using (\ref{eq:vrpf_weights}).
   \ENDFOR
@@
 
-The bootstrap versions of the VRPFs described in sections~\ref{sec:rb-vrpf} and~\ref{sec:pd-vrpf} may perform poorly if changepoints are not obvious until significantly after they occur.
+The bootstrap versions of the VRPF may perform poorly if changepoints are not obvious until significantly after they occur.
 For example, in a tracking example, if a jump occurs in the acceleration, but only the position is observed, then this change may not be obvious until a number of observations have arrived.
@@
 
-Filtering schemes which alter past changepoints -- whether using an SMC sampler or RM -- are computationally expensive, because many likelihood calculations must be conducted, for each observation from the time of the change onwards.
+%Filtering schemes which alter past changepoints -- whether using an SMC sampler or RM -- are computationally expensive, because many likelihood calculations must be conducted, for each observation from the time of the change onwards.
 In some cases, it may be simpler to just use a bootstrap filter with more particles.
@@
 A filter conducts inference sequentially as new observations are introduced.
-The purpose of a smoother is to produce a second estimate once all the observations have been made, using future values to improve upon the filter performance.
+The purpose of a smoother is to produce a revised estimate once all the observations have been made, using future values to improve upon the filter performance.
 Estimating changepoints online is a challenging task because the presence of a change may not be obvious until after it has happened.
-Thus, it is anticipated that a smoothing algorithm will provide significantly improved performance at changepoint estimation.
+Thus, a smoothing algorithm is expected to achieve significantly improved performance at changepoint estimation.
 
 The target distribution for a variable rate smoothing algorithm is the posterior distribution over the entire changepoint sequence, $p(\theta|y_{1:N})$.
-A particle approximation to this distribution is generated by the final step of the VRPF.
-However, in the same manner as the fixed rate filter-smoother of \cite{Kitagawa1996}, this approximation is likely to lack path-space diversity -- because of the necessary resampling step in the filtering algorithm, the particles all share the same set of changepoints before a particular time, with variation only appearing for changepoints closer to $T$.
-For a good characterisation of the smoothing distribution, it is necessary to rejuvenate the set of particles.
+This is the distribution approximated by the final step of the VRPF.
+However, in the same manner as the fixed rate filter-smoother of \cite{Kitagawa1996}, this approximation is likely to lack path-space diversity --- because of the necessary resampling step in the filtering algorithm, the particles all share the same set of changepoints before a particular time, with variation only appearing for changepoints closer to $T$.
+For a better characterisation of the smoothing distribution, it is necessary to rejuvenate the set of particles.
 This is achieved with a backward pass through the observations in a similar manner to the forward-backward method described in \cite{Godsill2004}.
@@
 It remains to devise a method for drawing samples from the backwards conditional distribution, $p(\theta_{n}|\theta_{n}^{+}, y_{1:N})$.
-Such methods are discussed in the following sections for the two classes of models.
+The simplifications used by the fixed rate particle smoother which exploit the Markovian nature of the state sequence \cite{Godsill2004} are of no help, because past changepoints are not independent of future observations (i.e.
+$p(\theta_{n}|\theta_{n}^{+}, y_{1:N}) \ne p(\theta_{n}|\theta_{n}^{+}, y_{1:n}))$.
+In the following sections, alternative methods for sampling this backwards conditional distribution are discussed for the two classes of model.
 
@@
 \begin{IEEEeqnarray}{rCl}
- \tilde{w}_n & \propto & \int p(y_{n+1:N}|x_n, \theta_{n}^+) \nonumber \\
+ \tilde{w}_n & \propto & w_n \int p(y_{n+1:N}|x_n, \theta_{n}^+) \nonumber \\
              &         & \times p(x_n|\theta_{n}^{(i)}, y_{1:n}) dx_n p(\theta_{n}^+|\theta_{n}^{(i)}) \label{eq:rb-vrps_back_cond_weight}
@@
  p(\theta_{n}^+|\theta_{n}) &=      & p(\tau_{K_n+1:K}, u_{K_n+1:K}|\tau_{K_n}, u_{K_n}, \tau_{K_n+1}>t_n) \nonumber \\
-                                    &\propto& p(u_{K_n+1}|\tau_{K_n+1}, \tau_{K_n}, u_{K_n}) p(\tau_{K_n+1}|\tau_{K_n}, u_{K_n})     .
+                                    &\propto& p(\tau_{K_n+1}, u_{K_n+1}|\tau_{K_n}, u_{K_n}, \tau_{K_n+1}>t_n)     .
 \end{IEEEeqnarray}
@@
 The second state term in (\ref{eq:rb-vrps_back_cond_weight}) is the likelihood, $p(y_{n+1:N}|x_n, \tilde{\theta}_{n}^+)$.
-This may be considered to be an improper density over $x_n$, and may be calculated analytically using a backwards Kalman filter, in a similar manner to that used in the two-filter smoother \cite{Fraser1969,Anderson1979,Sarkka2012}.
+This may be considered to be an improper density over $x_n$, and may be calculated analytically using a backwards Kalman filter, in a similar manner to that used in the two-filter smoother \cite{Fraser1969,Kitagawa1994,Sarkka2012,Gelb1974}.
 Such a backwards Kalman filter uses the following recursions.
-Details are provided in appendix~\ref{app:backward_filter}, and in the aforesaid references.
+Derivations may be found in the aforesaid references.
 %
@@
 \end{IEEEeqnarray}
-
+%
 \begin{IEEEeqnarray}{rCl}
@@
 
-Exact methods for initialising this recursion are discussed in appendix~\ref{app:backward_filter}.
+Exact methods for initialising this recursion are discussed in \cite{Kitagawa1994}.
 However, it is often sufficient to use an approximation, for example $\tilde{m}_N^{(i)} = m_N^{(i)}$ and $\tilde{P}_N^{(i)} = P_N^{(i)}$.
@@
 \begin{equation}
- \tilde{w}_n \propto p(\theta_{n}^+|\theta_{n}^{(i)}) \mathcal{N}(\tilde{m}_n^-|m_n, \tilde{P}_n^- + P_n)     .
+ \tilde{w}_n \propto w_n p(\theta_{n}^+|\theta_{n}^{(i)}) \mathcal{N}(\tilde{m}_n^-|m_n, \tilde{P}_n^- + P_n)     .
 \label{eq:rb-vrps_back_cond_weight2}
@@
 The algorithmic complexity of this variable rate particle smoother for conditionally linear-Gaussian models is $O(N_F \times N_S \times N)$.
-It is possible to reduce this by using an MCMC sampling scheme in the style of \cite{Bunch2012}, in which case it is no longer necessary to calculate the backwards sampling weights for every filter particle.
+It is possible to reduce this by using an MCMC sampling scheme in the style of \cite{Bunch2012}, which avoids the necessity of calculating the sampling weights for every filter particle.
 
@@
 
-In a piecewise-deterministic system, the trick of introducing a state variable to split the set of observations into past and future does not work.
-Furthermore, the simplifications used by the fixed rate particle smoother which exploit the Markovian nature of the state sequence \cite{Godsill2004} are of no help, because past changepoints are not independent of future observations (i.e.
-$p(\theta_{n}|\theta_{n}^{+}, y_{1:N}) \ne p(\theta_{n}|\theta_{n}^{+}, y_{1:n}))$.
-A new approach is required.
+In a conditionally deterministic system, the entire continuous time state trajectory, $x(t)$, is a function of the changepoint sequence $\theta$ and the initial state, $x_0$.
+When the changepoint sequence is split into past, $\theta_n$, and future $\theta_n^+$, and the past is altered, this will alter the state trajectory in the future.
+This is undesirable, as each sampling operation will require the recalculation of state values and observations likelihoods for all future observation times.
 
-Consider the augmented MPP consisting of $\tilde{\theta} = \{\tau_{0:K}, u_{0:K}, x_{0:K}\}$ ($\tilde{\theta}_n$ and $\tilde{\theta}_n^+$ defined in the same manner as before).
-The mark corresponding to $\tau_k$ is now the parameter-state pair $(u_k, x_k)$, i.e.
-the state at the time of the change plus the parameters for the next division.
-The smoother is formulated by factorising the augmented sequence posterior distribution as before,
+In order to achieve conditional independence between past and future observations, consider the augmented changepoint sequence, $\tilde{\theta} = \{\tau_{0:K}, u_{0:K}, x_{0:K}\}$ ($\tilde{\theta}_n$ and $\tilde{\theta}_n^+$ defined in the same manner as before).
+The changepoint state values, $x_{0:K}$ may be calculated deterministically from $\theta$ for each particle.
+The smoother is then formulated by factorising the augmented sequence posterior distribution as before,
 %
@@
 
-The backwards conditional term may be expanded as,
+Again, this suggests a backwards sequential sampling algorithm for smoothing. The backwards conditional term may be expanded as,
 %
 \begin{IEEEeqnarray}{rCl}
-p(\tilde{\theta}_{n}|\tilde{\theta}_{n}^{+}, y_{1:N}) & \propto & p(y_{1:N}|\tilde{\theta}_{n}, \tilde{\theta}_{n}^{+}) p(\tilde{\theta}_{n}^{+}|\tilde{\theta}_{n}) p(\tilde{\theta}_{n})    .
+p(\tilde{\theta}_{n}|\tilde{\theta}_{n}^{+}, y_{1:N}) & \propto & p(y_{n+1:N}|\tilde{\theta}_{n}, \tilde{\theta}_{n}^{+}) p(\tilde{\theta}_{n}^{+}|\tilde{\theta}_{n}) p(\tilde{\theta}_{n}|y_{1:n})    .
 \end{IEEEeqnarray}
 
-For each particle in the filtering approximations, the changepoint state values $x_{0:K_n}$ may be calculated deterministically from $\theta_n$ and stored during the filtering procedure.
-The obvious strategy then is to recursively sample $\tilde{\theta}_{n}$ from the backwards conditional term, using the re-weighted particles of the $n$th filtering approximation.
-However, the conditionally deterministic structure prevents this from working.
-$x(t)$ is fully determined by $\tilde{\theta}_{n}$ up to $\tau_{K_n+1}$, but this will not usually match up with $x_{K_n+1}$ as specified in $\tilde{\theta}_{n}^+$.
-The result is that $p(\tilde{\theta}_{n}^{+}|\tilde{\theta}_{n})=0$ for (almost) all particles.
-In other words, we are trying to join two deterministic state trajectories together, but they do not meet up in the middle.
+This presents a problem. We cannot substitute the particle approximation for the filtering distribution because the transition density term for the augmented sequences is given by,
+%
+\begin{IEEEeqnarray}{rCl}
+p(\tilde{\theta}_{n}^+ | \tilde{\theta}_{n}) & \propto & p(\tau_{K_n+1}, u_{K_n+1}| \tau_{K_n}, u_{K_n}) \delta_{f(x_{K_n}, u_{K_n}, \tau_{K_n}, \tau_{K_n+1})}(x_{K_n+1}),
+\end{IEEEeqnarray}
+
+which will be $0$ for almost all filtering particles. Intuitively, we are trying to join two deterministic state trajectories together, but they do not meet up in the middle.
 
 The solution to this problem is provided by an idea from the Sequential Monte Carlo (SMC) samplers of \cite{DelMoral2006}.
-Rather than just using a value of $\tilde{\theta}_{n}$ from the filtering particles, a replacement, $u_{K_n}'$, is proposed for the existing parameter corresponding to the final changepoint with a value chosen such that the past and future paths meet up.
+Rather than sampling $p(\tilde{\theta}_{n}|\tilde{\theta}_{n}^{+}, y_{1:N})$ directly, a value of $\tilde{\theta}_{n}$ is proposed from the filtering particles, after which a replacement, $u_{K_n}'$, is proposed for the existing parameter corresponding to the final changepoint, with a value chosen such that the past and future paths meet up.
 The modified sequence is denoted $\tilde{\theta}_{n}'$.
-Such a proposal may be written as,
-
+Such a proposal density may be written as,
+%
 \begin{IEEEeqnarray}{rCl}
@@
 
-For this proposal to work, a new condition to be imposed on the state dynamics.
+The target distribution is augmented by introduction of an artificial density to cover the new changepoint sequence and the discarded parameter.
+%
+\begin{IEEEeqnarray}{rCl}
+\IEEEeqnarraymulticol{3}{l}{ p(\tilde{\theta}_{n}'|\tilde{\theta}_{n}^{+}, y_{1:N}) L(u_{K_n}|\tilde{\theta}_{n}', \tilde{\theta}_{n}^{+}, y_{1:N}) } \nonumber \\
+ & \propto & p(y_{1:N}|\tilde{\theta}_{n}', \tilde{\theta}_{n}^+) p(\tilde{\theta}_{n}^{+}|\tilde{\theta}_{n}') p(\tilde{\theta}_{n}') L(u_{K_n}|\tilde{\theta}_{n}', \tilde{\theta}_{n}^{+}, y_{1:N})  \label{eq:pd-vrps_target}     .
+\end{IEEEeqnarray}
+
+Clearly this new target distribution admits the desired posterior as a marginal.
+
+For this method to work, a new condition must be imposed on the state dynamics.
 For any pair of adjacent changepoints, $(\tau_k,x_k)$, and $(\tau_{k+1},x_{k+1})$ it must be possible to calculate a parameter value which results in the transition from the former to the latter.
@@
 
-The target distribution is augmented by introduction of an artificial density to cover the new changepoint sequence and the discarded parameter.
-
-\begin{IEEEeqnarray}{rCl}
- p(\tilde{\theta}_{n}'|\tilde{\theta}_{n}^{+}, y_{1:N}) L(u_{K_n}|\tilde{\theta}_{n}', \tilde{\theta}_{n}^{+}, y_{1:N}) \label{eq:pd-vrps_target}     ,
-\end{IEEEeqnarray}
-
-It is clear that this new target distribution admits the desired posterior term as a marginal.
-
-The augmented target distribution may be sampled using Markov chain Monte Carlo (MCMC) \cite{Gilks1996}, in a similar manner to the fixed rate smoother of \cite{Bunch2012}.
+The augmented target distribution may be sampled using Markov chain Monte Carlo (MCMC) \cite{Gilks1996}, in a similar manner to the fixed rate smoother of \cite{Bunch2012} (IS could be used, but this would require resampling steps, which would lead to a loss of path-space diversity).
 Metropolis-Hastings (MH) steps are conducted to draw samples, $\{\tilde{\theta}_{n}, u_{K_n}'\}$, from the target distribution by sampling the proposal and accepting the new values with a given probability.
@@
 The ratio of target (\ref{eq:pd-vrps_target}) and proposal (\ref{eq:pd-vrps_proposal}) densities is given by,
-
+%
 \begin{IEEEeqnarray}{rCl}
@@
 
-The transition density term will include a delta function with unit probability mass at the point for which the past and future state trajectories match up.
+%The transition density term will include a delta function with unit probability mass at the point for which the past and future state trajectories match up.
 
-\begin{IEEEeqnarray}{rCl}
-\IEEEeqnarraymulticol{3}{l}{p(\tilde{\theta}_{n}^{+}|\tilde{\theta}_{n}') \propto p(u_{K_n+1}|\tau_{K_n+1}, \tau_{K_n}, u_{K_n})} \nonumber \\ \times p(\tau_{K_n+1}|\tau_{K_n}, u_{K_n}) \delta_{f(x_{K_n}, v_{K_n}, \tau_{K_n}, \tau_{K_n+1})}(x_{K_n+1})
-\end{IEEEeqnarray}
+%\begin{IEEEeqnarray}{rCl}
+%\IEEEeqnarraymulticol{3}{l}{p(\tilde{\theta}_{n}^{+}|\tilde{\theta}_{n}') \propto p(u_{K_n+1}|\tau_{K_n+1}, \tau_{K_n}, u_{K_n})} \nonumber \\ \times p(\tau_{K_n+1}|\tau_{K_n}, u_{K_n}) \delta_{f(x_{K_n}, v_{K_n}, \tau_{K_n}, \tau_{K_n+1})}(x_{K_n+1})
+%\end{IEEEeqnarray}
 
-If there is only one value of $u_{K_n}'$ for which the two trajectories match up, then the proposal density $q(u_{K_n}'|\tilde{\theta}_{n}, \tilde{\theta}_{n}^{+}, y_{1:N})$ is also a delta function, with unit probability mass at the one valid value.
+The proposal density $q(u_{K_n}'|\tilde{\theta}_{n}, \tilde{\theta}_{n}^{+}, y_{1:N})$ is constructed so as to be non-zero only where $p(\tilde{\theta}_{n}^{+}|\tilde{\theta}_{n}')$ is non-zero.
+This may require it to be a delta function if there is only one feasible value.
+The artificial conditional term, $L(u_{K_n}|\tilde{\theta}_{n}, \tilde{\theta}_{n}^{+}, y_{1:N})$, may be any arbitrary density, but the simplest choice is simply as a uniform distribution so that it cancels out in the acceptance probability.
 The remaining terms are likelihoods and parameter transition density terms, defined by the system equations.
-The simplest choice for $L(.|.)$ is simply as a uniform distribution so that it cancels out in the acceptance probability.
 
 If $\beta_n^{(m-1)}$ is the target-proposal ratio for the current state in the chain, and $\beta_n^*$ that for the proposed state, then the acceptance probability is given by,
-
+%
 \begin{IEEEeqnarray}{rCl}
@@
 
-The Markov chains require no burn-in, because the original value is itself a sample from the target distribution.
+The Markov chains require no burn-in, because the starting value is itself a sample from the target distribution.
 The number of MH steps, $M$, at each observation time, $n$, allows a trade-off of performance against time.
 Larger values of $M$ will ensure more unique samples in the smoothing approximation, but will also take longer to execute.
+See \cite{Bunch2012} for analysis of this trade-off on a fixed rate model.
 In the simulations discussed in this report, $M=1$ was used throughout.
 
-\section{Simulations}
+\section{Simulations} \label{sec:simulations}
 
@@
 \begin{IEEEeqnarray}{c}
- d\mathbf{x}(t) = \begin{bmatrix}0 & 1 \\ 0 & -\lambda \end{bmatrix} \mathbf{x}(t) dt + \begin{bmatrix}0 \\ \sigma \end{bmatrix} d\mathbf{\beta}(t) + d\mathbf{J}(t)
+ d\mathbf{x}(t) = \begin{bmatrix}0 & 1 \\ 0 & -\lambda \end{bmatrix} \mathbf{x}(t) dt + \begin{bmatrix}0 \\ \sigma \end{bmatrix} d\beta(t) + d\mathbf{J}(t)
 \end{IEEEeqnarray}
@@
 
-where the $\mathbf{w}_b$ and $v_n$ are Gaussian random variables with covariance matrixes $Q_n$ and $R$ respectively. The time between observations times is denotes $\Delta t = t_{n} - t_{n-1}$.
+where the $\mathbf{w}_n$ and $v_n$ are Gaussian random variables with covariance matrixes $Q_n$ and $R$ respectively. The time between observations times is denotes $\Delta t = t_{n} - t_{n-1}$.
 %
@@
 %
-\begin{IEEEeqnarray}{rCl}
- \IEEEeqnarraymulticol{3}{l}{p(\tau_k, u_k|\tau_{k-1}, u_{k-1}) = P(u_k) p(\tau_k|\tau_{k-1})} \\
- p(\tau_k|\tau_{k-1})               &=& \begin{cases}\frac{1}{\alpha} \exp ( \alpha (\tau_k-\tau_{k-1}) ) & \tau_k>\tau_{k-1} \\
-                                                     0 & \tau_k < \tau_{k-1} \end{cases} \\
- P(u_k)                             &=& \begin{cases}0.5 & u_k = 1 \\ 0.5 & u_k = 2\end{cases}
-\end{IEEEeqnarray}
+%\begin{IEEEeqnarray}{rCl}
+% \IEEEeqnarraymulticol{3}{l}{p(\tau_k, u_k|\tau_{k-1}, u_{k-1}) = P(u_k) p(\tau_k|\tau_{k-1})} \\
+% p(\tau_k|\tau_{k-1})               &=& \begin{cases}\frac{1}{\alpha} \exp ( \alpha (\tau_k-\tau_{k-1}) ) & \tau_k>\tau_{k-1} \\
+%                                                     0 & \tau_k < \tau_{k-1} \end{cases} \\
+% P(u_k)                             &=& \begin{cases}0.5 & u_k = 1 \\ 0.5 & u_k = 2\end{cases}
+%\end{IEEEeqnarray}
 
@@
 
-The results are shown in tables~\ref{tab:cp_performance} and ~\ref{tab:state_performance}.% The OSPA parameters are: exponent $p=1$, threshold $c=0.01$.
+The results are shown in tables~\ref{tab:finance_cp_performance} and ~\ref{tab:finance_state_performance}.% The OSPA parameters are: exponent $p=1$, threshold $c=0.01$.
 
@@
 \caption{Changepoint sequence estimation performance.}
-\label{tab:cp_performance}
+\label{tab:finance_cp_performance}
 \centering
@@
 \caption{State estimation performance.}
-\label{tab:state_performance}
+\label{tab:finance_state_performance}
 \centering
@@
 
-Finally, we demonstrate the algorithms on a real financial data set, 1000 data points representing 1.7s of USD-GBP foreign exchange rate data from April 2008, using the same parameters as for the previous simulations.
+%Finally, we demonstrate the algorithms on a real financial data set, 1000 data points representing 1.7s of USD-GBP foreign exchange rate data from April 2008, using the same parameters as for the previous simulations.
 The data shown in Fig.~\ref{fig:fx_results}, along with jump time kernel density estimates from the filter and smoother.
@@
 
+%\begin{figure}[!t]
+%\centering
+%\subfloat[]{\includegraphics[width=0.95\columnwidth]{fx_data.pdf}} \\
+%\subfloat[]{\includegraphics[width=0.95\columnwidth]{fx_filter_kdest.pdf}} \\
+%\subfloat[]{\includegraphics[width=0.95\columnwidth]{fx_smoother_kdest.pdf}}
+%\caption{Foreign exchange data (a), along with filter (b) and smoother (c) kernel density estimates for the state and trend jump times.}
+%\label{fig:fx_results}
+%\end{figure}
+
+\subsection{Conditionally Deterministic Model}
+
+The piecewise-deterministic VRPS was tested using a 2-dimensional dynamic model for manoeuvring vehicles \cite{Bunch2012a}.
+The target is modelled as a particle subject to perpendicular forces tangential and normal to the direction of motion.
+These forces are held constant between changepoint times, $\tau_{1:K}$, which represent the beginnings and ends of manoeuvres.
+
+The state vector consists of cartesian position, $x(t)$ and $y(t)$, plus bearing, $\psi(t)$, and speed, $\dot{s}(t)$.
+%
+\begin{equation}
+\mathbf{x}(t) = [x(t), y(t), \psi(t), \dot{s}(t)]^T
+\end{equation}
+
+In addition to the tangential and normal accelerations, $a_{T,k}$ and $a_{N,k}$, we introduce two linear velocity terms, $d_{X,k}$ and $d_{Y,k}$.
+Without these, the model could not be used for smoothing, because for an arbitrary pair of adjacent changepoints ($\tau_{k}$, $\mathbf{x}_{k}$) and ($\tau_{k+1}$, $\mathbf{x}_{k+1}$), there would not in general be a pair of accelerations which achieved the required transition.
+These linear velocity terms may be considered to be merely relaxation terms, allowing arbitrary past and future trajectories to be joined, or they may represent real physical effects, such as wind (for aircraft) or currents (for boats).
+Together, these four variables make up the vector of motion parameters,
+%
+\begin{equation}
+\mathbf{u}_k = [a_{T,k}, a_{N,k}, d_{X,k}, d_{Y,k}]^T
+\end{equation}
+
+The target dynamics are described by four differential equations.
+%
+\begin{IEEEeqnarray}{rCl}
+\ddot{s}_t & = & a_{T,K(t)} \\
+\dot{s}_t \dot{\psi}_t & = & a_{N,K(t)} \\
+\dot{x}_t & = & \dot{s}_t \cos(\psi_t) + d_{X,K(t)} \\
+\dot{y}_t & = & \dot{s}_t \sin(\psi_t) + d_{Y,K(t)}     .
+\end{IEEEeqnarray}
+
+Solving these yields the following state equations (where $\Delta\tau = \tau_{k+1} - \tau_k$):
+%
+\begin{IEEEeqnarray}{rCl}
+\dot{s}(t) & = & \dot{s}_{K(t)} + a_{T,k} \Delta t \label{eq:2D_ICmodel_start} \\
+\psi(t) & = & \psi_{K(t)} + \frac{a_{N,k}}{a_{T,k}} \log \left( \frac{\dot{s}(t)}{\dot{s}_{K(t)}} \right) \\
+x(t) & = & x_{K(t)} + d_{X,K(t)} \Delta t \\
+     \IEEEeqnarraymulticol{3}{l}{ \quad + \: \frac{ \dot{s}(t)^2 }{ 4 a_{T,k}^2 + a_{N,k}^2 } \left[  a_{N,k} \sin(\psi(t)) + 2 a_{T,k} \cos(\psi(t))  \right]} \nonumber \\
+     \IEEEeqnarraymulticol{3}{l}{ \quad - \: \frac{\dot{s}_{K(t)}^2}{4 a_{T,k}^2 + a_{N,k}^2} \left[  a_{N,k} \sin(\psi_{K(t)}) + 2 a_{T,k} \cos(\psi_{K(t)})  \right]} \nonumber \\
+y(t) & = & y_{K(t)} + d_{Y,K(t)} \Delta t \\
+     \IEEEeqnarraymulticol{3}{l}{ \quad + \: \frac{ \dot{s}(t)^2 }{ 4 a_{T,k}^2 + a_{N,k}^2 } \left[ -a_{N,k} \cos(\psi(t)) + 2 a_{T,k} \sin(\psi(t))  \right]} \nonumber \\
+     \IEEEeqnarraymulticol{3}{l}{ \quad - \: \frac{\dot{s}_{K(t)}^2}{4 a_{T,k}^2 + a_{N,k}^2} \left[  -a_{N,k} \cos(\psi_{K(t)}) + 2 a_{T,k} \sin(\psi_{K(t)})  \right]} \nonumber      .
+\end{IEEEeqnarray}
+
+The changepoint motion parameters are assumed to be independent and zero-mean Gaussian distributed.
+The times between changepoints are modelled as gamma distributed.
+Observations are made via a radar-style bearing and range measurement model with Gaussian noise.
+
+The filtering and smoothing algorithms were tested on artificial data simulated from the model.
+Observations were generated every $0.1$.
+The standard deviations of tangential and normal accelerations were set to $0.01$ and $1$ respectively, and those for the drift velocities to $1$.
+The parameters of the Gamma distribution for inter-changepoint times were $5$ and $1$ for the shape and scale respectively.
+Observation noise standard deviations were $0.1$ and $0.25^{\circ}$ respectively.
+
+The filter and smoother were each used to generate $50$ particles, with the filter employing resample-move steps with proposals based on an unscented transform \cite{Julier2004} approximation to the optimal proposal density.
+
+An example trajectory is shown in Fig.~\ref{fig:simulated_trajectory}, with the results of the VRPF and VRPS algorithms in Fig.\ref{fig:2D_particle_results}.
+The improved particle diversity achieved by the smoothing algorithm is apparent.
+
 \begin{figure}[!t]
 \centering
-\subfloat[]{\includegraphics[width=0.95\columnwidth]{fx_data.pdf}} \\
-\subfloat[]{\includegraphics[width=0.95\columnwidth]{fx_filter_kdest.pdf}} \\
-\subfloat[]{\includegraphics[width=0.95\columnwidth]{fx_smoother_kdest.pdf}}
-\caption{Foreign exchange data (a), along with filter (a) and smoother (b) kernel density estimates for the state and trend jump times.}
-\label{fig:fx_results}
+\includegraphics[width=0.9\columnwidth]{simulated_problem.pdf}
+\caption{An example simulated data set, showing true position (dashed line), observations (dots) and changepoint positions (crosses). }
+\label{fig:simulated_trajectory}
 \end{figure}
 
+\begin{figure}[!t]
+\centering
+\subfloat[]{\includegraphics[width=0.95\columnwidth]{2DFilter.pdf}} \\
+\subfloat[]{\includegraphics[width=0.95\columnwidth]{2DSmoother.pdf}}
+\caption{Filter (a) and smoother (b) particle estimates of position (solid lines), with particle changepoint positions (crosses). }
+\label{fig:2D_particle_results}
+\end{figure}
+
+Performance was evaluated by testing on 10 scenarios, each with 500 observations.
+At each observation time, an estimate of the position and velocity is made by taking the average of values from the array of particles.
+The resulting root mean square errors (RMSEs) are displayed in table~\ref{tab:tracking_state_performance}, demonstrating an improvement in accuracy from the smoother.
+The smoother also provides a more accurate estimate of the number of changepoints in each test, with an average error of 0.6 compared to 2.0 from the final frame filter approximation.
+In addition, the mean number of unique particle estimates at each time step is shown in Fig.~\ref{fig:unique_particles}.
+This illustrates the main advantage of the smoother algorithm --- the increased number of unique particles in the approximation means a better characterisation of the state distribution, allowing, for example, calculation of state covariance measures.
+
+\begin{table}%
+\caption{State estimation performance.}
+\label{tab:tracking_state_performance}
+\centering
+\renewcommand{\arraystretch}{1.5}
+\begin{tabular}{|c|c|c|c|}
+\hline
+ &\raisebox{0cm}[0.4cm][0.3cm]{\parbox[c]{1.5cm}{\centering VRPF}} & \raisebox{0cm}[0.4cm][0.3cm]{\parbox[c]{1.5cm}{\centering Final frame VRPF}} & \raisebox{0cm}[0.4cm][0.3cm]{\parbox[c]{1.5cm}{\centering VRPS}} \\
+% & \begin{minipage}[c]{1.5cm}\centering RBVRPF + KF\end{minipage} & \begin{minipage}[c]{1.5cm}\centering RBVRPF + RTS\end{minipage} & \begin{minipage}[c]{1.5cm}\centering RBVRPS + RTS\end{minipage} \\
+\hline \hline
+\raisebox{0cm}[0.4cm][0.3cm]{\parbox[c]{2cm}{\centering mean position estimate RMSE}} & 2.14 & 1.36 & 1.12 \\
+\hline
+\raisebox{0cm}[0.4cm][0.3cm]{\parbox[c]{2cm}{\centering mean velocity estimate RMSE}} & 1.89 & 1.03 & 0.85 \\
+\hline
+\end{tabular}
+\end{table}
+
+\begin{figure}[!t]
+\centering
+\includegraphics[width=0.9\columnwidth]{unique_particles.pdf}
+\caption{Mean number of unique particles in the final filter (dashed) and smoother (solid) estimates.}
+\label{fig:unique_particles}
+\end{figure}
+
+\section{Conclusions}
+
+New smoothing algorithms have been introduced for variable rate models which have either linear-Gaussian or deterministic dynamics conditional on a set of unknown changepoints.
+For the linear-Gaussian case, the algorithm employs Rao-Blackwellisation, using particle methods to estimate the nonlinear changepoint sequence, and Kalman filtering/smoothing to estimate the linear state components.
+For the deterministic case, smoothing is achieved by sampling an augmented target distribution using MCMC.
+Simulations demonstrate that the smoothers generate approximation with improved particle diversity and accuracy when compared to the filters.
+
+%\appendices
+%\section{Initialisation of Backwards Kalman Filter} \label{app:init_backwards_KF}
+%
+%The backwards Kalman filter recursion may be initialised exactly by inversion of the observation model.
+By stacking up the last $L+1$ observations into a single vector, it is possible to write down a joint Gaussian distribution conditional on the $(N-L)$th state.
+This may be inverted to produce an improper Gaussian density for $x_{N-L}$,
+%%
+%\begin{IEEEeqnarray}{rCl}
+% \IEEEeqnarraymulticol{3}{l}{p(y_{N-L:N}|x_{N-L}) = \mathcal{N}( \mathbf{y}_{N-L:N} | H_{N-L} x_{N-L}, \Gamma_{N-L} ) } \\
+% & \propto & \mathcal{N}(x_{N-L} | (H_{N-L}^T \Gamma_{N-L} H_{N-L})^{-1} H_{N-L}^T \Gamma_{N-L}, (H_{N-L}^T \Gamma_{N-L} H_{N-L})^{-1} )      ,
+%\end{IEEEeqnarray}
+%
+%where $H_{N-L}$ and $\Gamma_{N-L}$ may be constructed from the system models and,
+%%
+%\begin{IEEEeqnarray}{rCl}
+%\mathbf{y}_{N-L:N} & = & \begin{bmatrix}y_N \\ y_{N-1} \\ \vdots \\ y_{N-L} \end{bmatrix}     .
+%\end{IEEEeqnarray}
+%
+%This should effected for the smallest value of $L$ such that the matrix inversions possible.
+
+%\newpage
 \bibliographystyle{IEEEtran}
 \bibliography{D:/pb404/Dropbox/PhD/OTbib}
+\input{D:/pb404/Dropbox/PhD/Biographies/IEEE-Bunch.tex}
+\input{D:/pb404/Dropbox/PhD/Biographies/IEEE-Godsill.tex}
 \end{document}
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-A Markov chain is initialised by sampling a particle from the approximation generated by the final filtering step.
-Because this is an approximation of the target distribution, no burn-in is requiired.
-Proceeding backwards in time, MH steps are conducted by proposing a replacement for the changepoint sequence history from the appropriate filtering approximation.
-At the $n$th step, the future changepoints are considered to be fixed, $\tilde{\zeta}_{[t_n,T]}$.
-A replacement for the past changepoints is proposed from the filtering particles.
-
-\begin{equation}
- \zeta_{[0,t_n]}^* \sim q(\zeta_{[0,t_n]}|y_{1:n}) = \sum_j \gamma_n^{(j)} \delta_{(\zeta_{[0,t_n]}^{(j)})}(\zeta_{[0,t_n]})
-\end{equation}
-
-This proposed move is accepted with the following acceptance probability, where $\zeta_{[0,t_n]}^{(m)}$ is the previous sequence of past changepoints in the Markov chain.
-
-\begin{IEEEeqnarray}{rCl}
- \alpha & = & \frac{ p(\zeta_{[0,t_n]}^*, \tilde{\zeta}_{[t_n,T]}|y_{1:N}) q(\zeta_{[0,t_n]}^{(m)}|y_{1:n}) }{ p(\zeta_{[0,t_n]}^{(m)}, \tilde{\zeta}_{[t_n,T]}|y_{1:N}) q(\zeta_{[0,t_n]}^*|y_{1:n}) } \nonumber \\
-        & = & \frac{ \gamma_n^{(m)} }{ \gamma_n^* } \times \frac{ p(\tilde{\zeta}_{[t_n,T]}|\zeta_{[0,t_n]}^*) p(\zeta_{[0,t_n]}^*) }{ p(\tilde{\zeta}_{[t_n,T]}|\zeta_{[0,t_n]}^{(m)}) p(\zeta_{[0,t_n]}^{(m)}) } \nonumber \\
-        &   &  \times \frac{ p(y_{r_k^-:r_k^+}|\hat{x}_{r_k^-:r_k^+}^*) p(y_{1:r_k^- -1}|\hat{x}_{1:r_k^- -1}^*) }{ p(y_{r_k^-:r_k^+}|\hat{x}_{r_k^-:r_k^+}^{(m)}) p(y_{1:r_k^- -1}|\hat{x}_{1:r_k^- -1}^{(m)}) } \nonumber \\
-        & = & \frac{ \gamma_n^{(m)} }{ \gamma_n^* } \times \frac{ p(\zeta_{[0,t_n]}^*) p(y_{1:r_k^- -1}|\hat{x}_{1:r_k^- -1}^*) }{ p(\zeta_{[0,t_n]}^{(m)}) p(y_{1:r_k^- -1}|\hat{x}_{1:r_k^- -1}^{(m)}) } \nonumber \\
-        &   &  \times \frac{ p(\tilde{\zeta}_{k+1}|\zeta_{k}^*) p(y_{r_k^-:r_k^+}|\hat{x}_{r_k^-:r_k^+}^*) }{ p(\tilde{\zeta}_{k+1}|\zeta_{k}^{(m)}) p(y_{r_k^-:r_k^+}|\hat{x}_{r_k^-:r_k^+}^{(m)}) }     \label{eq:MCMC-VRPS_ap}
-\end{IEEEeqnarray}
-
-The first fraction is simply a ratio of proposal weights for the current and proposed changepoint histories.
-The second ratio comprises probabilities which depend solely on the past changepoints, and which are thus the same as values calculated during the filtering stage.
-These may be stored when first calculated, meaning they can simply be fetched from memory when calculating the acceptance probability.
-
-The third ratio consists of the transition probability between two known changepoints, $\zeta_k$ and $\zeta_{k+1}$, and the likelihood of observations occurring between them.
-These may be found by calculating the innovation, $v_k$, to get from $\zeta_k$ to $\zeta_{k+1}$.
-The continuous state trajectory may then be calculated allowing the likelihood terms to be evaluated.
-The transition density term is given by:
-
-\begin{IEEEeqnarray}{rCl}
- p(\zeta_{k+1}|\zeta_{k}) & = & p(\tau_{k+1}|\tau_{k}) p(v_{k}|v_{k-1})
-\end{IEEEeqnarray}
-
-The variable rate particle smoother (VRPS) is summarised below.%in algorithm~\ref{alg:MCMC-VRPS}.
-
-%\begin{algorithm}
- \begin{algorithmic}
-  \STATE Run a particle filter to approximate $p(\theta_{[0,t_n]}, x_0|y_{1:n})$ for each $n$.
-  \STATE Convert each particle innovation sequence to a changepoint state sequence, to produce approximations of $p(\zeta_{[0,t_n]}|y_{1:n})$.
-  \FOR{$i=1 \dots S$}
-    \STATE Initialise sampler with $\tilde{\zeta}_{[0,T]}^{(i)} \sim \sum_{j} w_N^{(j)} \delta_{\zeta_{[0,T]}^{(j)}}(\zeta_{[0,T]})$
-    \FOR{$n=N \dots 1$}
-      \STATE $\zeta_{[0,T]}^{(i)(0)} \gets \tilde{\zeta}_{[0,T]}^{(i)}$.
-      \FOR{$m=1 \dots M$}
-        \STATE Propose a new past $\zeta_{[0,t_n]}^{(i)*} \sim \sum_j \gamma_n^{(j)} \delta_{(\zeta_{[0,t_n]}^{(j)})}(\zeta_{[0,t_n]})$.
-	      \STATE Calculate $\alpha$ using equation~\ref{eq:MCMC-VRPS_ap}
-	      \STATE With probability $\alpha$, $\zeta_{[0,t_n]}^{(i)(m)} \gets \zeta_{[0,t_n]}^{(i)*}$. Otherwise $\zeta_{[0,t_n]}^{(i)(m)} \gets \zeta_{[0,t_n]}^{(i)(m-1)}$.
-      \ENDFOR
-      \STATE $\tilde{\zeta}_{[0,t_n]}^{(i)} \gets \zeta_{[0,t]}^{(i)(M)}$.
-    \ENDFOR
-  \ENDFOR
- \end{algorithmic}
-%\label{alg:MCMC-VRPS}
-%\caption{MCMC variable rate particle smoother}
-%\end{algorithm}
-
-The number of MH steps, $M$, at each observation time, $n$, allows a trade-off of performance against time.
-Larger values of $M$ will ensure more unique samples in the smoothing approximation, but will also take longer to execute.
-In the simulations discussed in this report, $M=1$ was used throughout.
-
-The recursion is initialised by inversion of the observation model.
-By stacking up the last $L+1$ observations into a single vector, it is possible to write down a joint Gaussian distribution conditional on the $(N-L)$th state.
-This may be inverted to produce an improper Gaussian density for $x_{N-L}$,
-%
-\begin{IEEEeqnarray}{rCl}
- \IEEEeqnarraymulticol{3}{l}{p(y_{N-L:N}|x_{N-L}) = \mathcal{N}( \mathbf{y}_{N-L:N} | H_{N-L} x_{N-L}, \Gamma_{N-L} ) } \\
- & \propto & \mathcal{N}(x_{N-L} | (H_{N-L}^T \Gamma_{N-L} H_{N-L})^{-1} H_{N-L}^T \Gamma_{N-L}, (H_{N-L}^T \Gamma_{N-L} H_{N-L})^{-1} )      ,
-\end{IEEEeqnarray}
-
-where $H_{N-L}$ and $\Gamma_{N-L}$ may be constructed from the system models and,
-%
-\begin{IEEEeqnarray}{rCl}
-\mathbf{y}_{N-L:N} & = & \begin{bmatrix}y_N \\ y_{N-1} \\ \vdots \\ y_{N-L} \end{bmatrix}     .
-\end{IEEEeqnarray}
-
-This should effected for the smallest value of $L$ such that the matrix inversions possible.
-Further details are provided in appendix~\ref{app:backward_filter}.
-Alternatively, the recursion could be initialised with an approximation.

### 2012-04-06 draft2_VRsmoothing.tex
A version I'm happy with. Complete except for making the results section better and shorter, and lacking an abstract.

@@
 
-The target distribution may be factorised as follows,
+\subsection{Conditionally Linear-Gaussian Smoothing} \label{sec:rb-vrps}
+
+At time $t_n$, the target distribution may be factorised as follows,
 %
@@
 
-This suggests a sequential strategy for smoothing.
-A particle drawn from $p(\theta_{n}^{+}|y_{1:N})$ may be extended backwards by sampling from the backwards conditional distribution, $p(\theta_{n}|\theta_{n}^{+}, y_{1:N})$ and concatenating the two partial sequences.
+This suggests a sequential strategy for particle smoothing.
+Starting with a low diversity set of particles from $p(\theta|y_{1:N})$, we first marginalise $\theta_{n}$ by simply discarding the changepoints before $t_n$.
+New values of $\theta_{n}$ are then sampled from the conditional distribution, $p(\theta_{n}|\theta_{n}^{+}, y_{1:N})$, which is approximated using the filter particles, and the past and future sequences are concatenated.
+By iterating backwards from $n = N \dots 1$, a diverse set of particles approximating $p(\theta|y_{1:N})$ will be generated.
+
+%A particle drawn from $p(\theta_{n}^{+}|y_{1:N})$ may be extended backwards by sampling from the conditional distribution, $p(\theta_{n}|\theta_{n}^{+}, y_{1:N})$ and concatenating the past and future sequences.
 The new particle will be a sample from the target distribution, but the resulting approximation will still suffer from low path-space diversity before time $t_{n-1}$.
@@
 
-At step $n$ for a given particle, the future changepoint sequence may be considered to be fixed.
+%At step $n$ for a given particle, the future changepoint sequence may be considered to be fixed.
 It remains to devise a method for drawing samples from the backwards conditional distribution, $p(\theta_{n}|\theta_{n}^{+}, y_{1:N})$.
+
+It remains to devise a scheme for sampling the conditional distribution, $p(\theta_{n}|\theta_{n}^{+}, y_{1:N})$.
 The simplifications used by the fixed rate particle smoother which exploit the Markovian nature of the state sequence \cite{Godsill2004} are of no help, because past changepoints are not independent of future observations (i.e.
 $p(\theta_{n}|\theta_{n}^{+}, y_{1:N}) \ne p(\theta_{n}|\theta_{n}^{+}, y_{1:n}))$.
-In the following sections, alternative methods for sampling this backwards conditional distribution are discussed for the two classes of model.
-
-\subsection{Conditionally Linear-Gaussian Smoothing} \label{sec:rb-vrps}
-
-When the state dynamics are linear-Gaussian conditional on the changepoint sequence, it is possible split the dependence on the observations into past and future by introducing the current state as an additional variable.
+However, when the state dynamics are linear-Gaussian conditional on the changepoint sequence, it is possible split the dependence on the observations into past and future by introducing the current state as an additional variable.
 This trick was used in \cite{Sarkka2012} in the derivation of the Rao-Blackwellised particle filter for fixed rate conditionally linear-Gaussian models.
@@
 \begin{algorithmic}[1]
+\linespread{1.7} \selectfont
   \STATE Run Rao-Blackwellised variable rate particle filter to approximate $p(\theta_{n}|y_{1:n})$ with weighted particles $\{\theta_{n}^{(i)}, w_{n}^{(i)}\}$ and $\{p(x_n|\theta_{n}^{(i)},y_{1:n})\}$ as normal distributions with moments $\{m_{n}^{(i)}\}$ and $\{P_{n}^{(i)}\}$.
@@
     \STATE Initialise sufficient statistics $\tilde{m}_N^{(i)}$ and $\tilde{P}_N^{(i)}$ (see text).
-    \FOR{$n=N \dots 1$}
+    \FOR{$n=N-1 \dots 1$}
+      \STATE Discard $\theta_{n}^{(i)}$.
       \STATE Predict state mean and covariance $\tilde{m}_n^{-(i)}$ and $\tilde{P}_n^{-(i)}$ using \ref{eq:backward_kf_predict_start} to \ref{eq:backward_kf_predict_stop}.
@@
       \STATE Sample $\theta_{n}^{(i)} \sim \sum_j \tilde{w}_n^{(j)} \delta_{\theta_{n}^{(j)}}(\theta_{n})$.
+      \STATE Join $\theta^{(i)} \gets \theta_{n}^{(i)} \cup \theta_{n}^{+(i)}$.
       \STATE Update state mean and covariance $\tilde{m}_n^{(i)}$ and $\tilde{P}_n^{(i)}$ using \ref{eq:backward_kf_update_start} to \ref{eq:backward_kf_update_stop}.
-      \STATE Discard $\theta_{n-1}^{(i)}$.
     \ENDFOR
@@
 
-Again, this suggests a backwards sequential sampling algorithm for smoothing. The backwards conditional term may be expanded as,
+Again, this suggests a backwards sequential sampling algorithm for smoothing. The conditional term may be expanded as,
 %
 \begin{IEEEeqnarray}{rCl}
-p(\tilde{\theta}_{n}|\tilde{\theta}_{n}^{+}, y_{1:N}) & \propto & p(y_{n+1:N}|\tilde{\theta}_{n}, \tilde{\theta}_{n}^{+}) p(\tilde{\theta}_{n}^{+}|\tilde{\theta}_{n}) p(\tilde{\theta}_{n}|y_{1:n})    .
+p(\tilde{\theta}_{n}|\tilde{\theta}_{n}^{+}, y_{1:N}) & \propto & p(y_{n+1:N}|\tilde{\theta}_{n}, \tilde{\theta}_{n}^{+}) \nonumber \\
+& & \times p(\tilde{\theta}_{n}^{+}|\tilde{\theta}_{n}) p(\tilde{\theta}_{n}|y_{1:n})    .
 \end{IEEEeqnarray}
 
-This presents a problem. We cannot substitute the particle approximation for the filtering distribution because the transition density term for the augmented sequences is given by,
+This presents a problem. We cannot directly substitute the particle approximation for the filtering distribution because the transition density term for the augmented sequences is given by,
 %
 \begin{IEEEeqnarray}{rCl}
-p(\tilde{\theta}_{n}^+ | \tilde{\theta}_{n}) & \propto & p(\tau_{K_n+1}, u_{K_n+1}| \tau_{K_n}, u_{K_n}) \delta_{f(x_{K_n}, u_{K_n}, \tau_{K_n}, \tau_{K_n+1})}(x_{K_n+1}),
+p(\tilde{\theta}_{n}^+ | \tilde{\theta}_{n}) & \propto & p(\tau_{K_n+1}, u_{K_n+1}| \tau_{K_n}, u_{K_n}) \nonumber \\
+ & & \times \delta_{f(x_{K_n}, u_{K_n}, \tau_{K_n}, \tau_{K_n+1})}(x_{K_n+1}),
 \end{IEEEeqnarray}
@@
 The solution to this problem is provided by an idea from the Sequential Monte Carlo (SMC) samplers of \cite{DelMoral2006}.
-Rather than sampling $p(\tilde{\theta}_{n}|\tilde{\theta}_{n}^{+}, y_{1:N})$ directly, a value of $\tilde{\theta}_{n}$ is proposed from the filtering particles, after which a replacement, $u_{K_n}'$, is proposed for the existing parameter corresponding to the final changepoint, with a value chosen such that the past and future paths meet up.
+Rather than sampling $p(\tilde{\theta}_{n}|\tilde{\theta}_{n}^{+}, y_{1:N})$ directly, a value of $\tilde{\theta}_{n}$ is proposed from the filtering particles, after which a replacement, $u_{K_n}'$, is proposed for the existing final changepoint parameter, with a value chosen such that the past and future paths meet up.
 The modified sequence is denoted $\tilde{\theta}_{n}'$.
@@
 
-The augmented target distribution may be sampled using Markov chain Monte Carlo (MCMC) \cite{Gilks1996}, in a similar manner to the fixed rate smoother of \cite{Bunch2012} (IS could be used, but this would require resampling steps, which would lead to a loss of path-space diversity).
+The augmented target distribution may be sampled using Markov chain Monte Carlo (MCMC) \cite{Gilks1996}, in a similar manner to the fixed rate smoother of \cite{Bunch2012}.
 Metropolis-Hastings (MH) steps are conducted to draw samples, $\{\tilde{\theta}_{n}, u_{K_n}'\}$, from the target distribution by sampling the proposal and accepting the new values with a given probability.
+% (IS could be used, but this would require resampling steps, which would lead to a loss of path-space diversity)
 
@@
 \begin{IEEEeqnarray}{rCl}
-\beta_n &\propto& \frac{ p(y_{r_n^-:r_n^+}|\tilde{\theta}_{n}') }{ p(y_{r_n^-:n}|\tilde{\theta}_{n}) } \nonumber \\
-  & & \times \frac{ p(\tilde{\theta}_{n}^{+}|\tilde{\theta}_{n}') p(u_{K_n}'|\tilde{\theta}_n \setminus u_{K_n}) }{ p(u_{K_n}|\tilde{\theta}_n \setminus u_{K_n}) } \nonumber \\
-  & & \times \frac{ L(u_{K_n}|\tilde{\theta}_{n}, \tilde{\theta}_{n}^{+}, y_{1:N}) }{ q(u_{K_n}'|\tilde{\theta}_{n}, \tilde{\theta}_{n}^{+}, y_{1:N}) } \label{eq:pd-vrps_tpr}    ,
+\beta_n & \propto & \frac{ p(y_{r_n^-:r_n^+}|\tilde{\theta}_{n}') }{ p(y_{r_n^-:n}|\tilde{\theta}_{n}) } \times \frac{ p(u_{K_n}'|\tau_{K_n}, \tau_{K_n-1}, u_{K_n-1}) }{ p(u_{K_n}|\tau_{K_n}, \tau_{K_n-1}, u_{K_n-1}) } \nonumber \\
+  & & \times p(\tilde{\theta}_{n}^{+}|\tilde{\theta}_{n}') \times \frac{ L(u_{K_n}|\tilde{\theta}_{n}', \tilde{\theta}_{n}^{+}, y_{1:N}) }{ q(u_{K_n}'|\tilde{\theta}_{n}, \tilde{\theta}_{n}^{+}, y_{1:N}) } \label{eq:pd-vrps_tpr}    ,
 \end{IEEEeqnarray}
 
-where $r_n^- = \min( m : t_m > \tau_{K_n} )$ and $r_n^+ = \max( m : t_m < \tau_{K_n+1} )$.
+where $r_n^- = \min( m : t_m > \tau_{K_n} )$ and $r_n^+ = \max( m : t_m < \tau_{K_n+1} )$ are the indexes of the earliest and latest observations to fall between the changepoints each side of $t_n$.
 
@@
 This may require it to be a delta function if there is only one feasible value.
-The artificial conditional term, $L(u_{K_n}|\tilde{\theta}_{n}, \tilde{\theta}_{n}^{+}, y_{1:N})$, may be any arbitrary density, but the simplest choice is simply as a uniform distribution so that it cancels out in the acceptance probability.
+The artificial conditional term, $L(u_{K_n}|\tilde{\theta}_{n}, \tilde{\theta}_{n}^{+}, y_{1:N})$, may be any arbitrary density, but the simplest choice is simply as a uniform distribution so that it cancels out in the MH acceptance probability.
 The remaining terms are likelihoods and parameter transition density terms, defined by the system equations.
 
-If $\beta_n^{(m-1)}$ is the target-proposal ratio for the current state in the chain, and $\beta_n^*$ that for the proposed state, then the acceptance probability is given by,
+If $\beta_n^{(m-1)}$ is the target-proposal ratio for the current state in the chain, and $\beta_n^*$ that for the new state, then the acceptance probability is given by,
 %
@@
 \end{IEEEeqnarray}
+
+Each Markov chain may be initialised with the output from the previous stage.
+This value is itself a particle from the target distribution (albeit one from an approximation with poor diversity), meaning that no burn-in period is required.
 
@@
 \begin{algorithmic}[1]
+\linespread{1.7} \selectfont
   \STATE Run variable rate particle filter to approximate $p(\tilde{\theta}_{n}|y_{1:n})$ with weighted particles $\{\tilde{\theta}_{n}^{(i)}, w_{n}^{(i)}\}$. Store all results.
   \FOR{$i=1 \dots N_S$}
-  	\STATE Initialise particle from $N$th filter approximation $\tilde{\theta}^{(i)} \sim \sum_j w_N^{(j)} \delta_{\tilde{\theta}^{(j)}}(\tilde{\theta})$.
+  	\STATE Initialise particle using $\tilde{\theta}^{(i)} \sim \sum_j w_N^{(j)} \delta_{\tilde{\theta}^{(j)}}(\tilde{\theta})$.
     \FOR{$n=N \dots 1$}
-      \STATE Get the current changepoint history $\tilde{\theta}_n^{(i)(0)} \gets \tilde{\theta}_n^{(i)}$.
+      \STATE Initialise chain with $\tilde{\theta}_n^{(i)(0)} \gets \tilde{\theta}_n^{(i)}$.
       \FOR{$m=1 \dots M$}
-        \STATE Propose a new changepoint history from $n$th filter approximation $\tilde{\theta}_n^{(i)*} \sim \sum_j w_n^{(j)} \delta_{\tilde{\theta}_{n}^{(j)}}(\tilde{\theta}_{n})$.
-        \STATE Propose a replacement parameter $u_{K_n}'^* \sim q(u_{K_n}|\tilde{\theta}_{n}, \tilde{\theta}_{n}^{+}, y_{1:N})$.
-        \STATE Construct a new changepoint history, $\tilde{\theta}_n'^{(i)*}$, by replacing $u_{K_n}^*$ with $u_{K_n}'^*$.
-        \STATE Calculate acceptance probability $\alpha_n^{(m)}$ using (\ref{eq:pd-vrps_tpr}) and (\ref{eq:pd-vrps_ap}).
-        \STATE With probability $\alpha_n^{(m)}$, $\tilde{\theta}_n^{(i)(m)} \gets \tilde{\theta}_n^{(i)*}$. Otherwise, $\tilde{\theta}_n^{(i)(m)} \gets \tilde{\theta}_n^{(i)(m-1)}$.
+        \STATE Propose a new history $\tilde{\theta}_n^{(i)*} \sim \sum_j w_n^{(j)} \delta_{\tilde{\theta}_{n}^{(j)}}(\tilde{\theta}_{n})$.
+        \STATE Propose a parameter $u_{K_n}'^* \sim q(u_{K_n}|\tilde{\theta}_{n}, \tilde{\theta}_{n}^{+}, y_{1:N})$.
+        \STATE Replace $u_{K_n}^*$ with $u_{K_n}'^*$ to form $\tilde{\theta}_n'^{(i)*}$.
+        \STATE Calculate $\alpha_n^{(m)}$ using (\ref{eq:pd-vrps_tpr}) and (\ref{eq:pd-vrps_ap}).
+        \STATE With probability $\alpha_n^{(m)}$, $\tilde{\theta}_n^{(i)(m)} \gets \tilde{\theta}_n^{(i)*}$. \\ Otherwise, $\tilde{\theta}_n^{(i)(m)} \gets \tilde{\theta}_n^{(i)(m-1)}$.
       \ENDFOR
-      \STATE Keep final sample $\tilde{\theta}_n^{(i)} \gets \tilde{\theta}_n^{(i)(M)}$.
+      \STATE Store final sample $\tilde{\theta}_n^{(i)} \gets \tilde{\theta}_n^{(i)(M)}$.
     \ENDFOR
@@
 
-The Markov chains require no burn-in, because the starting value is itself a sample from the target distribution.
 The number of MH steps, $M$, at each observation time, $n$, allows a trade-off of performance against time.
@@
 
-Solving these yields the following state equations (where $\Delta\tau = \tau_{k+1} - \tau_k$):
+Solving these yields the following state equations (where $\Delta\ t = t - \tau_{K(t)}$):
 %

### 2012-04-09 draft3_VRsmoothing.tex
Added abstract

--- 2012-04-06 draft2_VRsmoothing.tex
@@
 \begin{abstract}
-The abstract goes here.
-Blah, blah, blah, blah, blah, blah, blah, blah, blah, blah.
-Blah, blah, blah, blah, blah, blah, blah, blah, blah, blah.
-Blah, blah, blah, blah, blah, blah, blah, blah, blah, blah.
-Blah, blah, blah, blah, blah, blah, blah, blah, blah, blah.
-Blah, blah, blah, blah, blah, blah, blah, blah, blah, blah.
-Blah, blah, blah, blah, blah, blah, blah, blah, blah, blah.
-Blah, blah, blah, blah, blah, blah, blah, blah, blah, blah.
-Blah, blah, blah, blah, blah, blah, blah, blah, blah, blah.
-Blah, blah, blah, blah, blah, blah, blah, blah, blah, blah.
-Blah, blah, blah, blah, blah, blah, blah, blah, blah, blah.
+Standard state-space methods assume that the latent state evolves uniformly over time, and can be modelled with a discrete-time process synchronous with the observations.
+This may be a poor representation of some systems in which the state evolution displays discontinuities in its behaviour.
+For such cases, a variable rate model may be more appropriate; the system dynamics are conditioned on a set of random changepoints which constitute a marked point process.
+In this paper, new particle smoothing algorithms are presented for use with conditionally linear-Gaussian and conditionally deterministic dynamics.
+These are demonstrated on problems of financial modelling and target tracking.
+Results indicate that the smoothing approximations provide more accurate and more diverse representations of the state posterior distributions.
 \end{abstract}
@@
 \begin{IEEEkeywords}
-
+Bayesian inference, state-space model, variable rate, particle filter, filtering, smoothing
 \end{IEEEkeywords}

### 2012-04-19 draft3_VRsmoothing.tex
Final draft.

@@
 
-The transition term, $p(\theta_{n \setminus n-1} | \theta_{n-1})$, has a similar form to the changepoint prior of (\ref{eq:cp_sequence_prior}) \cite{Jacobsen2006}, comprising a product of density terms (one for each new changepoint) and a survivor function term.% In the particular (but common) case that no new changepoints occur within the interval, the density consists of a probability mass on the empty set, with weight $P(\tau_{K_n+1} > t_n | \tau_{K_n+1} > t_{n-1}, \tau_{K_n} )$.
-%
-\begin{IEEEeqnarray}{rCl}
-\IEEEeqnarraymulticol{3}{l}{p(\theta_{n \setminus n-1} | \theta_{n-1})} \nonumber \\
-    & = & p(\tau_{K_{n-1}+1:K_n}, u_{K_{n-1}+1:K_n}, \tau_{K_n+1}>t_n|\tau_{K_{n-1}+1}>t_{n-1}, \tau_{K_{n-1}}, u_{K_{n-1}}) \nonumber \\
-    & = & \begin{cases} S(\tau_{K_n}, t_n) \prod\limits_{j:t_{n-1} \leq \tau_j < t_n} p(\tau_j, u_j| \tau_{j-1}, u_{j-1}, \tau_j>t_{n-1}) & K_n > K_{n-1} \\
-                        S(\tau_{K_n}, t_n) / S(\tau_{K_n}, t_{n-1}) & K_n = K_{n-1} \end{cases} \IEEEeqnarraynumspace \label{eq:cp_sequence_trandens}
-\end{IEEEeqnarray}
-
-For all but the first changepoint in the interval, the density is given by the prior model of (\ref{eq:cp_model}).
+The transition term, $p(\theta_{n \setminus n-1} | \theta_{n-1})$, has a similar form to the changepoint prior of (\ref{eq:cp_sequence_prior}) \cite{Jacobsen2006}.
+An additional survivor function is included to account for the condition that changepoints cannot occur before $t_{n-1}$.
+%, comprising a product of density terms (one for each new changepoint) and a survivor function term.% In the particular (but common) case that no new changepoints occur within the interval, the density consists of a probability mass on the empty set, with weight $P(\tau_{K_n+1} > t_n | \tau_{K_n+1} > t_{n-1}, \tau_{K_n} )$.
+%
+\begin{IEEEeqnarray}{rCl}
+\IEEEeqnarraymulticol{3}{l}{p(\theta_{n \setminus n-1} | \theta_{n-1}) = S(\tau_{K_n}, t_n) / S(\tau_{K_{n-1}+1}, t_{n-1}) } \nonumber \\
+ \qquad  \qquad & & \times \prod\limits_{j:t_{n-1} \leq \tau_j < t_n} p(\tau_j, u_j| \tau_{j-1}, u_{j-1})  \label{eq:cp_sequence_trandens}
+\end{IEEEeqnarray}
+%\begin{IEEEeqnarray}{rCl}
+%\IEEEeqnarraymulticol{3}{l}{p(\theta_{n \setminus n-1} | \theta_{n-1})} \nonumber \\
+%    & = & p(\tau_{K_{n-1}+1:K_n}, u_{K_{n-1}+1:K_n}, \tau_{K_n+1}>t_n|\tau_{K_{n-1}+1}>t_{n-1}, \tau_{K_{n-1}}, u_{K_{n-1}}) \nonumber \\
+%    & = & \begin{cases} S(\tau_{K_n}, t_n) \prod\limits_{j:t_{n-1} \leq \tau_j < t_n} p(\tau_j, u_j| \tau_{j-1}, u_{j-1}, \tau_j>t_{n-1}) & K_n > K_{n-1} \\
+%                        S(\tau_{K_n}, t_n) / S(\tau_{K_n}, t_{n-1}) & K_n = K_{n-1} \end{cases} \IEEEeqnarraynumspace \label{eq:cp_sequence_trandens}
+%\end{IEEEeqnarray}
+
+%For all but the first changepoint in the interval, the density is given by the prior model of (\ref{eq:cp_model}).
 For the first changepoint, indexed by $k=K_{n-1}+1$, we must account for the fact that it cannot occur before $t_{n-1}$,
-%
-\begin{IEEEeqnarray}{rCl}
-\IEEEeqnarraymulticol{3}{l}{p(\tau_{k}, u_{k}| \tau_{k-1}, u_{k-1}, \tau_{k}>t_{n-1})} \nonumber \\
-  & = & \frac{1}{S(\tau_{k-1}, t_{n-1})} \begin{cases} p(\tau_{k}, u_{k}| \tau_{k-1}, u_{k-1}) & \tau_{k} > t_{n-1} \\ 0 & \tau_{k} < t_{n-1} \end{cases}  \IEEEeqnarraynumspace \label{eq:cp_cond_model}   .
-\end{IEEEeqnarray}
+%%
+%\begin{IEEEeqnarray}{rCl}
+%\IEEEeqnarraymulticol{3}{l}{p(\tau_{k}, u_{k}| \tau_{k-1}, u_{k-1}, \tau_{k}>t_{n-1})} \nonumber \\
+%  & = & \frac{1}{S(\tau_{k-1}, t_{n-1})} \begin{cases} p(\tau_{k}, u_{k}| \tau_{k-1}, u_{k-1}) & \tau_{k} > t_{n-1} \\ 0 & \tau_{k} < t_{n-1} \end{cases}  \IEEEeqnarraynumspace \label{eq:cp_cond_model}   .
+%\end{IEEEeqnarray}
 
@@
 For the most basic ``bootstrap'' \cite{Gordon1993} form of the VRPF, $\theta_{n \setminus n-1}$ may be proposed from the prior transition density (\ref{eq:cp_sequence_trandens}).
-This can be achieved by sampling new changepoints sequentially from the transition model (\ref{eq:cp_model}) (apart from the first which is sampled from (\ref{eq:cp_cond_model})) until one falls after the current time, $t_n$.
+This can be achieved by sampling new changepoints sequentially from the transition model (\ref{eq:cp_model}) until one falls after the current time, $t_n$.
 This final future changepoint is discarded.
 (This process can be thought of as sampling the entire future changepoint sequence from $t_{n-1}$ onwards, and then marginalising those which fall after $t_n$.) The bootstrap proposal leads to the usual simplification of the weight formula.
+% (apart from the first which is sampled from (\ref{eq:cp_cond_model}))
 %
@@
 \begin{algorithmic}[1]
-\linespread{1.7} \selectfont
+\linespread{1.5} \selectfont
 \STATE For each $i$, initialise particle changepoint sequence with $\{\theta_{0}^{(i)}\} \gets \{\tau_0^{(i)}, u_0^{(i)}\}$, where $\tau_0^{(i)}=0$ and $u_0^{(i)} \sim p(u_0)$.
@@
 \begin{algorithmic}[1]
-\linespread{1.7} \selectfont
+\linespread{1.5} \selectfont
   \STATE Run Rao-Blackwellised variable rate particle filter to approximate $p(\theta_{n}|y_{1:n})$ with weighted particles $\{\theta_{n}^{(i)}, w_{n}^{(i)}\}$ and $\{p(x_n|\theta_{n}^{(i)},y_{1:n})\}$ as normal distributions with moments $\{m_{n}^{(i)}\}$ and $\{P_{n}^{(i)}\}$.
@@
 \begin{algorithmic}[1]
-\linespread{1.7} \selectfont
+\linespread{1.5} \selectfont
   \STATE Run variable rate particle filter to approximate $p(\tilde{\theta}_{n}|y_{1:n})$ with weighted particles $\{\tilde{\theta}_{n}^{(i)}, w_{n}^{(i)}\}$. Store all results.
@@
 
+The smoothers and the final step of the filters each produce a particle approximation to the changepoint sequence posterior, $p(\theta|y_{1:N})$.
+Calculating a single estimator from these is no simple task.
+Each particle consists of a list of changepoint times and parameters, and there is no obvious way to take an average over the set to produce a single estimate.
+Furthermore, if a best estimate could be calculated, it is not trivial to define a distance metric between the estimated and true sequences to assess relative accuracy.
+
+Comparisons of accuracy are most easily conducted by comparing state estimates.
+These can be calculated by simply taking a mean of the particle values (using the Gaussian means for the linear-Gaussian case).
+The accuracy is then assessed by comparing the root mean squared error (RMSE) of the resulting estimators.
+
+In the following examples, we compare the state estimates from the VRPF, the ``filter-smoother'' and the VRPS.
+Filter-smoother estimates use the particle approximation from the final step of the VRPF for the changepoint distribution.
+In the linear-Gaussian case, an RTS smoother is used to find the state posterior distribution for each particle, $p(x_n|\theta^{(i)}, y_{1:N})$, from which state estimates are derived.
+
+In addition, the benefit of employing a smoothing algorithm is that it ensures a more diverse particle representation of the posterior state distribution.
+To demonstrate this effect, the number of unique state estimates is plotted over time in the following examples.
+
 \subsection{Conditionally Linear-Gaussian Model} \label{sec:finance}
@@
 
-The jump covariance matrices are,
-%
-\begin{IEEEeqnarray}{c}
-Q_{J,u_k} = \begin{cases} \begin{bmatrix}\sigma_{J1}^2 & 0 \\ 0 & 0 \end{bmatrix} & u_k = 1 \\
-                          \begin{bmatrix}0 & 0 \\ 0 & \sigma_{J2}^2 \end{bmatrix} & u_k = 2  \end{cases}   .
-\end{IEEEeqnarray}
-
-This model may be discretised at the observation times using standard methods (see e.g.
-\cite{Grewal2002}).
-Assuming observations of value only and Gaussian observation noise with standard deviation $\sigma_y^2$, the resulting discrete time dynamics are described by the following equations:% (see appendix \ref{app:model_discretisation}):
+%The jump covariance matrices are,
+%%
+%\begin{IEEEeqnarray}{c}
+%Q_{J,u_k} = \begin{cases} \begin{bmatrix}\sigma_{J1}^2 & 0 \\ 0 & 0 \end{bmatrix} & u_k = 1 \\
+%                          \begin{bmatrix}0 & 0 \\ 0 & \sigma_{J2}^2 \end{bmatrix} & u_k = 2  \end{cases}   .
+%\end{IEEEeqnarray}
+
+This model may be discretised at the observation times using standard methods (see e.g. \cite{Grewal2002}), leading to,
+%
+%Assuming observations of value only and Gaussian observation noise with standard deviation $\sigma_y^2$, the resulting discrete time dynamics are described by the following equations:% (see appendix \ref{app:model_discretisation}):
 %
@@
 
-where the $\mathbf{w}_n$ and $v_n$ are Gaussian random variables with covariance matrixes $Q_n$ and $R$ respectively. The time between observations times is denotes $\Delta t = t_{n} - t_{n-1}$.
-%
-\begin{IEEEeqnarray}{rCl}
- A     &=& \begin{bmatrix}1 & \frac{1}{\lambda}(1-e^{(-\lambda \Delta t)} \\ 0 & e^{(-\lambda \Delta t)}\end{bmatrix} \\
- C     &=& \begin{bmatrix}1 & 0\end{bmatrix}
-\end{IEEEeqnarray}
-\begin{IEEEeqnarray}{rCl}
- Q_n   &=& \begin{cases}Q_{D} + Q_{J,u_k} & \exists k : \tau_k \in [t_{n-1},t_n]\\
-                        Q_{D}             & \text{otherwise} \end{cases} \\
- Q_{D} &=& \frac{\sigma^2}{2 \lambda}\begin{bmatrix} q_1 & q_2 \\ q_2 & q_3\end{bmatrix} \\
- q_1   &=& \frac{1}{\lambda^2}(2 \lambda \Delta t - (3 - e^{(-\lambda \Delta t)})(1 - e^{(-\lambda \Delta t)})~) \\
- q_2   &=& \frac{1}{\lambda} (1-e^{(-\lambda \Delta t)})^2 \\
- q_3   &=& 1-e^{(-2 \lambda \Delta t)} \\
- R     &=& [\sigma_y^2]
-\end{IEEEeqnarray}
+where the $\mathbf{w}_n$ and $v_n$ are Gaussian random variables with covariance matrixes $Q_n$ and $R$ respectively.% The time between observations times is denotes $\Delta t = t_{n} - t_{n-1}$.
+%
+%\begin{IEEEeqnarray}{rCl}
+% A     &=& \begin{bmatrix}1 & \frac{1}{\lambda}(1-e^{(-\lambda \Delta t)} \\ 0 & e^{(-\lambda \Delta t)}\end{bmatrix} \\
+% C     &=& \begin{bmatrix}1 & 0\end{bmatrix}
+%\end{IEEEeqnarray}
+%\begin{IEEEeqnarray}{rCl}
+% Q_n   &=& \begin{cases}Q_{D} + Q_{J,u_k} & \exists k : \tau_k \in [t_{n-1},t_n]\\
+%                        Q_{D}             & \text{otherwise} \end{cases} \\
+% Q_{D} &=& \frac{\sigma^2}{2 \lambda}\begin{bmatrix} q_1 & q_2 \\ q_2 & q_3\end{bmatrix} \\
+% q_1   &=& \frac{1}{\lambda^2}(2 \lambda \Delta t - (3 - e^{(-\lambda \Delta t)})(1 - e^{(-\lambda \Delta t)})~) \\
+% q_2   &=& \frac{1}{\lambda} (1-e^{(-\lambda \Delta t)})^2 \\
+% q_3   &=& 1-e^{(-2 \lambda \Delta t)} \\
+% R     &=& [\sigma_y^2]
+%\end{IEEEeqnarray}
 
@@
 
-The algorithms were first tested on artificial data simulated from this model.
-The following parameters were used: $\Delta t = 0.0017$, $N = 1000$, $\alpha = 20$, $\lambda = 5$, $\sigma = 0.05$, $\sigma_{J1} = 0.005$, $\sigma_{J2} = 0.05$, $\sigma_{y} = 0.001$.
-
-The filter used $N_P = 100$ particles, and the smoother resampled $N_S = 100$ sequences. Bootstrap proposals were used for the filter.
-
+The algorithms were tested on artificial data simulated from this model.
+Observations were made on a regular time grid with a spacing of $0.0017s$.
+The decay constant for the distribution of inter-changepoint times is 20, and the trend mean-reversion coefficient is 5.
+The diffusion constant is 0.05 and the observation noise standard deviation 0.001.
+The jump standard deviations are 0.005 and 0.05 respectively for value and trend.
+
+%The following parameters were used: $\Delta t = 0.0017$, $N = 1000$, $\alpha = 20$, $\lambda = 5$, $\sigma = 0.05$, $\sigma_{J1} = 0.005$, $\sigma_{J2} = 0.05$, $\sigma_{y} = 0.001$.
+
+The filter used $N_P = 100$ particles, and the smoother resampled $N_S = 100$ sequences.
+Bootstrap proposals were used for the filter.
 An example realisation simulated from the model is shown in figure~\ref{fig:example_data}.
+The trend estimates produced by the filter-smoother and VRPS are shown in figure~\ref{fig:example_state} (the equivalent graphs for value estimates are not particularly informative, as this quantity is observed).
 
@@
 
-Both the smoother and the final processing step of the filter produce a particle approximation to the changepoint sequence posterior, $p(\theta|y_{1:N})$.
-It is actually no simple task to compare the quality of these approximations.
-Each particle consists of a list of changepoint times and types, and there is no obvious way to take an average over the set to produce a single estimate.
-Furthermore, once a best estimate has been calculated, it is not obvious how the error should be assessed, as it is not trivial to define a distance metric between lists of varying lengths.
-
-For a first, qualitative comparison of the changepoint sequence distributions, it is possible to generate kernel density estimates for the changepoint times.
+\begin{figure}[!t]
+\centering
+\subfloat[]{\includegraphics[width=0.95\columnwidth]{example_filtersmoother_state.pdf}} \\
+\subfloat[]{\includegraphics[width=0.95\columnwidth]{example_smoother_state.pdf}}
+\caption{Trend estimates from (a) the filter-smoother and (b) the VRPS. Solid lines show means of each particle. Dashed lines show mean $\pm$2 standard deviations.}
+\label{fig:example_state}
+\end{figure}
+
+%For a first, qualitative comparison of the changepoint sequence distributions, it is possible to generate kernel density estimates for the changepoint times.
 These are constructed by simply adding a (unit amplitude) Gaussian kernel for each changepoint present in each particle and scaling by the number of particles.
@@
 
-\begin{figure}[!t]
-\centering
-\subfloat[]{\includegraphics[width=0.95\columnwidth]{example_filter_kdest.pdf}} \\
-\subfloat[]{\includegraphics[width=0.95\columnwidth]{example_smoother_kdest.pdf}}
-\caption{Filter (a) and smoother (b) kernel density estimates for value (top) and trend (bottom) jump times. Correct times overlayed as dotted vertical lines.}
-\label{fig:example_kdest}
-\end{figure}
-
-It is immediately apparent that the approximation produced by the filter lacks diversity.
+%\begin{figure}[!t]
+%\centering
+%\subfloat[]{\includegraphics[width=0.95\columnwidth]{example_filter_kdest.pdf}} \\
+%\subfloat[]{\includegraphics[width=0.95\columnwidth]{example_smoother_kdest.pdf}}
+%\caption{Filter (a) and smoother (b) kernel density estimates for value (top) and trend (bottom) jump times. Correct times overlayed as dotted vertical lines.}
+%\label{fig:example_kdest}
+%\end{figure}
+
+%It is immediately apparent that the approximation produced by the filter lacks diversity.
 The jump time kernel density estimate comprises peaks with magnitude $1$, with regions of $0$ between.
@@
 
-A further comparison of the algorithms is possible via the state estimates they generate.
+%A further comparison of the algorithms is possible via the state estimates they generate.
 Here we can compare three options: the filtering results, using the VRPF and a Kalman filter for the state estimates; the filter-smoother results, using the final VRPF approximation for the changepoint sequence and a Rauch-Tung-Striebel (RTS) smoother for the state estimates; and the full smoothing results, using the VRPS followed by an RTS smoother.
@@
 
-\begin{figure}[!t]
-\centering
-\subfloat[]{\includegraphics[width=0.95\columnwidth]{example_filter_state.pdf}} \\
-\subfloat[]{\includegraphics[width=0.95\columnwidth]{example_filtersmoother_state.pdf}} \\
-\subfloat[]{\includegraphics[width=0.95\columnwidth]{example_smoother_state.pdf}}
-\caption{Trend estimates using RBVRPF + Kalman filter (a) RBVRPF + RTS smoother (b) and RBVRPS + RTS smoother.
+%\begin{figure}[!t]
+%\centering
+%\subfloat[]{\includegraphics[width=0.95\columnwidth]{example_filter_state.pdf}} \\
+%\subfloat[]{\includegraphics[width=0.95\columnwidth]{example_filtersmoother_state.pdf}} \\
+%\subfloat[]{\includegraphics[width=0.95\columnwidth]{example_smoother_state.pdf}}
+%\caption{Trend estimates using RBVRPF + Kalman filter (a) RBVRPF + RTS smoother (b) and RBVRPS + RTS smoother.
 Solid lines show means of each particle.
 Dashed lines show mean $\pm$2 standard deviations.}
-\label{fig:example_state}
-\end{figure}
-
-For a quantitative comparison, the algorithms were tested on $10$ realisations from the model, each of $1000$ time steps, and the following statistics were calculated for each:
-\begin{itemize}
-	\item The number of unique changepoint sequences. This is a measure of particle diversity of the approximation.
-	\item The number of unique changepoint times. Another measure of particle diversity.
-	%\item The optimal sub-pattern assignment (OSPA) distance between the maximum a posteriori (MAP) changepoint sequence and the true sequence.
+%\label{fig:example_state}
+%\end{figure}
+
+%For a quantitative comparison, the algorithms were tested on $10$ realisations from the model, each of $1000$ time steps, and the following statistics were calculated for each:
+%\begin{itemize}
+%	\item The number of unique changepoint sequences. This is a measure of particle diversity of the approximation.
+%	\item The number of unique changepoint times. Another measure of particle diversity.
+%	%\item The optimal sub-pattern assignment (OSPA) distance between the maximum a posteriori (MAP) changepoint sequence and the true sequence.
 The OSPA is a concept from multiple target tracking introduced in \cite{Schuhmacher2008} to measure the distance between sets of points of varying cardinality.
@@
 For each approximation, the sequence from the MAP particle was used as a point estimate.
-	%\item The root-mean-square error (RMSE) of the MAP state estimate.
+%	%\item The root-mean-square error (RMSE) of the MAP state estimate.
 The MAP state estimate is taken to be the Gaussian mean of the conditional state distribution, from the Kalman smoother or RTS, selected from the particle with the highest posterior probability.
-	\item The root-mean-square error (RMSE) of the mean state estimate. The mean state estimate is the average of the Gaussian means from all the particles.
-\end{itemize}
+%	\item The root-mean-square error (RMSE) of the mean state estimate. The mean state estimate is the average of the Gaussian means from all the particles.
+%\end{itemize}
 
@@
 
-The results are shown in tables~\ref{tab:finance_cp_performance} and ~\ref{tab:finance_state_performance}.% The OSPA parameters are: exponent $p=1$, threshold $c=0.01$.
-
-\begin{table}%
-\caption{Changepoint sequence estimation performance.}
-\label{tab:finance_cp_performance}
-\centering
-\renewcommand{\arraystretch}{1.5}
-\begin{tabular}{|c|c|c|}
-\hline
- & RBVRPF & RBVRPS \\
-\hline \hline
-mean unique no. sequences   & 7.8      & 100       \\
-\hline
-mean unique no. jump times  & 47.8     & 1076.7    \\
-%MAP sequence OSPA    & $5.36 \times 10^{-3}$ & $4.98 \times 10^{-3}$  \\
-\hline
-\end{tabular}
-\end{table}
+Performance was evaluated by testing on 10 scenarios, each with 1000 observations.
+At each observation time, an estimate of the value and trend is made by taking the average of Gaussian means generated by an RTS smoother for each particle.
+The resulting root mean square errors (RMSEs) are displayed in table~\ref{tab:finance_state_performance}, demonstrating an improvement in accuracy from the smoother.
 
@@
 \hline
- &\raisebox{0cm}[0.4cm][0.3cm]{\parbox[c]{1.5cm}{\centering RBVRPF and KF}} & \raisebox{0cm}[0.4cm][0.3cm]{\parbox[c]{1.5cm}{\centering RBVRPF and RTS}} & \raisebox{0cm}[0.4cm][0.3cm]{\parbox[c]{1.5cm}{\centering RBVRPS and RTS}} \\
-% & \begin{minipage}[c]{1.5cm}\centering RBVRPF + KF\end{minipage} & \begin{minipage}[c]{1.5cm}\centering RBVRPF + RTS\end{minipage} & \begin{minipage}[c]{1.5cm}\centering RBVRPS + RTS\end{minipage} \\
+ &\raisebox{0cm}[0.4cm][0.3cm]{\parbox[c]{1.5cm}{\centering VRPF}} & \raisebox{0cm}[0.4cm][0.3cm]{\parbox[c]{1.5cm}{\centering Filter-smoother}} & \raisebox{0cm}[0.4cm][0.3cm]{\parbox[c]{1.5cm}{\centering VRPS}} \\
 \hline \hline
@@
 \raisebox{0cm}[0.4cm][0.3cm]{\parbox[c]{2cm}{\centering mean trend estimate RMSE}}   & $2.56 \times 10^{-2}$ & $1.79 \times 10^{-2}$ & $1.49 \times 10^{-2}$ \\
-%MAP value estimate RMSE    & $5.98 \times 10^{-4}$ & $4.48 \times 10^{-4}$ & $4.49 \times 10^{-4}$ \\
-%MAP trend estimate RMSE    & $3.13 \times 10^{-2}$ & $1.79 \times 10^{-2}$ & $1.74 \times 10^{-2}$ \\
 \hline
@@
 
-The new RBVRPS algorithm outperforms the filter in terms of both state estimation accuracy and particle diversity.
+%The results are shown in tables~\ref{tab:finance_cp_performance} and ~\ref{tab:finance_state_performance}.% The OSPA parameters are: exponent $p=1$, threshold $c=0.01$.
+
+%\begin{table}%
+%\caption{Changepoint sequence estimation performance.}
+%\label{tab:finance_cp_performance}
+%\centering
+%\renewcommand{\arraystretch}{1.5}
+%\begin{tabular}{|c|c|c|}
+%\hline
+% & RBVRPF & RBVRPS \\
+%\hline \hline
+%mean unique no. sequences   & 7.8      & 100       \\
+%\hline
+%mean unique no. jump times  & 47.8     & 1076.7    \\
+%%MAP sequence OSPA    & $5.36 \times 10^{-3}$ & $4.98 \times 10^{-3}$  \\
+%\hline
+%\end{tabular}
+%\end{table}
+
+%\begin{table}%
+%\caption{State estimation performance.}
+%\label{tab:finance_state_performance}
+%\centering
+%\renewcommand{\arraystretch}{1.5}
+%\begin{tabular}{|c|c|c|c|}
+%\hline
+% &\raisebox{0cm}[0.4cm][0.3cm]{\parbox[c]{1.5cm}{\centering RBVRPF and KF}} & \raisebox{0cm}[0.4cm][0.3cm]{\parbox[c]{1.5cm}{\centering RBVRPF and RTS}} & \raisebox{0cm}[0.4cm][0.3cm]{\parbox[c]{1.5cm}{\centering RBVRPS and RTS}} \\
+%% & \begin{minipage}[c]{1.5cm}\centering RBVRPF + KF\end{minipage} & \begin{minipage}[c]{1.5cm}\centering RBVRPF + RTS\end{minipage} & \begin{minipage}[c]{1.5cm}\centering RBVRPS + RTS\end{minipage} \\
+%\hline \hline
+%\raisebox{0cm}[0.4cm][0.3cm]{\parbox[c]{2cm}{\centering mean value estimate RMSE}}   & $5.31 \times 10^{-4}$ & $4.48 \times 10^{-4}$ & $4.16 \times 10^{-4}$ \\
+%\hline
+%\raisebox{0cm}[0.4cm][0.3cm]{\parbox[c]{2cm}{\centering mean trend estimate RMSE}}   & $2.56 \times 10^{-2}$ & $1.79 \times 10^{-2}$ & $1.49 \times 10^{-2}$ \\
+%%MAP value estimate RMSE    & $5.98 \times 10^{-4}$ & $4.48 \times 10^{-4}$ & $4.49 \times 10^{-4}$ \\
+%%MAP trend estimate RMSE    & $3.13 \times 10^{-2}$ & $1.79 \times 10^{-2}$ & $1.74 \times 10^{-2}$ \\
+%\hline
+%\end{tabular}
+%\end{table}
+
+%The new RBVRPS algorithm outperforms the filter in terms of both state estimation accuracy and particle diversity.
 
@@
 
+The mean number of unique particle estimates at each time step is shown in Fig.~\ref{fig:finance_unique_particles}.
+This illustrates the main advantage of the smoother algorithm --- the increased number of unique particles in the approximation means a better characterisation of the posterior changepoint distribution.
+
+\begin{figure}[!t]
+\centering
+\includegraphics[width=0.9\columnwidth]{finance_unique_particles.pdf}
+\caption{Mean number of unique particles in the filter-smoother (dashed) and smoother (solid) estimates.}
+\label{fig:finance_unique_particles}
+\end{figure}
+
 \subsection{Conditionally Deterministic Model}
@@
 The smoother also provides a more accurate estimate of the number of changepoints in each test, with an average error of 0.6 compared to 2.0 from the final frame filter approximation.
-In addition, the mean number of unique particle estimates at each time step is shown in Fig.~\ref{fig:unique_particles}.
-This illustrates the main advantage of the smoother algorithm --- the increased number of unique particles in the approximation means a better characterisation of the state distribution, allowing, for example, calculation of state covariance measures.
+
+The mean number of unique particle estimates at each time step is shown in Fig.~\ref{fig:tracking_unique_particles}.
+This illustrates the main advantage of the smoother algorithm --- the increased number of unique particles in the approximation means a better characterisation of the posterior changepoint distribution, allowing, for example, calculation of state covariance measures.
 
@@
 \hline
- &\raisebox{0cm}[0.4cm][0.3cm]{\parbox[c]{1.5cm}{\centering VRPF}} & \raisebox{0cm}[0.4cm][0.3cm]{\parbox[c]{1.5cm}{\centering Final frame VRPF}} & \raisebox{0cm}[0.4cm][0.3cm]{\parbox[c]{1.5cm}{\centering VRPS}} \\
+ &\raisebox{0cm}[0.4cm][0.3cm]{\parbox[c]{1.5cm}{\centering VRPF}} & \raisebox{0cm}[0.4cm][0.3cm]{\parbox[c]{1.5cm}{\centering Filter-smoother}} & \raisebox{0cm}[0.4cm][0.3cm]{\parbox[c]{1.5cm}{\centering VRPS}} \\
 % & \begin{minipage}[c]{1.5cm}\centering RBVRPF + KF\end{minipage} & \begin{minipage}[c]{1.5cm}\centering RBVRPF + RTS\end{minipage} & \begin{minipage}[c]{1.5cm}\centering RBVRPS + RTS\end{minipage} \\
@@
 \centering
-\includegraphics[width=0.9\columnwidth]{unique_particles.pdf}
+\includegraphics[width=0.9\columnwidth]{tracking_unique_particles.pdf}
 \caption{Mean number of unique particles in the final filter (dashed) and smoother (solid) estimates.}
-\label{fig:unique_particles}
+\label{fig:tracking_unique_particles}
 \end{figure}

### 2012-04-19 finaldraft_VRsmoothing.tex
Final draft.

--- 2012-04-19 draft3_VRsmoothing.tex
@@
 \graphicspath{{./images/}{./images/finance/}{./images/tracking/}}
+%\interdisplaylinepenalty=1000
 
@@
       \STATE Discard $\theta_{n}^{(i)}$.
-      \STATE Predict state mean and covariance $\tilde{m}_n^{-(i)}$ and $\tilde{P}_n^{-(i)}$ using \ref{eq:backward_kf_predict_start} to \ref{eq:backward_kf_predict_stop}.
+      \STATE Predict state mean and covariance $\tilde{m}_n^{-(i)}$ and $\tilde{P}_n^{-(i)}$ using (\ref{eq:backward_kf_predict_start}) to (\ref{eq:backward_kf_predict_stop}).
       \FOR{$j=1 \dots N_P$}
@@
 
-This model may be discretised at the observation times using standard methods (see e.g. \cite{Grewal2002}), leading to,
+This model may be discretised at the observation times using standard methods (see \cite{Godsill2007a,Christensen2012,Grewal2002}), leading to,
 %
@@
 \centering
-\includegraphics[width=0.9\columnwidth]{finance_unique_particles.pdf}
+\includegraphics[width=0.75\columnwidth]{finance_unique_particles.pdf}
 \caption{Mean number of unique particles in the filter-smoother (dashed) and smoother (solid) estimates.}
@@
 Solving these yields the following state equations (where $\Delta\ t = t - \tau_{K(t)}$):
+%The target dynamics are governed by the following state equations \cite{Bunch2012a} (where $\Delta\ t = t - \tau_{K(t)}$):
 %
@@
 \dot{s}(t) & = & \dot{s}_{K(t)} + a_{T,k} \Delta t \label{eq:2D_ICmodel_start} \\
-\psi(t) & = & \psi_{K(t)} + \frac{a_{N,k}}{a_{T,k}} \log \left( \frac{\dot{s}(t)}{\dot{s}_{K(t)}} \right) \\
+\psi(t) & = & \psi_{K(t)} + \frac{a_{N,k}}{a_{T,k}} \log \left( \frac{\dot{s}(t)}{\dot{s}_{K(t)}} \right)
+\end{IEEEeqnarray}
+\begin{IEEEeqnarray}{rCl}
 x(t) & = & x_{K(t)} + d_{X,K(t)} \Delta t \\
      \IEEEeqnarraymulticol{3}{l}{ \quad + \: \frac{ \dot{s}(t)^2 }{ 4 a_{T,k}^2 + a_{N,k}^2 } \left[  a_{N,k} \sin(\psi(t)) + 2 a_{T,k} \cos(\psi(t))  \right]} \nonumber \\
-     \IEEEeqnarraymulticol{3}{l}{ \quad - \: \frac{\dot{s}_{K(t)}^2}{4 a_{T,k}^2 + a_{N,k}^2} \left[  a_{N,k} \sin(\psi_{K(t)}) + 2 a_{T,k} \cos(\psi_{K(t)})  \right]} \nonumber \\
+     \IEEEeqnarraymulticol{3}{l}{ \quad - \: \frac{\dot{s}_{K(t)}^2}{4 a_{T,k}^2 + a_{N,k}^2} \left[  a_{N,k} \sin(\psi_{K(t)}) + 2 a_{T,k} \cos(\psi_{K(t)})  \right]} \nonumber
+\end{IEEEeqnarray}
+\begin{IEEEeqnarray}{rCl}
 y(t) & = & y_{K(t)} + d_{Y,K(t)} \Delta t \\
@@
 \centering
-\includegraphics[width=0.9\columnwidth]{tracking_unique_particles.pdf}
+\includegraphics[width=0.75\columnwidth]{tracking_unique_particles.pdf}
 \caption{Mean number of unique particles in the final filter (dashed) and smoother (solid) estimates.}
@@
 \bibliographystyle{IEEEtran}
-\bibliography{D:/pb404/Dropbox/PhD/OTbib}
+\bibliography{D:/pb404/Dropbox/PhD/Cleanbib}
 \input{D:/pb404/Dropbox/PhD/Biographies/IEEE-Bunch.tex}

### 2012-04-19 full_finaldraft_VRsmoothing.tex
Final draft.

--- 2012-04-19 finaldraft_VRsmoothing.tex
@@
 \usepackage{mdwtab}
-%\usepackage{eqparbox}
 \usepackage[caption=false,font=footnotesize]{subfig}
@@
 An additional survivor function is included to account for the condition that changepoints cannot occur before $t_{n-1}$.
-%, comprising a product of density terms (one for each new changepoint) and a survivor function term.% In the particular (but common) case that no new changepoints occur within the interval, the density consists of a probability mass on the empty set, with weight $P(\tau_{K_n+1} > t_n | \tau_{K_n+1} > t_{n-1}, \tau_{K_n} )$.
 %
@@
 
-This completes the requirements for the particle filter, resulting in the final algorithm shown in Fig.~\ref{alg:RBVRPF}.% The concept of estimating the posterior for just the changepoint sequence with a particle filter rather than the joint posterior over changepoints and linear states is an example of Rao-Blackwellisation (see, e.g.
-\cite{Casella1996,Doucet2000}), so this algorithm is known as the Rao-Blackwellised variable rate particle filter (RB-VRPF) \cite{Godsill2007a,Christensen2012}.
+This completes the requirements for the particle filter, resulting in the final algorithm shown in Fig.~\ref{alg:RBVRPF}.
+
+The VRPF applied to a conditionally linear-Gaussian model estimates a posterior distribution over the changepoints only.
+The contribution of the linear state is calculated analytically using the Kalman recursions.
+This is an example of Rao-Blackwellisation (see, e.g.
+\cite{Casella1996,Doucet2000}).
 
@@
 
-This leads to a particle filter for variable rate models with conditionally deterministic dynamics, summarised in Fig.~\ref{alg:VRPF}.% The resulting trajectories $x(t)^{(i)}$ for each particle are realisations from a piecewise-deterministic process, so this algorithm is named the piecewise-deterministic variable rate particle filter (PD-VRPF).
+This leads to a particle filter for variable rate models with conditionally deterministic dynamics, summarised in Fig.~\ref{alg:VRPF}.
 
@@
 
-%Filtering schemes which alter past changepoints -- whether using an SMC sampler or RM -- are computationally expensive, because many likelihood calculations must be conducted, for each observation from the time of the change onwards.
-In some cases, it may be simpler to just use a bootstrap filter with more particles.
-
 \section{Variable Rate Particle Smoothing} \label{sec:vrps}
@@
 Estimating changepoints online is a challenging task because the presence of a change may not be obvious until after it has happened.
-Thus, a smoothing algorithm is expected to achieve significantly improved performance at changepoint estimation.
+Thus, a smoothing algorithm is expected to achieve improved performance at changepoint estimation.
 
 The target distribution for a variable rate smoothing algorithm is the posterior distribution over the entire changepoint sequence, $p(\theta|y_{1:N})$.
-This is the distribution approximated by the final step of the VRPF.
+This distribution is approximated by the final step of the VRPF.
 However, in the same manner as the fixed rate filter-smoother of \cite{Kitagawa1996}, this approximation is likely to lack path-space diversity --- because of the necessary resampling step in the filtering algorithm, the particles all share the same set of changepoints before a particular time, with variation only appearing for changepoints closer to $T$.
@@
 By iterating backwards from $n = N \dots 1$, a diverse set of particles approximating $p(\theta|y_{1:N})$ will be generated.
-
-%A particle drawn from $p(\theta_{n}^{+}|y_{1:N})$ may be extended backwards by sampling from the conditional distribution, $p(\theta_{n}|\theta_{n}^{+}, y_{1:N})$ and concatenating the past and future sequences.
-The new particle will be a sample from the target distribution, but the resulting approximation will still suffer from low path-space diversity before time $t_{n-1}$.
-Therefore, $\theta_{n-1}$ is marginalised by simply discarding the changepoints which occur before this time.
-This results in a particle drawn from $p(\theta_{n-1}^{+}|y_{1:N})$.
-The procedure then continues recursively.
-After a complete backwards pass through the observations, a single particle from the target distribution will have been generated.
-The process is then repeated until sufficient samples are obtained.
-
-%At step $n$ for a given particle, the future changepoint sequence may be considered to be fixed.
-It remains to devise a method for drawing samples from the backwards conditional distribution, $p(\theta_{n}|\theta_{n}^{+}, y_{1:N})$.
 
@@
 This may be considered to be an improper density over $x_n$, and may be calculated analytically using a backwards Kalman filter, in a similar manner to that used in the two-filter smoother \cite{Fraser1969,Kitagawa1994,Sarkka2012,Gelb1974}.
-Such a backwards Kalman filter uses the following recursions.
-Derivations may be found in the aforesaid references.
+Such a backwards Kalman filter uses the following recursions, which are derived in appendix~\ref{app:backwards_KF}.
 %
@@
 
-Exact methods for initialising this recursion are discussed in \cite{Kitagawa1994}.
+Exact methods for initialising this recursion are discussed in appendix~\ref{app:init_backwards_KF}, and also in \cite{Kitagawa1994}.
 However, it is often sufficient to use an approximation, for example $\tilde{m}_N^{(i)} = m_N^{(i)}$ and $\tilde{P}_N^{(i)} = P_N^{(i)}$.
@@
 Metropolis-Hastings (MH) steps are conducted to draw samples, $\{\tilde{\theta}_{n}, u_{K_n}'\}$, from the target distribution by sampling the proposal and accepting the new values with a given probability.
-% (IS could be used, but this would require resampling steps, which would lead to a loss of path-space diversity)
+Note that IS could be used instead of MCMC, but this would require resampling steps, which would lead to a loss of path-space diversity similar to that suffered by the particle filter.
 
@@
 where $r_n^- = \min( m : t_m > \tau_{K_n} )$ and $r_n^+ = \max( m : t_m < \tau_{K_n+1} )$ are the indexes of the earliest and latest observations to fall between the changepoints each side of $t_n$.
-
-%The transition density term will include a delta function with unit probability mass at the point for which the past and future state trajectories match up.
-
-%\begin{IEEEeqnarray}{rCl}
-%\IEEEeqnarraymulticol{3}{l}{p(\tilde{\theta}_{n}^{+}|\tilde{\theta}_{n}') \propto p(u_{K_n+1}|\tau_{K_n+1}, \tau_{K_n}, u_{K_n})} \nonumber \\ \times p(\tau_{K_n+1}|\tau_{K_n}, u_{K_n}) \delta_{f(x_{K_n}, v_{K_n}, \tau_{K_n}, \tau_{K_n+1})}(x_{K_n+1})
-%\end{IEEEeqnarray}
 
@@
 Each Markov chain may be initialised with the output from the previous stage.
-This value is itself a particle from the target distribution (albeit one from an approximation with poor diversity), meaning that no burn-in period is required.
+This value is itself a particle from the target distribution (albeit one from an approximation with poor path-space diversity), meaning that no burn-in period is required.
 
@@
 Each particle consists of a list of changepoint times and parameters, and there is no obvious way to take an average over the set to produce a single estimate.
-Furthermore, if a best estimate could be calculated, it is not trivial to define a distance metric between the estimated and true sequences to assess relative accuracy.
+Furthermore, if a best estimate could be calculated (e.g.
+by finding the maximum a posteriori sequence), it is not trivial to define a distance metric between the estimated and true sequences to assess relative accuracy.
 
@@
 
-In the following examples, we compare the state estimates from the VRPF, the ``filter-smoother'' and the VRPS.
+In the following examples, we compare the state estimates from the ``filter-smoother'' and the VRPS.
 Filter-smoother estimates use the particle approximation from the final step of the VRPF for the changepoint distribution.
@@
 where $\lambda$ introduces a mean regression effect on the trend and $\mathbf{\beta}(t)$ is standard Brownian motion (with unit diffusion constant).
-The jump term, $d\mathbf{J}(t)$, is zero everywhere except where jumps occur.
+The jump term, $d\mathbf{J}(t)$, is zero everywhere except when jumps occur.
 %
@@
 
-Jumps occur at a random set of times, $\{\tau_k\}$, and are one of two types: value jumps, indicated by $u_k = 1$, and trend jumps, indicated by $u_k=2$.
+$\{\tau_k\}$ is the set of random jump times, each of which is one of two types: a value jump, indicated by $u_k = 1$, or a trend jump, indicated by $u_k=2$.
 
@@
 
-This model may be discretised at the observation times using standard methods (see \cite{Godsill2007a,Christensen2012,Grewal2002}), leading to,
+This model may be discretised at the observation times using standard methods, detailed in appendix~\ref{app:lg_model_discretisation}, leading to,% (see \cite{Godsill2007a,Christensen2012,Grewal2002}), leading to,
 %
@@
 
-where the $\mathbf{w}_n$ and $v_n$ are Gaussian random variables with covariance matrixes $Q_n$ and $R$ respectively.% The time between observations times is denotes $\Delta t = t_{n} - t_{n-1}$.
+where the $\mathbf{w}_n$ and $v_n$ are Gaussian random variables with covariance matrixes $Q_n$ and $R$ respectively.
+$Q_n$ is a function of the sequence of jump times and their types, giving rise to the desired conditionally linear-Gaussian structure.
+% The time between observations times is denotes $\Delta t = t_{n} - t_{n-1}$.
 %
@@
 
-The times between changepoints were assumed to be exponentially distributed, with equal probability of value and trend jumps.
+The times between changepoints are assumed to be exponentially distributed, with equal probability of value and trend jumps.
 %
@@
 The jump standard deviations are 0.005 and 0.05 respectively for value and trend.
+(Terms defined in appendix~\ref{app:lg_model_discretisation})
 
@@
 
-The filter used $N_P = 100$ particles, and the smoother resampled $N_S = 100$ sequences.
+The filter used $N_F = 100$ particles, and the smoother resampled $N_S = 100$ sequences.
 Bootstrap proposals were used for the filter.
@@
 
-%\appendices
-%\section{Initialisation of Backwards Kalman Filter} \label{app:init_backwards_KF}
-%
-%The backwards Kalman filter recursion may be initialised exactly by inversion of the observation model.
+\appendices
+\section{The Backwards Kalman Filter} \label{app:backwards_KF}
+
+A backwards Kalman filter \cite{Fraser1969,Kitagawa1994} is used to calculate the joint likelihood term, $p(y_{n+1:N}|x_n, \tilde{\theta}_{n}^+)$, when the state dynamics are linear-Gaussian.
+Assume that at time $t_{n+1}$ there is a Gaussian density such that,
+%
+\begin{equation}
+ p(y_{n+1:N}|x_{n+1}, \tilde{\theta}_{n}^+) = \tilde{Z}_{n+1} \mathcal{N}(x_{n+1}|\tilde{m}_{n+1}, \tilde{P}_{n+1})  .
+\end{equation}
+
+The state transition model is linear-Gaussian, so the backwards predictive density is given by,
+%
+\begin{IEEEeqnarray}{rCl}
+\IEEEeqnarraymulticol{3}{l}{p(y_{n+1:N}|x_n, \tilde{\theta}_{n}^+)} \nonumber \\
+  & = & \int p(y_{n+1:N}|x_{n+1}, \tilde{\theta}_{n}^+)  p(x_{n+1}|x_n, \tilde{\theta}_{n}^+) dx_{n+1} \nonumber \\
+  & = & \int \tilde{Z}_{n+1} \mathcal{N}(x_{n+1}|\tilde{m}_{n+1}, \tilde{P}_{n+1}) \nonumber \\
+  &   & \qquad \mathcal{N}(x_{n+1}|A_{n+1} x_{n}, Q_{n+1}) dx_{n+1} \nonumber \\
+  & = & \int \tilde{Z}_{n+1} ||A_{n+1}^{-1}|| \mathcal{N}(x_{n+1}|\tilde{m}_{n+1}, \tilde{P}_{n+1}) \nonumber \\
+  &   & \qquad \mathcal{N}(x_{n}|A_{n+1}^{-1} x_{n+1}, A_{n+1}^{-1} Q_{n+1} A_{n+1}^{-T}) dx_{n+1} \nonumber \\
+  & = & \tilde{Z}_{n}^- \mathcal{N}(x_n|\tilde{m}_n^-, \tilde{P}_n^-)
+\end{IEEEeqnarray}
+
+where,
+%
+\begin{IEEEeqnarray}{rCl}
+ \tilde{m}_n^- & = & A_{n+1}^{-1} \tilde{m}_{n+1} \\
+ \tilde{P}_n^- & = & A_{n+1}^{-1} (\tilde{P}_{n+1} + Q_{n+1}) A_{n+1}^{-T} \\
+ \tilde{Z}_n^- & = & \tilde{Z}_{n+1} ||A_{n+1}||^{-1}   .
+\end{IEEEeqnarray}
+
+$||A||$ is used to denote the magnitude of the determinant of $A$, and it is assumed that $A$ is non-singular.
+
+The update step simply involves multiplying by the likelihood of the observation at time $t_n$,
+%
+\begin{IEEEeqnarray}{rCl}
+\IEEEeqnarraymulticol{3}{l}{p(y_{n:N}|x_n, \tilde{\theta}_{n}^+)} \\
+ & = & p(y_{n+1:N}|x_n, \tilde{\theta}_{n}^+) p(y_{n}|x_n, \tilde{\theta}_{n}^+) \nonumber \\
+ & = & \tilde{Z}_n^- \mathcal{N}(x_n|\tilde{m}_n^-, \tilde{P}_n^-) \mathcal{N}(y_n|C_n x_n, R_n) \nonumber \\
+ & = & \tilde{Z}_n \mathcal{N}(x_n|\tilde{m}_n, \tilde{P}_n)
+\end{IEEEeqnarray}
+
+where,
+%
+\begin{IEEEeqnarray}{rCl}
+ \tilde{\mu}_n & = & C_n \tilde{m}_n^- \\
+ \tilde{S}_n   & = & C_n \tilde{P}_n^- C_n^T + R_n \\
+ \tilde{K}_n   & = & \tilde{P}_n^- C_n^T \tilde{S}_n^{-1} \\
+ \tilde{m}_n   & = & \tilde{m}_n^- + \tilde{K}_n (y_n - \tilde{\mu}_n) \\
+ \tilde{P}_n   & = & \tilde{P}_n^- - \tilde{K}_n \tilde{S}_n \tilde{K}_n^T \\
+ \tilde{Z}_n   & = & \tilde{Z}_n^- \mathcal{N}(y_n|\tilde{\mu}_n, \tilde{S}_n)   .
+\end{IEEEeqnarray}
+
+All steps use only standard Gaussian identities.
+
+\section{Initialisation of the Backwards Kalman Filter} \label{app:init_backwards_KF}
+
+The backwards Kalman filter recursion may be initialised exactly by inversion of the observation model using the method of \cite{Kitagawa1994}.
 By stacking up the last $L+1$ observations into a single vector, it is possible to write down a joint Gaussian distribution conditional on the $(N-L)$th state.
 This may be inverted to produce an improper Gaussian density for $x_{N-L}$,
-%%
+%
 %\begin{IEEEeqnarray}{rCl}
-% \IEEEeqnarraymulticol{3}{l}{p(y_{N-L:N}|x_{N-L}) = \mathcal{N}( \mathbf{y}_{N-L:N} | H_{N-L} x_{N-L}, \Gamma_{N-L} ) } \\
-% & \propto & \mathcal{N}(x_{N-L} | (H_{N-L}^T \Gamma_{N-L} H_{N-L})^{-1} H_{N-L}^T \Gamma_{N-L}, (H_{N-L}^T \Gamma_{N-L} H_{N-L})^{-1} )      ,
+% \IEEEeqnarraymulticol{3}{l}{p(y_{N-L:N}|x_{N-L}) = \mathcal{N}( \mathbf{y}_{N-L:N} | H_{N-L} x_{N-L}, \Gamma_{N-L} ) } \nonumber \\
+% & = & \sqrt{\frac{|(H_{N-L}^T \Gamma_{N-L}^{-1} H_{N-L})^{-1}|}{|\Gamma_{N-L}|}} \nonumber \\
+% & & \times \mathcal{N}(x_{N-L} | (H_{N-L}^T \Gamma_{N-L}^{-1} H_{N-L})^{-1} H_{N-L}^T \Gamma_{N-L}^{-1} \mathbf{y}_{N-L:N}, \nonumber \\
+% & & (H_{N-L}^T \Gamma_{N-L}^{-1} H_{N-L})^{-1} )      ,
 %\end{IEEEeqnarray}
-%
-%where $H_{N-L}$ and $\Gamma_{N-L}$ may be constructed from the system models and,
-%%
-%\begin{IEEEeqnarray}{rCl}
-%\mathbf{y}_{N-L:N} & = & \begin{bmatrix}y_N \\ y_{N-1} \\ \vdots \\ y_{N-L} \end{bmatrix}     .
+\begin{IEEEeqnarray}{rCl}
+ \IEEEeqnarraymulticol{3}{l}{p(y_{N-L:N}|x_{N-L}) = \mathcal{N}( \mathbf{y}_{N-L:N} | H_{N-L} x_{N-L}, \Gamma_{N-L} ) } \nonumber \\
+ & = & \sqrt{\frac{|V_{N-L}|}{|\Gamma_{N-L}|}} \mathcal{N}(x_{N-L} | \nu_{N-L}, V_{N-L} )
+\end{IEEEeqnarray}
+\begin{IEEEeqnarray}{rCl}
+ V_{N-L} & = & (H_{N-L}^T \Gamma_{N-L}^{-1} H_{N-L})^{-1} \\
+ \nu_{N-L} & = & V_{N-L} H_{N-L}^T \Gamma_{N-L}^{-1} \mathbf{y}_{N-L:N}     ,
+\end{IEEEeqnarray}
+
+where $\mathbf{y}_{N-L:N}$ is a vector of concatenated observations, and $H_{N-L}$ and $\Gamma_{N-L}$ may be constructed from the system models,
+%
+%\begin{IEEEeqnarray}{rClCrCl}
+%\mathbf{y}_{N-L:N} & = & \begin{bmatrix}y_N \\ y_{N-1} \\ \vdots \\ y_{N-L} \end{bmatrix} & , \qquad & H_{N-L} & = & \begin{bmatrix} C A^L \\ C A^{L-1} \\ \vdots \\ C \end{bmatrix}
 %\end{IEEEeqnarray}
-%
-%This should effected for the smallest value of $L$ such that the matrix inversions possible.
-
-%\newpage
+\begin{IEEEeqnarray}{rCl}
+\mathbf{y}_{N-L:N} & = & \begin{bmatrix}y_N \\ y_{N-1} \\ \vdots \\ y_{N-L} \end{bmatrix}
+\end{IEEEeqnarray}
+\begin{IEEEeqnarray}{rCl}
+H_{N-L} & = & \begin{bmatrix} C A^L \\ C A^{L-1} \\ \vdots \\ C \end{bmatrix}
+\end{IEEEeqnarray}
+\begin{IEEEeqnarray}{rCl}
+ \Gamma_{N-1} & = & K_{N-L} \begin{bmatrix} Q_N & 0 & \dots & 0 \\ 0 & Q_{N-1} & \dots & 0 \\ \vdots & \vdots & \ddots & \vdots \\ 0 & 0 & \dots & Q_{N-L+1} \end{bmatrix} K_{N-L}^T \nonumber \\
+  & & + \: \begin{bmatrix} R_N & 0 & \dots & 0 \\ 0 & R_{N-1} & \dots & 0 \\ \vdots & \vdots & \ddots & \vdots \\ 0 & 0 & \dots & R_{N-L} \end{bmatrix}
+\end{IEEEeqnarray}
+\begin{IEEEeqnarray}{rCl}
+ K_{N-L} & = & \begin{bmatrix} C & CA & \dots & CA^{L-1} \\ 0 & C & \dots & CA^{L-2} \\ \vdots & \vdots & \ddots & \vdots \\ 0 & 0 & \dots & C \\ 0 & 0 & \dots & 0 \end{bmatrix}
+\end{IEEEeqnarray}
+
+In the above, it is assumed for clarity that $A_n = A$ and $C_n = C$ are constant matrices.
+This assumption is easily relaxed.
+This inversion should be effected for the smallest value of $L$ such that the matrix inverses exist.
+
+\section{Discretisation of the Jump-Diffusion Model} \label{app:lg_model_discretisation}
+
+State evolution is governed by the following stochastic differential equation,
+%
+\begin{IEEEeqnarray}{c}
+ d\mathbf{x}(t) = \begin{bmatrix}0 & 1 \\ 0 & -\lambda \end{bmatrix} \mathbf{x}(t) dt + \begin{bmatrix}0 \\ \sigma \end{bmatrix} d\beta(t) + d\mathbf{J}(t)
+\end{IEEEeqnarray}
+
+where $\lambda$ is the mean regression constant, $\sigma$ is the diffusion constant, $\mathbf{\beta}(t)$ is standard Brownian motion, $d\mathbf{J}(t)$ is a jump term defined according to,
+%
+\begin{IEEEeqnarray}{rCl}
+ d\mathbf{J}(t) & = & \begin{cases} \mathbf{J}_k & t \in \{\tau_k\} \\ 0 & \text{elsewhere} \end{cases} \\
+ \mathbf{J}_k  & \sim & \mathcal{N}(\mathbf{J}_k| \mathbf{0}, Q_{J,u_k})
+\end{IEEEeqnarray}
+
+$\{\tau_k\}$ is the set of jump times, and $\{u_k\}$ the set of jump types, where $u_k = 1$ indicates a value jump, and $u_k=2$ a trend jump. The jump covariance matrices are,
+%
+\begin{IEEEeqnarray}{c}
+Q_{J,u_k} = \begin{cases} \begin{bmatrix}\sigma_{J1}^2 & 0 \\ 0 & 0 \end{bmatrix} & u_k = 1 \\
+                          \begin{bmatrix}0 & 0 \\ 0 & \sigma_{J2}^2 \end{bmatrix} & u_k = 2  \end{cases}   .
+\end{IEEEeqnarray}
+
+Discretisation proceeds by separating the diffusion and jump terms and integrating separately.
+The jump term is straightforward to integrate.
+The diffusion term is a standard linear time invariant stochastic differential equation, and may be solved using the methods described in, for example, \cite{Oksendal2003,Grewal2002}.
+Matrix fraction decomposition is used for solving the covariance propagation equation \cite{Grewal2002,Sarkka2006}.
+The resulting discrete-time system is governed by a difference equation,
+%
+\begin{IEEEeqnarray}{rCl}
+ \mathbf{x}_n &=& A \mathbf{x}_{n-1} + \mathbf{w}_n
+\end{IEEEeqnarray}
+
+where the $\mathbf{w}_n$ is a Gaussian random variable with covariance matrix $Q_n$. The inter-observation time is denoted $\Delta t = t_{n} - t_{n-1}$.
+
+\begin{IEEEeqnarray}{rCl}
+ A     &=& \begin{bmatrix}1 & \frac{1}{\lambda}(1-e^{(-\lambda \Delta t)} \\ 0 & e^{(-\lambda \Delta t)}\end{bmatrix} \\
+ Q_n   &=& \begin{cases}Q_{D} + Q_{J,u_k} & \exists k : \tau_k \in [t_{n-1},t_n]\\
+                        Q_{D}             & \text{otherwise} \end{cases} \\
+ Q_{D} &=& \frac{\sigma^2}{2 \lambda}\begin{bmatrix} q_1 & q_2 \\ q_2 & q_3\end{bmatrix} \\
+ q_1   &=& \frac{1}{\lambda^2}(2 \lambda \Delta t - (3 - e^{(-\lambda \Delta t)})(1 - e^{(-\lambda \Delta t)})~) \\
+ q_2   &=& \frac{1}{\lambda} (1-e^{(-\lambda \Delta t)})^2 \\
+ q_3   &=& 1-e^{(-2 \lambda \Delta t)}     .
+\end{IEEEeqnarray}
+
+Assuming only the value is observed, corrupted by Gaussian noise with standard deviation $\sigma_y^2$, the measurement model is defined by,
+%
+\begin{IEEEeqnarray}{rCl}
+ y_n &=& C \mathbf{x}_{n} + v_n
+\end{IEEEeqnarray}
+
+where $v_n$ is a Gaussian random variable with covariance matrix $R$.
+%
+\begin{IEEEeqnarray}{rCl}
+ C &=& \begin{bmatrix}1 & 0\end{bmatrix} \\
+ R &=& \sigma_y^2     .
+\end{IEEEeqnarray}
+
+\newpage
 \bibliographystyle{IEEEtran}

### 2012-04-20 full_finaldraft_VRsmoothing_withSJGcomments.tex
Simon's mods

--- 2012-04-19 full_finaldraft_VRsmoothing.tex
@@
 
-\title{Smoothing Algorithms for Variable Rate Models}
+\title{Particle Smoothing Algorithms for Variable Rate Models}
 
@@
 \markboth{IEEE Transaction in Signal Processing,~Vol.~1, No.~1, January~1901}%
-{Bunch \& Godsill: Smoothing Algorithms for Variable Rate Models}
+{Bunch \& Godsill: Particle Smoothing Algorithms for Variable Rate Models}
 
@@
 However, estimates can often be improved later once further observations have been made.
-In this paper, we address the problem of smoothing in variable models, i.e.
+In this paper, we address the problem of smoothing in variable rate models, i.e.
 the estimation of the changepoint sequence and latent state given all the observations.
-This is achieved with an efficient backward sweep through the observations, in a similar manner to the method for standard HMMs described in \cite{Godsill2004}.
+This is achieved in this paper with an efficient backward sweep through the observations, in a new version of the method for standard HMMs described in \cite{Godsill2004}.
 Two new schemes are introduced: one for conditionally linear-Gaussian models which exploits the method of Rao-Blackwellisation; and a second for use with conditionally-deterministic models which uses an augmented target distribution in the style of an SMC sampler \cite{DelMoral2006}.
@@
 During this period, an unknown number of changepoints, $K$, occur at times $\{\tau_0 = 0, \tau_1 \dots \tau_K \}$, each with associated changepoint parameters, $\{ u_0, u_1 \dots u_K \}$.
-The pairs $\{\tau_k, u_k\}$ are the elements of an marked point process (MPP).
+The pairs $\{\tau_k, u_k\}$ are the elements of a marked point process (MPP).
 The latent state is a continuous-time process denoted $x(t)$.
@@
 
-The changepoint sequence is assumed to be a Markov process.
-%
-\begin{IEEEeqnarray}{rCl}
- \{\tau_k, u_k\} & \sim & p(\tau_k, u_k|\tau_{k-1}, u_{k-1}) \label{eq:cp_model}
+The changepoint sequence is assumed to be a Markov process,
+%
+\begin{IEEEeqnarray}{rCl}
+ \{\tau_k, u_k\} & \sim & p(\tau_k, u_k|\tau_{k-1}, u_{k-1}) \label{eq:cp_model}     .
 \end{IEEEeqnarray}
@@
 
+This comprises a density term for each changepoint in the sequence and a survivor function term accounting for the probability that no other changepoints occur within the time interval.
+
 \subsection{Conditionally Linear-Gaussian Models}
 
-The first class of variable rate models to be considered is those whose state dynamics are linear-Gaussian conditional on the changepoint sequence.
-Such a model may be discretised onto the set of observation times in exactly the same manner as a standard the standard HMM.
+We first consider the class of variable rate models whose state dynamics are linear-Gaussian conditional on the changepoint sequence.
+Such a model may be discretised onto the set of observation times in exactly the same manner as a standard the standard HMM,
 %
@@
 
-where,
+where $w_n$ and $v_n$ are independent Gaussian variables (although the independence condition may be relaxed \cite{Grewal2002}),
 %
@@
 
-In addition, the prior state distribution should be Gaussian, with known mean and variance.
+In addition, the prior state distribution should be Gaussian, with known mean and covariance,
 %
@@
 
-If the changepoint sequence is known, or has been estimated, then the state values, $x_n$ may be inferred using optimal Kalman filtering and smoothing recursions.
-As well as the basic Kalman filter \cite{Kalman1960}, the Rauch-Tung-Striebel (RTS) smoother \cite{Rauch1965} and two-filter smoother \cite{Fraser1969} will prove useful for this step.
-
-Conditionally linear-Gaussian variable rate models were introduced in \cite{Godsill2007a} for a financial inference algorithm.
-Changepoints correspond to jumps in the value or trend of a security, at which points the process covariance, $Q_n(\theta_n)$, is inflated.
+If the changepoint sequence is known, or has been estimated, then the state values, $x_n$ may be inferred conditionally using optimal Kalman filtering and smoothing recursions.
+As well as the basic Kalman filter \cite{Kalman1960}, the Rauch-Tung-Striebel (RTS) smoother \cite{Rauch1965} and two-filter smoother \cite{Fraser1969} will prove useful for this step later in the paper.
+
+Conditionally linear-Gaussian models for use with variable rate particle filters were introduced in \cite{Godsill2007a,Christensen2012} for financial and tracking problems.
+For example, changepoints might correspond to jumps in the value or trend of a security, at which points the process covariance, $Q_n(\theta_n)$, is inflated in an optimal fashion.
 
@@
 
-In general, the state dynamics will be governed by a differential equation which may depend on the entire changepoint sequence. Here we assume that only the most recent changepoint is significant.
+In general, the state dynamics will be governed by a differential equation which may depend on the entire changepoint sequence.
+Here we assume that only the most recent changepoint determines the future state trajectory,
 %
@@
 By introducing a new sequence, $\{ x_0, x_1 \dots x_K \}$, which denotes the value of the state at each changepoint (i.e.
-$x(\tau_k)$), and assuming that an analytic solution exists, a state transition function may be found,
+$x_k = x(\tau_k)$), and assuming that an analytic solution exists, a piecewise function describing the path may be found,
 %
@@
 By choosing $t = \tau_{K_{n}+1}$, this equation specifies the state at the next changepoint time.
-Similarly, by choosing $t=t_n$, the state at the observation times may be calculated --- these points will be denoted $\hat{x}_n$.
+Similarly, by choosing $t=t_n$, the state at the observation times may be calculated --- these latter points will be denoted $\hat{x}_n$.
 To complete the model, a probabilistic measurement model must be devised for the observation process, $p(y_n|\hat{x}_n)$.
@@
 
-The transition term, $p(\theta_{n \setminus n-1} | \theta_{n-1})$, has a similar form to the changepoint prior of (\ref{eq:cp_sequence_prior}) \cite{Jacobsen2006}.
-An additional survivor function is included to account for the condition that changepoints cannot occur before $t_{n-1}$.
+The transition term, $p(\theta_{n \setminus n-1} | \theta_{n-1})$, has a similar form to the changepoint prior of (\ref{eq:cp_sequence_prior}) \cite{Jacobsen2006}, but an additional survivor function is included to account for the condition that changepoints cannot occur before $t_{n-1}$,
 %
@@
 \IEEEeqnarraymulticol{3}{l}{p(\theta_{n \setminus n-1} | \theta_{n-1}) = S(\tau_{K_n}, t_n) / S(\tau_{K_{n-1}+1}, t_{n-1}) } \nonumber \\
- \qquad  \qquad & & \times \prod\limits_{j:t_{n-1} \leq \tau_j < t_n} p(\tau_j, u_j| \tau_{j-1}, u_{j-1})  \label{eq:cp_sequence_trandens}
+ \qquad  \qquad & & \times \prod\limits_{j:t_{n-1} \leq \tau_j < t_n} p(\tau_j, u_j| \tau_{j-1}, u_{j-1})  \label{eq:cp_sequence_trandens}     .
 \end{IEEEeqnarray}
@@
 
-Practically, because changepoints will be relatively rare events, it is not likely that more than one new changepoint will occur between $t_{n-1}$ and $t_n$.
-
-The target distribution of (\ref{eq:vrpf_target}) cannot be calculated analytically, but may be approximated numerically.
+Practically, because changepoints will be relatively rare events in most models, it is not likely that more than one new changepoint will occur between $t_{n-1}$ and $t_n$.
+
+The target distribution of (\ref{eq:vrpf_target}) cannot be calculated analytically, but may be approximated numerically, for example, by using a particle filter.
+
 A particle filter is an algorithm for approximating a probability distribution using a set of weighted samples (or ``particles'') drawn from that distribution.
@@
 \begin{equation}
- \hat{p}(\theta_{n}|y_{1:n}) = \sum_j w_n^{(j)} \delta_{\theta_{n}^{(j)}}(\theta_{n})
+ \hat{p}(\theta_{n}|y_{1:n}) = \sum_j w_n^{(j)} \delta_{\theta_{n}^{(j)}}(\theta_{n}) \label{eq:vrpf}
 \end{equation}
@@
 The particle filter works recursively.
-At the $n$th step, a particle, $\theta_{n-1}^{(i)}$, is first resampled from those approximating the filtering distribution at the $(n-1)$th step, using an appropriately chosen set of proposal weights.
+At the $n$th step, a particle, $\theta_{n-1}^{(i)}$, is first resampled from those approximating the filtering distribution at the $(n-1)$th step, using an appropriately chosen set of proposal weights, $\{v_{n-1}^{(j)}\}$ (where $\sum_j v_{n-1}^{(j)} = 1$),
 %
 \begin{equation}
- q(\theta_{n-1}) = \sum_j v_{n-1}^{(j)} \delta_{\theta_{n-1}^{(j)}}(\theta_{n-1})
+ q(\theta_{n-1}) = \sum_j v_{n-1}^{(j)} \delta_{\theta_{n-1}^{(j)}}(\theta_{n-1})     .
 \end{equation}
@@
 
-Next, an extension to the changepoint sequence, $\theta_{n \setminus n-1}^{(i)}$, is proposed from an importance distribution, $q(\theta_{n \setminus n-1}|\theta_{n-1}^{(i)}, y_n)$, and concatenated with $\theta_{n-1}$ to create an estimate of $\theta_n$.
-Finally, the particle is weighted according to the ratio of the target and proposal densities.
+Next, an extension to the changepoint sequence, $\theta_{n \setminus n-1}^{(i)}$, is sampled from an importance distribution, $q(\theta_{n \setminus n-1}|\theta_{n-1}^{(i)}, y_n)$, and concatenated with $\theta_{n-1}$ to create a proposal for $\theta_n$.
+Finally, the particle is weighted according to the ratio of the target and proposal densities,
 %
@@
     & \propto & \frac{ p(y_n|\theta_n, y_{1:n-1}) p(\theta_{n \setminus n-1}|\theta_{n-1}) p(\theta_{n-1}|y_{1:n-1}) }{ q(\theta_{n-1}) q(\theta_{n \setminus n-1}|\theta_{n-1}^{(i)}, y_n) } \nonumber \\
-    & =       & \frac{w_{n-1}^{(i)}}{v_{n-1}^{(i)}} \times \frac{ p(y_n|\theta_n, y_{1:n-1}) p(\theta_{n \setminus n-1}|\theta_{n-1}) }{ q(\theta_{n \setminus n-1}|\theta_{n-1}^{(i)}, y_n) } \label{eq:vrpf_weights}
+    & =       & \frac{w_{n-1}^{(i)}}{v_{n-1}^{(i)}} \times \frac{ p(y_n|\theta_n, y_{1:n-1}) p(\theta_{n \setminus n-1}|\theta_{n-1}) }{ q(\theta_{n \setminus n-1}|\theta_{n-1}^{(i)}, y_n) } \label{eq:vrpf_weights}     .
 \end{IEEEeqnarray}
@@
 This final future changepoint is discarded.
-(This process can be thought of as sampling the entire future changepoint sequence from $t_{n-1}$ onwards, and then marginalising those which fall after $t_n$.) The bootstrap proposal leads to the usual simplification of the weight formula.
-% (apart from the first which is sampled from (\ref{eq:cp_cond_model}))
-%
-\begin{IEEEeqnarray}{rCl}
-w_n^{(i)} & = & \frac{w_{n-1}^{(i)}}{v_{n-1}^{(i)}} \times p(y_n|\hat{x}_n) \label{eq:bootstrap_vrpf_weights}
+(This process can be thought of as sampling the entire future changepoint sequence from $t_{n-1}$ onwards, and then marginalising those which fall after $t_n$.) The bootstrap proposal leads to the usual simplification of the weight formula, % (apart from the first which is sampled from (\ref{eq:cp_cond_model}))
+%
+\begin{IEEEeqnarray}{rCl}
+w_n^{(i)} & = & \frac{w_{n-1}^{(i)}}{v_{n-1}^{(i)}} \times p(y_n|\hat{x}_n) \label{eq:bootstrap_vrpf_weights}     .
 \end{IEEEeqnarray}
@@
 The bootstrap versions of the VRPF may perform poorly if changepoints are not obvious until significantly after they occur.
-For example, in a tracking example, if a jump occurs in the acceleration, but only the position is observed, then this change may not be obvious until a number of observations have arrived.
+For example, in a tracking example, if a jump occurs in the acceleration, but only the position is observed, then this change may not be obvious until a number of additional observations have arrived.
 In this case, the estimation may be improved by the introduction of resample-move (RM) steps \cite{Gilks2001}.
@@
 The purpose of a smoother is to produce a revised estimate once all the observations have been made, using future values to improve upon the filter performance.
-Estimating changepoints online is a challenging task because the presence of a change may not be obvious until after it has happened.
-Thus, a smoothing algorithm is expected to achieve improved performance at changepoint estimation.
+Estimating changepoints online is a challenging task because the presence of a change may not be obvious until after it has happened, as discussed above.
+Thus, a smoothing algorithm is expected to achieve improved performance in changepoint and state estimation.
 
@@
 For a better characterisation of the smoothing distribution, it is necessary to rejuvenate the set of particles.
-This is achieved with a backward pass through the observations in a similar manner to the forward-backward method described in \cite{Godsill2004}.
+This is achieved here with a backward pass through the observations in an extension of the forward-backward method described in \cite{Godsill2004}.
 
@@
 It remains to devise a scheme for sampling the conditional distribution, $p(\theta_{n}|\theta_{n}^{+}, y_{1:N})$.
-The simplifications used by the fixed rate particle smoother which exploit the Markovian nature of the state sequence \cite{Godsill2004} are of no help, because past changepoints are not independent of future observations (i.e.
+The simplifications used by the fixed rate particle smoother which exploit the Markovian nature of the state sequence \cite{Godsill2004} are of no help, because past changepoints are not independent of future observations in variable rate models (i.e.
 $p(\theta_{n}|\theta_{n}^{+}, y_{1:N}) \ne p(\theta_{n}|\theta_{n}^{+}, y_{1:n}))$.
 However, when the state dynamics are linear-Gaussian conditional on the changepoint sequence, it is possible split the dependence on the observations into past and future by introducing the current state as an additional variable.
-This trick was used in \cite{Sarkka2012} in the derivation of the Rao-Blackwellised particle filter for fixed rate conditionally linear-Gaussian models.
+This device was used in \cite{Sarkka2012} in the derivation of the Rao-Blackwellised particle filter for fixed rate conditionally linear-Gaussian models,
 %
@@
        & = & \int p(y_{n+1:N}|x_n, \theta_{n}^+) p(x_n|\theta_{n}, y_{1:n}) dx_n \nonumber \\
-       &   & \times p(\theta_{n}^+|\theta_{n}) p(\theta_{n}|y_{1:n})
-\end{IEEEeqnarray}
-
-Finally, the particle approximation is substituted for the filtering distribution.
-%
-\begin{IEEEeqnarray}{rCl}
-\hat{p}(\theta_{n}|\theta_{n}^+, y_{1:N}) & = & \sum_i \tilde{w}_{n}^{(i)} \delta_{\theta_{n}^{(i)}}(\theta_{n})  \label{eq:rb-vrps_back_cond}
-\end{IEEEeqnarray}
-
-where the backwards conditional weights are given by
+       &   & \times p(\theta_{n}^+|\theta_{n}) p(\theta_{n}|y_{1:n})     .
+\end{IEEEeqnarray}
+
+Finally, the particle approximation of (\ref{eq:vrpf}) is substituted for the filtering distribution, $p(\theta_{n}|y_{1:n})$,
+%
+\begin{IEEEeqnarray}{rCl}
+\hat{p}(\theta_{n}|\theta_{n}^+, y_{1:N}) & = & \sum_i \tilde{w}_{n}^{(i)} \delta_{\theta_{n}^{(i)}}(\theta_{n}) \label{eq:rb-vrps_back_cond}     ,
+\end{IEEEeqnarray}
+
+where the backwards conditional weights are given by,
 %
@@
  \tilde{w}_n & \propto & w_n \int p(y_{n+1:N}|x_n, \theta_{n}^+) \nonumber \\
-             &         & \times p(x_n|\theta_{n}^{(i)}, y_{1:n}) dx_n p(\theta_{n}^+|\theta_{n}^{(i)}) \label{eq:rb-vrps_back_cond_weight}
+             &         & \times p(x_n|\theta_{n}^{(i)}, y_{1:n}) dx_n p(\theta_{n}^+|\theta_{n}^{(i)}) \label{eq:rb-vrps_back_cond_weight}     .
 \end{IEEEeqnarray}
@@
 The integral in~(\ref{eq:rb-vrps_back_cond_weight}) contains two terms involving the current state, $x_n$.
-The first is the familiar state posterior generated by the Kalman filter using (\ref{eq:kf_predict_start}--\ref{eq:kf_update_stop}).
-%
-\begin{IEEEeqnarray}{rCl}
-p(x_n|\theta_{n}^{(i)}, y_{1:n}) & = & \mathcal{N}(x_n|m_n^{(i)}, P_n^{(i)})
+The first is the familiar state posterior generated by the Kalman filter using (\ref{eq:kf_predict_start})--(\ref{eq:kf_update_stop}),
+%
+\begin{IEEEeqnarray}{rCl}
+p(x_n|\theta_{n}^{(i)}, y_{1:n}) & = & \mathcal{N}(x_n|m_n^{(i)}, P_n^{(i)})     .
 \end{IEEEeqnarray}
@@
 
-The second state term in (\ref{eq:rb-vrps_back_cond_weight}) is the likelihood, $p(y_{n+1:N}|x_n, \tilde{\theta}_{n}^+)$.
+The second state-dependent term in (\ref{eq:rb-vrps_back_cond_weight}) is the likelihood, $p(y_{n+1:N}|x_n, \tilde{\theta}_{n}^+)$.
 This may be considered to be an improper density over $x_n$, and may be calculated analytically using a backwards Kalman filter, in a similar manner to that used in the two-filter smoother \cite{Fraser1969,Kitagawa1994,Sarkka2012,Gelb1974}.
-Such a backwards Kalman filter uses the following recursions, which are derived in appendix~\ref{app:backwards_KF}.
+Such a backwards Kalman filter uses the following recursions, which are derived in appendix~\ref{app:backwards_KF},
 %
@@
  \tilde{m}_n   & = & \tilde{m}_n^- + \tilde{K}_n (y_n - \tilde{\mu}_n) \\
- \tilde{P}_n   & = & \tilde{P}_n^- - \tilde{K}_n \tilde{S}_n \tilde{K}_n^T \label{eq:backward_kf_update_stop}
+ \tilde{P}_n   & = & \tilde{P}_n^- - \tilde{K}_n \tilde{S}_n \tilde{K}_n^T \label{eq:backward_kf_update_stop}     .
 \end{IEEEeqnarray}
@@
 
-Substituting into (\ref{eq:rb-vrps_back_cond_weight}), the backwards conditional weights are given by,
+Substituting the Gaussian densities into (\ref{eq:rb-vrps_back_cond_weight}) and performing the integral yields the following backwards conditional weights,
 %
@@
 
-Using these weights, a sample of $\theta_{n}$ may be drawn from the particle distribution of (\ref{eq:rb-vrps_back_cond}), completing the smoothing algorithm.
+Using these weights, a sample of $\theta_{n}$ may be drawn from the particle distribution of (\ref{eq:rb-vrps_back_cond}), completing one step of the smoothing algorithm.
 Once sampling has progressed backwards from $n=N \dots 1$, a particle from the smoothing distribution will have been generated.
@@
 
-In a conditionally deterministic system, the entire continuous time state trajectory, $x(t)$, is a function of the changepoint sequence $\theta$ and the initial state, $x_0$.
+In a conditionally deterministic system, the entire continuous-time state trajectory, $x(t)$, is a function of the changepoint sequence $\theta$ and the initial state, $x_0$.
 When the changepoint sequence is split into past, $\theta_n$, and future $\theta_n^+$, and the past is altered, this will alter the state trajectory in the future.
-This is undesirable, as each sampling operation will require the recalculation of state values and observations likelihoods for all future observation times.
+This is undesirable in practice, as each sampling operation will require the recalculation of state values and observations likelihoods for all future observation times.
 
@@
 \begin{IEEEeqnarray}{rCl}
- p(\tilde{\theta}|y_{1:N}) = p(\tilde{\theta_{n}}^{+}|y_{1:N}) p(\tilde{\theta}_{n}|\tilde{\theta}_{n}^{+}, y_{1:N})     .
-\end{IEEEeqnarray}
-
-Again, this suggests a backwards sequential sampling algorithm for smoothing. The conditional term may be expanded as,
+ p(\tilde{\theta}|y_{1:N}) = p(\tilde{\theta}_{n}^{+}|y_{1:N}) p(\tilde{\theta}_{n}|\tilde{\theta}_{n}^{+}, y_{1:N})     .
+\end{IEEEeqnarray}
+
+Again, this suggests a backwards sequential sampling algorithm for smoothing. The second term may be expanded as,
 %
@@
 
-which will be $0$ for almost all filtering particles. Intuitively, we are trying to join two deterministic state trajectories together, but they do not meet up in the middle.
+which will be $0$ for all but one unique filtering particle. Intuitively, we are trying to join two deterministic state trajectories together, but they do not meet up in the middle.
 
@@
 \begin{IEEEeqnarray}{rCl}
-\IEEEeqnarraymulticol{3}{l}{p(\tilde{\theta}_{n}|y_{1:n}) q(u_{K_n}|\tilde{\theta}_{n}, \tilde{\theta}_{n}^{+}, y_{1:N})} \nonumber \\
+\IEEEeqnarraymulticol{3}{l}{p(\tilde{\theta}_{n}|y_{1:n}) q(u_{K_n}'|\tilde{\theta}_{n}, \tilde{\theta}_{n}^{+}, y_{1:N})} \nonumber \\
  \qquad & \propto & p(y_{1:n}|\tilde{\theta}_{n}) p(\tilde{\theta}_{n}) q(u_{K_n}'|\tilde{\theta}_{n}, \tilde{\theta}_{n}^{+}, y_{1:N}) \label{eq:pd-vrps_proposal}     .
@@
 
-Clearly this new target distribution admits the desired posterior as a marginal.
+Clearly this new target distribution admits the desired conditional posterior, $p(\tilde{\theta}_{n}'|\tilde{\theta}_{n}^{+}, y_{1:N})$, as a marginal.
 
 For this method to work, a new condition must be imposed on the state dynamics.
-For any pair of adjacent changepoints, $(\tau_k,x_k)$, and $(\tau_{k+1},x_{k+1})$ it must be possible to calculate a parameter value which results in the transition from the former to the latter.
-As a rule of thumb, this requires the number of dimensions of $u_k$ to be greater or equal to the number of dimensions of $x_k$.
+For any pair of adjacent changepoints, $(\tau_k,x_k)$, and $(\tau_{k+1},x_{k+1})$ it must be possible to calculate a parameter value, $u_k$, which results in the transition from the former to the latter.
+Generally, this requires there to be as many degrees of freedom in the selection of $u_k$ as there are state dimensions of $x_k$.
 
@@
 The proposal density $q(u_{K_n}'|\tilde{\theta}_{n}, \tilde{\theta}_{n}^{+}, y_{1:N})$ is constructed so as to be non-zero only where $p(\tilde{\theta}_{n}^{+}|\tilde{\theta}_{n}')$ is non-zero.
-This may require it to be a delta function if there is only one feasible value.
+This could require it to be a delta function if there is only one feasible value.
 The artificial conditional term, $L(u_{K_n}|\tilde{\theta}_{n}, \tilde{\theta}_{n}^{+}, y_{1:N})$, may be any arbitrary density, but the simplest choice is simply as a uniform distribution so that it cancels out in the MH acceptance probability.
@@
 This value is itself a particle from the target distribution (albeit one from an approximation with poor path-space diversity), meaning that no burn-in period is required.
+This is the same reasoning which justifies the absence of a burn-in when using a resample-move scheme \cite{Gilks2001}.
 
@@
 See \cite{Bunch2012} for analysis of this trade-off on a fixed rate model.
-In the simulations discussed in this report, $M=1$ was used throughout.
+In the simulations discussed in this paper, $M=1$ was used throughout.
 
@@
 
-In the following examples, we compare the state estimates from the ``filter-smoother'' and the VRPS.
+In the following examples, we compare the state estimates from the ``filter-smoother'' \cite{Kitagawa1996} and the VRPS.
 Filter-smoother estimates use the particle approximation from the final step of the VRPF for the changepoint distribution.
@@
 In addition, the benefit of employing a smoothing algorithm is that it ensures a more diverse particle representation of the posterior state distribution.
-To demonstrate this effect, the number of unique state estimates is plotted over time in the following examples.
+To demonstrate this effect, the number of unique state samples is plotted over time in the following examples.
 
@@
 
-The latent state is a vector with two elements, the underlying value of the asset, and the trend followed by this value.
+The latent state is a vector with two elements, the underlying value of the asset, and the trend followed by this value,
 %
 \begin{equation}
- \mathbf{x}(t) = [ x(t), \dot{x}(t)]^T
+ \mathbf{x}(t) = [ x(t), \dot{x}(t)]^T     .
 \end{equation}
@@
 
-where $\lambda$ introduces a mean regression effect on the trend and $\mathbf{\beta}(t)$ is standard Brownian motion (with unit diffusion constant).
-The jump term, $d\mathbf{J}(t)$, is zero everywhere except when jumps occur.
+where $\lambda$ introduces a mean reversion effect on the trend and $\mathbf{\beta}(t)$ is standard Brownian motion (with unit diffusion constant).
+The jump term, $d\mathbf{J}(t)$, is zero everywhere except when jumps occur,
 %
@@
  d\mathbf{J}(t) & = & \begin{cases} \mathbf{J}_k & t \in \{\tau_k\} \\ 0 & \text{elsewhere} \end{cases} \\
- \mathbf{J}_k  & \sim & \mathcal{N}(\mathbf{J}_k| \mathbf{0}, Q_{J,u_k})
-\end{IEEEeqnarray}
-
-$\{\tau_k\}$ is the set of random jump times, each of which is one of two types: a value jump, indicated by $u_k = 1$, or a trend jump, indicated by $u_k=2$.
-
-%The jump covariance matrices are,
-%%
-%\begin{IEEEeqnarray}{c}
-%Q_{J,u_k} = \begin{cases} \begin{bmatrix}\sigma_{J1}^2 & 0 \\ 0 & 0 \end{bmatrix} & u_k = 1 \\
-%                          \begin{bmatrix}0 & 0 \\ 0 & \sigma_{J2}^2 \end{bmatrix} & u_k = 2  \end{cases}   .
-%\end{IEEEeqnarray}
+ \mathbf{J}_k  & \sim & \mathcal{N}(\mathbf{J}_k| \mathbf{0}, Q_{J,u_k})     .
+\end{IEEEeqnarray}
+
+$\{\tau_k\}$ is the set of random jump times, each of which is one of two types: a value jump, indicated by $u_k = 1$, or a trend jump, indicated by $u_k=2$. The jump covariance matrices are,
+%
+\begin{IEEEeqnarray}{c}
+Q_{J,u_k} = \begin{cases} \begin{bmatrix}\sigma_{J1}^2 & 0 \\ 0 & 0 \end{bmatrix} & u_k = 1 \\
+                          \begin{bmatrix}0 & 0 \\ 0 & \sigma_{J2}^2 \end{bmatrix} & u_k = 2  \end{cases}   .
+\end{IEEEeqnarray}
+
+Only the value is observed, and that corrupted by additive Gaussian noise with standard deviation $\sigma_y^2$.
 
@@
 where the $\mathbf{w}_n$ and $v_n$ are Gaussian random variables with covariance matrixes $Q_n$ and $R$ respectively.
-$Q_n$ is a function of the sequence of jump times and their types, giving rise to the desired conditionally linear-Gaussian structure.
-% The time between observations times is denotes $\Delta t = t_{n} - t_{n-1}$.
+$Q_n$ is a function of the sequence of jump times and their types, giving rise to the desired conditionally linear-Gaussian structure.% The time between observations times is denotes $\Delta t = t_{n} - t_{n-1}$.
 %
@@
 
-The times between changepoints are assumed to be exponentially distributed, with equal probability of value and trend jumps.
+The times between changepoints are assumed to be exponentially distributed with mean $\alpha$, and equal probability of value and trend jumps.
 %
@@
 Observations were made on a regular time grid with a spacing of $0.0017s$.
-The decay constant for the distribution of inter-changepoint times is 20, and the trend mean-reversion coefficient is 5.
+Other parameters used were: $\alpha = 20$, $\lambda = 5$, $\sigma = 0.05$, $\sigma_{J1} = 0.005$, $\sigma_{J2} = 0.05$, $\sigma_{y} = 0.001$.
+
+% The decay constant for the distribution of inter-changepoint times is 20, and the trend mean-reversion coefficient is 5.
 The diffusion constant is 0.05 and the observation noise standard deviation 0.001.
@@
 Bootstrap proposals were used for the filter.
-An example realisation simulated from the model is shown in figure~\ref{fig:example_data}.
-The trend estimates produced by the filter-smoother and VRPS are shown in figure~\ref{fig:example_state} (the equivalent graphs for value estimates are not particularly informative, as this quantity is observed).
+An example forward realisation simulated from the model is shown in figure~\ref{fig:example_data}.
+The sampled trend trajectories produced by the filter-smoother and VRPS are shown in figure~\ref{fig:example_state} (the equivalent graphs for value are not particularly informative as this quantity is observed, and the errors produced by both algorithms are small).
+Notice that the path diversity is dramatically improved by the VRPS.
 
@@
 \centering
-\subfloat[]{\includegraphics[width=0.95\columnwidth]{example_filtersmoother_state.pdf}} \\
-\subfloat[]{\includegraphics[width=0.95\columnwidth]{example_smoother_state.pdf}}
+\subfloat[]{\includegraphics[width=0.9\columnwidth]{example_filtersmoother_state.pdf}} \\
+\subfloat[]{\includegraphics[width=0.9\columnwidth]{example_smoother_state.pdf}}
 \caption{Trend estimates from (a) the filter-smoother and (b) the VRPS. Solid lines show means of each particle. Dashed lines show mean $\pm$2 standard deviations.}
@@
 \centering
-\includegraphics[width=0.75\columnwidth]{finance_unique_particles.pdf}
+\includegraphics[width=0.5\columnwidth,angle=90]{finance_unique_particles.pdf}
 \caption{Mean number of unique particles in the filter-smoother (dashed) and smoother (solid) estimates.}
@@
 
-The state vector consists of cartesian position, $x(t)$ and $y(t)$, plus bearing, $\psi(t)$, and speed, $\dot{s}(t)$.
+The state vector consists of Cartesian position, $x(t)$ and $y(t)$, plus heading angle, $\psi(t)$, and speed, $\dot{s}(t)$.
 %
@@
 The filtering and smoothing algorithms were tested on artificial data simulated from the model.
-Observations were generated every $0.1$.
-The standard deviations of tangential and normal accelerations were set to $0.01$ and $1$ respectively, and those for the drift velocities to $1$.
-The parameters of the Gamma distribution for inter-changepoint times were $5$ and $1$ for the shape and scale respectively.
-Observation noise standard deviations were $0.1$ and $0.25^{\circ}$ respectively.
+Observations were generated every $0.1s$.
+The standard deviations of tangential and normal accelerations were set to $0.01m/s^2$ and $1m/s^2$ respectively, and those for the drift velocities to $1m/s$.
+The parameters of the Gamma distribution for inter-changepoint times were $5$ and $1s$ for the shape and scale respectively.
+Observation noise standard deviations were $0.1m$ and $0.25^{\circ}$ respectively.
 
@@
 An example trajectory is shown in Fig.~\ref{fig:simulated_trajectory}, with the results of the VRPF and VRPS algorithms in Fig.\ref{fig:2D_particle_results}.
-The improved particle diversity achieved by the smoothing algorithm is apparent.
+The improved particle diversity achieved by the smoothing algorithm is readily apparent.
 
@@
 \centering
-\subfloat[]{\includegraphics[width=0.95\columnwidth]{2DFilter.pdf}} \\
-\subfloat[]{\includegraphics[width=0.95\columnwidth]{2DSmoother.pdf}}
+\subfloat[]{\includegraphics[width=0.9\columnwidth]{2DFilter.pdf}} \\
+\subfloat[]{\includegraphics[width=0.9\columnwidth]{2DSmoother.pdf}}
 \caption{Filter (a) and smoother (b) particle estimates of position (solid lines), with particle changepoint positions (crosses). }
@@
 
-Performance was evaluated by testing on 10 scenarios, each with 500 observations.
+Performance was evaluated by testing on 10 independently generated scenarios, each with 500 observations.
 At each observation time, an estimate of the position and velocity is made by taking the average of values from the array of particles.
@@
 The mean number of unique particle estimates at each time step is shown in Fig.~\ref{fig:tracking_unique_particles}.
-This illustrates the main advantage of the smoother algorithm --- the increased number of unique particles in the approximation means a better characterisation of the posterior changepoint distribution, allowing, for example, calculation of state covariance measures.
+This illustrates again the principal advantage of the smoother algorithm --- the increased number of unique particles in the approximation means a better characterisation of the posterior changepoint distribution, allowing, for example, calculation of state covariance measures.
 
@@
 \centering
-\includegraphics[width=0.75\columnwidth]{tracking_unique_particles.pdf}
+\includegraphics[width=0.5\columnwidth,angle=90]{tracking_unique_particles.pdf}
 \caption{Mean number of unique particles in the final filter (dashed) and smoother (solid) estimates.}

### 2012-04-25 full_finaldraft_VRsmoothing_withSJGcomments.tex
Numerous small changes.
More edits. Added some bits about space of theta. About to make \theta_n into \theta_n^-

@@
 
-In a variable rate model, the set of changepoints and associated parameters are modelled as a marked point process (MPP), the mathematical properties of which are set out thoroughly in \cite{Jacobsen2006}.
+In a variable rate setting, changepoints and associated parameters are modelled as a marked point process (MPP), the mathematical properties of which are explored comprehensively in \cite{Jacobsen2006}.
 Conditional upon this MPP the state evolves according to some benign dynamics.
@@
 the estimation of the changepoint sequence and latent state given all the observations.
-This is achieved in this paper with an efficient backward sweep through the observations, in a new version of the method for standard HMMs described in \cite{Godsill2004}.
+We achieve this with an efficient backward sweep through the observations, in a new version of the method for standard HMMs described in \cite{Godsill2004}.
 Two new schemes are introduced: one for conditionally linear-Gaussian models which exploits the method of Rao-Blackwellisation; and a second for use with conditionally-deterministic models which uses an augmented target distribution in the style of an SMC sampler \cite{DelMoral2006}.
@@
 This density will be constructed such that $P(\tau_k < \tau_{k-1}) = 0$.
-
-In the manner of \cite{Whiteley2011}, a survivor function is defined as the probability that no new changepoint occurs before a given time,
+As in \cite{Whiteley2011}, a survivor function is defined as the probability that no new changepoint occurs before a given time,
 %
@@
 
-It is now possible to write down a prior for the changepoint sequence, where we use the convention that $\tau_0 = 0$. The existence of such a density for a MPP is addressed in \cite{Jacobsen2006}.
+It is now possible to write down a prior for the changepoint sequence.
+This comprises a density term for each changepoint in the sequence and a survivor function term accounting for the probability that no additional changepoints occur within the time interval.
+We use the convention that $\tau_0 = 0$.
 
@@
 
-This comprises a density term for each changepoint in the sequence and a survivor function term accounting for the probability that no other changepoints occur within the time interval.
-
-\subsection{Conditionally Linear-Gaussian Models}
+The existence of such a density for a MPP is addressed in \cite{Jacobsen2006}.
+The number of changepoints, $K_n$, is not fixed.
+If $u_k$ varies over $\mathcal{U}$, and $\tau_k$ varies over the real line $\mathbb{R}$, then the space spanned by $\theta_n$ is the union,
+%
+\begin{IEEEeqnarray}{rCl}
+ \Theta & = & \bigcup_k \Theta_k \label{eq:theta_space} \\
+ \Theta_k & = & \{k\} \times \mathbb{R}^k \times \mathcal{U}^k     .
+\end{IEEEeqnarray}
+
+This closely resembles the variable-dimension spaces employed by reversible jump Markov chain Monte Carlo algorithms \cite{Green1995}.% or in multiple target tracking scenarios \cite{Vo2005}.
+
+\subsection{Conditionally Linear-Gaussian Models} \label{sec:clg_models}
 
@@
 
-\subsection{Conditionally Deterministic Models}
-
-The second class of variable rate models for consideration is those in which the state is completely specified by the changepoint sequence, with no additional random components.
+\subsection{Conditionally Deterministic Models} \label{sec:cd_models}
+
+Next we examine the class of variable rate models in which the state is completely specified by the changepoint sequence, with no additional random components.
 Such a process is commonly referred to as ``piecewise-deterministic'', as the latent state follows a deterministic path between changepoints.
-In this case, it is not necessary to discretise the state -- it may be kept as a continuous variable.
+In this case, it is not necessary to discretise the state --- it may be kept as a continuous variable.
 
@@
 \begin{IEEEeqnarray}{rCl}
- dx(t) & = & \mathrm{f}(x(t), \tau_{K(t)}, u_{K(t)})     .
+ dx(t) & = & \varphi(x(t), \tau_{K(t)}, u_{K(t)}) dt     .
 \end{IEEEeqnarray}
@@
 By choosing $t = \tau_{K_{n}+1}$, this equation specifies the state at the next changepoint time.
-Similarly, by choosing $t=t_n$, the state at the observation times may be calculated --- these latter points will be denoted $\hat{x}_n$.
-To complete the model, a probabilistic measurement model must be devised for the observation process, $p(y_n|\hat{x}_n)$.
+Similarly, by choosing $t=t_n$, the state at the observation times may be evaluated --- these latter points will be denoted $\hat{x}_n$.
+To complete the framework, a probabilistic measurement model must be devised for the observation process, $p(y_n|\hat{x}_n)$.
 
@@
 The variable rate particle filter (VRPF) is described in \cite{Godsill2007,Godsill2007a,Whiteley2011}.
-The objective of the algorithm is to sequentially estimate the posterior distribution of the changepoint sequence, $p(\theta_{n}| y_{1:n})$, at each time $t_n$.
+The objective of the algorithm is to sequentially estimate the posterior distribution of the changepoint sequence, $p(\theta_{n}| y_{1:n})$, at each time $t_n$, which is defined on the same space as the prior density (\ref{eq:theta_space}).
 This distribution may be expanded using Bayes' rule,
@@
 
-where $\delta_x(X)$ is a dirac probability mass at $X=x$.
+where $\delta_x(X)$ is a unit probability mass at $X=x$.
 
@@
 
-Next, an extension to the changepoint sequence, $\theta_{n \setminus n-1}^{(i)}$, is sampled from an importance distribution, $q(\theta_{n \setminus n-1}|\theta_{n-1}^{(i)}, y_n)$, and concatenated with $\theta_{n-1}$ to create a proposal for $\theta_n$.
+Next, an extension to the changepoint sequence, $\theta_{n \setminus n-1}^{(i)}$, is sampled from an importance distribution, $q(\theta_{n \setminus n-1}|\theta_{n-1}^{(i)}, y_n)$, and concatenated with $\theta_{n-1}^{(i)}$ to create a proposal for $\theta_n^{(i)}$.
 Finally, the particle is weighted according to the ratio of the target and proposal densities,
@@
 \begin{IEEEeqnarray}{rCl}
-w_n^{(i)} & = & \frac{ p(\theta_{n}^{(i)}|y_{1:n}) }{ q(\theta_{n}) } \nonumber \\
-    & \propto & \frac{ p(y_n|\theta_n, y_{1:n-1}) p(\theta_{n \setminus n-1}|\theta_{n-1}) p(\theta_{n-1}|y_{1:n-1}) }{ q(\theta_{n-1}) q(\theta_{n \setminus n-1}|\theta_{n-1}^{(i)}, y_n) } \nonumber \\
-    & =       & \frac{w_{n-1}^{(i)}}{v_{n-1}^{(i)}} \times \frac{ p(y_n|\theta_n, y_{1:n-1}) p(\theta_{n \setminus n-1}|\theta_{n-1}) }{ q(\theta_{n \setminus n-1}|\theta_{n-1}^{(i)}, y_n) } \label{eq:vrpf_weights}     .
+w_n^{(i)} & = & \frac{ p(\theta_{n}^{(i)}|y_{1:n}) }{ q(\theta_{n}^{(i)}) } \nonumber \\
+    & \propto & \frac{ p(y_n|\theta_n^{(i)}, y_{1:n-1}) p(\theta_{n \setminus n-1}^{(i)}|\theta_{n-1}^{(i)}) p(\theta_{n-1}^{(i)}|y_{1:n-1}) }{ q(\theta_{n-1}^{(i)}) q(\theta_{n \setminus n-1}^{(i)}|\theta_{n-1}^{(i)}, y_n) } \nonumber \\
+    & =       & \frac{w_{n-1}^{(i)}}{v_{n-1}^{(i)}} \times \frac{ p(y_n|\theta_n^{(i)}, y_{1:n-1}) p(\theta_{n \setminus n-1}^{(i)}|\theta_{n-1}^{(i)}) }{ q(\theta_{n \setminus n-1}^{(i)}|\theta_{n-1}^{(i)}, y_n) } \label{eq:vrpf_weights}     .
 \end{IEEEeqnarray}
@@
 This final future changepoint is discarded.
-(This process can be thought of as sampling the entire future changepoint sequence from $t_{n-1}$ onwards, and then marginalising those which fall after $t_n$.) The bootstrap proposal leads to the usual simplification of the weight formula, % (apart from the first which is sampled from (\ref{eq:cp_cond_model}))
+(This process can be thought of as sampling an entire future changepoint sequence from $t_{n-1}$ onwards, and then marginalising those which fall after $t_n$.) The bootstrap proposal leads to the usual simplification of the weight formula, % (apart from the first which is sampled from (\ref{eq:cp_cond_model}))
 %
@@
 
-For a conditionally linear-Gaussian model, the required likelihood term, $p(y_n|\theta_n, y_{1:n-1})$ is the predictive distribution estimated by the Kalman filter.
+For the conditionally linear-Gaussian models of section~\ref{sec:clg_models}, the required likelihood term, $p(y_n|\theta_n, y_{1:n-1})$ is the predictive distribution estimated by the Kalman filter.
 Conveniently, the Kalman filter also provides us with an estimate of the current state given the changepoint sequence and all the preceding observations.
@@
 
-When a conditionally deterministic model is used, the state at observation time $t_n$ is specified by the changepoint sequence $\theta_n$ (plus the initial state, $x_0$), using (\ref{eq:disc_time_state_diff_eq}).
+When the conditionally deterministic models of section~\ref{sec:cd_models} are used, the state at observation time $t_n$ is specified by the changepoint sequence $\theta_n$ (plus the initial state, $x_0$), using (\ref{eq:disc_time_state_diff_eq}).
 Thus, the required likelihood term is simply given by,
@@
 
-This leads to a particle filter for variable rate models with conditionally deterministic dynamics, summarised in Fig.~\ref{alg:VRPF}.
+This leads to a piecewise-deterministic variable rate particle filter, as summarised in Fig.~\ref{alg:VRPF}.
 
@@
 
-At time $t_n$, the target distribution may be factorised as follows,
+For the conditionally linear-Gaussian models of section~\ref{sec:clg_models}, the target distribution may be factorised as follows at time $t_n$,
 %
@@
 New values of $\theta_{n}$ are then sampled from the conditional distribution, $p(\theta_{n}|\theta_{n}^{+}, y_{1:N})$, which is approximated using the filter particles, and the past and future sequences are concatenated.
-By iterating backwards from $n = N \dots 1$, a diverse set of particles approximating $p(\theta|y_{1:N})$ will be generated.
+By iterating backwards over $n = N \dots 1$, a diverse set of particles approximating $p(\theta|y_{1:N})$ will be generated.
 
@@
 
-Finally, the particle approximation of (\ref{eq:vrpf}) is substituted for the filtering distribution, $p(\theta_{n}|y_{1:n})$,
+The particle approximation of (\ref{eq:vrpf}) is substituted for the filtering distribution, $p(\theta_{n}|y_{1:n})$, yielding,
 %
@@
 The second state-dependent term in (\ref{eq:rb-vrps_back_cond_weight}) is the likelihood, $p(y_{n+1:N}|x_n, \tilde{\theta}_{n}^+)$.
-This may be considered to be an improper density over $x_n$, and may be calculated analytically using a backwards Kalman filter, in a similar manner to that used in the two-filter smoother \cite{Fraser1969,Kitagawa1994,Sarkka2012,Gelb1974}.
+This may be regarded as an improper density over $x_n$, and may be calculated analytically using a backwards Kalman filter, in a similar manner to that used in the two-filter smoother \cite{Fraser1969,Kitagawa1994,Sarkka2012,Gelb1974}.
 Such a backwards Kalman filter uses the following recursions, which are derived in appendix~\ref{app:backwards_KF},
@@
 Exact methods for initialising this recursion are discussed in appendix~\ref{app:init_backwards_KF}, and also in \cite{Kitagawa1994}.
-However, it is often sufficient to use an approximation, for example $\tilde{m}_N^{(i)} = m_N^{(i)}$ and $\tilde{P}_N^{(i)} = P_N^{(i)}$.
-
-Substituting the Gaussian densities into (\ref{eq:rb-vrps_back_cond_weight}) and performing the integral yields the following backwards conditional weights,
+However, it is often sufficient to use an approximation, for example $\tilde{m}_N^{(i)} \gets m_N^{(i)}$ and $\tilde{P}_N^{(i)} \gets P_N^{(i)}$.
+
+Substituting the Gaussian densities into (\ref{eq:rb-vrps_back_cond_weight}) and integrating yields the following backwards conditional weights,
 %
@@
 Using these weights, a sample of $\theta_{n}$ may be drawn from the particle distribution of (\ref{eq:rb-vrps_back_cond}), completing one step of the smoothing algorithm.
-Once sampling has progressed backwards from $n=N \dots 1$, a particle from the smoothing distribution will have been generated.
+Once sampling has progressed backwards over $n=N \dots 1$, a particle from the smoothing distribution will have been generated.
 This procedure may then be repeated until sufficient particles have been obtained.
@@
 
-In a conditionally deterministic system, the entire continuous-time state trajectory, $x(t)$, is a function of the changepoint sequence $\theta$ and the initial state, $x_0$.
+In the conditionally deterministic models of section~\ref{sec:cd_models}, the entire continuous-time state trajectory, $x(t)$, is a function of the changepoint sequence $\theta$ and the initial state, $x_0$.
 When the changepoint sequence is split into past, $\theta_n$, and future $\theta_n^+$, and the past is altered, this will alter the state trajectory in the future.
@@
 
-Again, this suggests a backwards sequential sampling algorithm for smoothing. The second term may be expanded as,
+As before, this suggests a backwards sequential sampling algorithm for smoothing. The second term may be expanded as,
 %
@@
 
-This presents a problem. We cannot directly substitute the particle approximation for the filtering distribution because the transition density term for the augmented sequences is given by,
+A difficulty is now encountered. We cannot directly substitute the particle approximation for the filtering distribution because the transition density term for the augmented sequences is given by,
 %
@@
 
-which will be $0$ for all but one unique filtering particle. Intuitively, we are trying to join two deterministic state trajectories together, but they do not meet up in the middle.
+which will be $0$ for all but one unique filtering particle, the one from which the next selected future changepoint originated.
+Intuitively, we are trying to join two deterministic state trajectories together, but they do not meet up in the middle.
 
 The solution to this problem is provided by an idea from the Sequential Monte Carlo (SMC) samplers of \cite{DelMoral2006}.
-Rather than sampling $p(\tilde{\theta}_{n}|\tilde{\theta}_{n}^{+}, y_{1:N})$ directly, a value of $\tilde{\theta}_{n}$ is proposed from the filtering particles, after which a replacement, $u_{K_n}'$, is proposed for the existing final changepoint parameter, with a value chosen such that the past and future paths meet up.
+Rather than sampling $p(\tilde{\theta}_{n}|\tilde{\theta}_{n}^{+}, y_{1:N})$ directly, an augmented distribution is targeted.
+A value of $\tilde{\theta}_{n}$ can be proposed from the filtering particles, after which a replacement, $u_{K_n}'$, is proposed for the existing final changepoint parameter, with a value chosen such that the past and future paths meet up.
 The modified sequence is denoted $\tilde{\theta}_{n}'$.
@@
 
-The target distribution is augmented by introduction of an artificial density to cover the new changepoint sequence and the discarded parameter.
+The augmented target distribution over the the new sequence, $\tilde{\theta}_{n}'$, and the discarded parameter, $u_{K_n}$, is constructed by introducing an artificial density,
 %
@@
 This could require it to be a delta function if there is only one feasible value.
-The artificial conditional term, $L(u_{K_n}|\tilde{\theta}_{n}, \tilde{\theta}_{n}^{+}, y_{1:N})$, may be any arbitrary density, but the simplest choice is simply as a uniform distribution so that it cancels out in the MH acceptance probability.
+The artificial conditional term, $L(u_{K_n}|\tilde{\theta}_{n}, \tilde{\theta}_{n}^{+}, y_{1:N})$, may be any arbitrary density, but the simplest choice is a uniform distribution over the set of parameter values, $\{u_{K_n}^{(i)}\}$, which appear in the filtering approximation.
+With this choice, the artificial density terms will always cancel out in the MH acceptance probability.
 The remaining terms are likelihoods and parameter transition density terms, defined by the system equations.
@@
 \end{IEEEeqnarray}
+
+The algorithm is summarised in Fig.~\ref{alg:VRPS}.
 
@@
 This is the same reasoning which justifies the absence of a burn-in when using a resample-move scheme \cite{Gilks2001}.
-
-The algorithm is summarised in Fig.~\ref{alg:VRPS}.
 
@@
 by finding the maximum a posteriori sequence), it is not trivial to define a distance metric between the estimated and true sequences to assess relative accuracy.
+(One possible choice of metric has been considered for this comparison, the optimal sub-pattern assignment (OSPA) distance of \cite{Schuhmacher2008}.
+However, the relative performance of the algorithms is liable to vary significantly with the parameters of the metric.)
 
@@
 
-The Rao-Blackwellised VRPS algorithm was tested on the financial time series model of \cite{Godsill2007a,Christensen2012}, in which prices of an asset are treated as noisy observations of a latent state, which evolves according to a drift-diffusion with occasional jumps.
+The Rao-Blackwellised VRPS algorithm was tested on the financial time series model of \cite{Christensen2012}, in which prices of an asset are treated as noisy observations of a latent state, which evolves according to a drift-diffusion with occasional jumps.
 
@@
 
-This evolves continuously according to a jump-diffusion model:
+This evolves continuously according to a jump-diffusion model,
 %
@@
 
-The times between changepoints are assumed to be exponentially distributed with mean $\alpha$, and equal probability of value and trend jumps.
+The times between changepoints are assumed to be exponentially distributed with mean $1/\alpha$, and equal probability of value and trend jumps.
 %
@@
 The mean number of unique particle estimates at each time step is shown in Fig.~\ref{fig:finance_unique_particles}.
-This illustrates the main advantage of the smoother algorithm --- the increased number of unique particles in the approximation means a better characterisation of the posterior changepoint distribution.
+This illustrates the main advantage of the smoother algorithm --- the increased number of unique particles in the approximation means a better characterisation of the posterior changepoint and state distributions.
 
@@
 \centering
-\includegraphics[width=0.5\columnwidth,angle=90]{finance_unique_particles.pdf}
+\includegraphics[width=0.75\columnwidth]{finance_unique_particles.pdf}
 \caption{Mean number of unique particles in the filter-smoother (dashed) and smoother (solid) estimates.}
@@
 
-The state vector consists of Cartesian position, $x(t)$ and $y(t)$, plus heading angle, $\psi(t)$, and speed, $\dot{s}(t)$.
+The state vector consists of Cartesian position, $x(t)$ and $y(t)$, plus heading angle, $\psi(t)$, and speed, $\dot{s}(t)$,
 %
 \begin{equation}
-\mathbf{x}(t) = [x(t), y(t), \psi(t), \dot{s}(t)]^T
+\mathbf{x}(t) = [x(t), y(t), \psi(t), \dot{s}(t)]^T     .
 \end{equation}
@@
 In addition to the tangential and normal accelerations, $a_{T,k}$ and $a_{N,k}$, we introduce two linear velocity terms, $d_{X,k}$ and $d_{Y,k}$.
-Without these, the model could not be used for smoothing, because for an arbitrary pair of adjacent changepoints ($\tau_{k}$, $\mathbf{x}_{k}$) and ($\tau_{k+1}$, $\mathbf{x}_{k+1}$), there would not in general be a pair of accelerations which achieved the required transition.
+Without these, the model could not be used for smoothing, because for an arbitrary pair of adjacent changepoints ($\tau_{k}$, $\mathbf{x}_{k}$) and ($\tau_{k+1}$, $\mathbf{x}_{k+1}$), there would not, in general, be a pair of accelerations which achieved the required transition.
 These linear velocity terms may be considered to be merely relaxation terms, allowing arbitrary past and future trajectories to be joined, or they may represent real physical effects, such as wind (for aircraft) or currents (for boats).
@@
 \begin{equation}
-\mathbf{u}_k = [a_{T,k}, a_{N,k}, d_{X,k}, d_{Y,k}]^T
+\mathbf{u}_k = [a_{T,k}, a_{N,k}, d_{X,k}, d_{Y,k}]^T     .
 \end{equation}
 
-The target dynamics are described by four differential equations.
+The target dynamics are described by four differential equations,
 %
@@
 
-Solving these yields the following state equations (where $\Delta\ t = t - \tau_{K(t)}$):
+Solving these yields the following state equations (where $\Delta\ t = t - \tau_{K(t)}$),
 %The target dynamics are governed by the following state equations \cite{Bunch2012a} (where $\Delta\ t = t - \tau_{K(t)}$):
@@
 The resulting root mean square errors (RMSEs) are displayed in table~\ref{tab:tracking_state_performance}, demonstrating an improvement in accuracy from the smoother.
-The smoother also provides a more accurate estimate of the number of changepoints in each test, with an average error of 0.6 compared to 2.0 from the final frame filter approximation.
+The VRPS also provides a more accurate estimate of the number of changepoints in each test, with an average error of 0.6 compared to 2.0 from the filter-smoother.
 
 The mean number of unique particle estimates at each time step is shown in Fig.~\ref{fig:tracking_unique_particles}.
-This illustrates again the principal advantage of the smoother algorithm --- the increased number of unique particles in the approximation means a better characterisation of the posterior changepoint distribution, allowing, for example, calculation of state covariance measures.
+This illustrates again the principal advantage of the VRPS algorithm --- the increased number of unique particles in the approximation means a better characterisation of the posterior changepoint distribution, allowing, for example, calculation of state covariances.
 
@@
 \centering
-\includegraphics[width=0.5\columnwidth,angle=90]{tracking_unique_particles.pdf}
+\includegraphics[width=0.75\columnwidth]{tracking_unique_particles.pdf}
 \caption{Mean number of unique particles in the final filter (dashed) and smoother (solid) estimates.}
@@
 For the deterministic case, smoothing is achieved by sampling an augmented target distribution using MCMC.
-Simulations demonstrate that the smoothers generate approximation with improved particle diversity and accuracy when compared to the filters.
+Simulations demonstrate that the smoothers generate approximations with improved particle diversity and accuracy when compared to the filters.
 
@@
 
-$||A||$ is used to denote the magnitude of the determinant of $A$, and it is assumed that $A$ is non-singular.
+$||A||$ denotes the magnitude of the determinant of $A$, and it is assumed that $A$ is non-singular.
 
@@
  & = & \tilde{Z}_n^- \mathcal{N}(x_n|\tilde{m}_n^-, \tilde{P}_n^-) \mathcal{N}(y_n|C_n x_n, R_n) \nonumber \\
- & = & \tilde{Z}_n \mathcal{N}(x_n|\tilde{m}_n, \tilde{P}_n)
+ & = & \tilde{Z}_n \mathcal{N}(x_n|\tilde{m}_n, \tilde{P}_n)     ,
 \end{IEEEeqnarray}
@@
 
-where $\lambda$ is the mean regression constant, $\sigma$ is the diffusion constant, $\mathbf{\beta}(t)$ is standard Brownian motion, $d\mathbf{J}(t)$ is a jump term defined according to,
+where $\lambda$ is the mean reversion constant, $\sigma$ is the diffusion constant, $\mathbf{\beta}(t)$ is standard Brownian motion, $d\mathbf{J}(t)$ is a jump term defined according to,
 %
@@
  d\mathbf{J}(t) & = & \begin{cases} \mathbf{J}_k & t \in \{\tau_k\} \\ 0 & \text{elsewhere} \end{cases} \\
- \mathbf{J}_k  & \sim & \mathcal{N}(\mathbf{J}_k| \mathbf{0}, Q_{J,u_k})
+ \mathbf{J}_k  & \sim & \mathcal{N}(\mathbf{J}_k| \mathbf{0}, Q_{J,u_k})    .
 \end{IEEEeqnarray}
@@
 
-where the $\mathbf{w}_n$ is a Gaussian random variable with covariance matrix $Q_n$. The inter-observation time is denoted $\Delta t = t_{n} - t_{n-1}$.
+where the $\mathbf{w}_n$ is a Gaussian random variable with covariance matrix $Q_n$. The inter-observation time is denoted $\Delta t = t_{n} - t_{n-1}$, and,
 
@@
 
-\newpage
+%\newpage
 \bibliographystyle{IEEEtran}

### 2012-04-25 full_finaldraft_VRsmoothing_withSJGcomments.tex
Done Simon's suggestions.

@@
 This will be denoted as $\theta = \{\tau_{0:K}, u_{0:K}\}$.
-At a particular time $t_n$, the sequence will be divided into past $\theta_n = \{\tau_{j}, u_{j} \forall j : 0 \leq \tau_j < t_n \}$, and future $\theta_n^+ = \{\tau_{j}, u_{j} \forall j : t_n \leq \tau_j < T \}$.
+At a particular time $t_n$, the sequence will be divided into past $\theta_n^- = \{\tau_{j}, u_{j} \forall j : 0 \leq \tau_j < t_n \}$, and future $\theta_n^+ = \{\tau_{j}, u_{j} \forall j : t_n \leq \tau_j < T \}$.
 It will also be useful to define a variable for the changepoints which occur in the interval $[t_{n-1},t_n)$, $\theta_{n \setminus n-1} = \{\tau_{j}, u_{j} \forall j : t_{n-1} \leq \tau_j < t_n \}$.
@@
 \begin{IEEEeqnarray}{rCl}
-p(\theta_n) & = & S(\tau_{K_n},t_n) p(u_0) \prod_{k=1}^{K_n} p(\tau_k, u_k| \tau_{k-1}, u_{k-1}) \label{eq:cp_sequence_prior}
+p(\theta_n^-) & = & S(\tau_{K_n},t_n) p(u_0) \prod_{k=1}^{K_n} p(\tau_k, u_k| \tau_{k-1}, u_{k-1}) \label{eq:cp_sequence_prior}
 \end{IEEEeqnarray}
@@
 The number of changepoints, $K_n$, is not fixed.
-If $u_k$ varies over $\mathcal{U}$, and $\tau_k$ varies over the real line $\mathbb{R}$, then the space spanned by $\theta_n$ is the union,
+If $u_k$ varies over $\mathcal{U}$, and $\tau_k$ varies over the real line $\mathbb{R}$, then the space spanned by $\theta_n^-$ is the union,
 %
@@
 \begin{IEEEeqnarray}{rCl}
- x_n & = & A_n(\theta_{n})x_{n-1} + w_n \\
- y_n & = & C_n(\theta_{n})x_n + v_n       ,
+ x_n & = & A_n(\theta_{n}^-)x_{n-1} + w_n \\
+ y_n & = & C_n(\theta_{n}^-)x_n + v_n       ,
 \end{IEEEeqnarray}
@@
 \begin{IEEEeqnarray}{rCl}
- w_n & \sim & \mathcal{N}(w_n|0, Q_n(\theta_{n})) \\
- v_n & \sim & \mathcal{N}(v_n|0, R_n(\theta_{n}))       .
+ w_n & \sim & \mathcal{N}(w_n|0, Q_n(\theta_{n}^-)) \\
+ v_n & \sim & \mathcal{N}(v_n|0, R_n(\theta_{n}^-))       .
 \end{IEEEeqnarray}
@@
 Conditionally linear-Gaussian models for use with variable rate particle filters were introduced in \cite{Godsill2007a,Christensen2012} for financial and tracking problems.
-For example, changepoints might correspond to jumps in the value or trend of a security, at which points the process covariance, $Q_n(\theta_n)$, is inflated in an optimal fashion.
+For example, changepoints might correspond to jumps in the value or trend of a security, at which points the process covariance, $Q_n(\theta_n^-)$, is inflated in an optimal fashion.
 
@@
 The variable rate particle filter (VRPF) is described in \cite{Godsill2007,Godsill2007a,Whiteley2011}.
-The objective of the algorithm is to sequentially estimate the posterior distribution of the changepoint sequence, $p(\theta_{n}| y_{1:n})$, at each time $t_n$, which is defined on the same space as the prior density (\ref{eq:theta_space}).
+The objective of the algorithm is to sequentially estimate the posterior distribution of the changepoint sequence, $p(\theta_{n}^-| y_{1:n})$, at each time $t_n$, which is defined on the same space as the prior density (\ref{eq:theta_space}).
 This distribution may be expanded using Bayes' rule,
@@
 \begin{IEEEeqnarray}{rCl}
-\IEEEeqnarraymulticol{3}{l}{ p(\theta_{n}|y_{1:n}) \propto p(y_n|\theta_{n}, y_{1:n-1}) } \nonumber \\
- \qquad & & \times p(\theta_{n \setminus n-1}|\theta_{n-1}) p(\theta_{n-1}|y_{1:n-1}) \label{eq:vrpf_target}     .
-\end{IEEEeqnarray}
-
-The transition term, $p(\theta_{n \setminus n-1} | \theta_{n-1})$, has a similar form to the changepoint prior of (\ref{eq:cp_sequence_prior}) \cite{Jacobsen2006}, but an additional survivor function is included to account for the condition that changepoints cannot occur before $t_{n-1}$,
-%
-\begin{IEEEeqnarray}{rCl}
-\IEEEeqnarraymulticol{3}{l}{p(\theta_{n \setminus n-1} | \theta_{n-1}) = S(\tau_{K_n}, t_n) / S(\tau_{K_{n-1}+1}, t_{n-1}) } \nonumber \\
+\IEEEeqnarraymulticol{3}{l}{ p(\theta_{n}^-|y_{1:n}) \propto p(y_n|\theta_{n}^-, y_{1:n-1}) } \nonumber \\
+ \qquad & & \times p(\theta_{n \setminus n-1}|\theta_{n-1}^-) p(\theta_{n-1}^-|y_{1:n-1}) \label{eq:vrpf_target}     .
+\end{IEEEeqnarray}
+
+The transition term, $p(\theta_{n \setminus n-1} | \theta_{n-1}^-)$, has a similar form to the changepoint prior of (\ref{eq:cp_sequence_prior}) \cite{Jacobsen2006}, but an additional survivor function is included to account for the condition that changepoints cannot occur before $t_{n-1}$,
+%
+\begin{IEEEeqnarray}{rCl}
+\IEEEeqnarraymulticol{3}{l}{p(\theta_{n \setminus n-1} | \theta_{n-1}^-) = S(\tau_{K_n}, t_n) / S(\tau_{K_{n-1}+1}, t_{n-1}) } \nonumber \\
  \qquad  \qquad & & \times \prod\limits_{j:t_{n-1} \leq \tau_j < t_n} p(\tau_j, u_j| \tau_{j-1}, u_{j-1})  \label{eq:cp_sequence_trandens}     .
@@
 
-A particle filter is an algorithm for approximating a probability distribution using a set of weighted samples (or ``particles'') drawn from that distribution.
+A particle filter is an algorithm for approximating a probability distribution using a set of weighted samples (or ``particles'') drawn from that distribution using importance sampling (IS).
 In this case, each particle will be a set of changepoint times and parameters.
@@
 \begin{equation}
- \hat{p}(\theta_{n}|y_{1:n}) = \sum_j w_n^{(j)} \delta_{\theta_{n}^{(j)}}(\theta_{n}) \label{eq:vrpf}
+ \hat{p}(\theta_{n}^-|y_{1:n}) = \sum_j w_n^{(j)} \delta_{\theta_{n}^{-(j)}}(\theta_{n}^-) \label{eq:vrpf}
 \end{equation}
@@
 where $\delta_x(X)$ is a unit probability mass at $X=x$.
-
 The particle filter works recursively.
-At the $n$th step, a particle, $\theta_{n-1}^{(i)}$, is first resampled from those approximating the filtering distribution at the $(n-1)$th step, using an appropriately chosen set of proposal weights, $\{v_{n-1}^{(j)}\}$ (where $\sum_j v_{n-1}^{(j)} = 1$),
+At the $n$th step, a particle, $\theta_{n-1}^{-(i)}$, is first resampled from those approximating the filtering distribution at the $(n-1)$th step, using an appropriately chosen set of proposal weights, $\{v_{n-1}^{(j)}\}$ (where $\sum_j v_{n-1}^{(j)} = 1$),
 %
 \begin{equation}
- q(\theta_{n-1}) = \sum_j v_{n-1}^{(j)} \delta_{\theta_{n-1}^{(j)}}(\theta_{n-1})     .
+ q(\theta_{n-1}^-) = \sum_j v_{n-1}^{(j)} \delta_{\theta_{n-1}^{-(j)}}(\theta_{n-1}^-)     .
 \end{equation}
@@
 The choice of weights determines the type of resampling used.
-The simplest choice, $v_{n-1}^{(j)} = 1/N_F$ (where $N_F$ is the number of filter particles) may be achieved by simply omitting this step all together and using the particles of $\hat{p}(\theta_{n-1}|y_{1:n-1})$.
+The simplest choice, $v_{n-1}^{(j)} = 1/N_F$ (where $N_F$ is the number of filter particles) may be achieved by simply omitting this step all together and using the particles of $\hat{p}(\theta_{n-1}^-|y_{1:n-1})$.
 This, however, leads to degeneracy of the particle weights over time.
@@
 
-Next, an extension to the changepoint sequence, $\theta_{n \setminus n-1}^{(i)}$, is sampled from an importance distribution, $q(\theta_{n \setminus n-1}|\theta_{n-1}^{(i)}, y_n)$, and concatenated with $\theta_{n-1}^{(i)}$ to create a proposal for $\theta_n^{(i)}$.
+Next, an extension to the changepoint sequence, $\theta_{n \setminus n-1}^{(i)}$, is sampled from an importance distribution, $q(\theta_{n \setminus n-1}|\theta_{n-1}^{-(i)}, y_n)$, and concatenated with $\theta_{n-1}^{-(i)}$ to create a proposal for $\theta_n^-{(i)}$.
 Finally, the particle is weighted according to the ratio of the target and proposal densities,
@@
 \begin{IEEEeqnarray}{rCl}
-w_n^{(i)} & = & \frac{ p(\theta_{n}^{(i)}|y_{1:n}) }{ q(\theta_{n}^{(i)}) } \nonumber \\
-    & \propto & \frac{ p(y_n|\theta_n^{(i)}, y_{1:n-1}) p(\theta_{n \setminus n-1}^{(i)}|\theta_{n-1}^{(i)}) p(\theta_{n-1}^{(i)}|y_{1:n-1}) }{ q(\theta_{n-1}^{(i)}) q(\theta_{n \setminus n-1}^{(i)}|\theta_{n-1}^{(i)}, y_n) } \nonumber \\
-    & =       & \frac{w_{n-1}^{(i)}}{v_{n-1}^{(i)}} \times \frac{ p(y_n|\theta_n^{(i)}, y_{1:n-1}) p(\theta_{n \setminus n-1}^{(i)}|\theta_{n-1}^{(i)}) }{ q(\theta_{n \setminus n-1}^{(i)}|\theta_{n-1}^{(i)}, y_n) } \label{eq:vrpf_weights}     .
+w_n^{(i)} & = & \frac{ p(\theta_{n}^{-(i)}|y_{1:n}) }{ q(\theta_{n}^{-(i)}) } \nonumber \\
+    & \propto & \frac{ p(y_n|\theta_n^{-(i)}, y_{1:n-1}) p(\theta_{n \setminus n-1}^{(i)}|\theta_{n-1}^{-(i)}) p(\theta_{n-1}^{-(i)}|y_{1:n-1}) }{ q(\theta_{n-1}^{-(i)}) q(\theta_{n \setminus n-1}^{(i)}|\theta_{n-1}^{-(i)}, y_n) } \nonumber \\
+    & =       & \frac{w_{n-1}^{(i)}}{v_{n-1}^{(i)}} \times \frac{ p(y_n|\theta_n^{-(i)}, y_{1:n-1}) p(\theta_{n \setminus n-1}^{(i)}|\theta_{n-1}^{-(i)}) }{ q(\theta_{n \setminus n-1}^{(i)}|\theta_{n-1}^{-(i)}, y_n) } \label{eq:vrpf_weights}     .
 \end{IEEEeqnarray}
@@
 
-It only remains to consider the likelihood term required for evaluation of the importance weights, $p(y_n|\theta_n, y_{1:n-1})$.
+It only remains to consider the likelihood term required for evaluation of the importance weights, $p(y_n|\theta_n^-, y_{1:n-1})$.
 The form of this term depends on the particular model under consideration.
@@
 
-For the conditionally linear-Gaussian models of section~\ref{sec:clg_models}, the required likelihood term, $p(y_n|\theta_n, y_{1:n-1})$ is the predictive distribution estimated by the Kalman filter.
+For the conditionally linear-Gaussian models of section~\ref{sec:clg_models}, the required likelihood term, $p(y_n|\theta_n^-, y_{1:n-1})$ is the predictive distribution estimated by the Kalman filter.
 Conveniently, the Kalman filter also provides us with an estimate of the current state given the changepoint sequence and all the preceding observations.
@@
 \begin{IEEEeqnarray}{rCl}
- p(x_n|\theta_{n}, y_{1:n-1}) & = & \mathcal{N}(x_n|m_n^-, P_n^-) \\
- p(x_n|\theta_{n}, y_{1:n}) & = & \mathcal{N}(x_n|m_n, P_n) \\
- p(y_n|\theta_{n}, y_{1:n-1}) & = & \mathcal{N}(y_n|\mu_n, S_n)     ,
-\end{IEEEeqnarray}
-
-with means and variances given by the following standard recursions (dependence on $\theta_{n}$ suppressed for clarity),
+ p(x_n|\theta_{n}^-, y_{1:n-1}) & = & \mathcal{N}(x_n|m_n^-, P_n^-) \\
+ p(x_n|\theta_{n}^-, y_{1:n}) & = & \mathcal{N}(x_n|m_n, P_n) \\
+ p(y_n|\theta_{n}^-, y_{1:n-1}) & = & \mathcal{N}(y_n|\mu_n, S_n)     ,
+\end{IEEEeqnarray}
+
+with means and variances given by the following standard recursions (dependence on $\theta_{n}^-$ suppressed for clarity),
 %
@@
 \linespread{1.5} \selectfont
-\STATE For each $i$, initialise particle changepoint sequence with $\{\theta_{0}^{(i)}\} \gets \{\tau_0^{(i)}, u_0^{(i)}\}$, where $\tau_0^{(i)}=0$ and $u_0^{(i)} \sim p(u_0)$.
+\STATE For each $i$, initialise particle changepoint sequence with $\{\theta_{0}^{-(i)}\} \gets \{\tau_0^{(i)}, u_0^{(i)}\}$, where $\tau_0^{(i)}=0$ and $u_0^{(i)} \sim p(u_0)$.
 \STATE For each $i$, initialise particle sufficient statistics, $m_0^{(i)}$ and $P_0^{(i)}$ with prior values.
@@
   \FOR{$i=1 \dots N_F$}
-  	\STATE Sample a history $\theta_{n-1}^{(i)} \sim \sum_j v_{n-1}^{(j)} \delta_{\theta_{n-1}^{(j)}}(\theta_{n-1})$.
-    \STATE Propose an extension $\theta_{n \setminus n-1}^{(i)} \sim q(\theta_{n \setminus n-1} | \theta_{n-1}^{(i)})$.
-    \STATE Add extension to sequence $\theta_n^{(i)} \gets \theta_{n-1}^{(i)} \cup \theta_{n \setminus n-1}^{(i)}$.
+  	\STATE Sample a history $\theta_{n-1}^{-(i)} \sim \sum_j v_{n-1}^{(j)} \delta_{\theta_{n-1}^{-(j)}}(\theta_{n-1}^-)$.
+    \STATE Propose an extension $\theta_{n \setminus n-1}^{(i)} \sim q(\theta_{n \setminus n-1} | \theta_{n-1}^{-(i)})$.
+    \STATE Add extension to sequence $\theta_n^{-(i)} \gets \theta_{n-1}^{-(i)} \cup \theta_{n \setminus n-1}^{(i)}$.
     \STATE Predict observation mean and covariance $\mu_n^{(i)}$ and $S_n^{(i)}$ using (\ref{eq:kf_predict_start})~to~(\ref{eq:kf_predict_stop}).
@@
 
-When the conditionally deterministic models of section~\ref{sec:cd_models} are used, the state at observation time $t_n$ is specified by the changepoint sequence $\theta_n$ (plus the initial state, $x_0$), using (\ref{eq:disc_time_state_diff_eq}).
+When the conditionally deterministic models of section~\ref{sec:cd_models} are used, the state at observation time $t_n$ is specified by the changepoint sequence $\theta_n^-$ (plus the initial state, $x_0$), using (\ref{eq:disc_time_state_diff_eq}).
 Thus, the required likelihood term is simply given by,
@@
 \begin{IEEEeqnarray}{rCl}
- p(y_n|\theta_{n}, y_{1:n-1}) & = & p(y_n|\hat{x}_n)     .
+ p(y_n|\theta_{n}^-, y_{1:n-1}) & = & p(y_n|\hat{x}_n)     .
 \end{IEEEeqnarray}
@@
   \FOR{$i=1 \dots N_F$}
-  	\STATE Sample a history $\theta_{n-1}^{(i)} \sim \sum_j v_{n-1}^{(j)} \delta_{\theta_{n-1}^{(j)}}(\theta_{n-1})$.
-    \STATE Propose an extension $\theta_{n \setminus n-1}^{(i)} \sim q(\theta_{n \setminus n-1}|\theta_{n-1}^{(i)}, y_{n})$.
-    \STATE Add extension to sequence $\theta_n^{(i)} \gets \theta_{n-1}^{(i)} \cup \theta_{n \setminus n-1}^{(i)}$.
+  	\STATE Sample a history $\theta_{n-1}^{-(i)} \sim \sum_j v_{n-1}^{(j)} \delta_{\theta_{n-1}^{-(j)}}(\theta_{n-1}^-)$.
+    \STATE Propose an extension $\theta_{n \setminus n-1}^{(i)} \sim q(\theta_{n \setminus n-1}|\theta_{n-1}^{-(i)}, y_{n})$.
+    \STATE Add extension to sequence $\theta_n^{-(i)} \gets \theta_{n-1}^{-(i)} \cup \theta_{n \setminus n-1}^{(i)}$.
     \STATE Calculate state $\hat{x}_n$ using (\ref{eq:disc_time_state_diff_eq}).
@@
 \begin{IEEEeqnarray}{rCl}
- p(\theta|y_{1:N}) = p(\theta_{n}^{+}|y_{1:N}) p(\theta_{n}|\theta_{n}^{+}, y_{1:N})     .
+ p(\theta|y_{1:N}) = p(\theta_{n}^{+}|y_{1:N}) p(\theta_{n}^-|\theta_{n}^{+}, y_{1:N})     .
 \end{IEEEeqnarray}
@@
 This suggests a sequential strategy for particle smoothing.
-Starting with a low diversity set of particles from $p(\theta|y_{1:N})$, we first marginalise $\theta_{n}$ by simply discarding the changepoints before $t_n$.
-New values of $\theta_{n}$ are then sampled from the conditional distribution, $p(\theta_{n}|\theta_{n}^{+}, y_{1:N})$, which is approximated using the filter particles, and the past and future sequences are concatenated.
+Starting with a low diversity set of particles from $p(\theta|y_{1:N})$, we first marginalise $\theta_{n}^-$ by simply discarding the changepoints before $t_n$.
+New values of $\theta_{n}^-$ are then sampled from the conditional distribution, $p(\theta_{n}^-|\theta_{n}^{+}, y_{1:N})$, which is approximated using the filter particles, and the past and future sequences are concatenated.
 By iterating backwards over $n = N \dots 1$, a diverse set of particles approximating $p(\theta|y_{1:N})$ will be generated.
 
-It remains to devise a scheme for sampling the conditional distribution, $p(\theta_{n}|\theta_{n}^{+}, y_{1:N})$.
+It remains to devise a scheme for sampling the conditional distribution, $p(\theta_{n}^-|\theta_{n}^{+}, y_{1:N})$.
 The simplifications used by the fixed rate particle smoother which exploit the Markovian nature of the state sequence \cite{Godsill2004} are of no help, because past changepoints are not independent of future observations in variable rate models (i.e.
-$p(\theta_{n}|\theta_{n}^{+}, y_{1:N}) \ne p(\theta_{n}|\theta_{n}^{+}, y_{1:n}))$.
+$p(\theta_{n}^-|\theta_{n}^{+}, y_{1:N}) \ne p(\theta_{n}^-|\theta_{n}^{+}, y_{1:n}))$.
 However, when the state dynamics are linear-Gaussian conditional on the changepoint sequence, it is possible split the dependence on the observations into past and future by introducing the current state as an additional variable.
@@
 \begin{IEEEeqnarray}{rCl}
- \IEEEeqnarraymulticol{3}{l}{p(\theta_{n}|\theta_{n}^+, y_{1:N})} \nonumber \\
-\qquad & \propto & p(\theta_{n}, \theta_{n}^+| y_{1:N}) \nonumber  \\
-       & =       & \int p(x_n, \theta_{n}, \theta_{n}^+| y_{1:N}) dx_n \nonumber  \\
-       & \propto & \int p(y_{n+1:N}|x_n, \theta_{n}, \theta_{n}^+, y_{1:n}) p(x_n, \theta_{n}, \theta_{n}^+| y_{1:n}) dx_n \nonumber \\
-       & = & \int p(y_{n+1:N}|x_n, \theta_{n}^+) p(x_n|\theta_{n}, y_{1:n}) dx_n \nonumber \\
-       &   & \times p(\theta_{n}^+|\theta_{n}) p(\theta_{n}|y_{1:n})     .
-\end{IEEEeqnarray}
-
-The particle approximation of (\ref{eq:vrpf}) is substituted for the filtering distribution, $p(\theta_{n}|y_{1:n})$, yielding,
-%
-\begin{IEEEeqnarray}{rCl}
-\hat{p}(\theta_{n}|\theta_{n}^+, y_{1:N}) & = & \sum_i \tilde{w}_{n}^{(i)} \delta_{\theta_{n}^{(i)}}(\theta_{n}) \label{eq:rb-vrps_back_cond}     ,
+ \IEEEeqnarraymulticol{3}{l}{p(\theta_{n}^-|\theta_{n}^+, y_{1:N})} \nonumber \\
+\qquad & \propto & p(\theta_{n}^-, \theta_{n}^+| y_{1:N}) \nonumber  \\
+       & =       & \int p(x_n, \theta_{n}^-, \theta_{n}^+| y_{1:N}) dx_n \nonumber  \\
+       & \propto & \int p(y_{n+1:N}|x_n, \theta_{n}^-, \theta_{n}^+, y_{1:n}) p(x_n, \theta_{n}^-, \theta_{n}^+| y_{1:n}) dx_n \nonumber \\
+       & = & \int p(y_{n+1:N}|x_n, \theta_{n}^+) p(x_n|\theta_{n}^-, y_{1:n}) dx_n \nonumber \\
+       &   & \times p(\theta_{n}^+|\theta_{n}^-) p(\theta_{n}^-|y_{1:n})     .
+\end{IEEEeqnarray}
+
+The particle approximation of (\ref{eq:vrpf}) is substituted for the filtering distribution, $p(\theta_{n}^-|y_{1:n})$, yielding,
+%
+\begin{IEEEeqnarray}{rCl}
+\hat{p}(\theta_{n}^-|\theta_{n}^+, y_{1:N}) & = & \sum_i \tilde{w}_{n}^{(i)} \delta_{\theta_{n}^{-(i)}}(\theta_{n}^-) \label{eq:rb-vrps_back_cond}     ,
 \end{IEEEeqnarray}
@@
  \tilde{w}_n & \propto & w_n \int p(y_{n+1:N}|x_n, \theta_{n}^+) \nonumber \\
-             &         & \times p(x_n|\theta_{n}^{(i)}, y_{1:n}) dx_n p(\theta_{n}^+|\theta_{n}^{(i)}) \label{eq:rb-vrps_back_cond_weight}     .
+             &         & \times p(x_n|\theta_{n}^{-(i)}, y_{1:n}) dx_n p(\theta_{n}^+|\theta_{n}^{-(i)}) \label{eq:rb-vrps_back_cond_weight}     .
 \end{IEEEeqnarray}
@@
 \begin{IEEEeqnarray}{rCl}
- p(\theta_{n}^+|\theta_{n}) &=      & p(\tau_{K_n+1:K}, u_{K_n+1:K}|\tau_{K_n}, u_{K_n}, \tau_{K_n+1}>t_n) \nonumber \\
+ p(\theta_{n}^+|\theta_{n}^-) &=      & p(\tau_{K_n+1:K}, u_{K_n+1:K}|\tau_{K_n}, u_{K_n}, \tau_{K_n+1}>t_n) \nonumber \\
                                     &\propto& p(\tau_{K_n+1}, u_{K_n+1}|\tau_{K_n}, u_{K_n}, \tau_{K_n+1}>t_n)     .
@@
 \begin{IEEEeqnarray}{rCl}
-p(x_n|\theta_{n}^{(i)}, y_{1:n}) & = & \mathcal{N}(x_n|m_n^{(i)}, P_n^{(i)})     .
+p(x_n|\theta_{n}^{-(i)}, y_{1:n}) & = & \mathcal{N}(x_n|m_n^{(i)}, P_n^{(i)})     .
 \end{IEEEeqnarray}
@@
 \begin{equation}
- \tilde{w}_n \propto w_n p(\theta_{n}^+|\theta_{n}^{(i)}) \mathcal{N}(\tilde{m}_n^-|m_n, \tilde{P}_n^- + P_n)     .
+ \tilde{w}_n \propto w_n p(\theta_{n}^+|\theta_{n}^{-(i)}) \mathcal{N}(\tilde{m}_n^-|m_n, \tilde{P}_n^- + P_n)     .
 \label{eq:rb-vrps_back_cond_weight2}
@@
 
-Using these weights, a sample of $\theta_{n}$ may be drawn from the particle distribution of (\ref{eq:rb-vrps_back_cond}), completing one step of the smoothing algorithm.
+Using these weights, a sample of $\theta_{n}^-$ may be drawn from the particle distribution of (\ref{eq:rb-vrps_back_cond}), completing one step of the smoothing algorithm.
 Once sampling has progressed backwards over $n=N \dots 1$, a particle from the smoothing distribution will have been generated.
@@
 \linespread{1.5} \selectfont
-  \STATE Run Rao-Blackwellised variable rate particle filter to approximate $p(\theta_{n}|y_{1:n})$ with weighted particles $\{\theta_{n}^{(i)}, w_{n}^{(i)}\}$ and $\{p(x_n|\theta_{n}^{(i)},y_{1:n})\}$ as normal distributions with moments $\{m_{n}^{(i)}\}$ and $\{P_{n}^{(i)}\}$.
+  \STATE Run Rao-Blackwellised variable rate particle filter to approximate $p(\theta_{n}^-|y_{1:n})$ with weighted particles $\{\theta_{n}^{-(i)}, w_{n}^{(i)}\}$ and $\{p(x_n|\theta_{n}^{-(i)},y_{1:n})\}$ as normal distributions with moments $\{m_{n}^{(i)}\}$ and $\{P_{n}^{(i)}\}$.
 Store all results.
@@
     \FOR{$n=N-1 \dots 1$}
-      \STATE Discard $\theta_{n}^{(i)}$.
+      \STATE Discard $\theta_{n}^{-(i)}$.
       \STATE Predict state mean and covariance $\tilde{m}_n^{-(i)}$ and $\tilde{P}_n^{-(i)}$ using (\ref{eq:backward_kf_predict_start}) to (\ref{eq:backward_kf_predict_stop}).
@@
       \ENDFOR
-      \STATE Sample $\theta_{n}^{(i)} \sim \sum_j \tilde{w}_n^{(j)} \delta_{\theta_{n}^{(j)}}(\theta_{n})$.
-      \STATE Join $\theta^{(i)} \gets \theta_{n}^{(i)} \cup \theta_{n}^{+(i)}$.
-      \STATE Update state mean and covariance $\tilde{m}_n^{(i)}$ and $\tilde{P}_n^{(i)}$ using \ref{eq:backward_kf_update_start} to \ref{eq:backward_kf_update_stop}.
+      \STATE Sample $\theta_{n}^{-(i)} \sim \sum_j \tilde{w}_n^{(j)} \delta_{\theta_{n}^{-(j)}}(\theta_{n}^-)$.
+      \STATE Join $\theta^{(i)} \gets \theta_{n}^{-(i)} \cup \theta_{n}^{+(i)}$.
+      \STATE Update state mean and covariance $\tilde{m}_n^{(i)}$ and $\tilde{P}_n^{(i)}$ using (\ref{eq:backward_kf_update_start}) to (\ref{eq:backward_kf_update_stop}).
     \ENDFOR
@@
 In the conditionally deterministic models of section~\ref{sec:cd_models}, the entire continuous-time state trajectory, $x(t)$, is a function of the changepoint sequence $\theta$ and the initial state, $x_0$.
-When the changepoint sequence is split into past, $\theta_n$, and future $\theta_n^+$, and the past is altered, this will alter the state trajectory in the future.
+When the changepoint sequence is split into past, $\theta_n^-$, and future $\theta_n^+$, and the past is altered, this will alter the state trajectory in the future.
 This is undesirable in practice, as each sampling operation will require the recalculation of state values and observations likelihoods for all future observation times.
 
-In order to achieve conditional independence between past and future observations, consider the augmented changepoint sequence, $\tilde{\theta} = \{\tau_{0:K}, u_{0:K}, x_{0:K}\}$ ($\tilde{\theta}_n$ and $\tilde{\theta}_n^+$ defined in the same manner as before).
+In order to achieve conditional independence between past and future observations, consider the augmented changepoint sequence, $\tilde{\theta} = \{\tau_{0:K}, u_{0:K}, x_{0:K}\}$ ($\tilde{\theta}_n^-$ and $\tilde{\theta}_n^+$ defined in the same manner as before).
 The changepoint state values, $x_{0:K}$ may be calculated deterministically from $\theta$ for each particle.
@@
 \begin{IEEEeqnarray}{rCl}
- p(\tilde{\theta}|y_{1:N}) = p(\tilde{\theta}_{n}^{+}|y_{1:N}) p(\tilde{\theta}_{n}|\tilde{\theta}_{n}^{+}, y_{1:N})     .
+ p(\tilde{\theta}|y_{1:N}) = p(\tilde{\theta}_{n}^{+}|y_{1:N}) p(\tilde{\theta}_{n}^-|\tilde{\theta}_{n}^{+}, y_{1:N})     .
 \end{IEEEeqnarray}
@@
 \begin{IEEEeqnarray}{rCl}
-p(\tilde{\theta}_{n}|\tilde{\theta}_{n}^{+}, y_{1:N}) & \propto & p(y_{n+1:N}|\tilde{\theta}_{n}, \tilde{\theta}_{n}^{+}) \nonumber \\
-& & \times p(\tilde{\theta}_{n}^{+}|\tilde{\theta}_{n}) p(\tilde{\theta}_{n}|y_{1:n})    .
+p(\tilde{\theta}_{n}^-|\tilde{\theta}_{n}^{+}, y_{1:N}) & \propto & p(y_{n+1:N}|\tilde{\theta}_{n}^-, \tilde{\theta}_{n}^{+}) \nonumber \\
+& & \times p(\tilde{\theta}_{n}^{+}|\tilde{\theta}_{n}^-) p(\tilde{\theta}_{n}^-|y_{1:n})    .
 \end{IEEEeqnarray}
@@
 \begin{IEEEeqnarray}{rCl}
-p(\tilde{\theta}_{n}^+ | \tilde{\theta}_{n}) & \propto & p(\tau_{K_n+1}, u_{K_n+1}| \tau_{K_n}, u_{K_n}) \nonumber \\
+p(\tilde{\theta}_{n}^+ | \tilde{\theta}_{n}^-) & \propto & p(\tau_{K_n+1}, u_{K_n+1}| \tau_{K_n}, u_{K_n}) \nonumber \\
  & & \times \delta_{f(x_{K_n}, u_{K_n}, \tau_{K_n}, \tau_{K_n+1})}(x_{K_n+1}),
@@
 The solution to this problem is provided by an idea from the Sequential Monte Carlo (SMC) samplers of \cite{DelMoral2006}.
-Rather than sampling $p(\tilde{\theta}_{n}|\tilde{\theta}_{n}^{+}, y_{1:N})$ directly, an augmented distribution is targeted.
-A value of $\tilde{\theta}_{n}$ can be proposed from the filtering particles, after which a replacement, $u_{K_n}'$, is proposed for the existing final changepoint parameter, with a value chosen such that the past and future paths meet up.
-The modified sequence is denoted $\tilde{\theta}_{n}'$.
+Rather than sampling $p(\tilde{\theta}_{n}^-|\tilde{\theta}_{n}^{+}, y_{1:N})$ directly, an augmented distribution is targeted.
+A value of $\tilde{\theta}_{n}^-$ can be proposed from the filtering particles, after which a replacement, $u_{K_n}'$, is proposed for the existing final changepoint parameter, with a value chosen such that the past and future paths meet up.
+The modified sequence is denoted $\tilde{\theta}_{n}^{-'}$.
 Such a proposal density may be written as,
@@
 \begin{IEEEeqnarray}{rCl}
-\IEEEeqnarraymulticol{3}{l}{p(\tilde{\theta}_{n}|y_{1:n}) q(u_{K_n}'|\tilde{\theta}_{n}, \tilde{\theta}_{n}^{+}, y_{1:N})} \nonumber \\
- \qquad & \propto & p(y_{1:n}|\tilde{\theta}_{n}) p(\tilde{\theta}_{n}) q(u_{K_n}'|\tilde{\theta}_{n}, \tilde{\theta}_{n}^{+}, y_{1:N}) \label{eq:pd-vrps_proposal}     .
-\end{IEEEeqnarray}
-
-The augmented target distribution over the the new sequence, $\tilde{\theta}_{n}'$, and the discarded parameter, $u_{K_n}$, is constructed by introducing an artificial density,
-%
-\begin{IEEEeqnarray}{rCl}
-\IEEEeqnarraymulticol{3}{l}{ p(\tilde{\theta}_{n}'|\tilde{\theta}_{n}^{+}, y_{1:N}) L(u_{K_n}|\tilde{\theta}_{n}', \tilde{\theta}_{n}^{+}, y_{1:N}) } \nonumber \\
- & \propto & p(y_{1:N}|\tilde{\theta}_{n}', \tilde{\theta}_{n}^+) p(\tilde{\theta}_{n}^{+}|\tilde{\theta}_{n}') p(\tilde{\theta}_{n}') L(u_{K_n}|\tilde{\theta}_{n}', \tilde{\theta}_{n}^{+}, y_{1:N})  \label{eq:pd-vrps_target}     .
-\end{IEEEeqnarray}
-
-Clearly this new target distribution admits the desired conditional posterior, $p(\tilde{\theta}_{n}'|\tilde{\theta}_{n}^{+}, y_{1:N})$, as a marginal.
+\IEEEeqnarraymulticol{3}{l}{p(\tilde{\theta}_{n}^-|y_{1:n}) q(u_{K_n}'|\tilde{\theta}_{n}^-, \tilde{\theta}_{n}^{+}, y_{1:N})} \nonumber \\
+ \qquad & \propto & p(y_{1:n}|\tilde{\theta}_{n}^-) p(\tilde{\theta}_{n}^-) q(u_{K_n}'|\tilde{\theta}_{n}^-, \tilde{\theta}_{n}^{+}, y_{1:N}) \label{eq:pd-vrps_proposal}     .
+\end{IEEEeqnarray}
+
+The augmented target distribution over the the new sequence, $\tilde{\theta}_{n}^{-'}$, and the discarded parameter, $u_{K_n}$, is constructed by introducing an artificial density,
+%
+\begin{IEEEeqnarray}{rCl}
+\IEEEeqnarraymulticol{3}{l}{ p(\tilde{\theta}_{n}^{-'}|\tilde{\theta}_{n}^{+}, y_{1:N}) L(u_{K_n}|\tilde{\theta}_{n}^{-'}, \tilde{\theta}_{n}^{+}, y_{1:N}) } \nonumber \\
+ \qquad & \propto & p(y_{1:N}|\tilde{\theta}_{n}^{-'}, \tilde{\theta}_{n}^+) p(\tilde{\theta}_{n}^{+}|\tilde{\theta}_{n}^{-'}) p(\tilde{\theta}_{n}^{-'}) \nonumber \\
+ & & \times L(u_{K_n}|\tilde{\theta}_{n}^{-'}, \tilde{\theta}_{n}^{+}, y_{1:N})  \label{eq:pd-vrps_target}     .
+\end{IEEEeqnarray}
+
+Clearly this new target distribution admits the desired conditional posterior, $p(\tilde{\theta}_{n}^{-'}|\tilde{\theta}_{n}^{+}, y_{1:N})$, as a marginal.
 
@@
 The augmented target distribution may be sampled using Markov chain Monte Carlo (MCMC) \cite{Gilks1996}, in a similar manner to the fixed rate smoother of \cite{Bunch2012}.
-Metropolis-Hastings (MH) steps are conducted to draw samples, $\{\tilde{\theta}_{n}, u_{K_n}'\}$, from the target distribution by sampling the proposal and accepting the new values with a given probability.
+Metropolis-Hastings (MH) steps are conducted to draw samples, $\{\tilde{\theta}_{n}^-, u_{K_n}'\}$, from the target distribution by sampling the proposal and accepting the new values with a given probability.
 Note that IS could be used instead of MCMC, but this would require resampling steps, which would lead to a loss of path-space diversity similar to that suffered by the particle filter.
@@
 \begin{IEEEeqnarray}{rCl}
-\beta_n & \propto & \frac{ p(y_{r_n^-:r_n^+}|\tilde{\theta}_{n}') }{ p(y_{r_n^-:n}|\tilde{\theta}_{n}) } \times \frac{ p(u_{K_n}'|\tau_{K_n}, \tau_{K_n-1}, u_{K_n-1}) }{ p(u_{K_n}|\tau_{K_n}, \tau_{K_n-1}, u_{K_n-1}) } \nonumber \\
-  & & \times p(\tilde{\theta}_{n}^{+}|\tilde{\theta}_{n}') \times \frac{ L(u_{K_n}|\tilde{\theta}_{n}', \tilde{\theta}_{n}^{+}, y_{1:N}) }{ q(u_{K_n}'|\tilde{\theta}_{n}, \tilde{\theta}_{n}^{+}, y_{1:N}) } \label{eq:pd-vrps_tpr}    ,
+\beta_n & \propto & \frac{ p(y_{r_n^-:r_n^+}|\tilde{\theta}_{n}^{-'}) }{ p(y_{r_n^-:n}|\tilde{\theta}_{n}^-) } \times \frac{ p(u_{K_n}'|\tau_{K_n}, \tau_{K_n-1}, u_{K_n-1}) }{ p(u_{K_n}|\tau_{K_n}, \tau_{K_n-1}, u_{K_n-1}) } \nonumber \\
+  & & \times p(\tilde{\theta}_{n}^{+}|\tilde{\theta}_{n}^{-'}) \times \frac{ L(u_{K_n}|\tilde{\theta}_{n}^{-'}, \tilde{\theta}_{n}^{+}, y_{1:N}) }{ q(u_{K_n}'|\tilde{\theta}_{n}^-, \tilde{\theta}_{n}^{+}, y_{1:N}) } \label{eq:pd-vrps_tpr}    ,
 \end{IEEEeqnarray}
@@
 
-The proposal density $q(u_{K_n}'|\tilde{\theta}_{n}, \tilde{\theta}_{n}^{+}, y_{1:N})$ is constructed so as to be non-zero only where $p(\tilde{\theta}_{n}^{+}|\tilde{\theta}_{n}')$ is non-zero.
+The proposal density $q(u_{K_n}'|\tilde{\theta}_{n}^-, \tilde{\theta}_{n}^{+}, y_{1:N})$ is constructed so as to be non-zero only where $p(\tilde{\theta}_{n}^{+}|\tilde{\theta}_{n}^{-'})$ is non-zero.
 This could require it to be a delta function if there is only one feasible value.
-The artificial conditional term, $L(u_{K_n}|\tilde{\theta}_{n}, \tilde{\theta}_{n}^{+}, y_{1:N})$, may be any arbitrary density, but the simplest choice is a uniform distribution over the set of parameter values, $\{u_{K_n}^{(i)}\}$, which appear in the filtering approximation.
+The artificial conditional term, $L(u_{K_n}|\tilde{\theta}_{n}^-, \tilde{\theta}_{n}^{+}, y_{1:N})$, may be any arbitrary density, but the simplest choice is a uniform distribution over the set of parameter values, $\{u_{K_n}^{(i)}\}$, which appear in the filtering approximation.
 With this choice, the artificial density terms will always cancel out in the MH acceptance probability.
@@
 \linespread{1.5} \selectfont
-  \STATE Run variable rate particle filter to approximate $p(\tilde{\theta}_{n}|y_{1:n})$ with weighted particles $\{\tilde{\theta}_{n}^{(i)}, w_{n}^{(i)}\}$. Store all results.
+  \STATE Run variable rate particle filter to approximate $p(\tilde{\theta}_{n}^-|y_{1:n})$ with weighted particles $\{\tilde{\theta}_{n}^{-(i)}, w_{n}^{(i)}\}$. Store all results.
   \FOR{$i=1 \dots N_S$}
@@
     \FOR{$n=N \dots 1$}
-      \STATE Initialise chain with $\tilde{\theta}_n^{(i)(0)} \gets \tilde{\theta}_n^{(i)}$.
+      \STATE Initialise chain with $\tilde{\theta}_n^{-(i)(0)} \gets \tilde{\theta}_n^{-(i)}$.
       \FOR{$m=1 \dots M$}
-        \STATE Propose a new history $\tilde{\theta}_n^{(i)*} \sim \sum_j w_n^{(j)} \delta_{\tilde{\theta}_{n}^{(j)}}(\tilde{\theta}_{n})$.
-        \STATE Propose a parameter $u_{K_n}'^* \sim q(u_{K_n}|\tilde{\theta}_{n}, \tilde{\theta}_{n}^{+}, y_{1:N})$.
-        \STATE Replace $u_{K_n}^*$ with $u_{K_n}'^*$ to form $\tilde{\theta}_n'^{(i)*}$.
+        \STATE Propose a new history $\tilde{\theta}_n^{-(i)*} \sim \sum_j w_n^{(j)} \delta_{\tilde{\theta}_{n}^{-(j)}}(\tilde{\theta}_{n}^-)$.
+        \STATE Propose a parameter $u_{K_n}'^* \sim q(u_{K_n}|\tilde{\theta}_{n}^-, \tilde{\theta}_{n}^{+}, y_{1:N})$.
+        \STATE Replace $u_{K_n}^*$ with $u_{K_n}'^*$ to form $\tilde{\theta}_n^{-'(i)*}$.
         \STATE Calculate $\alpha_n^{(m)}$ using (\ref{eq:pd-vrps_tpr}) and (\ref{eq:pd-vrps_ap}).
-        \STATE With probability $\alpha_n^{(m)}$, $\tilde{\theta}_n^{(i)(m)} \gets \tilde{\theta}_n^{(i)*}$. \\ Otherwise, $\tilde{\theta}_n^{(i)(m)} \gets \tilde{\theta}_n^{(i)(m-1)}$.
+        \STATE With probability $\alpha_n^{(m)}$, $\tilde{\theta}_n^{-(i)(m)} \gets \tilde{\theta}_n^{-(i)*}$. \\ Otherwise, $\tilde{\theta}_n^{-(i)(m)} \gets \tilde{\theta}_n^{-(i)(m-1)}$.
       \ENDFOR
-      \STATE Store final sample $\tilde{\theta}_n^{(i)} \gets \tilde{\theta}_n^{(i)(M)}$.
+      \STATE Store final sample $\tilde{\theta}_n^{-(i)} \gets \tilde{\theta}_n^{-(i)(M)}$.
     \ENDFOR

### 2012-06-01 full_finaldraft_VRsmoothing_withSJGcomments.tex
moved models to section 2. lots of minor tidying.

@@
 
-Conditionally linear-Gaussian models for use with variable rate particle filters were introduced in \cite{Godsill2007a,Christensen2012} for financial and tracking problems.
-For example, changepoints might correspond to jumps in the value or trend of a security, at which points the process covariance, $Q_n(\theta_n^-)$, is inflated in an optimal fashion.
+\subsubsection*{A Jump-Diffusion model for Financial Assets} \label{sec:financial_model}
+
+Conditionally linear-Gaussian models for use with variable rate particle filters were introduced in \cite{Godsill2007a,Christensen2012} for financial applications.
+A version of these models is used for testing the new smoothing algorithm.
+Prices of an asset are treated as noisy observations of a latent state, which evolves according to a drift-diffusion with occasional jumps.
+
+The state is a vector with two elements, the underlying value of the asset, and the trend followed by this value,
+%
+\begin{equation}
+ x(t) = [ z(t), \dot{z}(t)]^T     .
+\end{equation}
+
+This evolves continuously according to a jump-diffusion model,
+%
+\begin{IEEEeqnarray}{c}
+ dx(t) = \begin{bmatrix}0 & 1 \\ 0 & -\lambda \end{bmatrix} x(t) dt + \begin{bmatrix}0 \\ \sigma \end{bmatrix} d\beta(t) + dJ(t)
+\end{IEEEeqnarray}
+
+where $\lambda$ introduces a mean reversion effect on the trend and $\beta(t)$ is standard Brownian motion (with unit diffusion constant).
+The jump term, $dJ(t)$, is zero everywhere except when jumps occur,
+%
+\begin{IEEEeqnarray}{rCl}
+ dJ(t) & = & \begin{cases} J_k & t \in \{\tau_k\} \\ 0 & \text{elsewhere} \end{cases} \\
+ J_k  & \sim & \mathcal{N}(J_k| 0, Q_{J,u_k})     .
+\end{IEEEeqnarray}
+
+$\{\tau_k\}$ is the set of random jump times, each of which is one of two types: a value jump, indicated by $u_k = 1$, or a trend jump, indicated by $u_k=2$. The jump covariance matrices are,
+%
+\begin{IEEEeqnarray}{c}
+Q_{J,u_k} = \begin{cases} \begin{bmatrix}\sigma_{J1}^2 & 0 \\ 0 & 0 \end{bmatrix} & u_k = 1 \\
+                          \begin{bmatrix}0 & 0 \\ 0 & \sigma_{J2}^2 \end{bmatrix} & u_k = 2  \end{cases}   .
+\end{IEEEeqnarray}
+
+Only the value is observed, and that corrupted by additive Gaussian noise with standard deviation $\sigma_y^2$.
+
+This model may be discretised at the observation times using standard methods, detailed in appendix~\ref{app:lg_model_discretisation}, leading to,% (see \cite{Godsill2007a,Christensen2012,Grewal2002}), leading to,
+%
+%Assuming observations of value only and Gaussian observation noise with standard deviation $\sigma_y^2$, the resulting discrete time dynamics are described by the following equations:% (see appendix \ref{app:model_discretisation}):
+%
+\begin{IEEEeqnarray}{rCl}
+ x_n & = & A x_{n-1} + w_n \\
+ y_n & = & C x_{n} + v_n
+\end{IEEEeqnarray}
+
+where the $w_n$ and $v_n$ are Gaussian random variables with covariance matrixes $Q_n$ and $R$ respectively.
+$Q_n$ is a function of the sequence of jump times and their types, giving rise to the desired conditionally linear-Gaussian structure.
 
@@
 
+\subsubsection*{An Intrinsic Coordinate Model for Manoeuvring Vehicles} \label{sec:tracking_model}
+
 Target tracking algorithms are commonly based upon fixed rate models (see, e.g.
@@
 radar measurements) are made.
-In \cite{Godsill2007a,Godsill2007,Whiteley2011}, variable rate models were introduced for tracking, in which the state trajectory is divided up by a set of changepoints between which the motion follows a deterministic path governed by motion parameters (accelerations, etc.) which are fixed for that division.
+In \cite{Godsill2007a,Godsill2007,Whiteley2011,Bunch2012a}, variable rate models were introduced for tracking.
+The state trajectory is divided up by a set of changepoints, between which the motion follows a deterministic path governed by motion parameters (accelerations, etc.) which are fixed for that division.
+In this paper we use the 2-dimensional intrinsic coordinate model of \cite{Bunch2012a} to test the new smoothing algorithm.
+
+The target is modelled as a particle subject to perpendicular forces tangential and normal to the direction of motion.
+These forces are held constant between changepoint times, $\tau_{1:K}$, which represent the beginnings and ends of manoeuvres.
+The state vector consists of Cartesian position, $x_1(t)$ and $x_2(t)$, plus heading angle, $\psi(t)$, and speed, $\dot{s}(t)$,
+%
+\begin{equation}
+x(t) = [x_1(t), x_2(t), \psi(t), \dot{s}(t)]^T     .
+\end{equation}
+
+In addition to the tangential and normal accelerations, $a_{T,k}$ and $a_{N,k}$, we introduce two linear velocity terms, $d_{X,k}$ and $d_{Y,k}$.
+As will become apparent later, these are necessary for the operation of an efficient smoothing algorithm.
+They provide us with enough degrees of freedom to match the state trajectory between an arbitrary past changepoint ($\tau_{K(t)}$, $\mathbf{x}_{K(t)}$) and an arbitrary future changepoint ($\tau_{K(t)+1}$, $\mathbf{x}_{K(t)+1}$).
+These linear velocity parameters may be considered to be merely relaxation terms, or they may represent real physical effects, such as wind (for aircraft) or currents (for boats).
+Together, these four variables make up the vector of motion parameters,
+%
+\begin{equation}
+u_k = [a_{T,k}, a_{N,k}, d_{X,k}, d_{Y,k}]^T     .
+\end{equation}
+
+The target dynamics are described by four differential equations,
+%
+\begin{IEEEeqnarray}{rCl}
+\ddot{s}(t) & = & a_{T,K(t)} \\
+\dot{s}(t) \dot{\psi}(t) & = & a_{N,K(t)} \\
+\dot{x}_{1}(t) & = & \dot{s}(t) \cos(\psi(t)) + d_{X,K(t)} \\
+\dot{x}_{2}(t) & = & \dot{s}(t) \sin(\psi(t)) + d_{Y,K(t)}     .
+\end{IEEEeqnarray}
+
+Solving these yields the following state equations (where $\Delta\ t = t - \tau_{K(t)}$),
+%
+\begin{IEEEeqnarray}{rCl}
+\dot{s}(t) & = & \dot{s}_{K(t)} + a_{T,k} \Delta t \label{eq:2D_ICmodel_1} \\
+\psi(t) & = & \psi_{K(t)} + \frac{a_{N,k}}{a_{T,k}} \log \left( \frac{\dot{s}(t)}{\dot{s}_{K(t)}} \right) \label{eq:2D_ICmodel_2}
+\end{IEEEeqnarray}
+\begin{IEEEeqnarray}{rCl}
+x_1(t) & = & x_{1,K(t)} + d_{X,K(t)} \Delta t \label{eq:2D_ICmodel_3} \\
+     \IEEEeqnarraymulticol{3}{l}{ \quad + \: \frac{ \dot{s}(t)^2 }{ 4 a_{T,k}^2 + a_{N,k}^2 } \left[  a_{N,k} \sin(\psi(t)) + 2 a_{T,k} \cos(\psi(t))  \right]} \nonumber \\
+     \IEEEeqnarraymulticol{3}{l}{ \quad - \: \frac{\dot{s}_{K(t)}^2}{4 a_{T,k}^2 + a_{N,k}^2} \left[  a_{N,k} \sin(\psi_{K(t)}) + 2 a_{T,k} \cos(\psi_{K(t)})  \right]} \nonumber
+\end{IEEEeqnarray}
+\begin{IEEEeqnarray}{rCl}
+x_2(t) & = & x_{2,K(t)} + d_{Y,K(t)} \Delta t \label{eq:2D_ICmodel_4} \\
+     \IEEEeqnarraymulticol{3}{l}{ \quad + \: \frac{ \dot{s}(t)^2 }{ 4 a_{T,k}^2 + a_{N,k}^2 } \left[ -a_{N,k} \cos(\psi(t)) + 2 a_{T,k} \sin(\psi(t))  \right]} \nonumber \\
+     \IEEEeqnarraymulticol{3}{l}{ \quad - \: \frac{\dot{s}_{K(t)}^2}{4 a_{T,k}^2 + a_{N,k}^2} \left[  -a_{N,k} \cos(\psi_{K(t)}) + 2 a_{T,k} \sin(\psi_{K(t)})  \right]} \nonumber      .
+\end{IEEEeqnarray}
+
+When formulating the smoothing algorithm, it will be necessary that $u_k$ can be calculated for any pair of adjacent changepoints, $(\tau_k,x_k)$, and $(\tau_{k+1},x_{k+1})$.
+This condition is met by the preceding model --- the equations are solved in the order specified to find $a_{T,k}$ from (\ref{eq:2D_ICmodel_1}), $a_{N,k}$ from (\ref{eq:2D_ICmodel_2}), and finally $d_{X,K(t)}$ and $d_{Y,K(t)}$ from (\ref{eq:2D_ICmodel_3}) and (\ref{eq:2D_ICmodel_4}).
 
@@
 \begin{IEEEeqnarray}{rCl}
- m_n^- & = & A_n m_{n-1} \label{eq:kf_predict_start} \\
- P_n^- & = & A_n P_{n-1} A_n^T + Q_n \\
- \mu_n & = & C_n m_n^- \\
- S_n   & = & C_n P_n^- C_n^T + R_n \label{eq:kf_predict_stop} \\
- K_n   & = & P_n^- C_n^T S_n^{-1} \label{eq:kf_update_start}\\
- m_n   & = & m_n^- + K_n (y_n - \mu_n) \\
- P_n   & = & P_n^- - K_n S_n K_n^T \label{eq:kf_update_stop}    .
+ m_n^- & = & A_n m_{n-1} \nonumber \\
+ P_n^- & = & A_n P_{n-1} A_n^T + Q_n \nonumber \\
+ \mu_n & = & C_n m_n^- \nonumber \\
+ S_n   & = & C_n P_n^- C_n^T + R_n \label{eq:kf_predict} \\
+ \nonumber \\
+ G_n   & = & P_n^- C_n^T S_n^{-1} \nonumber \\
+ m_n   & = & m_n^- + G_n (y_n - \mu_n) \nonumber \\
+ P_n   & = & P_n^- - G_n S_n G_n^T \label{eq:kf_update}    .
 \end{IEEEeqnarray}
@@
     \STATE Add extension to sequence $\theta_n^{-(i)} \gets \theta_{n-1}^{-(i)} \cup \theta_{n \setminus n-1}^{(i)}$.
-    \STATE Predict observation mean and covariance $\mu_n^{(i)}$ and $S_n^{(i)}$ using (\ref{eq:kf_predict_start})~to~(\ref{eq:kf_predict_stop}).
+    \STATE Predict observation mean and covariance $\mu_n^{(i)}$ and $S_n^{(i)}$ using (\ref{eq:kf_predict}).
     \STATE Calculate weight $w_n^{(i)}$ using (\ref{eq:vrpf_weights}).
-    \STATE Update state mean and covariance $m_n^{(i)}$ and $P_n^{(i)}$ using (\ref{eq:kf_update_start})~to~(\ref{eq:kf_update_stop}).
+    \STATE Update state mean and covariance $m_n^{(i)}$ and $P_n^{(i)}$ using (\ref{eq:kf_update}).
   \ENDFOR
@@
 
-The changepoint transition density may be expressed as,
-%
-\begin{IEEEeqnarray}{rCl}
- p(\theta_{n}^+|\theta_{n}^-) &=      & p(\tau_{K_n+1:K}, u_{K_n+1:K}|\tau_{K_n}, u_{K_n}, \tau_{K_n+1}>t_n) \nonumber \\
+When calculating the changepoint transition density of the future sequence, only the terms depending on the past sequence need to be considered, because the future sequence is fixed,
+%
+\begin{IEEEeqnarray}{rCl}
+ p(\theta_{n}^+|\theta_{n}^-) & =     & p(\tau_{K_n+1:K}, u_{K_n+1:K}|\tau_{K_n}, u_{K_n}, \tau_{K_n+1}>t_n) \nonumber \\
                               &\propto& p(\tau_{K_n+1}, u_{K_n+1}|\tau_{K_n}, u_{K_n}, \tau_{K_n+1}>t_n)     .
@@
 The integral in~(\ref{eq:rb-vrps_back_cond_weight}) contains two terms involving the current state, $x_n$.
-The first is the familiar state posterior generated by the Kalman filter using (\ref{eq:kf_predict_start})--(\ref{eq:kf_update_stop}),
+The first is the familiar state posterior generated by the Kalman filter using (\ref{eq:kf_predict}) and (\ref{eq:kf_update}),
 %
@@
 \begin{IEEEeqnarray}{rCl}
- \tilde{m}_n^- & = & A_{n+1}^{-1} \tilde{m}_{n+1} \label{eq:backward_kf_predict_start} \\
- \tilde{P}_n^- & = & A_{n+1}^{-1} (\tilde{P}_{n+1} + Q_{n+1}) A_{n+1}^{-T} \label{eq:backward_kf_predict_stop} \\
- \tilde{\mu}_n & = & C_n \tilde{m}_n^- \label{eq:backward_kf_update_start} \\
- \tilde{S}_n   & = & C_n \tilde{P}_n^- C_n^T + R_n \\
- \tilde{K}_n   & = & \tilde{P}_n^- C_n^T \tilde{S}_n^{-1} \\
- \tilde{m}_n   & = & \tilde{m}_n^- + \tilde{K}_n (y_n - \tilde{\mu}_n) \\
- \tilde{P}_n   & = & \tilde{P}_n^- - \tilde{K}_n \tilde{S}_n \tilde{K}_n^T \label{eq:backward_kf_update_stop}     .
+ \tilde{m}_n^- & = & A_{n+1}^{-1} \tilde{m}_{n+1} \nonumber \\
+ \tilde{P}_n^- & = & A_{n+1}^{-1} (\tilde{P}_{n+1} + Q_{n+1}) A_{n+1}^{-T} \label{eq:backward_kf_predict} \\
+ \nonumber \\
+ \tilde{\mu}_n & = & C_n \tilde{m}_n^- \nonumber \\
+ \tilde{S}_n   & = & C_n \tilde{P}_n^- C_n^T + R_n \nonumber \\
+ \tilde{G}_n   & = & \tilde{P}_n^- C_n^T \tilde{S}_n^{-1} \nonumber \\
+ \tilde{m}_n   & = & \tilde{m}_n^- + \tilde{G}_n (y_n - \tilde{\mu}_n) \nonumber \\
+ \tilde{P}_n   & = & \tilde{P}_n^- - \tilde{G}_n \tilde{S}_n \tilde{G}_n^T \label{eq:backward_kf_update}     .
 \end{IEEEeqnarray}
@@
       \STATE Discard $\theta_{n}^{-(i)}$.
-      \STATE Predict state mean and covariance $\tilde{m}_n^{-(i)}$ and $\tilde{P}_n^{-(i)}$ using (\ref{eq:backward_kf_predict_start}) to (\ref{eq:backward_kf_predict_stop}).
-      \FOR{$j=1 \dots N_P$}
+      \STATE Predict state mean and covariance $\tilde{m}_n^{-(i)}$ and $\tilde{P}_n^{-(i)}$ using (\ref{eq:backward_kf_predict}).
+      \FOR{$j=1 \dots N_F$}
 	      \STATE Calculate weight $\tilde{w}_n^{(j)}$ using (\ref{eq:rb-vrps_back_cond_weight2}).
@@
       \STATE Join $\theta^{(i)} \gets \theta_{n}^{-(i)} \cup \theta_{n}^{+(i)}$.
-      \STATE Update state mean and covariance $\tilde{m}_n^{(i)}$ and $\tilde{P}_n^{(i)}$ using (\ref{eq:backward_kf_update_start}) to (\ref{eq:backward_kf_update_stop}).
+      \STATE Update state mean and covariance $\tilde{m}_n^{(i)}$ and $\tilde{P}_n^{(i)}$ using (\ref{eq:backward_kf_update}).
     \ENDFOR
@@
 The changepoint state values, $x_{0:K}$ may be calculated deterministically from $\theta$ for each particle.
-The smoother is then formulated by factorising the augmented sequence posterior distribution as before,
+The smoother is then formulated by factorising the augmented sequence posterior distribution,
 %
@@
 Generally, this requires there to be as many degrees of freedom in the selection of $u_k$ as there are state dimensions of $x_k$.
+This condition is satisfied by the tracking model of section~\ref{sec:tracking_model} only through the introduction of the linear drift parameters.
 
@@
 This value is itself a particle from the target distribution (albeit one from an approximation with poor path-space diversity), meaning that no burn-in period is required.
+However, longer chains are likely to result in less dependence between samples and a reduction in Monte Carlo error.
 This is the same reasoning which justifies the absence of a burn-in when using a resample-move scheme \cite{Gilks2001}.
@@
 
-The Rao-Blackwellised VRPS algorithm was tested on the financial time series model of \cite{Christensen2012}, in which prices of an asset are treated as noisy observations of a latent state, which evolves according to a drift-diffusion with occasional jumps.
-
-The latent state is a vector with two elements, the underlying value of the asset, and the trend followed by this value,
-%
-\begin{equation}
- \mathbf{x}(t) = [ x(t), \dot{x}(t)]^T     .
-\end{equation}
-
-This evolves continuously according to a jump-diffusion model,
-%
-\begin{IEEEeqnarray}{c}
- d\mathbf{x}(t) = \begin{bmatrix}0 & 1 \\ 0 & -\lambda \end{bmatrix} \mathbf{x}(t) dt + \begin{bmatrix}0 \\ \sigma \end{bmatrix} d\beta(t) + d\mathbf{J}(t)
-\end{IEEEeqnarray}
-
-where $\lambda$ introduces a mean reversion effect on the trend and $\mathbf{\beta}(t)$ is standard Brownian motion (with unit diffusion constant).
-The jump term, $d\mathbf{J}(t)$, is zero everywhere except when jumps occur,
-%
-\begin{IEEEeqnarray}{rCl}
- d\mathbf{J}(t) & = & \begin{cases} \mathbf{J}_k & t \in \{\tau_k\} \\ 0 & \text{elsewhere} \end{cases} \\
- \mathbf{J}_k  & \sim & \mathcal{N}(\mathbf{J}_k| \mathbf{0}, Q_{J,u_k})     .
-\end{IEEEeqnarray}
-
-$\{\tau_k\}$ is the set of random jump times, each of which is one of two types: a value jump, indicated by $u_k = 1$, or a trend jump, indicated by $u_k=2$. The jump covariance matrices are,
-%
-\begin{IEEEeqnarray}{c}
-Q_{J,u_k} = \begin{cases} \begin{bmatrix}\sigma_{J1}^2 & 0 \\ 0 & 0 \end{bmatrix} & u_k = 1 \\
-                          \begin{bmatrix}0 & 0 \\ 0 & \sigma_{J2}^2 \end{bmatrix} & u_k = 2  \end{cases}   .
-\end{IEEEeqnarray}
-
-Only the value is observed, and that corrupted by additive Gaussian noise with standard deviation $\sigma_y^2$.
-
-This model may be discretised at the observation times using standard methods, detailed in appendix~\ref{app:lg_model_discretisation}, leading to,% (see \cite{Godsill2007a,Christensen2012,Grewal2002}), leading to,
-%
-%Assuming observations of value only and Gaussian observation noise with standard deviation $\sigma_y^2$, the resulting discrete time dynamics are described by the following equations:% (see appendix \ref{app:model_discretisation}):
-%
-\begin{IEEEeqnarray}{rCl}
- \mathbf{x}_n &=& A \mathbf{x}_{n-1} + \mathbf{w}_n \\
- y_n          &=& C \mathbf{x}_{n} + v_n
-\end{IEEEeqnarray}
-
-where the $\mathbf{w}_n$ and $v_n$ are Gaussian random variables with covariance matrixes $Q_n$ and $R$ respectively.
-$Q_n$ is a function of the sequence of jump times and their types, giving rise to the desired conditionally linear-Gaussian structure.% The time between observations times is denotes $\Delta t = t_{n} - t_{n-1}$.
-%
-%\begin{IEEEeqnarray}{rCl}
-% A     &=& \begin{bmatrix}1 & \frac{1}{\lambda}(1-e^{(-\lambda \Delta t)} \\ 0 & e^{(-\lambda \Delta t)}\end{bmatrix} \\
-% C     &=& \begin{bmatrix}1 & 0\end{bmatrix}
-%\end{IEEEeqnarray}
-%\begin{IEEEeqnarray}{rCl}
-% Q_n   &=& \begin{cases}Q_{D} + Q_{J,u_k} & \exists k : \tau_k \in [t_{n-1},t_n]\\
-%                        Q_{D}             & \text{otherwise} \end{cases} \\
-% Q_{D} &=& \frac{\sigma^2}{2 \lambda}\begin{bmatrix} q_1 & q_2 \\ q_2 & q_3\end{bmatrix} \\
-% q_1   &=& \frac{1}{\lambda^2}(2 \lambda \Delta t - (3 - e^{(-\lambda \Delta t)})(1 - e^{(-\lambda \Delta t)})~) \\
-% q_2   &=& \frac{1}{\lambda} (1-e^{(-\lambda \Delta t)})^2 \\
-% q_3   &=& 1-e^{(-2 \lambda \Delta t)} \\
-% R     &=& [\sigma_y^2]
-%\end{IEEEeqnarray}
-
+The Rao-Blackwellised variable rate algorithms were tested on the jump-diffusion model outlined in section~\ref{sec:financial_model} using artificial simulated data.
 The times between changepoints are assumed to be exponentially distributed with mean $1/\alpha$, and equal probability of value and trend jumps.
-%
-%\begin{IEEEeqnarray}{rCl}
-% \IEEEeqnarraymulticol{3}{l}{p(\tau_k, u_k|\tau_{k-1}, u_{k-1}) = P(u_k) p(\tau_k|\tau_{k-1})} \\
-% p(\tau_k|\tau_{k-1})               &=& \begin{cases}\frac{1}{\alpha} \exp ( \alpha (\tau_k-\tau_{k-1}) ) & \tau_k>\tau_{k-1} \\
-%                                                     0 & \tau_k < \tau_{k-1} \end{cases} \\
-% P(u_k)                             &=& \begin{cases}0.5 & u_k = 1 \\ 0.5 & u_k = 2\end{cases}
-%\end{IEEEeqnarray}
-
-The algorithms were tested on artificial data simulated from this model.
 Observations were made on a regular time grid with a spacing of $0.0017s$.
 Other parameters used were: $\alpha = 20$, $\lambda = 5$, $\sigma = 0.05$, $\sigma_{J1} = 0.005$, $\sigma_{J2} = 0.05$, $\sigma_{y} = 0.001$.
-
-% The decay constant for the distribution of inter-changepoint times is 20, and the trend mean-reversion coefficient is 5.
-The diffusion constant is 0.05 and the observation noise standard deviation 0.001.
-The jump standard deviations are 0.005 and 0.05 respectively for value and trend.
-(Terms defined in appendix~\ref{app:lg_model_discretisation})
-
-%The following parameters were used: $\Delta t = 0.0017$, $N = 1000$, $\alpha = 20$, $\lambda = 5$, $\sigma = 0.05$, $\sigma_{J1} = 0.005$, $\sigma_{J2} = 0.05$, $\sigma_{y} = 0.001$.
 
@@
 \end{figure}
-
-%For a first, qualitative comparison of the changepoint sequence distributions, it is possible to generate kernel density estimates for the changepoint times.
-These are constructed by simply adding a (unit amplitude) Gaussian kernel for each changepoint present in each particle and scaling by the number of particles.
-The resulting function provides an approximate measure of the likelihood of finding a jump at a given time.
-These are shown in figure~\ref{fig:example_kdest} for the example data set.
-
-%\begin{figure}[!t]
-%\centering
-%\subfloat[]{\includegraphics[width=0.95\columnwidth]{example_filter_kdest.pdf}} \\
-%\subfloat[]{\includegraphics[width=0.95\columnwidth]{example_smoother_kdest.pdf}}
-%\caption{Filter (a) and smoother (b) kernel density estimates for value (top) and trend (bottom) jump times. Correct times overlayed as dotted vertical lines.}
-%\label{fig:example_kdest}
-%\end{figure}
-
-%It is immediately apparent that the approximation produced by the filter lacks diversity.
-The jump time kernel density estimate comprises peaks with magnitude $1$, with regions of $0$ between.
-This is a result of every particle containing an almost identical list of changepoints, as a result of resampling.
-In contrast, the smoother is clearly providing a less degenerate particle representation, and generally contains larger kernel density peaks at the times of larger jumps.
-Furthermore, the kernel density for the trend jump times contains broader peaks, capturing the fact that because the trend is not observed directly, the jumps are harder to localise precisely.
-
-%A further comparison of the algorithms is possible via the state estimates they generate.
-Here we can compare three options: the filtering results, using the VRPF and a Kalman filter for the state estimates; the filter-smoother results, using the final VRPF approximation for the changepoint sequence and a Rauch-Tung-Striebel (RTS) smoother for the state estimates; and the full smoothing results, using the VRPS followed by an RTS smoother.
-For the example data set, the trend estimates are shown in figure~\ref{fig:example_state} (The estimates of value are less informative, as this quantity is observed).
-Again, the improved particle diversity of the smoother is apparent.
-
-%\begin{figure}[!t]
-%\centering
-%\subfloat[]{\includegraphics[width=0.95\columnwidth]{example_filter_state.pdf}} \\
-%\subfloat[]{\includegraphics[width=0.95\columnwidth]{example_filtersmoother_state.pdf}} \\
-%\subfloat[]{\includegraphics[width=0.95\columnwidth]{example_smoother_state.pdf}}
-%\caption{Trend estimates using RBVRPF + Kalman filter (a) RBVRPF + RTS smoother (b) and RBVRPS + RTS smoother.
-Solid lines show means of each particle.
-Dashed lines show mean $\pm$2 standard deviations.}
-%\label{fig:example_state}
-%\end{figure}
-
-%For a quantitative comparison, the algorithms were tested on $10$ realisations from the model, each of $1000$ time steps, and the following statistics were calculated for each:
-%\begin{itemize}
-%	\item The number of unique changepoint sequences. This is a measure of particle diversity of the approximation.
-%	\item The number of unique changepoint times. Another measure of particle diversity.
-%	%\item The optimal sub-pattern assignment (OSPA) distance between the maximum a posteriori (MAP) changepoint sequence and the true sequence.
-The OSPA is a concept from multiple target tracking introduced in \cite{Schuhmacher2008} to measure the distance between sets of points of varying cardinality.
-It is thus well suited to measuring the error between changepoint sequences of varying length.
-For each approximation, the sequence from the MAP particle was used as a point estimate.
-%	%\item The root-mean-square error (RMSE) of the MAP state estimate.
-The MAP state estimate is taken to be the Gaussian mean of the conditional state distribution, from the Kalman smoother or RTS, selected from the particle with the highest posterior probability.
-%	\item The root-mean-square error (RMSE) of the mean state estimate. The mean state estimate is the average of the Gaussian means from all the particles.
-%\end{itemize}
-
-%Note that the MAP particle may be found by calculating the posterior probability of each. The terms required for this calculation will already have been evaluated during the filtering procedure.
-
-%\begin{IEEEeqnarray}{rCl}
-% p(\theta|y_{1:N}) & \propto & p(y_{1:N}|\theta) p(\theta) \nonumber \\
-%                         &         & S(\tau_K, T) \prod_{k=1}^K p(u_k) p(\tau_k|\tau_{k-1}) \prod_{n=1}^{N} p(y_n|\theta, y_{1:n-1})
-%\end{IEEEeqnarray}
 
@@
 
-%The results are shown in tables~\ref{tab:finance_cp_performance} and ~\ref{tab:finance_state_performance}.% The OSPA parameters are: exponent $p=1$, threshold $c=0.01$.
-
-%\begin{table}%
-%\caption{Changepoint sequence estimation performance.}
-%\label{tab:finance_cp_performance}
-%\centering
-%\renewcommand{\arraystretch}{1.5}
-%\begin{tabular}{|c|c|c|}
-%\hline
-% & RBVRPF & RBVRPS \\
-%\hline \hline
-%mean unique no. sequences   & 7.8      & 100       \\
-%\hline
-%mean unique no. jump times  & 47.8     & 1076.7    \\
-%%MAP sequence OSPA    & $5.36 \times 10^{-3}$ & $4.98 \times 10^{-3}$  \\
-%\hline
-%\end{tabular}
-%\end{table}
-
-%\begin{table}%
-%\caption{State estimation performance.}
-%\label{tab:finance_state_performance}
-%\centering
-%\renewcommand{\arraystretch}{1.5}
-%\begin{tabular}{|c|c|c|c|}
-%\hline
-% &\raisebox{0cm}[0.4cm][0.3cm]{\parbox[c]{1.5cm}{\centering RBVRPF and KF}} & \raisebox{0cm}[0.4cm][0.3cm]{\parbox[c]{1.5cm}{\centering RBVRPF and RTS}} & \raisebox{0cm}[0.4cm][0.3cm]{\parbox[c]{1.5cm}{\centering RBVRPS and RTS}} \\
-%% & \begin{minipage}[c]{1.5cm}\centering RBVRPF + KF\end{minipage} & \begin{minipage}[c]{1.5cm}\centering RBVRPF + RTS\end{minipage} & \begin{minipage}[c]{1.5cm}\centering RBVRPS + RTS\end{minipage} \\
-%\hline \hline
-%\raisebox{0cm}[0.4cm][0.3cm]{\parbox[c]{2cm}{\centering mean value estimate RMSE}}   & $5.31 \times 10^{-4}$ & $4.48 \times 10^{-4}$ & $4.16 \times 10^{-4}$ \\
-%\hline
-%\raisebox{0cm}[0.4cm][0.3cm]{\parbox[c]{2cm}{\centering mean trend estimate RMSE}}   & $2.56 \times 10^{-2}$ & $1.79 \times 10^{-2}$ & $1.49 \times 10^{-2}$ \\
-%%MAP value estimate RMSE    & $5.98 \times 10^{-4}$ & $4.48 \times 10^{-4}$ & $4.49 \times 10^{-4}$ \\
-%%MAP trend estimate RMSE    & $3.13 \times 10^{-2}$ & $1.79 \times 10^{-2}$ & $1.74 \times 10^{-2}$ \\
-%\hline
-%\end{tabular}
-%\end{table}
-
-%The new RBVRPS algorithm outperforms the filter in terms of both state estimation accuracy and particle diversity.
-
-%\begin{meta}
-%Another interesting question is how the posterior distribution over the number of changepoints is expected to behave.
-The prior is poisson distributed with mean $\alpha T$.
-The means of both the filter and smoother particle posterior are consistently greater than the true number (by about 50\%).
-Is this expected and why?
-%\end{meta}
-
-%Finally, we demonstrate the algorithms on a real financial data set, 1000 data points representing 1.7s of USD-GBP foreign exchange rate data from April 2008, using the same parameters as for the previous simulations.
-The data shown in Fig.~\ref{fig:fx_results}, along with jump time kernel density estimates from the filter and smoother.
-In this example, there is no ground truth against which to judge the results.
-However, the smoother does a pleasing job of estimating jump times in the same areas that one would if analysing the data ``by eye''.
-We submit this simply to illustrate the practical applications of the model and algorithm.
-
-%\begin{figure}[!t]
-%\centering
-%\subfloat[]{\includegraphics[width=0.95\columnwidth]{fx_data.pdf}} \\
-%\subfloat[]{\includegraphics[width=0.95\columnwidth]{fx_filter_kdest.pdf}} \\
-%\subfloat[]{\includegraphics[width=0.95\columnwidth]{fx_smoother_kdest.pdf}}
-%\caption{Foreign exchange data (a), along with filter (b) and smoother (c) kernel density estimates for the state and trend jump times.}
-%\label{fig:fx_results}
-%\end{figure}
-
 The mean number of unique particle estimates at each time step is shown in Fig.~\ref{fig:finance_unique_particles}.
@@
 
-The piecewise-deterministic VRPS was tested using a 2-dimensional dynamic model for manoeuvring vehicles \cite{Bunch2012a}.
-The target is modelled as a particle subject to perpendicular forces tangential and normal to the direction of motion.
-These forces are held constant between changepoint times, $\tau_{1:K}$, which represent the beginnings and ends of manoeuvres.
-
-The state vector consists of Cartesian position, $x(t)$ and $y(t)$, plus heading angle, $\psi(t)$, and speed, $\dot{s}(t)$,
-%
-\begin{equation}
-\mathbf{x}(t) = [x(t), y(t), \psi(t), \dot{s}(t)]^T     .
-\end{equation}
-
-In addition to the tangential and normal accelerations, $a_{T,k}$ and $a_{N,k}$, we introduce two linear velocity terms, $d_{X,k}$ and $d_{Y,k}$.
-Without these, the model could not be used for smoothing, because for an arbitrary pair of adjacent changepoints ($\tau_{k}$, $\mathbf{x}_{k}$) and ($\tau_{k+1}$, $\mathbf{x}_{k+1}$), there would not, in general, be a pair of accelerations which achieved the required transition.
-These linear velocity terms may be considered to be merely relaxation terms, allowing arbitrary past and future trajectories to be joined, or they may represent real physical effects, such as wind (for aircraft) or currents (for boats).
-Together, these four variables make up the vector of motion parameters,
-%
-\begin{equation}
-\mathbf{u}_k = [a_{T,k}, a_{N,k}, d_{X,k}, d_{Y,k}]^T     .
-\end{equation}
-
-The target dynamics are described by four differential equations,
-%
-\begin{IEEEeqnarray}{rCl}
-\ddot{s}_t & = & a_{T,K(t)} \\
-\dot{s}_t \dot{\psi}_t & = & a_{N,K(t)} \\
-\dot{x}_t & = & \dot{s}_t \cos(\psi_t) + d_{X,K(t)} \\
-\dot{y}_t & = & \dot{s}_t \sin(\psi_t) + d_{Y,K(t)}     .
-\end{IEEEeqnarray}
-
-Solving these yields the following state equations (where $\Delta\ t = t - \tau_{K(t)}$),
-%The target dynamics are governed by the following state equations \cite{Bunch2012a} (where $\Delta\ t = t - \tau_{K(t)}$):
-%
-\begin{IEEEeqnarray}{rCl}
-\dot{s}(t) & = & \dot{s}_{K(t)} + a_{T,k} \Delta t \label{eq:2D_ICmodel_start} \\
-\psi(t) & = & \psi_{K(t)} + \frac{a_{N,k}}{a_{T,k}} \log \left( \frac{\dot{s}(t)}{\dot{s}_{K(t)}} \right)
-\end{IEEEeqnarray}
-\begin{IEEEeqnarray}{rCl}
-x(t) & = & x_{K(t)} + d_{X,K(t)} \Delta t \\
-     \IEEEeqnarraymulticol{3}{l}{ \quad + \: \frac{ \dot{s}(t)^2 }{ 4 a_{T,k}^2 + a_{N,k}^2 } \left[  a_{N,k} \sin(\psi(t)) + 2 a_{T,k} \cos(\psi(t))  \right]} \nonumber \\
-     \IEEEeqnarraymulticol{3}{l}{ \quad - \: \frac{\dot{s}_{K(t)}^2}{4 a_{T,k}^2 + a_{N,k}^2} \left[  a_{N,k} \sin(\psi_{K(t)}) + 2 a_{T,k} \cos(\psi_{K(t)})  \right]} \nonumber
-\end{IEEEeqnarray}
-\begin{IEEEeqnarray}{rCl}
-y(t) & = & y_{K(t)} + d_{Y,K(t)} \Delta t \\
-     \IEEEeqnarraymulticol{3}{l}{ \quad + \: \frac{ \dot{s}(t)^2 }{ 4 a_{T,k}^2 + a_{N,k}^2 } \left[ -a_{N,k} \cos(\psi(t)) + 2 a_{T,k} \sin(\psi(t))  \right]} \nonumber \\
-     \IEEEeqnarraymulticol{3}{l}{ \quad - \: \frac{\dot{s}_{K(t)}^2}{4 a_{T,k}^2 + a_{N,k}^2} \left[  -a_{N,k} \cos(\psi_{K(t)}) + 2 a_{T,k} \sin(\psi_{K(t)})  \right]} \nonumber      .
-\end{IEEEeqnarray}
-
+The piecewise-deterministic variable rate algorithms were tested on the intrinsic coordinate tracking model outlined in section~\ref{sec:tracking_model} using artificial simulated data.
 The changepoint motion parameters are assumed to be independent and zero-mean Gaussian distributed.
@@
 \begin{equation}
- p(y_{n+1:N}|x_{n+1}, \tilde{\theta}_{n}^+) = \tilde{Z}_{n+1} \mathcal{N}(x_{n+1}|\tilde{m}_{n+1}, \tilde{P}_{n+1})  .
+ p(y_{n+1:N}|x_{n+1}, \tilde{\theta}_{n}^+) = \tilde{Z}_{n+1} \mathcal{N}(x_{n+1}|\tilde{m}_{n+1}, \tilde{P}_{n+1}) \nonumber    .
 \end{equation}
@@
   &   & \qquad \mathcal{N}(x_{n}|A_{n+1}^{-1} x_{n+1}, A_{n+1}^{-1} Q_{n+1} A_{n+1}^{-T}) dx_{n+1} \nonumber \\
-  & = & \tilde{Z}_{n}^- \mathcal{N}(x_n|\tilde{m}_n^-, \tilde{P}_n^-)
+  & = & \tilde{Z}_{n}^- \mathcal{N}(x_n|\tilde{m}_n^-, \tilde{P}_n^-) \nonumber
 \end{IEEEeqnarray}
@@
 \begin{IEEEeqnarray}{rCl}
- \tilde{m}_n^- & = & A_{n+1}^{-1} \tilde{m}_{n+1} \\
- \tilde{P}_n^- & = & A_{n+1}^{-1} (\tilde{P}_{n+1} + Q_{n+1}) A_{n+1}^{-T} \\
- \tilde{Z}_n^- & = & \tilde{Z}_{n+1} ||A_{n+1}||^{-1}   .
+ \tilde{m}_n^- & = & A_{n+1}^{-1} \tilde{m}_{n+1} \nonumber \\
+ \tilde{P}_n^- & = & A_{n+1}^{-1} (\tilde{P}_{n+1} + Q_{n+1}) A_{n+1}^{-T} \nonumber \\
+ \tilde{Z}_n^- & = & \tilde{Z}_{n+1} ||A_{n+1}||^{-1} \nonumber   .
 \end{IEEEeqnarray}
@@
 \begin{IEEEeqnarray}{rCl}
-\IEEEeqnarraymulticol{3}{l}{p(y_{n:N}|x_n, \tilde{\theta}_{n}^+)} \\
+\IEEEeqnarraymulticol{3}{l}{p(y_{n:N}|x_n, \tilde{\theta}_{n}^+)} \nonumber \\
  & = & p(y_{n+1:N}|x_n, \tilde{\theta}_{n}^+) p(y_{n}|x_n, \tilde{\theta}_{n}^+) \nonumber \\
  & = & \tilde{Z}_n^- \mathcal{N}(x_n|\tilde{m}_n^-, \tilde{P}_n^-) \mathcal{N}(y_n|C_n x_n, R_n) \nonumber \\
- & = & \tilde{Z}_n \mathcal{N}(x_n|\tilde{m}_n, \tilde{P}_n)     ,
+ & = & \tilde{Z}_n \mathcal{N}(x_n|\tilde{m}_n, \tilde{P}_n) \nonumber     ,
 \end{IEEEeqnarray}
@@
 \begin{IEEEeqnarray}{rCl}
- \tilde{\mu}_n & = & C_n \tilde{m}_n^- \\
- \tilde{S}_n   & = & C_n \tilde{P}_n^- C_n^T + R_n \\
- \tilde{K}_n   & = & \tilde{P}_n^- C_n^T \tilde{S}_n^{-1} \\
- \tilde{m}_n   & = & \tilde{m}_n^- + \tilde{K}_n (y_n - \tilde{\mu}_n) \\
- \tilde{P}_n   & = & \tilde{P}_n^- - \tilde{K}_n \tilde{S}_n \tilde{K}_n^T \\
- \tilde{Z}_n   & = & \tilde{Z}_n^- \mathcal{N}(y_n|\tilde{\mu}_n, \tilde{S}_n)   .
+ \tilde{\mu}_n & = & C_n \tilde{m}_n^-  \nonumber \\
+ \tilde{S}_n   & = & C_n \tilde{P}_n^- C_n^T + R_n  \nonumber \\
+ \tilde{G}_n   & = & \tilde{P}_n^- C_n^T \tilde{S}_n^{-1} \nonumber \\
+ \tilde{m}_n   & = & \tilde{m}_n^- + \tilde{G}_n (y_n - \tilde{\mu}_n) \nonumber \\
+ \tilde{P}_n   & = & \tilde{P}_n^- - \tilde{G}_n \tilde{S}_n \tilde{G}_n^T \nonumber \\
+ \tilde{Z}_n   & = & \tilde{Z}_n^- \mathcal{N}(y_n|\tilde{\mu}_n, \tilde{S}_n) \nonumber   .
 \end{IEEEeqnarray}
@@
 %
+\begin{IEEEeqnarray}{rCl}
+ \IEEEeqnarraymulticol{3}{l}{p(y_{N-L:N}|x_{N-L}) = \mathcal{N}( \mathbf{y}_{N-L:N} | H_{N-L} x_{N-L}, \Gamma_{N-L} ) } \nonumber \\
+ & = & \sqrt{\frac{|V_{N-L}|}{|\Gamma_{N-L}|}} \mathcal{N}(x_{N-L} | \nu_{N-L}, V_{N-L} ) \nonumber
+\end{IEEEeqnarray}
+\begin{IEEEeqnarray}{rCl}
+ V_{N-L} & = & (H_{N-L}^T \Gamma_{N-L}^{-1} H_{N-L})^{-1} \nonumber \\
+ \nu_{N-L} & = & V_{N-L} H_{N-L}^T \Gamma_{N-L}^{-1} \mathbf{y}_{N-L:N} \nonumber     ,
+\end{IEEEeqnarray}
+
+where $\mathbf{y}_{N-L:N}$ is a vector of concatenated observations, and $H_{N-L}$ and $\Gamma_{N-L}$ may be constructed from the system models,
+%
+\begin{IEEEeqnarray}{rCl}
+\mathbf{y}_{N-L:N} & = & \begin{bmatrix}y_N \\ y_{N-1} \\ \vdots \\ y_{N-L} \end{bmatrix} \nonumber \\
+%\end{IEEEeqnarray}
 %\begin{IEEEeqnarray}{rCl}
-% \IEEEeqnarraymulticol{3}{l}{p(y_{N-L:N}|x_{N-L}) = \mathcal{N}( \mathbf{y}_{N-L:N} | H_{N-L} x_{N-L}, \Gamma_{N-L} ) } \nonumber \\
-% & = & \sqrt{\frac{|(H_{N-L}^T \Gamma_{N-L}^{-1} H_{N-L})^{-1}|}{|\Gamma_{N-L}|}} \nonumber \\
-% & & \times \mathcal{N}(x_{N-L} | (H_{N-L}^T \Gamma_{N-L}^{-1} H_{N-L})^{-1} H_{N-L}^T \Gamma_{N-L}^{-1} \mathbf{y}_{N-L:N}, \nonumber \\
-% & & (H_{N-L}^T \Gamma_{N-L}^{-1} H_{N-L})^{-1} )      ,
+H_{N-L} & = & \begin{bmatrix} C A^L \\ C A^{L-1} \\ \vdots \\ C \end{bmatrix} \nonumber \\
 %\end{IEEEeqnarray}
-\begin{IEEEeqnarray}{rCl}
- \IEEEeqnarraymulticol{3}{l}{p(y_{N-L:N}|x_{N-L}) = \mathcal{N}( \mathbf{y}_{N-L:N} | H_{N-L} x_{N-L}, \Gamma_{N-L} ) } \nonumber \\
- & = & \sqrt{\frac{|V_{N-L}|}{|\Gamma_{N-L}|}} \mathcal{N}(x_{N-L} | \nu_{N-L}, V_{N-L} )
-\end{IEEEeqnarray}
-\begin{IEEEeqnarray}{rCl}
- V_{N-L} & = & (H_{N-L}^T \Gamma_{N-L}^{-1} H_{N-L})^{-1} \\
- \nu_{N-L} & = & V_{N-L} H_{N-L}^T \Gamma_{N-L}^{-1} \mathbf{y}_{N-L:N}     ,
-\end{IEEEeqnarray}
-
-where $\mathbf{y}_{N-L:N}$ is a vector of concatenated observations, and $H_{N-L}$ and $\Gamma_{N-L}$ may be constructed from the system models,
-%
-%\begin{IEEEeqnarray}{rClCrCl}
-%\mathbf{y}_{N-L:N} & = & \begin{bmatrix}y_N \\ y_{N-1} \\ \vdots \\ y_{N-L} \end{bmatrix} & , \qquad & H_{N-L} & = & \begin{bmatrix} C A^L \\ C A^{L-1} \\ \vdots \\ C \end{bmatrix}
+%\begin{IEEEeqnarray}{rCl}
+ \Gamma_{N-1} & = & K_{N-L} \begin{bmatrix} Q_N & 0 & \dots & 0 \\ 0 & Q_{N-1} & \dots & 0 \\ \vdots & \vdots & \ddots & \vdots \\ 0 & 0 & \dots & Q_{N-L+1} \end{bmatrix} K_{N-L}^T \nonumber \\
+  & & + \: \begin{bmatrix} R_N & 0 & \dots & 0 \\ 0 & R_{N-1} & \dots & 0 \\ \vdots & \vdots & \ddots & \vdots \\ 0 & 0 & \dots & R_{N-L} \end{bmatrix} \nonumber \\
 %\end{IEEEeqnarray}
-\begin{IEEEeqnarray}{rCl}
-\mathbf{y}_{N-L:N} & = & \begin{bmatrix}y_N \\ y_{N-1} \\ \vdots \\ y_{N-L} \end{bmatrix}
-\end{IEEEeqnarray}
-\begin{IEEEeqnarray}{rCl}
-H_{N-L} & = & \begin{bmatrix} C A^L \\ C A^{L-1} \\ \vdots \\ C \end{bmatrix}
-\end{IEEEeqnarray}
-\begin{IEEEeqnarray}{rCl}
- \Gamma_{N-1} & = & K_{N-L} \begin{bmatrix} Q_N & 0 & \dots & 0 \\ 0 & Q_{N-1} & \dots & 0 \\ \vdots & \vdots & \ddots & \vdots \\ 0 & 0 & \dots & Q_{N-L+1} \end{bmatrix} K_{N-L}^T \nonumber \\
-  & & + \: \begin{bmatrix} R_N & 0 & \dots & 0 \\ 0 & R_{N-1} & \dots & 0 \\ \vdots & \vdots & \ddots & \vdots \\ 0 & 0 & \dots & R_{N-L} \end{bmatrix}
-\end{IEEEeqnarray}
-\begin{IEEEeqnarray}{rCl}
- K_{N-L} & = & \begin{bmatrix} C & CA & \dots & CA^{L-1} \\ 0 & C & \dots & CA^{L-2} \\ \vdots & \vdots & \ddots & \vdots \\ 0 & 0 & \dots & C \\ 0 & 0 & \dots & 0 \end{bmatrix}
+%\begin{IEEEeqnarray}{rCl}
+ K_{N-L} & = & \begin{bmatrix} C & CA & \dots & CA^{L-1} \\ 0 & C & \dots & CA^{L-2} \\ \vdots & \vdots & \ddots & \vdots \\ 0 & 0 & \dots & C \\ 0 & 0 & \dots & 0 \end{bmatrix} \nonumber
 \end{IEEEeqnarray}
@@
 \begin{IEEEeqnarray}{c}
- d\mathbf{x}(t) = \begin{bmatrix}0 & 1 \\ 0 & -\lambda \end{bmatrix} \mathbf{x}(t) dt + \begin{bmatrix}0 \\ \sigma \end{bmatrix} d\beta(t) + d\mathbf{J}(t)
-\end{IEEEeqnarray}
-
-where $\lambda$ is the mean reversion constant, $\sigma$ is the diffusion constant, $\mathbf{\beta}(t)$ is standard Brownian motion, $d\mathbf{J}(t)$ is a jump term defined according to,
-%
-\begin{IEEEeqnarray}{rCl}
- d\mathbf{J}(t) & = & \begin{cases} \mathbf{J}_k & t \in \{\tau_k\} \\ 0 & \text{elsewhere} \end{cases} \\
- \mathbf{J}_k  & \sim & \mathcal{N}(\mathbf{J}_k| \mathbf{0}, Q_{J,u_k})    .
+ dx(t) = \begin{bmatrix}0 & 1 \\ 0 & -\lambda \end{bmatrix} x(t) dt + \begin{bmatrix}0 \\ \sigma \end{bmatrix} d\beta(t) + dJ(t) \nonumber
+\end{IEEEeqnarray}
+
+where $\lambda$ is the mean reversion constant, $\sigma$ is the diffusion constant, $\beta(t)$ is standard Brownian motion, $dJ(t)$ is a jump term defined according to,
+%
+\begin{IEEEeqnarray}{rCl}
+ dJ(t) & = & \begin{cases} J_k & t \in \{\tau_k\} \\ 0 & \text{elsewhere} \end{cases} \nonumber \\
+ J_k  & \sim & \mathcal{N}(J_k| 0, Q_{J,u_k}) \nonumber    .
 \end{IEEEeqnarray}
@@
 Q_{J,u_k} = \begin{cases} \begin{bmatrix}\sigma_{J1}^2 & 0 \\ 0 & 0 \end{bmatrix} & u_k = 1 \\
-                          \begin{bmatrix}0 & 0 \\ 0 & \sigma_{J2}^2 \end{bmatrix} & u_k = 2  \end{cases}   .
+                          \begin{bmatrix}0 & 0 \\ 0 & \sigma_{J2}^2 \end{bmatrix} & u_k = 2  \end{cases} \nonumber    .
 \end{IEEEeqnarray}
@@
 \begin{IEEEeqnarray}{rCl}
- \mathbf{x}_n &=& A \mathbf{x}_{n-1} + \mathbf{w}_n
-\end{IEEEeqnarray}
-
-where the $\mathbf{w}_n$ is a Gaussian random variable with covariance matrix $Q_n$. The inter-observation time is denoted $\Delta t = t_{n} - t_{n-1}$, and,
-
-\begin{IEEEeqnarray}{rCl}
- A     &=& \begin{bmatrix}1 & \frac{1}{\lambda}(1-e^{(-\lambda \Delta t)} \\ 0 & e^{(-\lambda \Delta t)}\end{bmatrix} \\
+ x_n &=& A x_{n-1} + w_n \nonumber
+\end{IEEEeqnarray}
+
+where the $w_n$ is a Gaussian random variable with covariance matrix $Q_n$. The inter-observation time is denoted $\Delta t = t_{n} - t_{n-1}$, and,
+%
+\begin{IEEEeqnarray}{rCl}
+ A     &=& \begin{bmatrix}1 & \frac{1}{\lambda}(1-e^{(-\lambda \Delta t)} \\ 0 & e^{(-\lambda \Delta t)}\end{bmatrix} \nonumber \\
  Q_n   &=& \begin{cases}Q_{D} + Q_{J,u_k} & \exists k : \tau_k \in [t_{n-1},t_n]\\
-                        Q_{D}             & \text{otherwise} \end{cases} \\
- Q_{D} &=& \frac{\sigma^2}{2 \lambda}\begin{bmatrix} q_1 & q_2 \\ q_2 & q_3\end{bmatrix} \\
- q_1   &=& \frac{1}{\lambda^2}(2 \lambda \Delta t - (3 - e^{(-\lambda \Delta t)})(1 - e^{(-\lambda \Delta t)})~) \\
- q_2   &=& \frac{1}{\lambda} (1-e^{(-\lambda \Delta t)})^2 \\
- q_3   &=& 1-e^{(-2 \lambda \Delta t)}     .
+                        Q_{D}             & \text{otherwise} \end{cases} \nonumber \\
+ Q_{D} &=& \frac{\sigma^2}{2 \lambda}\begin{bmatrix} q_1 & q_2 \\ q_2 & q_3\end{bmatrix} \nonumber \\
+ q_1   &=& \frac{1}{\lambda^2}(2 \lambda \Delta t - (3 - e^{(-\lambda \Delta t)})(1 - e^{(-\lambda \Delta t)})~) \nonumber \\
+ q_2   &=& \frac{1}{\lambda} (1-e^{(-\lambda \Delta t)})^2 \nonumber \\
+ q_3   &=& 1-e^{(-2 \lambda \Delta t)} \nonumber     .
 \end{IEEEeqnarray}
@@
 \begin{IEEEeqnarray}{rCl}
- y_n &=& C \mathbf{x}_{n} + v_n
+ y_n &=& C x_{n} + v_n \nonumber
 \end{IEEEeqnarray}
@@
 \begin{IEEEeqnarray}{rCl}
- C &=& \begin{bmatrix}1 & 0\end{bmatrix} \\
- R &=& \sigma_y^2     .
+ C &=& \begin{bmatrix}1 & 0\end{bmatrix} \nonumber \\
+ R &=& \sigma_y^2  \nonumber    .
 \end{IEEEeqnarray}

### 2012-06-08 submit_doublecol_VRsmoothing.tex
Submitted for publication

--- 2012-06-01 full_finaldraft_VRsmoothing_withSJGcomments.tex
@@
-\documentclass[journal]{IEEEtran}
+\documentclass[10pt,twocolumn,twoside]{IEEEtran}
 
@@
 
-\author{Pete~Bunch,~\IEEEmembership{Member,~IEEE,}
+\author{Pete~Bunch*,~\IEEEmembership{Member,~IEEE,}
         Simon~Godsill,~\IEEEmembership{Member,~IEEE,}% <-this % stops a space
@@
 of Engineering, Cambridge University, UK. email: \{pb404,sjg30\}@cam.ac.uk}% <-this % stops a space
-\thanks{Manuscript received January 01, 1901; revised January 02, 1901.}}
+}%\thanks{Manuscript received January 01, 1901; revised January 02, 1901.}}
 
 % The paper headers
-\markboth{IEEE Transaction in Signal Processing,~Vol.~1, No.~1, January~1901}%
-{Bunch \& Godsill: Particle Smoothing Algorithms for Variable Rate Models}
+\markboth{IEEE Transaction in Signal Processing}{IEEE Transaction in Signal Processing}%,~Vol.~1, No.~1, January~1901}%
+%{Bunch \& Godsill: Particle Smoothing Algorithms for Variable Rate Models}
 
@@
 In this paper, new particle smoothing algorithms are presented for use with conditionally linear-Gaussian and conditionally deterministic dynamics.
-These are demonstrated on problems of financial modelling and target tracking.
+These are demonstrated on problems in financial modelling and target tracking.
 Results indicate that the smoothing approximations provide more accurate and more diverse representations of the state posterior distributions.
@@
 
-The VRPF allows the changepoint sequence -- and hence the current state -- to be estimated sequentially as observations are received.
+The VRPF allows the changepoint sequence --- and hence the current state --- to be estimated sequentially as observations are received.
 However, estimates can often be improved later once further observations have been made.
@@
 The VRPF is reviewed in section~\ref{sec:vrpf}, and new variable rate smoothing algorithms are described in section~\ref{sec:vrps}.
-In section~\ref{sec:simulations}, particular examples of variable rate models are presented and algorithm performance is demonstrated in a series of simulations.
+In section~\ref{sec:simulations}, algorithm performance is demonstrated through simulations.
 
@@
 \begin{IEEEeqnarray}{rCl}
-p(\theta_n^-) & = & S(\tau_{K_n},t_n) p(u_0) \prod_{k=1}^{K_n} p(\tau_k, u_k| \tau_{k-1}, u_{k-1}) \label{eq:cp_sequence_prior}
+p(\theta_n^-) & = & S(\tau_{K_n},u_{K_n},t_n) p(u_0) \prod_{k=1}^{K_n} p(\tau_k, u_k| \tau_{k-1}, u_{k-1}) \label{eq:cp_sequence_prior}
 \end{IEEEeqnarray}
@@
 
-This closely resembles the variable-dimension spaces employed by reversible jump Markov chain Monte Carlo algorithms \cite{Green1995}.% or in multiple target tracking scenarios \cite{Vo2005}.
+This closely resembles the variable-dimension spaces employed by reversible jump Markov chain Monte Carlo (MCMC) algorithms \cite{Green1995}.% or in multiple target tracking scenarios \cite{Vo2005}.
 
@@
 We first consider the class of variable rate models whose state dynamics are linear-Gaussian conditional on the changepoint sequence.
-Such a model may be discretised onto the set of observation times in exactly the same manner as a standard the standard HMM,
+Such a model may be discretised onto the set of observation times in exactly the same manner as a standard HMM,
 %
@@
 
-If the changepoint sequence is known, or has been estimated, then the state values, $x_n$ may be inferred conditionally using optimal Kalman filtering and smoothing recursions.
+If the changepoint sequence is known, or has been estimated, then the state posterior distributions may be inferred conditionally using optimal Kalman filtering and smoothing recursions.
 As well as the basic Kalman filter \cite{Kalman1960}, the Rauch-Tung-Striebel (RTS) smoother \cite{Rauch1965} and two-filter smoother \cite{Fraser1969} will prove useful for this step later in the paper.
@@
 \begin{IEEEeqnarray}{rCl}
-\IEEEeqnarraymulticol{3}{l}{p(\theta_{n \setminus n-1} | \theta_{n-1}^-) = S(\tau_{K_n}, t_n) / S(\tau_{K_{n-1}+1}, t_{n-1}) } \nonumber \\
- \qquad  \qquad & & \times \prod\limits_{j:t_{n-1} \leq \tau_j < t_n} p(\tau_j, u_j| \tau_{j-1}, u_{j-1})  \label{eq:cp_sequence_trandens}     .
+\IEEEeqnarraymulticol{3}{l}{p(\theta_{n \setminus n-1} | \theta_{n-1}^-)} \nonumber \\
+  &=& S(\tau_{K_n},u_{K_n}, t_n) / S(\tau_{K_{n-1}+1},u_{K_{n-1}+1}, t_{n-1}) \nonumber \\
+  & & \times \prod\limits_{j:t_{n-1} \leq \tau_j < t_n} p(\tau_j, u_j| \tau_{j-1}, u_{j-1})  \label{eq:cp_sequence_trandens}     .
 \end{IEEEeqnarray}
@@
 The choice of weights determines the type of resampling used.
-The simplest choice, $v_{n-1}^{(j)} = 1/N_F$ (where $N_F$ is the number of filter particles) may be achieved by simply omitting this step all together and using the particles of $\hat{p}(\theta_{n-1}^-|y_{1:n-1})$.
+The simplest choice, $v_{n-1}^{(j)} = 1/\aleph_F$ (where $\aleph_F$ is the number of filter particles) may be achieved by simply omitting this step all together and using the particles of $\hat{p}(\theta_{n-1}^-|y_{1:n-1})$.
 This, however, leads to degeneracy of the particle weights over time.
@@
 Any other choice results in an auxiliary particle filter \cite{Pitt1999}.
-For further discussion of resampling, see \cite{Cappe2007,Doucet2009,Douc2005}.
-
-Next, an extension to the changepoint sequence, $\theta_{n \setminus n-1}^{(i)}$, is sampled from an importance distribution, $q(\theta_{n \setminus n-1}|\theta_{n-1}^{-(i)}, y_n)$, and concatenated with $\theta_{n-1}^{-(i)}$ to create a proposal for $\theta_n^-{(i)}$.
+For further discussion of resampling, see \cite{Cappe2007,Doucet2009}.
+
+Next, an extension to the changepoint sequence, $\theta_{n \setminus n-1}^{(i)}$, is sampled from an importance distribution, $q(\theta_{n \setminus n-1}|\theta_{n-1}^{-(i)}, y_n)$, and concatenated with $\theta_{n-1}^{-(i)}$ to create a proposal for $\theta_n^{-(i)}$.
 Finally, the particle is weighted according to the ratio of the target and proposal densities,
@@
 \begin{IEEEeqnarray}{rCl}
-v_{n-1}^{(i)} & \propto & \max ( 1, N_F w_{n-1}^{(i)} )     .
+v_{n-1}^{(i)} & \propto & \max ( 1, \aleph_F w_{n-1}^{(i)} )     .
 \end{IEEEeqnarray}
@@
 Conveniently, the Kalman filter also provides us with an estimate of the current state given the changepoint sequence and all the preceding observations.
-It follows from the the Gaussian dynamics and prior that these distributions are all Gaussian as well, \cite{Grewal2002},
+It follows from the Gaussian dynamics and prior that these distributions are all Gaussian as well, \cite{Grewal2002},
 %
@@
 \FOR{$n=1 \dots N$}
-  \FOR{$i=1 \dots N_F$}
+  \FOR{$i=1 \dots \aleph_F$}
   	\STATE Sample a history $\theta_{n-1}^{-(i)} \sim \sum_j v_{n-1}^{(j)} \delta_{\theta_{n-1}^{-(j)}}(\theta_{n-1}^-)$.
@@
 This distribution is approximated by the final step of the VRPF.
-However, in the same manner as the fixed rate filter-smoother of \cite{Kitagawa1996}, this approximation is likely to lack path-space diversity --- because of the necessary resampling step in the filtering algorithm, the particles all share the same set of changepoints before a particular time, with variation only appearing for changepoints closer to $T$.
+However, in the same manner as the fixed rate filter-smoother of \cite{Kitagawa1996}, this approximation is likely to lack path-space diversity; because of the necessary resampling step in the filtering algorithm, the particles all share the same set of changepoints before a particular time, with variation only appearing for changepoints closer to $T$.
 For a better characterisation of the smoothing distribution, it is necessary to rejuvenate the set of particles.
@@
 However, when the state dynamics are linear-Gaussian conditional on the changepoint sequence, it is possible split the dependence on the observations into past and future by introducing the current state as an additional variable.
-This device was used in \cite{Sarkka2012} in the derivation of the Rao-Blackwellised particle filter for fixed rate conditionally linear-Gaussian models,
+This device was used in \cite{Sarkka2012} in the derivation of the Rao-Blackwellised particle smoother for fixed rate conditionally linear-Gaussian models,
 %
@@
 Store all results.
-  \FOR{$i=1 \dots N_S$}
+  \FOR{$i=1 \dots \aleph_S$}
   	\STATE Initialise particle using $\theta^{(i)} \sim \sum_j w_N^{(j)} \delta_{\theta^{(j)}}(\theta)$.
@@
       \STATE Predict state mean and covariance $\tilde{m}_n^{-(i)}$ and $\tilde{P}_n^{-(i)}$ using (\ref{eq:backward_kf_predict}).
-      \FOR{$j=1 \dots N_F$}
+      \FOR{$j=1 \dots \aleph_F$}
 	      \STATE Calculate weight $\tilde{w}_n^{(j)}$ using (\ref{eq:rb-vrps_back_cond_weight2}).
@@
 
-The algorithmic complexity of this variable rate particle smoother for conditionally linear-Gaussian models is $O(N_F \times N_S \times N)$.
+The algorithmic complexity of this variable rate particle smoother for conditionally linear-Gaussian models is $O(\aleph_F \times \aleph_S \times N)$.
 It is possible to reduce this by using an MCMC sampling scheme in the style of \cite{Bunch2012}, which avoids the necessity of calculating the sampling weights for every filter particle.
@@
 When the changepoint sequence is split into past, $\theta_n^-$, and future $\theta_n^+$, and the past is altered, this will alter the state trajectory in the future.
-This is undesirable in practice, as each sampling operation will require the recalculation of state values and observations likelihoods for all future observation times.
+This is undesirable in practice, as each sampling operation will require the recalculation of state values and likelihoods for all future observation times.
 
@@
 The augmented target distribution may be sampled using Markov chain Monte Carlo (MCMC) \cite{Gilks1996}, in a similar manner to the fixed rate smoother of \cite{Bunch2012}.
-Metropolis-Hastings (MH) steps are conducted to draw samples, $\{\tilde{\theta}_{n}^-, u_{K_n}'\}$, from the target distribution by sampling the proposal and accepting the new values with a given probability.
+Metropolis-Hastings (MH) steps are conducted to draw samples, $\{\tilde{\theta}_{n}^-, u_{K_n}\}$, from the target distribution by sampling the proposal and accepting the new values with a given probability.
 Note that IS could be used instead of MCMC, but this would require resampling steps, which would lead to a loss of path-space diversity similar to that suffered by the particle filter.
@@
   \STATE Run variable rate particle filter to approximate $p(\tilde{\theta}_{n}^-|y_{1:n})$ with weighted particles $\{\tilde{\theta}_{n}^{-(i)}, w_{n}^{(i)}\}$. Store all results.
-  \FOR{$i=1 \dots N_S$}
+  \FOR{$i=1 \dots \aleph_S$}
   	\STATE Initialise particle using $\tilde{\theta}^{(i)} \sim \sum_j w_N^{(j)} \delta_{\tilde{\theta}^{(j)}}(\tilde{\theta})$.
@@
 
-The filter used $N_F = 100$ particles, and the smoother resampled $N_S = 100$ sequences.
+The filter used $\aleph_F = 100$ particles, and the smoother resampled $\aleph_S = 100$ sequences.
 Bootstrap proposals were used for the filter.
@@
 Observations were generated every $0.1s$.
-The standard deviations of tangential and normal accelerations were set to $0.01m/s^2$ and $1m/s^2$ respectively, and those for the drift velocities to $1m/s$.
+The variances of tangential and normal accelerations were set to $0.01m/s^2$ and $1m/s^2$ respectively, and those for the drift velocities to $1m/s$.
 The parameters of the Gamma distribution for inter-changepoint times were $5$ and $1s$ for the shape and scale respectively.
@@
 \bibliography{D:/pb404/Dropbox/PhD/Cleanbib}
-\input{D:/pb404/Dropbox/PhD/Biographies/IEEE-Bunch.tex}
-\input{D:/pb404/Dropbox/PhD/Biographies/IEEE-Godsill.tex}
+
+%\input{D:/pb404/Dropbox/PhD/Biographies/IEEE-Bunch.tex}
+%\input{D:/pb404/Dropbox/PhD/Biographies/IEEE-Godsill.tex}
+
 \end{document}

### 2012-06-08 submit_singlecol_VRsmoothing.tex
Submitted for publication

--- 2012-06-08 submit_doublecol_VRsmoothing.tex
@@
-\documentclass[10pt,twocolumn,twoside]{IEEEtran}
+\documentclass[peerreview,11pt,draftcls,onecolumn]{IEEEtran}
 
@@
 of Engineering, Cambridge University, UK. email: \{pb404,sjg30\}@cam.ac.uk}% <-this % stops a space
-}%\thanks{Manuscript received January 01, 1901; revised January 02, 1901.}}
+}
+%\thanks{Manuscript received June 08, 2012.}}
 
@@
 
-\begin{IEEEkeywords}
-Bayesian inference, state-space model, variable rate, particle filter, filtering, smoothing
-\end{IEEEkeywords}
+%\begin{IEEEkeywords}
+%Bayesian inference, state-space model, variable rate, particle filter, filtering, smoothing
+%\end{IEEEkeywords}
+
+\ifCLASSOPTIONpeerreview
+\begin{center} \bfseries EDICS Categories: MLR-BAYL, SSP-NGAU, SSP-TRAC, SSP-FILT \end{center}
+\fi
+
+\newpage
+% For peerreview papers, this IEEEtran command inserts a page break and
+% creates the second title. It will be ignored for other modes.
+%\IEEEpeerreviewmaketitle
 
@@
 \centering
-\includegraphics[width=0.9\columnwidth]{example_data.pdf}
+\includegraphics[width=0.45\columnwidth]{example_data.pdf}
 \caption{An example simulated data set, showing value (top) and trend (bottom). Value observations are overlayed. Jump times are shown as dotted vertical lines.}
@@
 \centering
-\subfloat[]{\includegraphics[width=0.9\columnwidth]{example_filtersmoother_state.pdf}} \\
-\subfloat[]{\includegraphics[width=0.9\columnwidth]{example_smoother_state.pdf}}
+\subfloat[]{\includegraphics[width=0.45\columnwidth]{example_filtersmoother_state.pdf}} \\
+\subfloat[]{\includegraphics[width=0.45\columnwidth]{example_smoother_state.pdf}}
 \caption{Trend estimates from (a) the filter-smoother and (b) the VRPS. Solid lines show means of each particle. Dashed lines show mean $\pm$2 standard deviations.}
@@
 \centering
-\renewcommand{\arraystretch}{1.5}
+%\renewcommand{\arraystretch}{1.5}
 \begin{tabular}{|c|c|c|c|}
 \hline
- &\raisebox{0cm}[0.4cm][0.3cm]{\parbox[c]{1.5cm}{\centering VRPF}} & \raisebox{0cm}[0.4cm][0.3cm]{\parbox[c]{1.5cm}{\centering Filter-smoother}} & \raisebox{0cm}[0.4cm][0.3cm]{\parbox[c]{1.5cm}{\centering VRPS}} \\
+ & VRPF & Filter-smoother & VRPS \\
 \hline \hline
-\raisebox{0cm}[0.4cm][0.3cm]{\parbox[c]{2cm}{\centering mean value estimate RMSE}}   & $5.31 \times 10^{-4}$ & $4.48 \times 10^{-4}$ & $4.16 \times 10^{-4}$ \\
+\centering mean value estimate RMSE & $5.31 \times 10^{-4}$ & $4.48 \times 10^{-4}$ & $4.16 \times 10^{-4}$ \\
 \hline
-\raisebox{0cm}[0.4cm][0.3cm]{\parbox[c]{2cm}{\centering mean trend estimate RMSE}}   & $2.56 \times 10^{-2}$ & $1.79 \times 10^{-2}$ & $1.49 \times 10^{-2}$ \\
+\centering mean trend estimate RMSE & $2.56 \times 10^{-2}$ & $1.79 \times 10^{-2}$ & $1.49 \times 10^{-2}$ \\
 \hline
@@
 \centering
-\includegraphics[width=0.75\columnwidth]{finance_unique_particles.pdf}
+\includegraphics[width=0.45\columnwidth]{finance_unique_particles.pdf}
 \caption{Mean number of unique particles in the filter-smoother (dashed) and smoother (solid) estimates.}
@@
 \centering
-\includegraphics[width=0.9\columnwidth]{simulated_problem.pdf}
+\includegraphics[width=0.45\columnwidth]{simulated_problem.pdf}
 \caption{An example simulated data set, showing true position (dashed line), observations (dots) and changepoint positions (crosses). }
@@
 \centering
-\subfloat[]{\includegraphics[width=0.9\columnwidth]{2DFilter.pdf}} \\
-\subfloat[]{\includegraphics[width=0.9\columnwidth]{2DSmoother.pdf}}
+\subfloat[]{\includegraphics[width=0.45\columnwidth]{2DFilter.pdf}} \\
+\subfloat[]{\includegraphics[width=0.45\columnwidth]{2DSmoother.pdf}}
 \caption{Filter (a) and smoother (b) particle estimates of position (solid lines), with particle changepoint positions (crosses). }
@@
 \centering
-\renewcommand{\arraystretch}{1.5}
+%\renewcommand{\arraystretch}{1.5}
 \begin{tabular}{|c|c|c|c|}
 \hline
- &\raisebox{0cm}[0.4cm][0.3cm]{\parbox[c]{1.5cm}{\centering VRPF}} & \raisebox{0cm}[0.4cm][0.3cm]{\parbox[c]{1.5cm}{\centering Filter-smoother}} & \raisebox{0cm}[0.4cm][0.3cm]{\parbox[c]{1.5cm}{\centering VRPS}} \\
-% & \begin{minipage}[c]{1.5cm}\centering RBVRPF + KF\end{minipage} & \begin{minipage}[c]{1.5cm}\centering RBVRPF + RTS\end{minipage} & \begin{minipage}[c]{1.5cm}\centering RBVRPS + RTS\end{minipage} \\
+ &VRPF & Filter-smoother & VRPS \\
 \hline \hline
-\raisebox{0cm}[0.4cm][0.3cm]{\parbox[c]{2cm}{\centering mean position estimate RMSE}} & 2.14 & 1.36 & 1.12 \\
+mean position estimate RMSE & 2.14 & 1.36 & 1.12 \\
 \hline
-\raisebox{0cm}[0.4cm][0.3cm]{\parbox[c]{2cm}{\centering mean velocity estimate RMSE}} & 1.89 & 1.03 & 0.85 \\
+mean velocity estimate RMSE & 1.89 & 1.03 & 0.85 \\
 \hline
@@
 \centering
-\includegraphics[width=0.75\columnwidth]{tracking_unique_particles.pdf}
+\includegraphics[width=0.45\columnwidth]{tracking_unique_particles.pdf}
 \caption{Mean number of unique particles in the final filter (dashed) and smoother (solid) estimates.}
@@
 \begin{IEEEeqnarray}{rCl}
-\mathbf{y}_{N-L:N} & = & \begin{bmatrix}y_N \\ y_{N-1} \\ \vdots \\ y_{N-L} \end{bmatrix} \nonumber \\
-%\end{IEEEeqnarray}
-%\begin{IEEEeqnarray}{rCl}
-H_{N-L} & = & \begin{bmatrix} C A^L \\ C A^{L-1} \\ \vdots \\ C \end{bmatrix} \nonumber \\
-%\end{IEEEeqnarray}
-%\begin{IEEEeqnarray}{rCl}
+\mathbf{y}_{N-L:N} & = & \begin{bmatrix}y_N \\ y_{N-1} \\ \vdots \\ y_{N-L} \end{bmatrix} \nonumber
+\end{IEEEeqnarray}
+\begin{IEEEeqnarray}{rCl}
+H_{N-L} & = & \begin{bmatrix} C A^L \\ C A^{L-1} \\ \vdots \\ C \end{bmatrix} \nonumber
+\end{IEEEeqnarray}
+\begin{IEEEeqnarray}{rCl}
  \Gamma_{N-1} & = & K_{N-L} \begin{bmatrix} Q_N & 0 & \dots & 0 \\ 0 & Q_{N-1} & \dots & 0 \\ \vdots & \vdots & \ddots & \vdots \\ 0 & 0 & \dots & Q_{N-L+1} \end{bmatrix} K_{N-L}^T \nonumber \\
-  & & + \: \begin{bmatrix} R_N & 0 & \dots & 0 \\ 0 & R_{N-1} & \dots & 0 \\ \vdots & \vdots & \ddots & \vdots \\ 0 & 0 & \dots & R_{N-L} \end{bmatrix} \nonumber \\
-%\end{IEEEeqnarray}
-%\begin{IEEEeqnarray}{rCl}
+  & & + \: \begin{bmatrix} R_N & 0 & \dots & 0 \\ 0 & R_{N-1} & \dots & 0 \\ \vdots & \vdots & \ddots & \vdots \\ 0 & 0 & \dots & R_{N-L} \end{bmatrix} \nonumber
+\end{IEEEeqnarray}
+\begin{IEEEeqnarray}{rCl}
  K_{N-L} & = & \begin{bmatrix} C & CA & \dots & CA^{L-1} \\ 0 & C & \dots & CA^{L-2} \\ \vdots & \vdots & \ddots & \vdots \\ 0 & 0 & \dots & C \\ 0 & 0 & \dots & 0 \end{bmatrix} \nonumber

### 2012-09-13 response_to_reviewers.tex
Revisions

\documentclass{article}

\usepackage{amsmath}
\usepackage{IEEEtrantools}
\usepackage{color}

\newenvironment{meta}[0]{\color{red} \em}{}

\begin{document}

{\meta Many thanks for the comments, folks. I've made changes to address all but one, with which I have a bit of a philosophical disagreement. Reviewer 3's comments were particularly useful and prompted me to rethink carefully the purpose of the simulations section. Responses are inline with the individual points below. }

\section*{Reviewer Comments}

\subsection*{Reviewer: 1}

Recommendation: AQ - Publish In Minor, Required Changes

\subsubsection*{Comments}
This is an interesting, well written paper  that suggests new sampling strategies for smoothing  in variable rate models. The proposed method  rejuvenates the variable rate filtering particles that usually have poor smoothing performance due to the loss of particle diversity particularly when state discontinuities occur.

The paper relays on the key idea, using the previously generated filtering particles as a proposal  to the target conditional distributions, which is inherited from the previous work of the author [Godsill et al. 2004] . The proposed algorithms apply a backward pass by sampling the  target conditional distribution $P( \theta^{-}_{n} |  . )$  for $n=N…1$.   Two methodologies are proposed for sampling the target conditional distribution of the  variable rate structured "conditionally linear Gaussian models"  and "conditionally deterministic models". The former samples from a discrete set of re-weighted filtering particles according to well known Rao-Blackwellization approach and the other uses an  MCMC sampling scheme where the proposals are constructed by a  modified  filtering particle set.

Both algorithms are sophisticated techniques for rejuvenating the particles and can be  interpreted as a way of reconstructing the particle history by using the filtering particles sampled at each filtering step. Reference on the convergence property of the algorithms and a  discussion to understand the practical aspects of the the algorithm (Ex. whether is it  possible to increase the online filtering performance by using retrospective smoothing steps, the ability of the algorithm to  find the untracked modes of the posterior - this can also be suported with simulation results )  would be useful  for the reader in order to understand the limits of the algorithms.

{\meta I've added a brief note on convergence, which is inherited from the basic results for the forward-filtering-backward-sampling algorithm. I've also added an explicit example in the simulations of the smoother representing multi-modality where some particle contain a jump and others don't. Fixed interval smoothing won't help online filtering. Fixed-lag smoothing would, and this is covered elsewhere, especially in [Whitely et al. 2011]. }

In the simulations section, smoothing performance of the algorithms are compared to filtering and  the filtering- smoothing results.  In both tests improvement is achieved in means of RMSE. I see that in the  simulations filter and filter- smoother is able to represent the solution to the data. However it would be much more interesting and convincing to see an illustration of a scenario  where the filter is not able to represent the solution due to discarded particles in the resampling steps  (Ex.  poor estimate for the changepoint location or  parameters,  misses a jump or unable to track the maneuver due to multi-modality)  but  the smoother is able track the solution.

{\meta Because the smoother uses filter particles as a proposal, if the filter fails catastrophically (e.g. track loss) then the smoother will also fail. I have added an example where the multi-modality of the particle representation is lost due to resampling in the filter-smoother, but not the filter. }

Minor comments:

Page: 37 line 36 :  " $\theta$ " should be  "  $\theta'$ "
Fig: 4 step 11 " $\theta*$ " should be  " $\theta'*$  "

{\meta Well spotted. Now correct.}

\subsection*{Reviewer: 2}

Recommendation: A - Publish Unaltered

\subsubsection*{Comments}
(There are no comments. Please check to see if comments were included as a file attachment with this e-mail or as an attachment in your Author Center.)

{\meta No file attached, so I assume there are genuinely no comments. }

\subsection*{Reviewer: 3}

Recommendation: RQ - Review Again After Major Changes

\subsubsection*{Comments}
The authors propose a smoothing PF for variable rate models. The proposed method is essentially an adaptation of existing methods for fixed-rate models to the variable rate model. I feel there is enough substance in the adaptation process to merit publication provided some revisions are performed. My main concerns are with the performance analysis:

- It is never clarified but I'm assuming that the smoothing results are for fixed-interval smoothing, i.e., estimates at time $t_n, n=1,...,N$ use measurements from times $t_1,...,t_N$. This should be clearly stated.

{\meta I feel this is clear from the introduction and the maths, but it's now explicitly stated as well. }

- Only 10 realisations are used! It's difficult to draw conclusions with such a small sample size.

{\meta Ok. Now 100.}

- In Figs 7 and 10 the effective sample size would be a better measure of performance than the number of unique particles.

- The principal improvement offered by the more sophisticated particle smoothers compared to the filter-smoother is increased particle diversity. This should improve estimation performance but the improvements in consistency may be even greater. This is alluded to in the paper and partially covered by Figs 6 and 9 which show plots of sample trajectories. However, a more rigorous consistency analysis should be performed. For example, a simple approach would be to construct a Gaussian approximation to the posterior, establish a confidence ellipsoid and then count the number of times the true value falls within this ellipsoid.

{\meta These two comments concern the same question: How should particle approximations of a distribution be compared? There is, of course, much more to a distributional estimate than how close its mean is to the true value, so we do need something more than just the RMSE, as you suggest. Ideally, we would like to compare our set of particles to the true posterior distribution, using statistics such as the effective sample size (ESS), or normalised estimation error squared (NEES). However this true posterior is not available. As a substitute, either a Gaussian approximation or a particle approximation with loads of particles are sometimes used as ``truth''. This is far from ideal, and I find the latter particularly unsettling, because we end up testing our particle algorithms only against their own best performance.

My preferred alternative for measuring consistency is a mean ``empirical NEES'' statistic which takes a value between 0 and 1 and takes into account the estimation error, covariance estimate and diversity of the particle approximation. I've added details and plots of this metric to the paper. I think that together with the RMSE values, this provides sufficient evidence of the improvement provided by the smoother. }

- Comparing the proposed method and the filter-smoother for equal sample sizes doesn't really make sense given that the former has $O(N^2)$ expense while the latter has $O(N)$ expense. The filter-smoother should be implemented with a much larger sample size such that both algorithms have approximately equal expense. This approach, which is taken in (Fearnhead et al, 2010), seems much fairer.

{\meta I completely disagree. I think this would be misleading and is unnecessary for demonstrating the purpose of the new smoothers.

For a start, the filter-smoother and the smoother provide estimates with different qualities, and comparing them like this would be to compare apples with oranges. The principal reason for using a particle smoother is to rejuvenate a particle population with low path-space diversity. This is demonstrated by the increased number of unique particles and the improved consistency. (The reduced RMSE is a consequent bonus.) In contrast, [Fearnhead et al. 2010] are comparing a variety of smoothing algorithms all of which do the same thing, produce a diverse set of marginal smoothing particles. Their methods are not appropriate here.

Furthermore, if we were trying to solve a particular practical problem, such as producing the best state estimator possible for an offline tracker, then it would be a sensible test. However, the example simulations here are entirely fictional, cooked up simply to demonstrate that the new smoothing algorithm functions as expected. The problem is that whatever this test demonstrated about relative performance, the result could be reversed by changing the set-up of the simulation. }

- Results should be shown for a range of sample sizes so we can get an idea of the required sample size for (close to) optimal performance.

{\meta Yes. I've added graphs showing the effect of varying the number of smoothing trajectories for the tracking case. The finance results are similar and so omitted for space. }

Some minor comments:
- A Rao-Blackwellised VRPF was also developed in (Morelande and Gordon, 2009). Their formulation was more general than that used here since it permits the model to be partially linear-Gaussian.

{\meta Yes. This should be and now is referenced. Actually, their formulation is not more general, just different. The Rao-Blackwellisation in our paper is of the state variables when the dynamics are linear-Gaussian, while [Morelande and Gordon, 2009] Rao-Blackwellise the linear parts of the motion parameters, but in a conditionally deterministic model. We could extend our work by using their Rao-Blackwellisation scheme on the piecewise-deterministic smoother. }

- Recently, O(N) schemes for particle smoothing which improve upon the filter-smoother have been proposed by (Fearnhead et al, 2010) and (Douc et al, 2011). Perhaps these could be used for VR filtering. They should at least be mentioned.

{\meta I've added a reference to [Douc et al., 2011]. This is similar to our Metropolis-Hastings-based method which is already referenced (but unhelpfully so as it is still awaiting publication). ([Bunch and Godsill, 2012]). The Fearnhead method is for marginal (as opposed to joint) smoothing distributions only, so it doesn't really help. I've mentioned this in section IV. }

- My understanding is that in (33) $\tau_{K_{n-1}+1}$ should be drawn from a truncated distribution which forces it to be greater than $t_{n-1}$. According to the notation of (4), shouldn't the denominator then be $S(\tau_{K_{n-1}},...)$ rather than $S(\tau_{K_{n-1}+1},...)$?

{\meta Yes. Well spotted! Corrected. }

- On page 13, just before (51), there is a $\tilde{\theta}$ which should be a $\theta$.

{\meta Fixed. }

- In Fig. 9(a), are these results for the filter-smoother?

{\meta Yes. Clarified.}

- I'm not sure all the appendices are necessary since the results can be found in other, easily obtained sources. At least

{\meta I agree. I've cut the first two, giving the derivation and initialisation of the two-filter smoother. The discretisation of the finance model remains, as I can't find a complete and equivalent version in a published work. }

\end{document}

### 2012-09-13 revisions_VRsmoothing.tex
Revisions

--- 2012-06-08 submit_doublecol_VRsmoothing.tex
@@
 Conditional upon this MPP the state evolves according to some benign dynamics.
-In \cite{Godsill2007,Whiteley2011}, the conditional state evolution is treated as deterministic, while in \cite{Godsill2007a,Christensen2012} a conditionally linear-Gaussian state model is considered.
+In \cite{Godsill2007,Whiteley2011,Morelande2009a}, the conditional state evolution is treated as deterministic, while in \cite{Godsill2007a,Christensen2012} a conditionally linear-Gaussian state model is considered.
 
@@
 However, estimates can often be improved later once further observations have been made.
-In this paper, we address the problem of smoothing in variable rate models, i.e.
+In this paper, we address the problem of fixed-interval smoothing in variable rate models, i.e.
 the estimation of the changepoint sequence and latent state given all the observations.
@@
 radar measurements) are made.
-In \cite{Godsill2007a,Godsill2007,Whiteley2011,Bunch2012a}, variable rate models were introduced for tracking.
+In \cite{Godsill2007a,Godsill2007,Whiteley2011,Morelande2009a,Bunch2012a}, variable rate models were introduced for tracking.
 The state trajectory is divided up by a set of changepoints, between which the motion follows a deterministic path governed by motion parameters (accelerations, etc.) which are fixed for that division.
@@
 \IEEEeqnarraymulticol{3}{l}{p(\theta_{n \setminus n-1} | \theta_{n-1}^-)} \nonumber \\
-  &=& S(\tau_{K_n},u_{K_n}, t_n) / S(\tau_{K_{n-1}+1},u_{K_{n-1}+1}, t_{n-1}) \nonumber \\
+  &=& S(\tau_{K_n},u_{K_n}, t_n) / S(\tau_{K_{n-1}},u_{K_{n-1}}, t_{n-1}) \nonumber \\
   & & \times \prod\limits_{j:t_{n-1} \leq \tau_j < t_n} p(\tau_j, u_j| \tau_{j-1}, u_{j-1})  \label{eq:cp_sequence_trandens}     .
@@
 
+In \cite{Morelande2009a}, an improvement to this algorithm for conditionally-deterministic models was introduced in which the linear parts of the motion parameters are Rao-Blackwellised.
+Note that this is different to the scheme used in the previous section in which the state variables are Rao-Blackwellised.
+
 \subsection{Improving the Variable Rate Particle Filter}
@@
 For a better characterisation of the smoothing distribution, it is necessary to rejuvenate the set of particles.
-This is achieved here with a backward pass through the observations in an extension of the forward-backward method described in \cite{Godsill2004}.
+This is achieved here with a backward pass through the observations in an extension of the forward-filtering-backward-sampling (FFBS) method described in \cite{Godsill2004}.
+Since these algorithms are based on the FFBS algorithm, the convergence proofs of \cite{Godsill2004,Douc2011} are believed to apply.
+
+For standard HMMs, it is common to distinguish between joint smoothing distributions, $p(x_{1:N}|y_{1:N})$, and marginal smoothing distributions, $p(x_k|y_{1:N})$.
+Note that for variable rate models, the latter is ill defined --- we need the entire sequence of changepoints --- and thus algorithms for estimating marginal smoothing distributions, such as reweighting smoothers \cite{Doucet2000} and two-filter smoothers \cite{Briers2010,Fearnhead2010} will not be considered here.
 
@@
 
-The second state-dependent term in (\ref{eq:rb-vrps_back_cond_weight}) is the likelihood, $p(y_{n+1:N}|x_n, \tilde{\theta}_{n}^+)$.
-This may be regarded as an improper density over $x_n$, and may be calculated analytically using a backwards Kalman filter, in a similar manner to that used in the two-filter smoother \cite{Fraser1969,Kitagawa1994,Sarkka2012,Gelb1974}.
-Such a backwards Kalman filter uses the following recursions, which are derived in appendix~\ref{app:backwards_KF},
+The second state-dependent term in (\ref{eq:rb-vrps_back_cond_weight}) is the likelihood, $p(y_{n+1:N}|x_n, \theta_{n}^+)$.
+This may be regarded as an improper density over $x_n$, and may be calculated analytically using a backwards Kalman filter, in a similar manner to that used in the two-filter smoother \cite{Fraser1969,Kitagawa1994,Sarkka2012,Gelb1974,Sarkka2012}.
+Such a backwards Kalman filter uses the following recursions,% which are derived in appendix~\ref{app:backwards_KF},
 %
@@
 
-Exact methods for initialising this recursion are discussed in appendix~\ref{app:init_backwards_KF}, and also in \cite{Kitagawa1994}.
+Exact methods for initialising this recursion are discussed in \cite{Kitagawa1994,Sarkka2012}.
 However, it is often sufficient to use an approximation, for example $\tilde{m}_N^{(i)} \gets m_N^{(i)}$ and $\tilde{P}_N^{(i)} \gets P_N^{(i)}$.
+%appendix~\ref{app:init_backwards_KF}, and also in
 
@@
 The algorithmic complexity of this variable rate particle smoother for conditionally linear-Gaussian models is $O(\aleph_F \times \aleph_S \times N)$.
-It is possible to reduce this by using an MCMC sampling scheme in the style of \cite{Bunch2012}, which avoids the necessity of calculating the sampling weights for every filter particle.
+It is possible to reduce this by using a rejection sampling \cite{Douc2011} or MCMC \cite{Bunch2012} sampling scheme, which avoids the necessity of calculating the sampling weights for every filter particle.
 
@@
 The augmented target distribution may be sampled using Markov chain Monte Carlo (MCMC) \cite{Gilks1996}, in a similar manner to the fixed rate smoother of \cite{Bunch2012}.
-Metropolis-Hastings (MH) steps are conducted to draw samples, $\{\tilde{\theta}_{n}^-, u_{K_n}\}$, from the target distribution by sampling the proposal and accepting the new values with a given probability.
+Metropolis-Hastings (MH) steps are conducted to draw samples, $\{\tilde{\theta}_{n}^{-'}, u_{K_n}\}$, from the target distribution by sampling the proposal and accepting the new values with a given probability.
 Note that IS could be used instead of MCMC, but this would require resampling steps, which would lead to a loss of path-space diversity similar to that suffered by the particle filter.
@@
         \STATE Calculate $\alpha_n^{(m)}$ using (\ref{eq:pd-vrps_tpr}) and (\ref{eq:pd-vrps_ap}).
-        \STATE With probability $\alpha_n^{(m)}$, $\tilde{\theta}_n^{-(i)(m)} \gets \tilde{\theta}_n^{-(i)*}$. \\ Otherwise, $\tilde{\theta}_n^{-(i)(m)} \gets \tilde{\theta}_n^{-(i)(m-1)}$.
+        \STATE With probability $\alpha_n^{(m)}$, $\tilde{\theta}_n^{-(i)(m)} \gets \tilde{\theta}_n^{-'(i)*}$. \\ Otherwise, $\tilde{\theta}_n^{-(i)(m)} \gets \tilde{\theta}_n^{-(i)(m-1)}$.
       \ENDFOR
@@
 See \cite{Bunch2012} for analysis of this trade-off on a fixed rate model.
-In the simulations discussed in this paper, $M=1$ was used throughout.
+The resulting computational complexity of the algorithm is $O(M \times \aleph_S \times N)$.
 
@@
 These can be calculated by simply taking a mean of the particle values (using the Gaussian means for the linear-Gaussian case).
-The accuracy is then assessed by comparing the root mean squared error (RMSE) of the resulting estimators.
+The accuracy is then assessed by comparing the size of estimation errors of the resulting estimators.
+In addition, the benefit of employing a smoothing algorithm is that it ensures a more diverse particle representation of the posterior state distribution.
+To demonstrate this effect, a measure of estimator consistency is calculated by which to compare the algorithms, the mean empirical normalised estimation error squared (MENEES).
+This is defined as,  %the number of unique state samples is plotted over time in the following examples.
+A
+%
+\begin{IEEEeqnarray}{rCl}
+ \text{MENEES}_k & = & (\hat{x}_k - x_k^*)^T \hat{P}_k (\hat{x}_k - x_k^*) \nonumber \\ 
+ \hat{x}_k & = & \sum_i x_{k}^{(i)} \nonumber \\
+ \hat{P}_k & = & \sum_i (x_{k}^{(i)} - x_k^*) (x_{k}^{(i)} - x_k^*)^T \nonumber     ,
+\end{IEEEeqnarray}
+
+where $x_k^*$ is the true state, and $x_k^{(i)}$ are the individual particle states.
+This statistic takes a value between $0$ and $1$, where $1$ is scored by a degenerate particle estimate (fewer linearly independent particles than state dimensions).
+For a non-degenerate estimate, smaller values correspond to smaller estimation errors relative to the estimated covariance.
+Thus, it incorporates the diversity of the particle set as well the accuracy of the estimate and covariance.
 
@@
 In the linear-Gaussian case, an RTS smoother is used to find the state posterior distribution for each particle, $p(x_n|\theta^{(i)}, y_{1:N})$, from which state estimates are derived.
-
-In addition, the benefit of employing a smoothing algorithm is that it ensures a more diverse particle representation of the posterior state distribution.
-To demonstrate this effect, the number of unique state samples is plotted over time in the following examples.
 
@@
 Notice that the path diversity is dramatically improved by the VRPS.
-
+%
 \begin{figure}[!t]
@@
 \end{figure}
-
+%
 \begin{figure}[!t]
@@
 \subfloat[]{\includegraphics[width=0.9\columnwidth]{example_smoother_state.pdf}}
-\caption{Trend estimates from (a) the filter-smoother and (b) the VRPS. Solid lines show means of each particle. Dashed lines show mean $\pm$2 standard deviations.}
+\caption{Trend estimates from (a) the filter-smoother and (b) smoother. Solid lines show means of each particle. Dashed lines show mean $\pm$2 standard deviations.}
 \label{fig:example_state}
 \end{figure}
-
-Performance was evaluated by testing on 10 scenarios, each with 1000 observations.
+%
+Performance was evaluated by testing on 100 scenarios, each with 1000 observations.
 At each observation time, an estimate of the value and trend is made by taking the average of Gaussian means generated by an RTS smoother for each particle.
-The resulting root mean square errors (RMSEs) are displayed in table~\ref{tab:finance_state_performance}, demonstrating an improvement in accuracy from the smoother.
-
-\begin{table}%
-\caption{State estimation performance.}
-\label{tab:finance_state_performance}
-\centering
-\renewcommand{\arraystretch}{1.5}
-\begin{tabular}{|c|c|c|c|}
-\hline
- &\raisebox{0cm}[0.4cm][0.3cm]{\parbox[c]{1.5cm}{\centering VRPF}} & \raisebox{0cm}[0.4cm][0.3cm]{\parbox[c]{1.5cm}{\centering Filter-smoother}} & \raisebox{0cm}[0.4cm][0.3cm]{\parbox[c]{1.5cm}{\centering VRPS}} \\
-\hline \hline
-\raisebox{0cm}[0.4cm][0.3cm]{\parbox[c]{2cm}{\centering mean value estimate RMSE}}   & $5.31 \times 10^{-4}$ & $4.48 \times 10^{-4}$ & $4.16 \times 10^{-4}$ \\
-\hline
-\raisebox{0cm}[0.4cm][0.3cm]{\parbox[c]{2cm}{\centering mean trend estimate RMSE}}   & $2.56 \times 10^{-2}$ & $1.79 \times 10^{-2}$ & $1.49 \times 10^{-2}$ \\
-\hline
-\end{tabular}
-\end{table}
-
-The mean number of unique particle estimates at each time step is shown in Fig.~\ref{fig:finance_unique_particles}.
-This illustrates the main advantage of the smoother algorithm --- the increased number of unique particles in the approximation means a better characterisation of the posterior changepoint and state distributions.
-
+The resulting mean absolute error over all scenarios is shown in figure~\ref{fig:finance_rmse}, demonstrating a small improvement in accuracy from the smoother.
+More significantly, MENEES is shown in figure~\ref{fig:finance_menees}, showing a similar consistency between filter and smoother.
+The filter-smoother produces an MENEES of $1$ at almost all time instants, because there is only one unique particle in the set.% are displayed in table~\ref{tab:finance_state_performance}
+%
 \begin{figure}[!t]
 \centering
-\includegraphics[width=0.75\columnwidth]{finance_unique_particles.pdf}
-\caption{Mean number of unique particles in the filter-smoother (dashed) and smoother (solid) estimates.}
-\label{fig:finance_unique_particles}
+\subfloat[]{\includegraphics[width=0.9\columnwidth]{finance_value_rmse.pdf}} \\
+\subfloat[]{\includegraphics[width=0.9\columnwidth]{finance_trend_rmse.pdf}}
+\caption{Value (a) and trend (b) mean absolute error over 100 scenarios using filter, filter-smoother, and smoother.}
+\label{fig:finance_rmse}
+\end{figure}
+%
+\begin{figure}[!t]
+\centering
+\includegraphics[width=0.9\columnwidth]{finance_menees.pdf}
+\caption{MENEES over 100 scenarios using filter, filter-smoother, and smoother.}
+\label{fig:finance_menees}
+\end{figure}
+%
+%\begin{table}%
+%\caption{State estimation performance.}
+%\label{tab:finance_state_performance}
+%\centering
+%\renewcommand{\arraystretch}{1.5}
+%\begin{tabular}{|c|c|c|c|}
+%\hline
+% &\raisebox{0cm}[0.4cm][0.3cm]{\parbox[c]{1.5cm}{\centering VRPF}} & \raisebox{0cm}[0.4cm][0.3cm]{\parbox[c]{1.5cm}{\centering Filter-smoother}} & \raisebox{0cm}[0.4cm][0.3cm]{\parbox[c]{1.5cm}{\centering VRPS}} \\
+%\hline \hline
+%\raisebox{0cm}[0.4cm][0.3cm]{\parbox[c]{2cm}{\centering mean value estimate RMSE}}   & $5.31 \times 10^{-4}$ & $4.48 \times 10^{-4}$ & $4.16 \times 10^{-4}$ \\
+%\hline
+%\raisebox{0cm}[0.4cm][0.3cm]{\parbox[c]{2cm}{\centering mean trend estimate RMSE}}   & $2.56 \times 10^{-2}$ & $1.79 \times 10^{-2}$ & $1.49 \times 10^{-2}$ \\
+%\hline
+%\end{tabular}
+%\end{table}
+%
+%The mean number of unique particle estimates at each time step is shown in Fig.~\ref{fig:finance_unique_particles}.
+This illustrates the main advantage of the smoother algorithm --- the increased number of unique particles in the approximation means a better characterisation of the posterior changepoint and state distributions.
+%
+%\begin{figure}[!t]
+%\centering
+%\includegraphics[width=0.75\columnwidth]{finance_unique_particles.pdf}
+%\caption{Mean number of unique particles in the filter-smoother (dashed) and smoother (solid) estimates.}
+%\label{fig:finance_unique_particles}
+%\end{figure}
+%
+As well as providing more accurate and more diverse estimates, the smoother is also capable of preserving multi-modality in its representation of the posterior distribution, which is lost by the filter smoother due to the particle collapse to a single value.
+This is highlighted in figure~\ref{fig:finance_multimodality}, which shows a segment of the trend estimates from figure~\ref{fig:example_state} containing a small jump.
+In the filter-smoother results, none of the particles have a jump at this point.
+In contrast, a significant number of the smoother particles contain a jump around about the right place.
+This ability to represent multi-modality is a key advantage of good particle smoothing algorithms.
+%
+\begin{figure}[!t]
+\centering
+\subfloat[]{\includegraphics[width=0.45\columnwidth]{finance_filter_multimodality.pdf}}
+\subfloat[]{\includegraphics[width=0.45\columnwidth]{finance_smoother_multimodality.pdf}}
+\caption{Filter-smoother (a) and smoother (b) estimates of trend around a jump, highlighting the ability of the smoother to represent multi-modal distributions.}
+\label{fig:finance_multimodality}
 \end{figure}
@@
 
-The filter and smoother were each used to generate $50$ particles, with the filter employing resample-move steps with proposals based on an unscented transform \cite{Julier2004} approximation to the optimal proposal density.
-
-An example trajectory is shown in Fig.~\ref{fig:simulated_trajectory}, with the results of the VRPF and VRPS algorithms in Fig.\ref{fig:2D_particle_results}.
+The filter and smoother were each used to generate $50$ particles, with the filter employing resample-move steps with proposals based on an unscented transform \cite{Julier2004} approximation to the optimal proposal density, and the smoother using $M=5$ steps in each Markov chain.
+
+An example trajectory is shown in Fig.~\ref{fig:simulated_trajectory}, with the results from the filter-smoother and smoother in Fig.\ref{fig:2D_particle_results}.
 The improved particle diversity achieved by the smoothing algorithm is readily apparent.
-
+%
 \begin{figure}[!t]
@@
 \subfloat[]{\includegraphics[width=0.9\columnwidth]{2DSmoother.pdf}}
-\caption{Filter (a) and smoother (b) particle estimates of position (solid lines), with particle changepoint positions (crosses). }
+\caption{Filter-smoother (a) and smoother (b) particle estimates of position (solid lines), with particle changepoint positions (crosses). }
 \label{fig:2D_particle_results}
 \end{figure}
-
-Performance was evaluated by testing on 10 independently generated scenarios, each with 500 observations.
+%
+Performance was evaluated by testing on 100 independently generated scenarios, each with 500 observations.
 At each observation time, an estimate of the position and velocity is made by taking the average of values from the array of particles.
-The resulting root mean square errors (RMSEs) are displayed in table~\ref{tab:tracking_state_performance}, demonstrating an improvement in accuracy from the smoother.
-The VRPS also provides a more accurate estimate of the number of changepoints in each test, with an average error of 0.6 compared to 2.0 from the filter-smoother.
-
-The mean number of unique particle estimates at each time step is shown in Fig.~\ref{fig:tracking_unique_particles}.
-This illustrates again the principal advantage of the VRPS algorithm --- the increased number of unique particles in the approximation means a better characterisation of the posterior changepoint distribution, allowing, for example, calculation of state covariances.
-
-\begin{table}%
-\caption{State estimation performance.}
-\label{tab:tracking_state_performance}
-\centering
-\renewcommand{\arraystretch}{1.5}
-\begin{tabular}{|c|c|c|c|}
-\hline
- &\raisebox{0cm}[0.4cm][0.3cm]{\parbox[c]{1.5cm}{\centering VRPF}} & \raisebox{0cm}[0.4cm][0.3cm]{\parbox[c]{1.5cm}{\centering Filter-smoother}} & \raisebox{0cm}[0.4cm][0.3cm]{\parbox[c]{1.5cm}{\centering VRPS}} \\
-% & \begin{minipage}[c]{1.5cm}\centering RBVRPF + KF\end{minipage} & \begin{minipage}[c]{1.5cm}\centering RBVRPF + RTS\end{minipage} & \begin{minipage}[c]{1.5cm}\centering RBVRPS + RTS\end{minipage} \\
-\hline \hline
-\raisebox{0cm}[0.4cm][0.3cm]{\parbox[c]{2cm}{\centering mean position estimate RMSE}} & 2.14 & 1.36 & 1.12 \\
-\hline
-\raisebox{0cm}[0.4cm][0.3cm]{\parbox[c]{2cm}{\centering mean velocity estimate RMSE}} & 1.89 & 1.03 & 0.85 \\
-\hline
-\end{tabular}
-\end{table}
-
+The resulting mean absolute error over all scenarios is shown in figure~\ref{fig:finance_rmse}, again demonstrating a small improvement in accuracy from the smoother.
+The smoother also provides a more accurate estimate of the number of changepoints in each test, with an average error of 0.6 compared to 2.0 from the filter-smoother.
+MENEES is displayed in figure~\ref{fig:tracking_menees}, showing a similar consistency between filter and smoother.
+The filter-smoother produces an MENEES of $1$ at almost all time instants, because there is only one unique particle in the set.
+%in table~\ref{tab:tracking_state_performance}, demonstrating an improvement in accuracy from the smoother.
+%
 \begin{figure}[!t]
 \centering
-\includegraphics[width=0.75\columnwidth]{tracking_unique_particles.pdf}
-\caption{Mean number of unique particles in the final filter (dashed) and smoother (solid) estimates.}
-\label{fig:tracking_unique_particles}
+\subfloat[]{\includegraphics[width=0.9\columnwidth]{tracking_pos_rmse.pdf}} \\
+\subfloat[]{\includegraphics[width=0.9\columnwidth]{tracking_vel_rmse.pdf}}
+\caption{Position (a) and velocity (b) mean absolute error over 100 scenarios using filter, filter-smoother, and smoother.}
+\label{fig:tracking_rmse}
+\end{figure}
+%
+\begin{figure}[!t]
+\centering
+\includegraphics[width=0.9\columnwidth]{tracking_menees.pdf}
+\caption{MENEES over 100 scenarios using filter, filter-smoother, and smoother.}
+\label{fig:tracking_menees}
+\end{figure}
+%
+%The mean number of unique particle estimates at each time step is shown in Fig.~\ref{fig:tracking_unique_particles}.
+This illustrates again the principal advantage of the VRPS algorithm --- the increased number of unique particles in the approximation means a better characterisation of the posterior changepoint distribution, allowing, for example, calculation of state covariances.
+%
+%\begin{table}%
+%\caption{State estimation performance.}
+%\label{tab:tracking_state_performance}
+%\centering
+%\renewcommand{\arraystretch}{1.5}
+%\begin{tabular}{|c|c|c|c|}
+%\hline
+% &\raisebox{0cm}[0.4cm][0.3cm]{\parbox[c]{1.5cm}{\centering VRPF}} & \raisebox{0cm}[0.4cm][0.3cm]{\parbox[c]{1.5cm}{\centering Filter-smoother}} & \raisebox{0cm}[0.4cm][0.3cm]{\parbox[c]{1.5cm}{\centering VRPS}} \\
+%% & \begin{minipage}[c]{1.5cm}\centering RBVRPF + KF\end{minipage} & \begin{minipage}[c]{1.5cm}\centering RBVRPF + RTS\end{minipage} & \begin{minipage}[c]{1.5cm}\centering RBVRPS + RTS\end{minipage} \\
+%\hline \hline
+%\raisebox{0cm}[0.4cm][0.3cm]{\parbox[c]{2cm}{\centering mean position estimate RMSE}} & 2.14 & 1.36 & 1.12 \\
+%\hline
+%\raisebox{0cm}[0.4cm][0.3cm]{\parbox[c]{2cm}{\centering mean velocity estimate RMSE}} & 1.89 & 1.03 & 0.85 \\
+%\hline
+%\end{tabular}
+%\end{table}
+%
+%\begin{figure}[!t]
+%\centering
+%\includegraphics[width=0.75\columnwidth]{tracking_unique_particles.pdf}
+%\caption{Mean number of unique particles in the final filter (dashed) and smoother (solid) estimates.}
+%\label{fig:tracking_unique_particles}
+%\end{figure}
+%
+Finally, in figure~\ref{fig:varying_num_pts} the variation in accuracy and consistency of the smoothing particle approximation is demonstrated as the number of smoothing particles varies.
+As expected, the rate of improvement diminishes and becomes negligible as the number of smoothing particles exceeds the number of filtering particles.
+%
+\begin{figure}[!t]
+\centering
+\subfloat[]{\includegraphics[width=0.45\columnwidth]{pos_RMSE_vs_numpts.pdf}}
+\subfloat[]{\includegraphics[width=0.45\columnwidth]{MENEES_vs_numpts.pdf}}
+\caption{Average position error (a) and MENEES (b) over 100 scenarios with varying number of smoother particles and 50 filter particles.}
+\label{fig:varying_num_pts}
 \end{figure}
@@
 For the deterministic case, smoothing is achieved by sampling an augmented target distribution using MCMC.
-Simulations demonstrate that the smoothers generate approximations with improved particle diversity and accuracy when compared to the filters.
+Simulations demonstrate that the smoothers generate approximations with much improved particle diversity and small increases in accuracy when compared to the equivalent filter-smoothers, which produce only a degenerate particle representation.
 
 \appendices
-\section{The Backwards Kalman Filter} \label{app:backwards_KF}
-
-A backwards Kalman filter \cite{Fraser1969,Kitagawa1994} is used to calculate the joint likelihood term, $p(y_{n+1:N}|x_n, \tilde{\theta}_{n}^+)$, when the state dynamics are linear-Gaussian.
+%\section{The Backwards Kalman Filter} \label{app:backwards_KF}
+%
+%A backwards Kalman filter \cite{Fraser1969,Kitagawa1994} is used to calculate the joint likelihood term, $p(y_{n+1:N}|x_n, \tilde{\theta}_{n}^+)$, when the state dynamics are linear-Gaussian.
 Assume that at time $t_{n+1}$ there is a Gaussian density such that,
-%
-\begin{equation}
- p(y_{n+1:N}|x_{n+1}, \tilde{\theta}_{n}^+) = \tilde{Z}_{n+1} \mathcal{N}(x_{n+1}|\tilde{m}_{n+1}, \tilde{P}_{n+1}) \nonumber    .
-\end{equation}
-
-The state transition model is linear-Gaussian, so the backwards predictive density is given by,
-%
-\begin{IEEEeqnarray}{rCl}
-\IEEEeqnarraymulticol{3}{l}{p(y_{n+1:N}|x_n, \tilde{\theta}_{n}^+)} \nonumber \\
-  & = & \int p(y_{n+1:N}|x_{n+1}, \tilde{\theta}_{n}^+)  p(x_{n+1}|x_n, \tilde{\theta}_{n}^+) dx_{n+1} \nonumber \\
-  & = & \int \tilde{Z}_{n+1} \mathcal{N}(x_{n+1}|\tilde{m}_{n+1}, \tilde{P}_{n+1}) \nonumber \\
-  &   & \qquad \mathcal{N}(x_{n+1}|A_{n+1} x_{n}, Q_{n+1}) dx_{n+1} \nonumber \\
-  & = & \int \tilde{Z}_{n+1} ||A_{n+1}^{-1}|| \mathcal{N}(x_{n+1}|\tilde{m}_{n+1}, \tilde{P}_{n+1}) \nonumber \\
-  &   & \qquad \mathcal{N}(x_{n}|A_{n+1}^{-1} x_{n+1}, A_{n+1}^{-1} Q_{n+1} A_{n+1}^{-T}) dx_{n+1} \nonumber \\
-  & = & \tilde{Z}_{n}^- \mathcal{N}(x_n|\tilde{m}_n^-, \tilde{P}_n^-) \nonumber
-\end{IEEEeqnarray}
-
-where,
-%
-\begin{IEEEeqnarray}{rCl}
- \tilde{m}_n^- & = & A_{n+1}^{-1} \tilde{m}_{n+1} \nonumber \\
- \tilde{P}_n^- & = & A_{n+1}^{-1} (\tilde{P}_{n+1} + Q_{n+1}) A_{n+1}^{-T} \nonumber \\
- \tilde{Z}_n^- & = & \tilde{Z}_{n+1} ||A_{n+1}||^{-1} \nonumber   .
-\end{IEEEeqnarray}
-
-$||A||$ denotes the magnitude of the determinant of $A$, and it is assumed that $A$ is non-singular.
-
-The update step simply involves multiplying by the likelihood of the observation at time $t_n$,
-%
-\begin{IEEEeqnarray}{rCl}
-\IEEEeqnarraymulticol{3}{l}{p(y_{n:N}|x_n, \tilde{\theta}_{n}^+)} \nonumber \\
- & = & p(y_{n+1:N}|x_n, \tilde{\theta}_{n}^+) p(y_{n}|x_n, \tilde{\theta}_{n}^+) \nonumber \\
- & = & \tilde{Z}_n^- \mathcal{N}(x_n|\tilde{m}_n^-, \tilde{P}_n^-) \mathcal{N}(y_n|C_n x_n, R_n) \nonumber \\
- & = & \tilde{Z}_n \mathcal{N}(x_n|\tilde{m}_n, \tilde{P}_n) \nonumber     ,
-\end{IEEEeqnarray}
-
-where,
-%
-\begin{IEEEeqnarray}{rCl}
- \tilde{\mu}_n & = & C_n \tilde{m}_n^-  \nonumber \\
- \tilde{S}_n   & = & C_n \tilde{P}_n^- C_n^T + R_n  \nonumber \\
- \tilde{G}_n   & = & \tilde{P}_n^- C_n^T \tilde{S}_n^{-1} \nonumber \\
- \tilde{m}_n   & = & \tilde{m}_n^- + \tilde{G}_n (y_n - \tilde{\mu}_n) \nonumber \\
- \tilde{P}_n   & = & \tilde{P}_n^- - \tilde{G}_n \tilde{S}_n \tilde{G}_n^T \nonumber \\
- \tilde{Z}_n   & = & \tilde{Z}_n^- \mathcal{N}(y_n|\tilde{\mu}_n, \tilde{S}_n) \nonumber   .
-\end{IEEEeqnarray}
-
-All steps use only standard Gaussian identities.
-
-\section{Initialisation of the Backwards Kalman Filter} \label{app:init_backwards_KF}
-
-The backwards Kalman filter recursion may be initialised exactly by inversion of the observation model using the method of \cite{Kitagawa1994}.
+%%
+%\begin{equation}
+% p(y_{n+1:N}|x_{n+1}, \tilde{\theta}_{n}^+) = \tilde{Z}_{n+1} \mathcal{N}(x_{n+1}|\tilde{m}_{n+1}, \tilde{P}_{n+1}) \nonumber    .
+%\end{equation}
+%
+%The state transition model is linear-Gaussian, so the backwards predictive density is given by,
+%%
+%\begin{IEEEeqnarray}{rCl}
+%\IEEEeqnarraymulticol{3}{l}{p(y_{n+1:N}|x_n, \tilde{\theta}_{n}^+)} \nonumber \\
+%  & = & \int p(y_{n+1:N}|x_{n+1}, \tilde{\theta}_{n}^+)  p(x_{n+1}|x_n, \tilde{\theta}_{n}^+) dx_{n+1} \nonumber \\
+%  & = & \int \tilde{Z}_{n+1} \mathcal{N}(x_{n+1}|\tilde{m}_{n+1}, \tilde{P}_{n+1}) \nonumber \\
+%  &   & \qquad \mathcal{N}(x_{n+1}|A_{n+1} x_{n}, Q_{n+1}) dx_{n+1} \nonumber \\
+%  & = & \int \tilde{Z}_{n+1} ||A_{n+1}^{-1}|| \mathcal{N}(x_{n+1}|\tilde{m}_{n+1}, \tilde{P}_{n+1}) \nonumber \\
+%  &   & \qquad \mathcal{N}(x_{n}|A_{n+1}^{-1} x_{n+1}, A_{n+1}^{-1} Q_{n+1} A_{n+1}^{-T}) dx_{n+1} \nonumber \\
+%  & = & \tilde{Z}_{n}^- \mathcal{N}(x_n|\tilde{m}_n^-, \tilde{P}_n^-) \nonumber
+%\end{IEEEeqnarray}
+%
+%where,
+%%
+%\begin{IEEEeqnarray}{rCl}
+% \tilde{m}_n^- & = & A_{n+1}^{-1} \tilde{m}_{n+1} \nonumber \\
+% \tilde{P}_n^- & = & A_{n+1}^{-1} (\tilde{P}_{n+1} + Q_{n+1}) A_{n+1}^{-T} \nonumber \\
+% \tilde{Z}_n^- & = & \tilde{Z}_{n+1} ||A_{n+1}||^{-1} \nonumber   .
+%\end{IEEEeqnarray}
+%
+%$||A||$ denotes the magnitude of the determinant of $A$, and it is assumed that $A$ is non-singular.
+%
+%The update step simply involves multiplying by the likelihood of the observation at time $t_n$,
+%%
+%\begin{IEEEeqnarray}{rCl}
+%\IEEEeqnarraymulticol{3}{l}{p(y_{n:N}|x_n, \tilde{\theta}_{n}^+)} \nonumber \\
+% & = & p(y_{n+1:N}|x_n, \tilde{\theta}_{n}^+) p(y_{n}|x_n, \tilde{\theta}_{n}^+) \nonumber \\
+% & = & \tilde{Z}_n^- \mathcal{N}(x_n|\tilde{m}_n^-, \tilde{P}_n^-) \mathcal{N}(y_n|C_n x_n, R_n) \nonumber \\
+% & = & \tilde{Z}_n \mathcal{N}(x_n|\tilde{m}_n, \tilde{P}_n) \nonumber     ,
+%\end{IEEEeqnarray}
+%
+%where,
+%%
+%\begin{IEEEeqnarray}{rCl}
+% \tilde{\mu}_n & = & C_n \tilde{m}_n^-  \nonumber \\
+% \tilde{S}_n   & = & C_n \tilde{P}_n^- C_n^T + R_n  \nonumber \\
+% \tilde{G}_n   & = & \tilde{P}_n^- C_n^T \tilde{S}_n^{-1} \nonumber \\
+% \tilde{m}_n   & = & \tilde{m}_n^- + \tilde{G}_n (y_n - \tilde{\mu}_n) \nonumber \\
+% \tilde{P}_n   & = & \tilde{P}_n^- - \tilde{G}_n \tilde{S}_n \tilde{G}_n^T \nonumber \\
+% \tilde{Z}_n   & = & \tilde{Z}_n^- \mathcal{N}(y_n|\tilde{\mu}_n, \tilde{S}_n) \nonumber   .
+%\end{IEEEeqnarray}
+%
+%All steps use only standard Gaussian identities.
+
+%\section{Initialisation of the Backwards Kalman Filter} \label{app:init_backwards_KF}
+%
+%The backwards Kalman filter recursion may be initialised exactly by inversion of the observation model using the method of \cite{Kitagawa1994}.
 By stacking up the last $L+1$ observations into a single vector, it is possible to write down a joint Gaussian distribution conditional on the $(N-L)$th state.
 This may be inverted to produce an improper Gaussian density for $x_{N-L}$,
-%
-\begin{IEEEeqnarray}{rCl}
- \IEEEeqnarraymulticol{3}{l}{p(y_{N-L:N}|x_{N-L}) = \mathcal{N}( \mathbf{y}_{N-L:N} | H_{N-L} x_{N-L}, \Gamma_{N-L} ) } \nonumber \\
- & = & \sqrt{\frac{|V_{N-L}|}{|\Gamma_{N-L}|}} \mathcal{N}(x_{N-L} | \nu_{N-L}, V_{N-L} ) \nonumber
-\end{IEEEeqnarray}
-\begin{IEEEeqnarray}{rCl}
- V_{N-L} & = & (H_{N-L}^T \Gamma_{N-L}^{-1} H_{N-L})^{-1} \nonumber \\
- \nu_{N-L} & = & V_{N-L} H_{N-L}^T \Gamma_{N-L}^{-1} \mathbf{y}_{N-L:N} \nonumber     ,
-\end{IEEEeqnarray}
-
-where $\mathbf{y}_{N-L:N}$ is a vector of concatenated observations, and $H_{N-L}$ and $\Gamma_{N-L}$ may be constructed from the system models,
-%
-\begin{IEEEeqnarray}{rCl}
-\mathbf{y}_{N-L:N} & = & \begin{bmatrix}y_N \\ y_{N-1} \\ \vdots \\ y_{N-L} \end{bmatrix} \nonumber \\
+%%
+%\begin{IEEEeqnarray}{rCl}
+% \IEEEeqnarraymulticol{3}{l}{p(y_{N-L:N}|x_{N-L}) = \mathcal{N}( \mathbf{y}_{N-L:N} | H_{N-L} x_{N-L}, \Gamma_{N-L} ) } \nonumber \\
+% & = & \sqrt{\frac{|V_{N-L}|}{|\Gamma_{N-L}|}} \mathcal{N}(x_{N-L} | \nu_{N-L}, V_{N-L} ) \nonumber
 %\end{IEEEeqnarray}
 %\begin{IEEEeqnarray}{rCl}
-H_{N-L} & = & \begin{bmatrix} C A^L \\ C A^{L-1} \\ \vdots \\ C \end{bmatrix} \nonumber \\
+% V_{N-L} & = & (H_{N-L}^T \Gamma_{N-L}^{-1} H_{N-L})^{-1} \nonumber \\
+% \nu_{N-L} & = & V_{N-L} H_{N-L}^T \Gamma_{N-L}^{-1} \mathbf{y}_{N-L:N} \nonumber     ,
 %\end{IEEEeqnarray}
+%
+%where $\mathbf{y}_{N-L:N}$ is a vector of concatenated observations, and $H_{N-L}$ and $\Gamma_{N-L}$ may be constructed from the system models,
+%%
 %\begin{IEEEeqnarray}{rCl}
- \Gamma_{N-1} & = & K_{N-L} \begin{bmatrix} Q_N & 0 & \dots & 0 \\ 0 & Q_{N-1} & \dots & 0 \\ \vdots & \vdots & \ddots & \vdots \\ 0 & 0 & \dots & Q_{N-L+1} \end{bmatrix} K_{N-L}^T \nonumber \\
-  & & + \: \begin{bmatrix} R_N & 0 & \dots & 0 \\ 0 & R_{N-1} & \dots & 0 \\ \vdots & \vdots & \ddots & \vdots \\ 0 & 0 & \dots & R_{N-L} \end{bmatrix} \nonumber \\
+%\mathbf{y}_{N-L:N} & = & \begin{bmatrix}y_N \\ y_{N-1} \\ \vdots \\ y_{N-L} \end{bmatrix} \nonumber \\
+%%\end{IEEEeqnarray}
+%%\begin{IEEEeqnarray}{rCl}
+%H_{N-L} & = & \begin{bmatrix} C A^L \\ C A^{L-1} \\ \vdots \\ C \end{bmatrix} \nonumber \\
+%%\end{IEEEeqnarray}
+%%\begin{IEEEeqnarray}{rCl}
+% \Gamma_{N-1} & = & K_{N-L} \begin{bmatrix} Q_N & 0 & \dots & 0 \\ 0 & Q_{N-1} & \dots & 0 \\ \vdots & \vdots & \ddots & \vdots \\ 0 & 0 & \dots & Q_{N-L+1} \end{bmatrix} K_{N-L}^T \nonumber \\
+%  & & + \: \begin{bmatrix} R_N & 0 & \dots & 0 \\ 0 & R_{N-1} & \dots & 0 \\ \vdots & \vdots & \ddots & \vdots \\ 0 & 0 & \dots & R_{N-L} \end{bmatrix} \nonumber \\
+%%\end{IEEEeqnarray}
+%%\begin{IEEEeqnarray}{rCl}
+% K_{N-L} & = & \begin{bmatrix} C & CA & \dots & CA^{L-1} \\ 0 & C & \dots & CA^{L-2} \\ \vdots & \vdots & \ddots & \vdots \\ 0 & 0 & \dots & C \\ 0 & 0 & \dots & 0 \end{bmatrix} \nonumber
 %\end{IEEEeqnarray}
-%\begin{IEEEeqnarray}{rCl}
- K_{N-L} & = & \begin{bmatrix} C & CA & \dots & CA^{L-1} \\ 0 & C & \dots & CA^{L-2} \\ \vdots & \vdots & \ddots & \vdots \\ 0 & 0 & \dots & C \\ 0 & 0 & \dots & 0 \end{bmatrix} \nonumber
-\end{IEEEeqnarray}
-
-In the above, it is assumed for clarity that $A_n = A$ and $C_n = C$ are constant matrices.
+%
+%In the above, it is assumed for clarity that $A_n = A$ and $C_n = C$ are constant matrices.
 This assumption is easily relaxed.

### 2012-10-10 response_to_reviewers.tex
Provisionally final second version

@@
 
-{\meta Many thanks for the comments, folks.
-I've made changes to address all but one, with which I have a bit of a philosophical disagreement.
-Reviewer 3's comments were particularly useful and prompted me to rethink carefully the purpose of the simulations section.
-Responses are inline with the individual points below.
-}
+{\meta Many thanks for the comments, folks. I've made changes to address them all. Responses are inline with the individual points below. }
 
@@
 There is, of course, much more to a distributional estimate than how close its mean is to the true value, so we do need something more than just the RMSE, as you suggest.
-Ideally, we would like to compare our set of particles to the true posterior distribution, using statistics such as the effective sample size (ESS), or normalised estimation error squared (NEES).
-However this true posterior is not available.
-As a substitute, either a Gaussian approximation or a particle approximation with loads of particles are sometimes used as ``truth''.
-This is far from ideal, and I find the latter particularly unsettling, because we end up testing our particle algorithms only against their own best performance.
-
-My preferred alternative for measuring consistency is a mean ``empirical NEES'' statistic which takes a value between 0 and 1 and takes into account the estimation error, covariance estimate and diversity of the particle approximation.
-I've added details and plots of this metric to the paper.
+One useful and established metric for measuring efficiency is the normalised estimation error squared (NEES), but we can't use this directly, because when the particle estimate is degenerate the NEES is infinite.
+Therefore, I've added a transformed version of the NEES as an efficiency measure.
 I think that together with the RMSE values, this provides sufficient evidence of the improvement provided by the smoother.
@@
 
-{\meta I completely disagree. I think this would be misleading and is unnecessary for demonstrating the purpose of the new smoothers.
+{\meta I think this is a bit misleading really.
+Although it might be a fairer test for a given problem, the result can be changed by modifying the scenario.
+For example, the relative performance of the smoother will be improved by having a longer data set.
+All I want to demonstrate in the simulations section is that the smoother operates as expected --- producing rejuvenated particle sets.
+The better algorithm for any application will depend on the model, the length of the data, and the desired output (e.g.
+is a distributional estimate needed, or is a single sample enough?).
 
-For a start, the filter-smoother and the smoother provide estimates with different qualities, and comparing them like this would be to compare apples with oranges.
-The principal reason for using a particle smoother is to rejuvenate a particle population with low path-space diversity.
-This is demonstrated by the increased number of unique particles and the improved consistency.
-(The reduced RMSE is a consequent bonus.) In contrast, [Fearnhead et al.
+In contrast to us, [Fearnhead et al.
 2010] are comparing a variety of smoothing algorithms all of which do the same thing, produce a diverse set of marginal smoothing particles.
-Their methods are not appropriate here.
+Simple changes like altering the data length are less likely to affect their results, so their methods are appropriate to their purpose, but less so here.
 
-Furthermore, if we were trying to solve a particular practical problem, such as producing the best state estimator possible for an offline tracker, then it would be a sensible test.
-However, the example simulations here are entirely fictional, cooked up simply to demonstrate that the new smoothing algorithm functions as expected.
-The problem is that whatever this test demonstrated about relative performance, the result could be reversed by changing the set-up of the simulation.
-}
+However, Simon agrees with you. So i have actually changed the number of particles in the comparisons based on this equal-computation principle. }
 

### 2012-10-10 preprint_VRsmoothing.tex
Provisionally final second version

--- 2012-09-13 revisions_VRsmoothing.tex
@@
-\documentclass[10pt,twocolumn,twoside]{IEEEtran}
-
+\documentclass{article}
+
+\usepackage{IEEEtrantools}
 \usepackage{cite}
@@
 
-\author{Pete~Bunch*,~\IEEEmembership{Member,~IEEE,}
-        Simon~Godsill,~\IEEEmembership{Member,~IEEE,}% <-this % stops a space
-\thanks{P. Bunch and S. Godsill are with the Department
-of Engineering, Cambridge University, UK. email: \{pb404,sjg30\}@cam.ac.uk}% <-this % stops a space
-}%\thanks{Manuscript received January 01, 1901; revised January 02, 1901.}}
-
-% The paper headers
-\markboth{IEEE Transaction in Signal Processing}{IEEE Transaction in Signal Processing}%,~Vol.~1, No.~1, January~1901}%
-%{Bunch \& Godsill: Particle Smoothing Algorithms for Variable Rate Models}
+\author{Pete~Bunch and Simon~Godsill%
+\thanks{Department of Engineering, Cambridge University, UK. email: \{pb404,sjg30\}@cam.ac.uk}}
 
@@
 
-\begin{IEEEkeywords}
-Bayesian inference, state-space model, variable rate, particle filter, filtering, smoothing
-\end{IEEEkeywords}
-
 \section{Introduction}
 
-\IEEEPARstart{T}{he} objective of sequential Bayesian inference is to estimate an imperfectly observed quantity as it varies over time.
+The objective of sequential Bayesian inference is to estimate an imperfectly observed quantity as it varies over time.
 This is accomplished through the use of probabilistic models for the state evolution and measurement processes.
@@
 \begin{IEEEeqnarray}{rCl}
- \text{MENEES}_k & = & (\hat{x}_k - x_k^*)^T \hat{P}_k (\hat{x}_k - x_k^*) \nonumber \\ 
- \hat{x}_k & = & \sum_i x_{k}^{(i)} \nonumber \\
- \hat{P}_k & = & \sum_i (x_{k}^{(i)} - x_k^*) (x_{k}^{(i)} - x_k^*)^T \nonumber     ,
+ \text{MENEES}_k & = & (\hat{x}_k - x_k^*)^T \hat{P}_k^{-1} (\hat{x}_k - x_k^*) \nonumber \\
+ \hat{x}_k & = & \frac{1}{N_S} \sum_i x_{k}^{(i)} \nonumber \\
+ \hat{P}_k & = & \frac{1}{N_S} \sum_i (x_{k}^{(i)} - x_k^*) (x_{k}^{(i)} - x_k^*)^T \nonumber     ,
 \end{IEEEeqnarray}
@@
 
-\appendices
+\appendix
 %\section{The Backwards Kalman Filter} \label{app:backwards_KF}
@@
 \end{document}
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

### 2012-10-11 response_to_reviewers.tex
Updated response to reviewers

@@
 
-{\meta Many thanks for the comments, folks. I've made changes to address them all. Responses are inline with the individual points below. }
+{\meta Thankyou for the comments, all of which have been addressed. Responses are inline with the individual points below. }
 
@@
 
-{\meta I've added a brief note on convergence, which is inherited from the basic results for the forward-filtering-backward-sampling algorithm.
-I've also added an explicit example in the simulations of the smoother representing multi-modality where some particle contain a jump and others don't.
-Fixed interval smoothing won't help online filtering.
+{\meta Added a brief note on convergence, which is inherited from the basic results for the forward-filtering-backward-sampling algorithm.
+Added an explicit example in the simulations of the smoother representing multi-modality where some particle contain a jump and others do not.
+Fixed interval smoothing will not help online filtering.
 Fixed-lag smoothing would, and this is covered elsewhere, especially in [Whitely et al.
@@
 track loss) then the smoother will also fail.
-I have added an example where the multi-modality of the particle representation is lost due to resampling in the filter-smoother, but not the filter.
+Added an example where the multi-modality of the particle representation is lost due to resampling in the filter-smoother, but not the filter.
 }
@@
 
-{\meta No file attached, so I assume there are genuinely no comments. }
+{\meta No file attached. }
 
@@
 
-{\meta I feel this is clear from the introduction and the maths, but it's now explicitly stated as well. }
+{\meta This should be clear from the introduction and the maths, but it is now explicitly stated as well. }
 
@@
 
-{\meta These two comments concern the same question: How should particle approximations of a distribution be compared?
-There is, of course, much more to a distributional estimate than how close its mean is to the true value, so we do need something more than just the RMSE, as you suggest.
-One useful and established metric for measuring efficiency is the normalised estimation error squared (NEES), but we can't use this directly, because when the particle estimate is degenerate the NEES is infinite.
-Therefore, I've added a transformed version of the NEES as an efficiency measure.
-I think that together with the RMSE values, this provides sufficient evidence of the improvement provided by the smoother.
-}
+{\meta These two comments have been addressed by replacing the ``number of particles'' comparison with normalised estimation error squared (NEES) statistics.
+When the number of particles is very low, as is the case for the filter-smoother, covariance estimates become degenerate and the NEES is infinite.
+Therefore, a transformed version of the NEES is used.
+These show a better efficiency of the smoother relative to the filter-smoother.
+This is based on very similar principles to the test suggested (i.e.
+comparing Mahalanobis distances).}
 
@@
 
-{\meta I think this is a bit misleading really.
+{\meta %I disagree with this.
 Although it might be a fairer test for a given problem, the result can be changed by modifying the scenario.
 For example, the relative performance of the smoother will be improved by having a longer data set.
-All I want to demonstrate in the simulations section is that the smoother operates as expected --- producing rejuvenated particle sets.
+The objective of the simulations section is to demonstrate that the smoother operates as expected --- producing rejuvenated particle sets.
 The better algorithm for any application will depend on the model, the length of the data, and the desired output (e.g.
@@
 
-In contrast to us, [Fearnhead et al.
+%In contrast to us, [Fearnhead et al.
 2010] are comparing a variety of smoothing algorithms all of which do the same thing, produce a diverse set of marginal smoothing particles.
-Simple changes like altering the data length are less likely to affect their results, so their methods are appropriate to their purpose, but less so here.
+Simple changes like altering the data length are less likely to affect their comparisons, so their methods are appropriate to their purpose, but less so here.
 
-However, Simon agrees with you. So i have actually changed the number of particles in the comparisons based on this equal-computation principle. }
+%All this said, increasing the number of particles used by the filter-smoother so as to equalise processing time does very little to improve its performance, so we've done it anyway.
+
+Done.
+
+}
 
@@
 
-{\meta Yes. I've added graphs showing the effect of varying the number of smoothing trajectories for the tracking case. The finance results are similar and so omitted for space. }
+{\meta Yes. Added graphs showing the effect of varying the number of smoothing trajectories for the tracking case. The finance results are similar and so omitted for space. }
 
@@
 
-{\meta I've added a reference to [Douc et al., 2011].
-This is similar to our Metropolis-Hastings-based method which is already referenced (but unhelpfully so as it is still awaiting publication).
+{\meta Added a reference to [Douc et al., 2011].
+This is similar to our Metropolis-Hastings-based method which is already referenced (still awaiting publication).
 ([Bunch and Godsill, 2012]).
-The Fearnhead method is for marginal (as opposed to joint) smoothing distributions only, so it doesn't really help.
-I've mentioned this in section IV.
+The Fearnhead method is for marginal (as opposed to joint) smoothing distributions only, so it is not relevant here.
+This is now mentioned in section IV.
 }
@@
 {\meta I agree.
-I've cut the first two, giving the derivation and initialisation of the two-filter smoother.
+Removed the first two, giving the derivation and initialisation of the two-filter smoother.
 The discretisation of the finance model remains, as I can't find a complete and equivalent version in a published work.

### 2012-10-17 second_submit_doublecol_VR_smoothing.tex
Second submission

--- 2012-09-13 revisions_VRsmoothing.tex
@@
 For a better characterisation of the smoothing distribution, it is necessary to rejuvenate the set of particles.
-This is achieved here with a backward pass through the observations in an extension of the forward-filtering-backward-sampling (FFBS) method described in \cite{Godsill2004}.
-Since these algorithms are based on the FFBS algorithm, the convergence proofs of \cite{Godsill2004,Douc2011} are believed to apply.
+This is achieved here with a backward pass through the observations in an extension of the forward-filtering-backward-sampling (FFBS) method described in \cite{Godsill2004}.% Since these algorithms are based on the FFBS algorithm, the convergence results of \cite{Godsill2004,Douc2011} are believed to apply.
 
@@
 It is possible to reduce this by using a rejection sampling \cite{Douc2011} or MCMC \cite{Bunch2012} sampling scheme, which avoids the necessity of calculating the sampling weights for every filter particle.
+Since the algorithm is based on the FFBS algorithm, the convergence results of \cite{Godsill2004,Douc2011} are believed to apply.
 
@@
 These can be calculated by simply taking a mean of the particle values (using the Gaussian means for the linear-Gaussian case).
-The accuracy is then assessed by comparing the size of estimation errors of the resulting estimators.
+The accuracy is then assessed by comparing the size of errors of the resulting estimators.
 In addition, the benefit of employing a smoothing algorithm is that it ensures a more diverse particle representation of the posterior state distribution.
-To demonstrate this effect, a measure of estimator consistency is calculated by which to compare the algorithms, the mean empirical normalised estimation error squared (MENEES).
-This is defined as,  %the number of unique state samples is plotted over time in the following examples.
-A
-%
-\begin{IEEEeqnarray}{rCl}
- \text{MENEES}_k & = & (\hat{x}_k - x_k^*)^T \hat{P}_k (\hat{x}_k - x_k^*) \nonumber \\ 
- \hat{x}_k & = & \sum_i x_{k}^{(i)} \nonumber \\
- \hat{P}_k & = & \sum_i (x_{k}^{(i)} - x_k^*) (x_{k}^{(i)} - x_k^*)^T \nonumber     ,
-\end{IEEEeqnarray}
-
-where $x_k^*$ is the true state, and $x_k^{(i)}$ are the individual particle states.
-This statistic takes a value between $0$ and $1$, where $1$ is scored by a degenerate particle estimate (fewer linearly independent particles than state dimensions).
-For a non-degenerate estimate, smaller values correspond to smaller estimation errors relative to the estimated covariance.
-Thus, it incorporates the diversity of the particle set as well the accuracy of the estimate and covariance.
+To demonstrate this effect, a measure of efficiency is calculated by which to compare the algorithms.
+The normalised estimation error squared (NEES) \cite{Bar-Shalom1995,Briers2003} is a statistic derived from the chi-squared test.
+It is defined by,
+%
+\begin{IEEEeqnarray}{rCl}
+ \text{NEES}_k & = & (\hat{x}_k - x_k^*)^T \hat{P}_k^{-1} (\hat{x}_k - x_k^*) \nonumber \\
+ \hat{x}_k & = & \frac{1}{N_S} \sum_i x_{k}^{(i)} \nonumber \\
+ \hat{P}_k & = & \frac{1}{N_S} \sum_i (x_{k}^{(i)} - \hat{x}_k) (x_{k}^{(i)} - \hat{x}_k)^T \nonumber     ,
+\end{IEEEeqnarray}
+
+where $x_k^*$ is the true state value and $x_k^{(i)}$ are the individual particle states.
+Calculating NEES for particle approximations which might be degenerate is problematic, because $\hat{P}_k$ will not be invertible (resulting in an infinite value).
+Here we map the interval $(0, \infty)$ onto $(0, 1)$ using the following transformation, giving a transformed NEES (TNEES) score.
+%
+\begin{IEEEeqnarray}{rCl}
+ \text{TNEES}_k & = &\frac{\text{NEES}_k}{1+\text{NEES}_k}     .
+\end{IEEEeqnarray}
+
+Assuming Gaussian distributed estimation errors, it can be shown that NEES should obey a chi-squared distribution with the degrees of freedom parameter equal to the number of state dimensions.
+Under this assumption, an optimal value of mean TNEES can be found using the transformation above and compared to the average over a large number of Monte Carlo runs.
+The integral required to calculate this mean is intractable but may be accurately approximated by simulation.
 
@@
 
-The filter used $\aleph_F = 100$ particles, and the smoother resampled $\aleph_S = 100$ sequences.
+The filter and filter-smoother were run using using $\aleph_F = 1000$ particles, while the smoother used $\aleph_F = 100$ particles and was used to simulate $\aleph_S = 100$ smoothed paths.
+These values lead to roughly equal running time for the smoother and filter-smoother.
 Bootstrap proposals were used for the filter.
-An example forward realisation simulated from the model is shown in figure~\ref{fig:example_data}.
-The sampled trend trajectories produced by the filter-smoother and VRPS are shown in figure~\ref{fig:example_state} (the equivalent graphs for value are not particularly informative as this quantity is observed, and the errors produced by both algorithms are small).
+An example forward realisation simulated from the model is shown in Fig.~\ref{fig:example_data}.
+The sampled trend trajectories produced by the filter-smoother and VRPS are shown in Fig.~\ref{fig:example_state} (the equivalent graphs for value are not particularly informative as this quantity is observed, and the errors produced by both algorithms are small).
 Notice that the path diversity is dramatically improved by the VRPS.
@@
 At each observation time, an estimate of the value and trend is made by taking the average of Gaussian means generated by an RTS smoother for each particle.
-The resulting mean absolute error over all scenarios is shown in figure~\ref{fig:finance_rmse}, demonstrating a small improvement in accuracy from the smoother.
-More significantly, MENEES is shown in figure~\ref{fig:finance_menees}, showing a similar consistency between filter and smoother.
-The filter-smoother produces an MENEES of $1$ at almost all time instants, because there is only one unique particle in the set.% are displayed in table~\ref{tab:finance_state_performance}
+The resulting mean absolute error over all scenarios is shown in Fig.~\ref{fig:finance_rmse}, demonstrating a small improvement in accuracy from the smoother.
+Mean TNEES is shown in Fig.~\ref{fig:finance_mtnees}, showing a similar efficiency between filter and smoother.
+The filter-smoother produces a mean TNEES close to $1$ at most time instants, because there is often only one unique particle in the set.% are displayed in table~\ref{tab:finance_state_performance}
 %
@@
 \centering
-\includegraphics[width=0.9\columnwidth]{finance_menees.pdf}
-\caption{MENEES over 100 scenarios using filter, filter-smoother, and smoother.}
-\label{fig:finance_menees}
+\includegraphics[width=0.9\columnwidth]{finance_mtnees.pdf}
+\caption{Mean TNEES over 100 scenarios using filter, filter-smoother, and smoother. Optimum value: 0.54}
+\label{fig:finance_mtnees}
 \end{figure}
@@
 As well as providing more accurate and more diverse estimates, the smoother is also capable of preserving multi-modality in its representation of the posterior distribution, which is lost by the filter smoother due to the particle collapse to a single value.
-This is highlighted in figure~\ref{fig:finance_multimodality}, which shows a segment of the trend estimates from figure~\ref{fig:example_state} containing a small jump.
+This is highlighted in Fig.~\ref{fig:finance_multimodality}, which shows the trend estimates from a segment of simulated data containing a small jump.
 In the filter-smoother results, none of the particles have a jump at this point.
@@
 
-The filter and smoother were each used to generate $50$ particles, with the filter employing resample-move steps with proposals based on an unscented transform \cite{Julier2004} approximation to the optimal proposal density, and the smoother using $M=5$ steps in each Markov chain.
+The filter and filter-smoother were run using using $\aleph_F = 500$ particles, while the smoother used $\aleph_F = 50$ particles and was used to simulate $\aleph_S = 50$ smoothed paths.
+These values lead to roughly equal running time for the smoother and filter-smoother.
+The filter employed resample-move steps with proposals based on an unscented transform \cite{Julier2004} approximation to the optimal proposal density, and the smoother using $M=5$ steps in each Markov chain.
 
@@
 At each observation time, an estimate of the position and velocity is made by taking the average of values from the array of particles.
-The resulting mean absolute error over all scenarios is shown in figure~\ref{fig:finance_rmse}, again demonstrating a small improvement in accuracy from the smoother.
+The resulting mean absolute error over all scenarios is shown in Fig.~\ref{fig:tracking_rmse}, again demonstrating a small improvement in accuracy from the smoother.
 The smoother also provides a more accurate estimate of the number of changepoints in each test, with an average error of 0.6 compared to 2.0 from the filter-smoother.
-MENEES is displayed in figure~\ref{fig:tracking_menees}, showing a similar consistency between filter and smoother.
-The filter-smoother produces an MENEES of $1$ at almost all time instants, because there is only one unique particle in the set.
+Mean TNEES is displayed in Fig.~\ref{fig:tracking_mtnees}.
+It shows that while the smoother does not match the efficiency of the filter, it is a significant improvement over the filter-smoother, which has a Mean TNEES of $1$ at almost all time instants, because there is only one unique particle in the set.
 %in table~\ref{tab:tracking_state_performance}, demonstrating an improvement in accuracy from the smoother.
@@
 \centering
-\includegraphics[width=0.9\columnwidth]{tracking_menees.pdf}
-\caption{MENEES over 100 scenarios using filter, filter-smoother, and smoother.}
-\label{fig:tracking_menees}
+\includegraphics[width=0.9\columnwidth]{tracking_mtnees.pdf}
+\caption{Mean TNEES over 100 scenarios using filter, filter-smoother, and smoother. Optimum value: 0.73}
+\label{fig:tracking_mtnees}
 \end{figure}
@@
 %
-Finally, in figure~\ref{fig:varying_num_pts} the variation in accuracy and consistency of the smoothing particle approximation is demonstrated as the number of smoothing particles varies.
+Finally, in Fig.~\ref{fig:varying_num_pts} the variation in accuracy and efficiency of the smoothing particle approximation is demonstrated as the number of smoothing particles varies.
 As expected, the rate of improvement diminishes and becomes negligible as the number of smoothing particles exceeds the number of filtering particles.
 %
-\begin{figure}[!t]
+\begin{figure}
 \centering
 \subfloat[]{\includegraphics[width=0.45\columnwidth]{pos_RMSE_vs_numpts.pdf}}
-\subfloat[]{\includegraphics[width=0.45\columnwidth]{MENEES_vs_numpts.pdf}}
-\caption{Average position error (a) and MENEES (b) over 100 scenarios with varying number of smoother particles and 50 filter particles.}
+\subfloat[]{\includegraphics[width=0.45\columnwidth]{MTNEES_vs_numpts.pdf}}
+\caption{Average position error (a) and Mean TNEES (b) over 100 scenarios with varying number of smoother particles and 50 filter particles.}
 \label{fig:varying_num_pts}

### 2013-01-15 13-01-14_final_submission/VRsmoothing.tex
Final submitted version

--- 2012-10-17 second_submit_doublecol_VR_smoothing.tex
@@
-\documentclass[10pt,twocolumn,twoside]{IEEEtran}
+%\documentclass[10pt,twocolumn,twoside]{IEEEtran}
+\documentclass[peerreview,11pt,draftcls,onecolumn]{IEEEtran}
 
@@
 \graphicspath{{./images/}{./images/finance/}{./images/tracking/}}
-%\interdisplaylinepenalty=1000
 
@@
         Simon~Godsill,~\IEEEmembership{Member,~IEEE,}% <-this % stops a space
+\thanks{Copyright (c) 2012 IEEE.
+Personal use of this material is permitted.
+However, permission to use this material for any other purposes must be obtained from the IEEE by sending a request to pubs-permissions@ieee.org.}%
 \thanks{P. Bunch and S. Godsill are with the Department
@@
 
-This closely resembles the variable-dimension spaces employed by reversible jump Markov chain Monte Carlo (MCMC) algorithms \cite{Green1995}.% or in multiple target tracking scenarios \cite{Vo2005}.
+This closely resembles the variable-dimension spaces employed by reversible jump Markov chain Monte Carlo (MCMC) algorithms \cite{Green1995}.
 
@@
 
-This model may be discretised at the observation times using standard methods, detailed in appendix~\ref{app:lg_model_discretisation}, leading to,% (see \cite{Godsill2007a,Christensen2012,Grewal2002}), leading to,
-%
-%Assuming observations of value only and Gaussian observation noise with standard deviation $\sigma_y^2$, the resulting discrete time dynamics are described by the following equations:% (see appendix \ref{app:model_discretisation}):
+This model may be discretised at the observation times using standard methods, detailed in appendix~\ref{app:lg_model_discretisation}, leading to,
 %
@@
 \end{IEEEeqnarray}
-%\begin{IEEEeqnarray}{rCl}
-%\IEEEeqnarraymulticol{3}{l}{p(\theta_{n \setminus n-1} | \theta_{n-1})} \nonumber \\
-%    & = & p(\tau_{K_{n-1}+1:K_n}, u_{K_{n-1}+1:K_n}, \tau_{K_n+1}>t_n|\tau_{K_{n-1}+1}>t_{n-1}, \tau_{K_{n-1}}, u_{K_{n-1}}) \nonumber \\
-%    & = & \begin{cases} S(\tau_{K_n}, t_n) \prod\limits_{j:t_{n-1} \leq \tau_j < t_n} p(\tau_j, u_j| \tau_{j-1}, u_{j-1}, \tau_j>t_{n-1}) & K_n > K_{n-1} \\
-%                        S(\tau_{K_n}, t_n) / S(\tau_{K_n}, t_{n-1}) & K_n = K_{n-1} \end{cases} \IEEEeqnarraynumspace \label{eq:cp_sequence_trandens}
-%\end{IEEEeqnarray}
-
-%For all but the first changepoint in the interval, the density is given by the prior model of (\ref{eq:cp_model}).
-For the first changepoint, indexed by $k=K_{n-1}+1$, we must account for the fact that it cannot occur before $t_{n-1}$,
-%%
-%\begin{IEEEeqnarray}{rCl}
-%\IEEEeqnarraymulticol{3}{l}{p(\tau_{k}, u_{k}| \tau_{k-1}, u_{k-1}, \tau_{k}>t_{n-1})} \nonumber \\
-%  & = & \frac{1}{S(\tau_{k-1}, t_{n-1})} \begin{cases} p(\tau_{k}, u_{k}| \tau_{k-1}, u_{k-1}) & \tau_{k} > t_{n-1} \\ 0 & \tau_{k} < t_{n-1} \end{cases}  \IEEEeqnarraynumspace \label{eq:cp_cond_model}   .
-%\end{IEEEeqnarray}
 
@@
 This final future changepoint is discarded.
-(This process can be thought of as sampling an entire future changepoint sequence from $t_{n-1}$ onwards, and then marginalising those which fall after $t_n$.) The bootstrap proposal leads to the usual simplification of the weight formula, % (apart from the first which is sampled from (\ref{eq:cp_cond_model}))
-%
-\begin{IEEEeqnarray}{rCl}
-w_n^{(i)} & = & \frac{w_{n-1}^{(i)}}{v_{n-1}^{(i)}} \times p(y_n|\hat{x}_n) \label{eq:bootstrap_vrpf_weights}     .
-\end{IEEEeqnarray}
-
-%The choice of proposal weights, $\{v_{n-1}^{(i)}\}$, requires particular attention in the design of VRPFs.
-In some models a changepoint may not have an immediate effect on the observations, especially if a jump occurs in some quantity which is only observed via its integral, e.g.
-if there is a jump in the acceleration of a moving object, yet only the position is measured, the change will not be apparent until several more observations have been made.
-In the meantime, particles which contain a changepoint at the correct time may all have been removed by the resampling process.
-To avoid this loss of good particles, proposal weights should be chosen which preserve a significant number of low-weight particles.
-One scheme which has been found to work well is described in \cite{Godsill2007}, in which proposal weights are given by:
+(This process can be thought of as sampling an entire future changepoint sequence from $t_{n-1}$ onwards, and then marginalising those which fall after $t_n$.) The bootstrap proposal leads to the usual simplification of the weight formula,
+%
+\begin{IEEEeqnarray}{rCl}
+w_n^{(i)} & = & \frac{w_{n-1}^{(i)}}{v_{n-1}^{(i)}} \times p(y_n|\theta_{n}^-, y_{1:n-1}) \label{eq:bootstrap_vrpf_weights}     .
+\end{IEEEeqnarray}
 
@@
 For a better characterisation of the smoothing distribution, it is necessary to rejuvenate the set of particles.
-This is achieved here with a backward pass through the observations in an extension of the forward-filtering-backward-sampling (FFBS) method described in \cite{Godsill2004}.% Since these algorithms are based on the FFBS algorithm, the convergence results of \cite{Godsill2004,Douc2011} are believed to apply.
+This is achieved here with a backward pass through the observations in an extension of the forward-filtering-backward-sampling (FFBS) method described in \cite{Godsill2004}.
 
@@
 Mean TNEES is shown in Fig.~\ref{fig:finance_mtnees}, showing a similar efficiency between filter and smoother.
-The filter-smoother produces a mean TNEES close to $1$ at most time instants, because there is often only one unique particle in the set.% are displayed in table~\ref{tab:finance_state_performance}
+The filter-smoother produces a mean TNEES close to $1$ at most time instants, because there is often only one unique particle in the set.
 %
@@
 %
-%\begin{table}%
-%\caption{State estimation performance.}
-%\label{tab:finance_state_performance}
-%\centering
-%\renewcommand{\arraystretch}{1.5}
-%\begin{tabular}{|c|c|c|c|}
-%\hline
-% &\raisebox{0cm}[0.4cm][0.3cm]{\parbox[c]{1.5cm}{\centering VRPF}} & \raisebox{0cm}[0.4cm][0.3cm]{\parbox[c]{1.5cm}{\centering Filter-smoother}} & \raisebox{0cm}[0.4cm][0.3cm]{\parbox[c]{1.5cm}{\centering VRPS}} \\
-%\hline \hline
-%\raisebox{0cm}[0.4cm][0.3cm]{\parbox[c]{2cm}{\centering mean value estimate RMSE}}   & $5.31 \times 10^{-4}$ & $4.48 \times 10^{-4}$ & $4.16 \times 10^{-4}$ \\
-%\hline
-%\raisebox{0cm}[0.4cm][0.3cm]{\parbox[c]{2cm}{\centering mean trend estimate RMSE}}   & $2.56 \times 10^{-2}$ & $1.79 \times 10^{-2}$ & $1.49 \times 10^{-2}$ \\
-%\hline
-%\end{tabular}
-%\end{table}
-%
-%The mean number of unique particle estimates at each time step is shown in Fig.~\ref{fig:finance_unique_particles}.
-This illustrates the main advantage of the smoother algorithm --- the increased number of unique particles in the approximation means a better characterisation of the posterior changepoint and state distributions.
-%
-%\begin{figure}[!t]
-%\centering
-%\includegraphics[width=0.75\columnwidth]{finance_unique_particles.pdf}
-%\caption{Mean number of unique particles in the filter-smoother (dashed) and smoother (solid) estimates.}
-%\label{fig:finance_unique_particles}
-%\end{figure}
-%
 As well as providing more accurate and more diverse estimates, the smoother is also capable of preserving multi-modality in its representation of the posterior distribution, which is lost by the filter smoother due to the particle collapse to a single value.
@@
 It shows that while the smoother does not match the efficiency of the filter, it is a significant improvement over the filter-smoother, which has a Mean TNEES of $1$ at almost all time instants, because there is only one unique particle in the set.
-%in table~\ref{tab:tracking_state_performance}, demonstrating an improvement in accuracy from the smoother.
 %
@@
 %
-%The mean number of unique particle estimates at each time step is shown in Fig.~\ref{fig:tracking_unique_particles}.
-This illustrates again the principal advantage of the VRPS algorithm --- the increased number of unique particles in the approximation means a better characterisation of the posterior changepoint distribution, allowing, for example, calculation of state covariances.
-%
-%\begin{table}%
-%\caption{State estimation performance.}
-%\label{tab:tracking_state_performance}
-%\centering
-%\renewcommand{\arraystretch}{1.5}
-%\begin{tabular}{|c|c|c|c|}
-%\hline
-% &\raisebox{0cm}[0.4cm][0.3cm]{\parbox[c]{1.5cm}{\centering VRPF}} & \raisebox{0cm}[0.4cm][0.3cm]{\parbox[c]{1.5cm}{\centering Filter-smoother}} & \raisebox{0cm}[0.4cm][0.3cm]{\parbox[c]{1.5cm}{\centering VRPS}} \\
-%% & \begin{minipage}[c]{1.5cm}\centering RBVRPF + KF\end{minipage} & \begin{minipage}[c]{1.5cm}\centering RBVRPF + RTS\end{minipage} & \begin{minipage}[c]{1.5cm}\centering RBVRPS + RTS\end{minipage} \\
-%\hline \hline
-%\raisebox{0cm}[0.4cm][0.3cm]{\parbox[c]{2cm}{\centering mean position estimate RMSE}} & 2.14 & 1.36 & 1.12 \\
-%\hline
-%\raisebox{0cm}[0.4cm][0.3cm]{\parbox[c]{2cm}{\centering mean velocity estimate RMSE}} & 1.89 & 1.03 & 0.85 \\
-%\hline
-%\end{tabular}
-%\end{table}
-%
-%\begin{figure}[!t]
-%\centering
-%\includegraphics[width=0.75\columnwidth]{tracking_unique_particles.pdf}
-%\caption{Mean number of unique particles in the final filter (dashed) and smoother (solid) estimates.}
-%\label{fig:tracking_unique_particles}
-%\end{figure}
-%
 Finally, in Fig.~\ref{fig:varying_num_pts} the variation in accuracy and efficiency of the smoothing particle approximation is demonstrated as the number of smoothing particles varies.
@@
 \appendices
-%\section{The Backwards Kalman Filter} \label{app:backwards_KF}
-%
-%A backwards Kalman filter \cite{Fraser1969,Kitagawa1994} is used to calculate the joint likelihood term, $p(y_{n+1:N}|x_n, \tilde{\theta}_{n}^+)$, when the state dynamics are linear-Gaussian.
-Assume that at time $t_{n+1}$ there is a Gaussian density such that,
-%%
-%\begin{equation}
-% p(y_{n+1:N}|x_{n+1}, \tilde{\theta}_{n}^+) = \tilde{Z}_{n+1} \mathcal{N}(x_{n+1}|\tilde{m}_{n+1}, \tilde{P}_{n+1}) \nonumber    .
-%\end{equation}
-%
-%The state transition model is linear-Gaussian, so the backwards predictive density is given by,
-%%
-%\begin{IEEEeqnarray}{rCl}
-%\IEEEeqnarraymulticol{3}{l}{p(y_{n+1:N}|x_n, \tilde{\theta}_{n}^+)} \nonumber \\
-%  & = & \int p(y_{n+1:N}|x_{n+1}, \tilde{\theta}_{n}^+)  p(x_{n+1}|x_n, \tilde{\theta}_{n}^+) dx_{n+1} \nonumber \\
-%  & = & \int \tilde{Z}_{n+1} \mathcal{N}(x_{n+1}|\tilde{m}_{n+1}, \tilde{P}_{n+1}) \nonumber \\
-%  &   & \qquad \mathcal{N}(x_{n+1}|A_{n+1} x_{n}, Q_{n+1}) dx_{n+1} \nonumber \\
-%  & = & \int \tilde{Z}_{n+1} ||A_{n+1}^{-1}|| \mathcal{N}(x_{n+1}|\tilde{m}_{n+1}, \tilde{P}_{n+1}) \nonumber \\
-%  &   & \qquad \mathcal{N}(x_{n}|A_{n+1}^{-1} x_{n+1}, A_{n+1}^{-1} Q_{n+1} A_{n+1}^{-T}) dx_{n+1} \nonumber \\
-%  & = & \tilde{Z}_{n}^- \mathcal{N}(x_n|\tilde{m}_n^-, \tilde{P}_n^-) \nonumber
-%\end{IEEEeqnarray}
-%
-%where,
-%%
-%\begin{IEEEeqnarray}{rCl}
-% \tilde{m}_n^- & = & A_{n+1}^{-1} \tilde{m}_{n+1} \nonumber \\
-% \tilde{P}_n^- & = & A_{n+1}^{-1} (\tilde{P}_{n+1} + Q_{n+1}) A_{n+1}^{-T} \nonumber \\
-% \tilde{Z}_n^- & = & \tilde{Z}_{n+1} ||A_{n+1}||^{-1} \nonumber   .
-%\end{IEEEeqnarray}
-%
-%$||A||$ denotes the magnitude of the determinant of $A$, and it is assumed that $A$ is non-singular.
-%
-%The update step simply involves multiplying by the likelihood of the observation at time $t_n$,
-%%
-%\begin{IEEEeqnarray}{rCl}
-%\IEEEeqnarraymulticol{3}{l}{p(y_{n:N}|x_n, \tilde{\theta}_{n}^+)} \nonumber \\
-% & = & p(y_{n+1:N}|x_n, \tilde{\theta}_{n}^+) p(y_{n}|x_n, \tilde{\theta}_{n}^+) \nonumber \\
-% & = & \tilde{Z}_n^- \mathcal{N}(x_n|\tilde{m}_n^-, \tilde{P}_n^-) \mathcal{N}(y_n|C_n x_n, R_n) \nonumber \\
-% & = & \tilde{Z}_n \mathcal{N}(x_n|\tilde{m}_n, \tilde{P}_n) \nonumber     ,
-%\end{IEEEeqnarray}
-%
-%where,
-%%
-%\begin{IEEEeqnarray}{rCl}
-% \tilde{\mu}_n & = & C_n \tilde{m}_n^-  \nonumber \\
-% \tilde{S}_n   & = & C_n \tilde{P}_n^- C_n^T + R_n  \nonumber \\
-% \tilde{G}_n   & = & \tilde{P}_n^- C_n^T \tilde{S}_n^{-1} \nonumber \\
-% \tilde{m}_n   & = & \tilde{m}_n^- + \tilde{G}_n (y_n - \tilde{\mu}_n) \nonumber \\
-% \tilde{P}_n   & = & \tilde{P}_n^- - \tilde{G}_n \tilde{S}_n \tilde{G}_n^T \nonumber \\
-% \tilde{Z}_n   & = & \tilde{Z}_n^- \mathcal{N}(y_n|\tilde{\mu}_n, \tilde{S}_n) \nonumber   .
-%\end{IEEEeqnarray}
-%
-%All steps use only standard Gaussian identities.
-
-%\section{Initialisation of the Backwards Kalman Filter} \label{app:init_backwards_KF}
-%
-%The backwards Kalman filter recursion may be initialised exactly by inversion of the observation model using the method of \cite{Kitagawa1994}.
-By stacking up the last $L+1$ observations into a single vector, it is possible to write down a joint Gaussian distribution conditional on the $(N-L)$th state.
-This may be inverted to produce an improper Gaussian density for $x_{N-L}$,
-%%
-%\begin{IEEEeqnarray}{rCl}
-% \IEEEeqnarraymulticol{3}{l}{p(y_{N-L:N}|x_{N-L}) = \mathcal{N}( \mathbf{y}_{N-L:N} | H_{N-L} x_{N-L}, \Gamma_{N-L} ) } \nonumber \\
-% & = & \sqrt{\frac{|V_{N-L}|}{|\Gamma_{N-L}|}} \mathcal{N}(x_{N-L} | \nu_{N-L}, V_{N-L} ) \nonumber
-%\end{IEEEeqnarray}
-%\begin{IEEEeqnarray}{rCl}
-% V_{N-L} & = & (H_{N-L}^T \Gamma_{N-L}^{-1} H_{N-L})^{-1} \nonumber \\
-% \nu_{N-L} & = & V_{N-L} H_{N-L}^T \Gamma_{N-L}^{-1} \mathbf{y}_{N-L:N} \nonumber     ,
-%\end{IEEEeqnarray}
-%
-%where $\mathbf{y}_{N-L:N}$ is a vector of concatenated observations, and $H_{N-L}$ and $\Gamma_{N-L}$ may be constructed from the system models,
-%%
-%\begin{IEEEeqnarray}{rCl}
-%\mathbf{y}_{N-L:N} & = & \begin{bmatrix}y_N \\ y_{N-1} \\ \vdots \\ y_{N-L} \end{bmatrix} \nonumber \\
-%%\end{IEEEeqnarray}
-%%\begin{IEEEeqnarray}{rCl}
-%H_{N-L} & = & \begin{bmatrix} C A^L \\ C A^{L-1} \\ \vdots \\ C \end{bmatrix} \nonumber \\
-%%\end{IEEEeqnarray}
-%%\begin{IEEEeqnarray}{rCl}
-% \Gamma_{N-1} & = & K_{N-L} \begin{bmatrix} Q_N & 0 & \dots & 0 \\ 0 & Q_{N-1} & \dots & 0 \\ \vdots & \vdots & \ddots & \vdots \\ 0 & 0 & \dots & Q_{N-L+1} \end{bmatrix} K_{N-L}^T \nonumber \\
-%  & & + \: \begin{bmatrix} R_N & 0 & \dots & 0 \\ 0 & R_{N-1} & \dots & 0 \\ \vdots & \vdots & \ddots & \vdots \\ 0 & 0 & \dots & R_{N-L} \end{bmatrix} \nonumber \\
-%%\end{IEEEeqnarray}
-%%\begin{IEEEeqnarray}{rCl}
-% K_{N-L} & = & \begin{bmatrix} C & CA & \dots & CA^{L-1} \\ 0 & C & \dots & CA^{L-2} \\ \vdots & \vdots & \ddots & \vdots \\ 0 & 0 & \dots & C \\ 0 & 0 & \dots & 0 \end{bmatrix} \nonumber
-%\end{IEEEeqnarray}
-%
-%In the above, it is assumed for clarity that $A_n = A$ and $C_n = C$ are constant matrices.
-This assumption is easily relaxed.
-This inversion should be effected for the smallest value of $L$ such that the matrix inverses exist.
 
@@
 \bibliographystyle{IEEEtran}
-\bibliography{D:/pb404/Dropbox/PhD/Cleanbib}
-
-%\input{D:/pb404/Dropbox/PhD/Biographies/IEEE-Bunch.tex}
-%\input{D:/pb404/Dropbox/PhD/Biographies/IEEE-Godsill.tex}
+\bibliography{Cleanbib}
+
+\input{IEEE-Bunch.tex}
+\input{IEEE-Godsill.tex}
 
